---
title: "Council_Soil_Chamber_MW.NEE"  #breaking NEE apart by plot type for stats, adapted from NEE_statsmodels.Rmd #only 22 observations for complete cases of MW plot types 
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format, choose variables of interest 
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm, thawdepth) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
#Filter & make separate datasets for each plot_ID = "EC" "MW" and "BGC", and by flux type (GPP, NEE, RECO)

#EC - eddy covar tower plot types 
df_EClong <- df_long %>%
  filter(plot_type == "EC")

#MW - moisture warming plot types 
df_MWlong <- df_long %>%
  filter(plot_type == "MW")

#BGC - biogeochem plot types 
df_BGClong <- df_long %>%
  filter(plot_type == "BGC")

#GPP
df_GPP <-df_long %>%
  filter(flux_type == "GPP")

#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")

#RECO
df_RECO <-df_long %>%
  filter(flux_type == "RECO")



#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_EClong <- df_EClong %>% arrange(flux_type)
df_MWlong <- df_MWlong %>% arrange(flux_type)
df_BGClong <- df_BGClong %>% arrange(flux_type)

```

#Break NEE up by plot type 

# MW NEE - Variance and normality
```{r}
# Calculate variance within groups
df_MWlongNEE <- df_MWlong %>%
  filter(flux_type == "NEE")

sum(is.na(df_MWlongNEE$flux_value)) #check for 0's, in this one all the variables of interest have 0 NAs
# Look for missing values & remove / create complete cases dataset if needed
any(is.na(df_MWlongNEE[, c("flux_value", "landscape_position", "inundated", 
                           "soil_temp_10_cm", "thawdepth", "VWC")]))

#Flux and landscape pos
df_variance_landpos <- df_MWlongNEE %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland landscape position than lowland* --> no slope here 


#Flux and inundated - dropped 2 NAs in inundated
df_variance_inundated <- df_MWlongNEE %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)

# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance NOT* inundated plots 


# Shapiro-Wilk test for normality across entire dataset
shapiro_test <- shapiro.test(df_MWlongNEE$flux_value)
print(shapiro_test) #p=0.005, does NOT conform to normality assumptions 


# Test normality within each landscape position
df_MWlongNEE %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: lowland = yes; upland = no


# Test normality based on inundated
df_MWlongNEE %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: N = no, Y = yes


# Q-Q plot of full dataset 
qqnorm(df_MWlongNEE$flux_value)
qqline(df_MWlongNEE$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_MWlongNEE, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Landscape Position")


# histograms with density curves by inundated
ggplot(df_MWlongNEE, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for all main factors
leveneTest(flux_value ~ landscape_position, data = df_MWlongNEE) #p=0.15, homogeneity of var is ok
leveneTest(flux_value ~ inundated, data = df_MWlongNEE)#p = 0.15, ok


#for interactions
leveneTest(flux_value ~ landscape_position * inundated, data = df_MWlongNEE) #p=0.15, ok


```
#Testing models for MW NEE -- dataset already in complete cases for EC 
#Checking colinearity 
```{r}
#Couldn't run the full gls model due to singularity - checked for collinearity and it seems all upland plots are not inundated and all lowland plots are inundated, creating a perfect correlation. Need to simplify the model to remove inundated* 

# Check for correlations between numeric predictors
cor_matrix <- cor(df_MWlongNEE[c("soil_temp_10_cm", "thawdepth", "VWC")], 
                 use = "complete.obs")
print(cor_matrix)

# Check for separation in categorical variables
table(df_MWlongNEE$landscape_position, df_MWlongNEE$inundated)
```



####Testing whether random effect plot_ID should be included or not 
```{r}
#Make plot_ID, inundated, landscape_position as factor so they'll work with gls
df_MWlongNEE$plot_ID = factor(df_MWlongNEE$plot_ID)
df_MWlongNEE$landscape_position = factor(df_MWlongNEE$landscape_position)
df_MWlongNEE$inundated = factor(df_MWlongNEE$inundated)


#use 'ML' for comparing models 
library(nlme)
#gls - no random effMWt
gls.MW.NEE = gls(flux_value ~  landscape_position + soil_temp_10_cm + thawdepth + VWC, data = df_MWlongNEE, method = 'ML') #, na.action=na.exclude) --> don't need, already made dataset with complete cases
anova(gls.MW.NEE) #marginal sig diff in soil temp, p = 0.05

#lme - with random effect of plot_ID
lme.MW.NEE <- lme(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth + VWC, 
               random = ~1 | plot_ID, 
               data = df_MWlongNEE, # na.action=na.exclude, --> don't need, already made dataset with complete cases
               method = 'ML')
anova(lme.MW.NEE) #soil temp sig diff, p <0.01

#Comparing the models 
anova(gls.MW.NEE, lme.MW.NEE) #p = 0.09, no improvement to model with random effect - so plot_ID random effect *NOT needed here 

```

####Multicollinearity in model- EC.NEE
```{r}
#variance inflation factor (vif) - vif = 1 = variables are not correlated; between 1 & 5 = moderately correlated,  5 or higher indicates highly correlated & multicollinearity among variables. GVIF is a generalized VIF for categorical variables with more than two levels; the adj VIF gives values for comparing across variables with diff degrees of freedom* --> look at both when assessing collinearity in a model 
library(car)
vif(gls.MW.NEE) # landpos = 4.3, thaw depth = 3.6, VWC = 3.7, soil temp = 2.1 ....so some moderate collinearity here...but below 5 so should be ok...but may want to remove VWC since it and lanpos are on the high side and likely correlated
```

#### Variance Structre
```{r}
#testing variance structure - using only categorical variables here, not numeric (soil temp, thawdepth, VWC)

#Check sample size by group
table(df_MWlongNEE$landscape_position, df_MWlongNEE$inundated)

#based on inundation --> not using since it's highly correlated with landpos already 
# var.MW.NEE.inundated = update(gls.MW.NEE, weights = varIdent(form = ~1|inundated))
# anova(var.MW.NEE.inundated)
# anova(gls.MW.NEE, var.MW.NEE.inundated) #sig improvement for model, p<0.001


#based on landscape position
var.MW.NEE.landpos = update(gls.MW.NEE, weights = varIdent(form = ~1|landscape_position))
anova(var.MW.NEE.landpos) #all sig diff except landpos
anova(gls.MW.NEE, var.MW.NEE.landpos) #sig improvement for model, p<0.001


#now testing which models are best fit 
anova(gls.MW.NEE, var.MW.NEE.landpos) #p<0.001, so var structure = sig improvement, and moderate diffs in AIC/BIC
anova(var.MW.NEE.inundated, var.MW.NEE.landpos) #inundated and land pos are exactly the same 
```

#testing models - reducing variables, using landpos for variance  
```{r}
#Using the variance structure already determined - landpos 

library(nlme)

# full model.EC - all variables without inundated due to collinearity 
model.MW_full <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth + VWC, 
                  weights = varComb(varIdent(form = ~ 1| landscape_position)),
                 data = df_MWlongNEE, method = "ML", na.action = na.omit)
anova(model.MW_full)

# remove VWC
model.MW_noVWC <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth, 
                        weights = varComb(varIdent(form = ~ 1| landscape_position)),
                        data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove thawdepth
model.MW_nothaw <- gls(flux_value ~ landscape_position + soil_temp_10_cm + VWC, 
                          weights = varComb(varIdent(form = ~ 1| landscape_position)),
                         data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove thawdepth and VWC
model.MW_nothawVWC <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove soil temp
model.MW_nosoiltemp <- gls(flux_value ~ landscape_position + thawdepth + VWC, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove soil temp & VWC
model.MW_nosoiltempVWC <- gls(flux_value ~ landscape_position + thawdepth, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove soil temp & thawdepth
model.MW_nosoiltempthawdepth <- gls(flux_value ~ landscape_position + VWC, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_MWlongNEE, method = "ML", na.action = na.omit)

# remove soil temp & thawdepth & VWC
model.MW_nosoiltempVWCthawdepth <- gls(flux_value ~ landscape_position, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_MWlongNEE, method = "ML", na.action = na.omit)

# interaction of inundated and landpos, removing VWC as suggested by prev model.MW results 
#--> can't do interaction here as there are 0 upland plots that are inundated 



# Compare all variance structure model.MWs using AIC/BIC
model.MWs_list <- list(Full = model.MW_full, 
                   Nothaw = model.MW_nothaw,
                   NoVWC = model.MW_noVWC,
                   NothawVWC = model.MW_nothawVWC,
                   Nosoiltemp = model.MW_nosoiltemp,
                   NosoiltempVWC = model.MW_nosoiltempVWC,
                   Nosoiltempthawdepth = model.MW_nosoiltempthawdepth,
                   NosoiltempVWCthawdepth = model.MW_nosoiltempVWCthawdepth)

# Extract AIC and BIC
model.MW_comparison <- data.frame(
  model.MW = names(model.MWs_list),
  AIC = sapply(model.MWs_list, AIC),
  BIC = sapply(model.MWs_list, BIC)
)

# Sort by AIC
model.MW_comparison <- model.MW_comparison[order(model.MW_comparison$AIC),]
print(model.MW_comparison) #nothawVWC performs best 

# testing model.MWs 
anova(model.MW_full, model.MW_nothaw) #ML:p=0.007; full model better 
anova(model.MW_full, model.MW_noVWC) #ML:p<0.001; full model better
anova(model.MW_full, model.MW_nothawVWC) #ML: p<0.001; full model better 
anova(model.MW_full, model.MW_nosoiltemp) #ML: p<0.001; full model better 
anova(model.MW_full, model.MW_nosoiltempthawdepth)#ML: p<0.001; full model better 
anova(model.MW_full, model.MW_nosoiltempVWC)#p<0.001; full model better 
anova(model.MW_full, model.MW_nosoiltempVWCthawdepth) #p<0.001; full model better 

anova(model.MW_full, model.MW_nothaw, model.MW_noVWC, model.MW_nothawVWC, model.MW_nosoiltemp, model.MW_nosoiltempVWC, model.MW_nosoiltempthawdepth, model.MW_nosoiltempVWCthawdepth) 
#noVWC seems to be the best model.MW, but nothawVWC very close 
anova(model.MW_nothaw, model.MW_nothawVWC) #ML: p<0.001, nothaw better
anova(model.MW_noVWC, model.MW_nothawVWC) #ML:  p<0.001; noVWC better 

#Result: full model appears to be best

```
#Final model 
```{r}

# Final MW NEE model: keep all variables except for inundated -  refit with REML 
MW.NEE.final <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth + VWC, 
                           weights = varIdent(form = ~ 1| landscape_position),
                          data = df_MWlongNEE, method = "REML", na.action = na.omit)
anova(MW.NEE.final)
#sig diff in soil temp & thaw depth (p<0.001); land pos p=0.12; VWC p = 0.28)

```


####EC.NEE.final model - Plot model residuals and qqplot
```{r}
#plot model residuals (homogeneity of variance)
plot(MW.NEE.final)

#qqplot to verify normality
qqnorm(MW.NEE.final)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(MW.NEE.final), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(MW.NEE.final))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(MW.NEE.final)), 
             mean = 0, sd = sigma(MW.NEE.final))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#doesn't look awful, doesn't look great....have Kyle take a look 

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(MW.NEE.final))
hist(residuals(MW.NEE.final)) 
car::qqPlot(x = residuals(MW.NEE.final)) #shows where residuals breach normal distr

```

#Normality of Residuals 
```{r}
# Extract standardized/normalized residuals
std_resid <- residuals(MW.NEE.final, type = "normalized")

# Graphical assessment
par(mfrow = c(1, 2))
# Histogram of residuals
hist(std_resid, main = "Histogram of Standardized Residuals", 
     xlab = "Standardized Residuals", freq = FALSE)
curve(dnorm, add = TRUE, col = "red")

# QQ plot
qqnorm(std_resid, main = "Normal Q-Q Plot")
qqline(std_resid, col = "red")

# Formal test
shapiro_test <- shapiro.test(std_resid)
print(shapiro_test) #for gls final model: p=0.001; normality NOT supported 

```

#Homogeneity of variance 
```{r}
# Plot residuals vs fitted values
par(mfrow = c(1, 1))
plot(fitted(MW.NEE.final), std_resid,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# Check residuals by predictor variables
par(mfrow = c(2, 2))
boxplot(std_resid ~ df_MWlongNEE$landscape_position, 
        main = "Residuals by Landscape Position")
abline(h = 0, lty = 2)

boxplot(std_resid ~ df_MWlongNEE$inundated, 
        main = "Residuals by Inundation")
abline(h = 0, lty = 2)

plot(df_MWlongNEE$soil_temp_10_cm, std_resid,
     xlab = "Soil Temperature (10 cm)", ylab = "Standardized Residuals")
abline(h = 0, lty = 2)

#Result: spread isn't as good as we've seen...little wonky 


# Residuals vs. Fitted plot
plot(fitted(MW.NEE.final), std_resid, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lty = 2)

# Levene’s Test for Homogeneity of Variance
library(car)
leveneTest(std_resid ~ df_MWlongNEE$landscape_position) #p=0.7; this says it's ok 

```
#Need to ask kyle about the normality violation* 



#Stats with final model
```{r}

# Final MW NEE model: keep all variables except for inundated -  refit with REML 
MW.NEE.final <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth + VWC, 
                           weights = varIdent(form = ~ 1| landscape_position),
                          data = df_MWlongNEE, method = "REML", na.action = na.omit)
anova(MW.NEE.final)
#sig diff in soil temp & thaw depth (p<0.001); land pos p=0.12; VWC p = 0.28)

```
#Exploring thaw depth and NEE / binning?
```{r}
library(ggplot2)

# Basic scatterplot with smoothed fit line
ggplot(df_MWlongNEE, aes(x = thawdepth, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "NEE vs Thaw Depth in MW Plots",
       x = "Thaw Depth", y = "NEE Flux")

# Faceted by landscape position to see if the relationship varies
ggplot(df_MWlongNEE, aes(x = thawdepth, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "loess") +
  facet_wrap(~landscape_position) +
  theme_minimal()


library(effects)

# Get effect of thaw depth while holding other variables constant
thaw_effect <- Effect("thawdepth", MW.NEE.final)
plot(thaw_effect)
```

```{r}
# Create bins based on quantiles (equal number of observations)
df_MWlongNEE$thaw_bin <- cut(df_MWlongNEE$thawdepth, 
                            breaks = quantile(df_MWlongNEE$thawdepth, 
                                             probs = c(0, 0.33, 0.67, 1)),
                            labels = c("Shallow", "Medium", "Deep"),
                            include.lowest = TRUE)

# Test differences in NEE between bins
thaw_bin_model <- gls(flux_value ~ thaw_bin + landscape_position + 
                soil_temp_10_cm + VWC,
                 weights = varIdent(form = ~ 1| landscape_position),
                data = df_MWlongNEE, method = "REML")
anova(thaw_bin_model)
summary(thaw_bin_model)

# Visualize
boxplot(flux_value ~ thaw_bin, data = df_MWlongNEE)


```



