---
title: "Council Soil Chamber Analysis - STATS"#Breaking up the figures and data analysis codes 
#time is in AK daylight time** NEE/RECO/GPP are in gC/m2/s; FCO2 and FCH4 are in gC/m2/s, flux_CO2 is in umol/m2/s; flux_CH4 is in nano mol/m2/s
output: html_document
date: "2024-11-18"
---

#Working code - very far from finalized / still a mess 

#Note that for comparison purposes, both instruments 
were used to measure chamber fluxes on July 18, 2018 --> remove potential measurement duplicates from this date?

#measure the Net Ecosystem Exchange (NEE) with the transparent chamber during the day (when photosynthesis is occurring) and the Ecosystem Respiration (Reco) with the opaque chamber during the night (when only respiration is happening), then subtract the Reco value from the NEE value to get GPP: GPP = NEE (transparent chamber) - Reco (opaque chamber)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
#filtered for p<0.05; units umol/m2/s or nmol/m2/s
df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')

#fluxes and moisture/temp df merged; FCO2 in units g/m2/s
df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************
#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}

# library(dplyr)

#Filter the dataframe for plot_ID = "EC" "MW" and "BGC", and by flux type to create df for diff analysis options

#EC - eddy covar tower plot types 
df_EC <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "EC")

#EC - eddy covar tower plot types; just another way to subset - doing it this way saves you from having to make df_EC into d_long format, but below way allows you to choose variables of interest and simplify the df
# df_EC2 <- df_long %>%
#   filter(plot_type == "EC")

#MW - moisture warming plot types 
df_MW <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "MW")

#BGC - biogeochem plot types 
df_BGC <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "BGC")

#GPP
df_GPP <-df_long %>%
  filter(flux_type == "GPP")

#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")
#RECO
df_RECO <-df_long %>%
  filter(flux_type == "RECO")


#Reshape the dataframe to long format and choose variables of interest 

#EC
df_EClong <- df_EC %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>% *choosing variables of interest 
  select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")

#MW
df_MWlong <- df_MW %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>%
   select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")

#BGC
df_BGClong <- df_BGC %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>%
   select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_EClong <- df_EClong %>% arrange(flux_type)
df_MWlong <- df_MWlong %>% arrange(flux_type)
df_BGClong <- df_BGClong %>% arrange(flux_type)

```


#Variance spread in data

####Spread in dataset - NEE

```{r}
library(dplyr)
#looking at spread in the datasets

# Boxplots for categorical predictors
ggplot(df_NEE, aes(x = landscape_position, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Landscape Position")
#looks good, very little spread/difference here 

ggplot(df_NEE, aes(x = inundated, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Inundation")
#also looks good, very little spread here 

ggplot(df_NEE, aes(x = plot_type, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Plot Type")
#also looks good, very little spread 

# Scatterplot for continuous predictor
ggplot(df_NEE, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Flux Value vs Soil Temperature (10 cm)")
#also very little spread 

```
####Spread in Variance among diff variables - NEE

```{r}
# Calculate variance within groups

#exclude NA --> edited to be done beforehand in code chunks above 
# sum(is.na(df_NEE$flux_value)) #flux val has 0 NAs; inundated has 2
# df_NEE.c1 <- df_NEE[complete.cases(df_NEE[, c("inundated")]), ]


#Flux and landscape pos
df_variance_landpos <- df_NEE %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland landscape position than in slope and lowland* 

#Flux and plot type
df_variance_plottype <- df_NEE %>%
  group_by(plot_type) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_plottype)

# Variance plot - flux and plot type 
ggplot(df_variance_plottype, aes(x = plot_type, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Plot Type",
       y = "Variance")
#higher variance in BGC


#Flux and inundated 
#dropped NAs in inundated
df_variance_inundated <- df_NEE.c1 %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)


# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in inundated plots 


# Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(df_NEE$flux_value)
print(shapiro_test) #lower than 0.05, so does deviate from normality 

# Q-Q plot
qqnorm(df_NEE$flux_value)
qqline(df_NEE$flux_value, col = "red")

#homogeneity of variance - levene's test
# p > 0.05: Variances are homogenous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogenous.
library(car)
leveneTest(flux_value ~ landscape_position, data = df_NEE)
#p=0.43, homogeneity of var is ok

```
####Spread in dataset - GPP

```{r}
library(dplyr)
#looking at spread in the datasets

# Boxplots for categorical predictors
ggplot(df_GPP, aes(x = landscape_position, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Landscape Position")
#looks good, very little spread/difference here 

ggplot(df_GPP, aes(x = inundated, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Inundation")
#also looks good, very little spread here 

ggplot(df_GPP, aes(x = plot_type, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Plot Type")
#also looks good, very little spread 

# Scatterplot for continuous predictor
ggplot(df_GPP, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Flux Value vs Soil Temperature (10 cm)")
#also very little spread 

```

####Spread in Variance among diff variables - GPP

```{r}
# Calculate variance within groups

#exclude NA
sum(is.na(df_GPP$flux_value)) #flux val has 3 NAs
sum(is.na(df_GPP$inundated)) #inundated has 2 NAs
df_GPP.c1 <- df_GPP[complete.cases(df_GPP[, c("inundated")]), ]
df_GPP.c2 <- df_GPP[complete.cases(df_GPP[, c("flux_type")]), ]


#Flux and landscape pos
df_variance_landpos <- df_GPP %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland landscape position than in slope and lowland* 

#Flux and plot type
df_variance_plottype <- df_GPP %>%
  group_by(plot_type) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_plottype)

# Variance plot - flux and plot type 
ggplot(df_variance_plottype, aes(x = plot_type, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Plot Type",
       y = "Variance")
#higher variance in BGC


#Flux and inundated 
df_variance_inundated <- df_GPP %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)


# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in not inundated plots 


# Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(df_GPP$flux_value)
print(shapiro_test) #lower than 0.05, so does deviate from normality 

# Q-Q plot
qqnorm(df_GPP$flux_value)
qqline(df_GPP$flux_value, col = "red")

#homogeneity of variance - levene's test
# p > 0.05: Variances are homogenous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogenous.
library(car)
leveneTest(flux_value ~ landscape_position, data = df_GPP)
#p=0.68, homogeneity of var is ok

```
####Spread in dataset - RECO

```{r}
library(dplyr)
#looking at spread in the datasets

# Boxplots for categorical predictors
ggplot(df_RECO, aes(x = landscape_position, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Landscape Position")
#looks good, very little spread/difference here 

ggplot(df_RECO, aes(x = inundated, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Inundation")
#also looks good, very little spread here 

ggplot(df_RECO, aes(x = plot_type, y = flux_value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Flux Value by Plot Type")
#also looks good, very little spread 

# Scatterplot for continuous predictor
ggplot(df_RECO, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Flux Value vs Soil Temperature (10 cm)")
#also very little spread 

```

####Spread in Variance among diff variables - RECO

```{r}
# Calculate variance within groups

#exclude NA
sum(is.na(df_RECO$flux_value)) #flux val has 3 NAs
sum(is.na(df_RECO$inundated)) #inundated has 2 NAs
df_RECO.c1 <- df_RECO[complete.cases(df_RECO[, c("inundated")]), ]
df_RECO.c2 <- df_RECO[complete.cases(df_RECO[, c("flux_type")]), ]


#Flux and landscape pos
df_variance_landpos <- df_RECO %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland landscape position than in slope and lowland* 

#Flux and plot type
df_variance_plottype <- df_RECO %>%
  group_by(plot_type) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_plottype)

# Variance plot - flux and plot type 
ggplot(df_variance_plottype, aes(x = plot_type, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Plot Type",
       y = "Variance")
#all very similar 


#Flux and inundated 
df_variance_inundated <- df_RECO %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)


# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in not inundated plots 


# Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(df_RECO$flux_value)
print(shapiro_test) #lower than 0.05, so does deviate from normality 

# Q-Q plot
qqnorm(df_RECO$flux_value)
qqline(df_RECO$flux_value, col = "red") #bows upward at both ends

#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)
leveneTest(flux_value ~ landscape_position, data = df_RECO)
#p=0.78, homogeneity of var is ok

```


#Testing Models 


#NEE

#df_NEE: Testing df_NEE models 

```{r}

#Testing parameters in df_NEE as a whole  - using Kyle's "fluxes" code as guide 

#Make plot_ID, inundated, plot_type, landscape_position as factor so they'll work with gls

df_NEE$plot_ID = factor(df_NEE$plot_ID)
df_NEE$plot_type = factor (df_NEE$plot_type)
df_NEE$landscape_position = factor(df_NEE$landscape_position)
df_NEE$inundated = factor(df_NEE$inundated)
#df_NEE$soil_temp_10_cm = as.numeric (df_NEE$soil_temp_10_cm) #use this in case it reads it in as factor or character

#Tried with both 'ML' and 'REML' - no major diffs in results 
library(nlme)
#gls - no random effect
gls.NEE = gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, data = df_NEE, method = 'ML', na.action=na.exclude)
anova(gls.NEE) #this shows no sig diff in flux type among these 

#lme - with random effect of plot_ID
lme.NEE <- lme(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, 
               random = ~1 | plot_ID, 
               data = df_NEE, na.action=na.exclude , 
               method = 'ML')
anova(lme.NEE) #shows no sig diff in flux types among these 

#Comparing the models 
anova(gls.NEE, lme.NEE) #shows no diff in model with or without random effect - so random effect not needed here 

```

####Multicollinearity - NEE
```{r}
library(car)
vif(gls.NEE) # <2, no impactful multicollinearity 
```


####Checking each variable on its own 
```{r}
#double checking each var on its own 

gls.NEE.landscape = gls(flux_value ~ landscape_position, data = df_NEE, method = 'ML', na.action=na.exclude)
anova(gls.NEE.landscape) #p=0.11, not sig 

gls.NEE.plottype = gls(flux_value ~ plot_type, data = df_NEE, method = 'ML', na.action=na.exclude)
anova(gls.NEE.plottype) #p=0.0645, not sig 

gls.NEE.inundated = gls(flux_value ~ inundated, data = df_NEE, method = 'ML', na.action=na.exclude)
anova(gls.NEE.inundated) #p=0.32, not sig 

gls.NEE.soiltemp = gls(flux_value ~ soil_temp_10_cm, data = df_NEE, method = 'ML', na.action=na.exclude)
anova(gls.NEE.soiltemp) #p=0.28, not sig 

```

#### Variance Structre - df_NEE
```{r}
#testing variance structure 

#based on inundation
var.NEE.inundated = update(gls.NEE, weights = varIdent(form = ~1|inundated))
anova(var.NEE.inundated) #no diffs among variables 
anova(gls.NEE, var.NEE.inundated) #p = <0.04, adding in the variance by inundation improves the model fit (but only slightly lower AIC, lower BIC, and higher L.Ratio)

#based on landscape position
var.NEE.landpos = update(gls.NEE, weights = varIdent(form = ~1|landscape_position))
anova(var.NEE.landpos) #no diffs among variables 
anova(gls.NEE, var.NEE.landpos) #p = <0.0001, adding in the variance by inundation significantly improves the model fit (but only slightly lower AIC, lower BIC, and higher LogLik)

#based on plot type 
var.NEE.plottype = update(gls.NEE, weights = varIdent(form = ~1|plot_type))
anova(var.NEE.plottype) #now showing sig diffs for landscape pos (p=<0.04), plot type (p=0.03), marginally soil temp (<0.053)
anova(gls.NEE, var.NEE.plottype) #p = <0.0001, adding in the variance by inundation significantly improves the model fit (lower AIC, lower BIC, and higher LogLik)

#based on soil temp --> will not work for this model, "false convergence" --> leaving out, not categorical 
# var.NEE.soiltemp = update(gls.NEE, weights = varIdent(form = ~1|soil_temp_10_cm))
# anova(var.NEE.soiltemp) #"false convergence", will not work for this var structure 

#now testing which models are best fit 
anova(var.NEE.inundated, var.NEE.landpos, var.NEE.plottype) #sig improvement in landpos over inundated 
anova(var.NEE.inundated, var.NEE.landpos) #land pos better 
anova(var.NEE.inundated,var.NEE.plottype) #plot type better
anova(var.NEE.landpos, var.NEE.plottype) #no sig diff but plottype lower AIC/BIC, higher logLik
```

#### Multiple Var structure - NEE 
```{R}
#looking at models with multiple variance structures 
#inundation and land pos 
library(nlme)
var.NEE.landpos.inun = update(gls.NEE, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | landscape_position)))
anova(var.NEE.landpos.inun) #no sig diff among variables 

#inundation and plot type 
var.NEE.plottype.inun = update(gls.NEE, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | plot_type)))
anova(var.NEE.plottype.inun) #shows sig diff in landpos (p=0.004), soil temp (p=0.04), and marginally plot type (0.05)


#land pos and plot type 
var.NEE.landpos.plottype = update(gls.NEE, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type)))
anova(var.NEE.landpos.plottype) #all variables sig except inundated 


anova(var.NEE.landpos.inun,var.NEE.plottype.inun) #no sig diff here, but AIC/BIC lower & logLik higher in plottype.inun

anova(var.NEE.landpos.inun, var.NEE.landpos.plottype, var.NEE.plottype.inun) #appears landpos-plottype is best model of these three 


#land pos and plot type and inun
var.NEE.all = update(gls.NEE, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type), varIdent(form = ~1 |inundated)))
anova(var.NEE.all) #all variables sig except inundated 

#testing all multiple var models 
anova(var.NEE.plottype.inun, var.NEE.all) #NEE.all is better fit, sig 
anova(var.NEE.landpos.inun, var.NEE.landpos.plottype, var.NEE.plottype.inun, var.NEE.all) #sig diff, NEE.plottype.inun has slightly lower AIC/BIC, higher logLik --> seems to fit best, though NEE.all very close to being the same as NEE.plottype.inun

#testing multiple var models with single var models 
anova(gls.NEE, var.NEE.plottype.inun, var.NEE.inundated, var.NEE.landpos, var.NEE.plottype)
#NEE.plottype.inun has lowest AIC/BIC, highest logLik so it appears to be better model 


#looking at fixed effects with selected model of var.NEE.plottype.inun 
anova(var.NEE.plottype.inun, type = "marginal") # soil temp (p<0.04) is sig


#refit with REML
NEE.final = update(var.NEE.plottype.inun, method = "REML")
anova(NEE.final) #plottype marginal (p=0.049), landscape pos (p=0.006) and soil temp (p=0.046) are sig 


#checking colinearity 

gls.NEE2 = gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, data = df_NEE, method = 'REML', na.action=na.exclude)
anova(gls.NEE2)


library(car)
vif(gls.NEE2) #all values <2, so multicolinearity is not a problem here 
vif(gls.NEE)


```

####NEE.final model - Plot model residuals and qqplot
```{r}
#plot model residuals (homogeneity of variance)
plot(NEE.final)

#qqplot to verify normality
qqnorm(NEE.final)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(NEE.final), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(NEE.final))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(NEE.final)), 
             mean = 0, sd = sigma(NEE.final))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#doesn't look awful, doesn't look great....have Kyle take a look 

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(NEE.final))
hist(residuals(NEE.final)) #shows a bit of left skew
car::qqPlot(x = residuals(NEE.final)) #shows where residuals breach normal distr
```

#GPP

#df_GPP: Testing df_GPP models 
```{r}
#Testing parameters in df_GPP as a whole  - using Kyle's "fluxes" code as guide 

#Make plot_ID, inundated, plot_type, landscape_position as factor so they'll work with gls

df_GPP$plot_ID = factor(df_GPP$plot_ID)
df_GPP$plot_type = factor (df_GPP$plot_type)
df_GPP$landscape_position = factor(df_GPP$landscape_position)
df_GPP$inundated = factor(df_GPP$inundated)
#df_NEE$soil_temp_10_cm = as.numeric (df_NEE$soil_temp_10_cm)


#need to remove NA's - check for NAs and remove 
sum(is.na(df_GPP))
summary(df_GPP$soil_temp_10_cm)
# Omit rows with NA values in specific columns so it works with analyses below 
df_GPP.c1 <- df_GPP[complete.cases(df_GPP[, c("inundated")]), ]
df_GPP.c2 <- df_GPP[complete.cases(df_GPP[, c("soil_temp_10_cm")]), ]
df_GPP.c3 <- df_GPP[complete.cases(df_GPP[, c("flux_value")]), ]
sum(is.na(df_GPP.c3$flux_value)) #double check to make sure NAs are gone -> gone 
sum(is.infinite(df_GPP.c3$flux_value)) #and infinite values? --> no



#Tried with both 'ML' and 'REML' - no major diffs in results 
library(nlme)
#gls - no random effect
gls.GPP = gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, data = df_GPP, method = 'ML', na.action=na.exclude)
anova(gls.GPP) #only soil temp is sig 

#lme - with random effect of plot_ID
lme.GPP <- lme(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, 
               random = ~1 | plot_ID, 
               data = df_GPP, na.action=na.exclude , 
               method = 'ML')
anova(lme.GPP) #only soil temp is sig 

#Comparing the models 
anova(gls.GPP, lme.GPP) #shows no diff in model with or without random effect - so random effect not needed here 

```
####Multicollinearity - GPP
```{r}
library(car)
vif(gls.GPP) # <2, no impactful multicollinearity 
```

####Checking each variable on its own 
```{r}
#double checking each var on its own 

gls.GPP.landscape = gls(flux_value ~ landscape_position, data = df_GPP, method = 'ML', na.action=na.exclude)
anova(gls.GPP.landscape) #p=0.29, not sig 

gls.GPP.plottype = gls(flux_value ~ plot_type, data = df_GPP, method = 'ML', na.action=na.exclude)
anova(gls.GPP.plottype) #p=0.259, not sig 

gls.GPP.inundated = gls(flux_value ~ inundated, data = df_GPP, method = 'ML', na.action=na.exclude)
anova(gls.GPP.inundated) #p=0.65, not sig 

gls.GPP.soiltemp = gls(flux_value ~ soil_temp_10_cm, data = df_GPP, method = 'ML', na.action=na.exclude)
anova(gls.GPP.soiltemp) #p=0.053, not sig/borderline 

```

#****Variance Structure -GPP- inundation and soil temp not working 
```{r}

#select variance structure 

var.GPP.inundated = update(gls.GPP, weights = varIdent(form = ~1|inundated))
anova(var.GPP.inundated) #soil temp sig, p = 0.03

anova(gls.GPP, var.GPP.inundated) #p=0.03, GPP.inundated slightly better 

#based on landscape position
var.GPP.landpos = update(gls.GPP, weights = varIdent(form = ~1|landscape_position))
anova(var.GPP.landpos) #sig diffs for soil temp (<0.02)

anova(gls.GPP, var.GPP.landpos) #p = 0.002, GPP.landpos model slightly better 

#based on plot type 
var.GPP.plottype = update(gls.GPP, weights = varIdent(form = ~1|plot_type))
anova(var.GPP.plottype) #no sig diffs 
anova(gls.GPP, var.GPP.plottype) #p = 0.001, GPP.plottype slightly better 


#leaving out soil temp since it is not categorical 
ggplot(df_GPP)+
  geom_point(aes(soil_temp_10_cm,flux_value))


#now testing which models are best fit 
anova(var.GPP.inundated, var.GPP.landpos, var.GPP.plottype) #GPP.landpos slightly better 
anova(var.GPP.inundated, var.GPP.landpos)#lanpos slightly better p = 0.006
anova(var.GPP.inundated,var.GPP.plottype) #plottype slightly better, p = 0.03
anova(var.GPP.landpos, var.GPP.plottype) #no sig diff 


```

#### Multiple Var structure - GPP

```{r}
#looking at models with multiple variance structures 

#inundation and land pos 
library(nlme)
var.GPP.landpos.inun = update(gls.GPP, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | landscape_position)))
anova(var.GPP.landpos.inun) #shows sig diff in soil temp (p=0.017)

#inundation and plot type 
var.GPP.plottype.inun = update(gls.GPP, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | plot_type)))
anova(var.GPP.plottype.inun) #no sig diffs 



#land pos and plot type 
var.GPP.landpos.plottype = update(gls.GPP, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type)))
anova(var.GPP.landpos.plottype) #no sig diffs


anova(var.GPP.landpos.inun,var.GPP.plottype.inun) #no sig diff 

anova(var.GPP.landpos.inun, var.GPP.landpos.plottype, var.GPP.plottype.inun) #appears landpos-plottype is best model of these three 


#land pos and plot type and inun
var.GPP.all = update(gls.GPP, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type), varIdent(form = ~1 |inundated)))
anova(var.GPP.all) #no variables sig, soil temp borderline p=0.05

#testing all multiple var models 
anova(var.GPP.plottype.inun, var.GPP.all) #GPP.all is slightly better fit, p=0.001

anova(var.GPP.landpos.inun, var.GPP.landpos.plottype, var.GPP.plottype.inun, var.GPP.all) #sig diff, but landpost-plottype and GPP.all are nearly the same, landpos-plottype slightly better 

#testing multiple var models with single var models 
anova(gls.GPP, var.GPP.landpos.plottype, var.GPP.plottype.inun, var.GPP.landpos.inun, var.GPP.inundated, var.GPP.landpos, var.GPP.plottype)
#sig diffs but all very close, GPP.landpos-plottype appears best  


#looking at fixed effects with selected model of var.GPP.plottype.inun 
anova(var.GPP.landpos.plottype, type = "marginal") #soil temp is marginal (p=0.05), no others sig 


#refit with REML
GPP.final = update(var.GPP.landpos.plottype, method = "REML")
anova(GPP.final) #only soil temp is margincally sig p = 0.05



```


#RECO

#df_RECO: Testing df_RECO models
```{r}
#Testing parameters in df_RECO as a whole  - using Kyle's "fluxes" code as guide 

#Make plot_ID, inundated, plot_type, landscape_position factor so they'll work with gls

df_RECO$plot_ID = factor(df_RECO$plot_ID)
df_RECO$plot_type = factor (df_RECO$plot_type)
df_RECO$landscape_position = factor(df_RECO$landscape_position)
df_RECO$inundated = factor(df_RECO$inundated)
#df_NEE$soil_temp_10_cm = as.numeric (df_NEE$soil_temp_10_cm)

#Tried with both 'ML' and 'REML' - no major diffs in results 

#gls - no random effect
gls.RECO = gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, data = df_RECO, method = 'ML', na.action=na.exclude)
anova(gls.RECO) #sig diff in soil temp p<0.001

#lme - with random effect of plot_ID
lme.RECO <- lme(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, 
               random = ~1 | plot_ID, 
               data = df_RECO, na.action=na.exclude , 
               method = 'ML')
anova(lme.RECO) #sig diff in soil temp, p <0.001

#Comparing the models 
anova(gls.RECO, lme.RECO) #does show a sig diff, that lme.RECO is slightly better fit (p=0.02 but nearly identical AIC/BIC?logLik)

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(lme.RECO))
hist(residuals(lme.RECO)) #shows a bit of left skew
car::qqPlot(x = residuals(lme.RECO)) #shows where residuals breach normal distr

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(gls.RECO))
hist(residuals(gls.RECO)) #shows a bit of left skew
car::qqPlot(x = residuals(gls.RECO)) #shows where residuals breach normal distr

```
####Multico;llinearity - RECO
```{r}
library(car)
vif(gls.RECO) # <2, no impactful multicollinearity 
```


####Checking each variable on its own 
```{r}
#double checking each var on its own 

gls.RECO.landscape = gls(flux_value ~ landscape_position, data = df_RECO, method = 'ML', na.action=na.exclude)
anova(gls.RECO.landscape) #p=0.88, not sig 

gls.RECO.plottype = gls(flux_value ~ plot_type, data = df_RECO, method = 'ML', na.action=na.exclude)
anova(gls.RECO.plottype) #p=0.49, not sig 

gls.RECO.inundated = gls(flux_value ~ inundated, data = df_RECO, method = 'ML', na.action=na.exclude)
anova(gls.RECO.inundated) #p=0.51, not sig 

gls.RECO.soiltemp = gls(flux_value ~ soil_temp_10_cm, data = df_RECO, method = 'ML', na.action=na.exclude)
anova(gls.RECO.soiltemp) #p=0.006, Sig**

anova(gls.RECO, gls.RECO.soiltemp) #no sig difference

plot(df_RECO$flux_value, df_RECO$soil_temp_10_cm) + geom_point()

```
###Variance Structure - RECO 
```{r}
#select variance structure 

#based on inundation --> won't work, error: "Error in eigen(val, only.values = TRUE) : infinite or missing values in 'x'"
var.RECO.inundated = update(gls.RECO, weights = varIdent(form = ~1|inundated), na.action=na.exclude)
anova(var.RECO.inundated) #soil temp sig, p < 0.0001
anova(gls.RECO, var.RECO.inundated) #inundated better (p=0.04) but only slightly better AIC/BIC

#based on landscape position
var.RECO.landpos = update(gls.RECO, weights = varIdent(form = ~1|landscape_position))
anova(var.RECO.landpos) ##soil temp sig, p < 0.0001
anova(gls.RECO, var.RECO.landpos) #p = 0.7 - var does not improve model 

#based on plot type 
var.RECO.plottype = update(gls.RECO, weights = varIdent(form = ~1|plot_type))
anova(var.RECO.plottype) #soil temp sig, p < 0.0001
anova(gls.RECO, var.RECO.plottype) #p = <0.17, var does not improve model 

#based on soil temp --> "infinite or missing values in x" --> skipping, not categorical 
# var.RECO.soiltemp = update(gls.RECO, weights = varIdent(form = ~1|soil_temp_10_cm), na.action=na.exclude)
# anova(var.RECO.soiltemp) 

#now testing which models are best fit 
anova(var.RECO.inundated, var.RECO.landpos, var.RECO.plottype) #no sig diff 
anova(var.RECO.inundated, var.RECO.landpos) #no sig diff
anova(var.RECO.inundated,var.RECO.plottype) #no sig diff 
anova(var.RECO.landpos, var.RECO.plottype) #no sig diff 

#looks like models with variance structure do not make a significant improvement upon model, so not needed 

```

#Multiple Var Structure - RECO 
```{r}
#looking at models with multiple variance structures 
#inundation and land pos 
library(nlme)
var.RECO.landpos.inun = update(gls.RECO, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | landscape_position)))
anova(var.RECO.landpos.inun) #shows sig diff in soil temp (p<0.0001)

#inundation and plot type 
var.RECO.plottype.inun = update(gls.RECO, weights = varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | plot_type)))
anova(var.RECO.plottype.inun) #shows sig diff in soil temp (p<0.0001)


#testing landpos.inun vs plottype.inun models 
anova(var.RECO.landpos.inun,var.RECO.plottype.inun) #no sig diff here



#land pos and plot type 
var.RECO.landpos.plottype = update(gls.RECO, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type)))
anova(var.RECO.landpos.plottype) #shows sig diff in soil temp (p<0.0001)

#testing the three models 
anova(var.RECO.landpos.inun, var.RECO.landpos.plottype, var.RECO.plottype.inun) #no sig diffs


#land pos and plot type and inun
var.RECO.all = update(gls.RECO, weights = varComb(varIdent(form = ~ 1| landscape_position), varIdent(form = ~ 1 | plot_type), varIdent(form = ~1 |inundated)))
anova(var.RECO.all) #soil temp sig, p<0.001

#testing all multiple var models 
anova(var.RECO.plottype.inun, var.RECO.all) #no sig diff
anova(var.RECO.landpos.inun, var.RECO.landpos.plottype, var.RECO.plottype.inun, var.RECO.all) #no sig diff

#testing multiple var models with single var models 
anova(gls.RECO, var.RECO.plottype.inun, var.RECO.landpos.inun, var.RECO.landpos.plottype, var.RECO.all,    var.RECO.inundated, var.RECO.landpos, var.RECO.plottype)
#no sig diffs, use model without variance structure 


#looking at fixed effects with selected model of var.RECO.plottype.inun 
anova(gls.RECO, type = "marginal") #soil temp sig (p<0.0001) 


#refit with REML
RECO.final = update(gls.RECO, method = "REML")
anova(RECO.final) #soil temp sig, p<0.0001

#checking residuals 
plot(RECO.final)

#checking normality with qq plot
qqnorm(RECO.final)

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(RECO.final))
hist(residuals(RECO.final)) #shows a bit of left skew
car::qqPlot(x = residuals(RECO.final)) #shows where residuals breach normal distr

summary(RECO.final)

#slope and BGC plottype not showing up in the summary or stats....? or am I reading this wrong?

```
```{r}
#Tukey
library(lsmeans)
library(emmeans)



emmeans(RECO.final, specs = "landscape_position") # put those means/CIs in a data frame
emmeans(RECO.final, pairwise ~ landscape_position) #Tukeypairwise
#no sig diffs in flux among landscape positions 

#pairwise contrasts of flux among categorical factors / variables 
lsmeans(RECO.final, adjust = "Tukey", pairwise ~ landscape_position)
lsmeans(RECO.final, adjust = "Tukey", pairwise ~ plot_type) #no sig diffs in flux among plot types 
lsmeans(RECO.final, adjust = "Tukey", pairwise ~ inundated) #not sig 



```
####RECO vs temp relationship -lm with R2 and p 
```{r}
library(ggplot2)

# Create the plot with points and a linear regression line
soil_RECO <- ggplot(df_RECO, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship between RECO and Soil Temperature",
       x = "Soil Temperature at 10 cm",
       y = "RECO (Carbon Flux)")

# Print the plot
print(soil_RECO)



# Fit a linear model
model <- lm(flux_value ~ soil_temp_10_cm, data = df_RECO, na.action=na.exclude)

# Calculate R² value
r_squared <- summary(model)$r.squared
cat("R² value:", r_squared, "\n")

# Test for significance
summary(model)


#Add R2 to the figure 
# Fit a linear model
model <- lm(flux_value ~ soil_temp_10_cm, data = df_RECO)

# Calculate R² value
r_squared <- summary(model)$r.squared

# Create the plot with points, a linear regression line, and R² annotation
soil_RECO2 <- ggplot(df_RECO, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  annotate("text", x = Inf, y = Inf, label = paste("R² =", round(r_squared, 2)),
           hjust = 1.1, vjust = 1.1, size = 5, color = "red") +
  labs(title = "Relationship between RECO and Soil Temperature",
       x = "Soil Temperature at 10 cm",
       y = "RECO (Carbon Flux)")+
  theme_minimal()

# Print the plot
print(soil_RECO2)

```
#Interactions - RECO
```{r}
#landpos * inundated 
RECO.interaction <- gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm +
                          landscape_position * inundated,
                        data = df_RECO, method = 'REML', na.action = na.exclude)
anova(RECO.interaction) #landpos and inundated not a sig interaction 

anova(RECO.final, RECO.interaction) #RECO interaction better, p<0.0001

#plottype * inundated
RECO.interaction2 <- gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm + 
                           inundated * plot_type,
                        data = df_RECO, method = 'REML', na.action = na.exclude)
anova(RECO.interaction2) #plottype and inundated not a sig interaction 

anova(RECO.final, RECO.interaction2) #RECO interaction2 better, p<0.0001

#landpos * soil temp 
RECO.interaction3 <- gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm + 
                           landscape_position * soil_temp_10_cm,
                        data = df_RECO, method = 'REML', na.action = na.exclude)
anova(RECO.interaction3) #plottype and inundated not a sig interaction 

anova(RECO.final, RECO.interaction3) #RECO interaction2 better, p<0.0001, even though interaction effect not sig 


#plottype * soil temp 
RECO.interaction4 <- gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm + 
                           soil_temp_10_cm* plot_type,
                        data = df_RECO, method = 'REML', na.action = na.exclude)
anova(RECO.interaction4) #plottype and inundated not a sig interaction 

anova(RECO.final, RECO.interaction4) #RECO interaction2 better, p<0.0001, even though interaction effect not sig 

#plottype * soil temp 
RECO.interaction5 <- gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm + 
                           soil_temp_10_cm * inundated,
                        data = df_RECO, method = 'REML', na.action = na.exclude)
anova(RECO.interaction5) #plottype and inundated not a sig interaction 

anova(RECO.final, RECO.interaction5) #RECO interaction2 better, p<0.0001, even though interaction effect not sig 

####Simplifying the model 

#gls - no random effect
gls.RECO1 = gls(flux_value ~ plot_type + landscape_position + inundated + soil_temp_10_cm, data = df_RECO, method = 'REML', na.action=na.exclude)
anova(gls.RECO1)

gls.RECO2 = gls(flux_value ~ plot_type + inundated + soil_temp_10_cm, data = df_RECO, method = 'REML', na.action=na.exclude)
anova(gls.RECO2)


anova(gls.RECO1, gls.RECO2) #RECO2 better 


gls.RECO3 = gls(flux_value ~ plot_type + soil_temp_10_cm, data = df_RECO, method = 'REML', na.action=na.exclude)
anova(gls.RECO3)

anova(gls.RECO1, gls.RECO2, gls.RECO3) #RECO3 better



gls.RECO4 = gls(flux_value ~ soil_temp_10_cm, data = df_RECO, method = 'REML', na.action=na.exclude)
anova(gls.RECO4)

anova(gls.RECO1, gls.RECO2, gls.RECO3, gls.RECO4) #RECO4 better


```




#Older code, ignore for now 


#### EC plots - NEE 
```{r}
# Load nlme package
library(nlme)
library(lme4)
library(agricolae)
library(car)
library(emmeans)

df_EC_NEE <- df_EClong %>% filter(flux_type == "NEE")


#histogram of distr of data in df_NEE
hist(df_EC_NEE$flux_value)
#skewed a bit to the left

library(lme4)
EC_GPP_LMM <- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_GPP)
summary(EC_GPP_LMM, corr = F) #model with random effect 
Anova(EC_GPP_LMM, type = "II", test.statistic = "F", ddf = "Kenward-Roger") #Can only be used with REML
#p=0.157; not sig 

#testing which model is more appropriate, using plot_ID as a random effect 
library(lme4)
EC_GPP_LMM <- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_GPP, REML = FALSE )
EC_GPP_LMM_null <- lmer(flux_value ~ 1 + (1|plot_ID), data = df_EC_GPP, REML = FALSE )
#looking at if the random effect even matters, this seems to say it does not
#maybe try weighting by factors using varIdent 
anova(EC_GPP_LMM,EC_GPP_LMM_null) #compares models --> p = 0.09, no sig diff 




```

```{r}




#these below fit with ML
library(nlme)
model_NEE <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_NEE,method = 'ML')
summary(model_NEE)
anova(model_NEE) #p = 0.07 - can't customize ANOVA test stats in nlme 


#Fits with REML 
library(nlme)
EC_NEE_NLME <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_NEE)
summary(EC_NEE_NLME)
anova(EC_NEE_NLME)


library(lme4)
EC_NEE_LMM<- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_NEE )
summary(EC_NEE_LMM, corr = F)
Anova(EC_NEE_LMM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#p=0.087; not sig 

#testing the model without a random effect of plot_id
model_NEE_LM<- lm(flux_value ~ landscape_position, data = df_EC_NEE )
summary(model_NEE_LM, corr = F)
Anova(model_NEE_LM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#without plot_ID as a random effect, this is sig with p = 0.015

 #Testing which model is a beter fit: NEE_LMM = with random effect of plot_ID; NEE_LM = without random effect of plot_ID 
anova(EC_NEE_LMM, model_NEE_LM)
#Results show lower AIC and BIC for model_NEE2, with random effect of plot_ID, so we will proceed with this version 

#Testing with emmeans 
library(emmeans)
emmeans(EC_NEE_LMM, specs = "landscape_position") # 
emmeans(EC_NEE_LMM, pairwise ~ landscape_position) #Tukeypairwise --> not sig, p = 0.0872

```


#Residuals and normality of mixed-effects model - ECC plots NEE2
```{r}
#checking normality of residuals distribution 
plot(model_NEE2) # check constant variance
lattice::qqmath(model_NEE2) # check normality of residuals
plot(model_NEE2 , plot_ID ~ resid(., scaled=TRUE)) # equal var within Plots
#---------------------------------------------------------------------------------------------------
#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(model_NEE2), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(model_NEE2))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(model_NEE2)), 
             mean = 0, sd = sigma(model_NEE2))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#these look pretty good

#For further testing, if needed 
qqnorm(residuals(model_NEE2))
hist(residuals(model_NEE2)) #shows a bit of left skew

#brown forsythe test to test variance among groups, want p to be above 0.05 to show no sig diff
install.packages("onewaytests") #for a brown forscythe test
library(onewaytests)
##Testing for Homogeneity of variance with Brown-Forsythe test

#code syntax: #bf.test(dependent variable ~ independent variable, data = dataset) 
bf.test(flux_value ~ landscape_position, data=df_NEE) #this is variance among groups, so diff not being stat sig is a good thing -> but p is 0.00843, so it is sig 

#variance within groups test with Levene's test
#code syntax: leveneTest(dataset$dependent variable, dataset$independent variable)
leveneTest(df_NEE$flux_value, df_NEE$landscape_position) #p above 0.05 means there no sig variance within groups, so data is distr normally --> p = 0.57, so not sig 

#If I need to log transform:
#dataset$new_name of log dataset <- log(dataframe$dependent variable) 
#same for square rooting transformation, just use "sqrt"


#This not working, look into later 
#checking ratio of largest grp var to smallest group var, should be 3 or below 
grp_vars <- with(df_NEE, tapply(flux_type,landscape_position,var))
max(grp_vars)/min(grp_vars)


```

####EC plots - GPP

```{r}
library(nlme)
library(lme4)
library(agricolae)
library(car)
library(emmeans)

df_EC_GPP <- df_EClong %>% filter(flux_type == "GPP")

ggplot(df_EC_GPP, aes (x=landscape_position, y = flux_value))+geom_boxplot()

hist(df_EC_GPP$flux_value)
#skewed right 

library(nlme)
EC_GPP_NLME <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_GPP)
summary(EC_GPP_NLME)
anova(EC_GPP_NLME) #not sig, p = 0.1462; only works with chi-sqr


library(lme4)
EC_GPP_LMM <- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_GPP, REML = FALSE )
EC_GPP_LMM_null <- lmer(flux_value ~ 1 + (1|plot_ID), data = df_EC_GPP, REML = FALSE )
#looking at if the random effect even matters, this seems to say it does not
#maybe try weighting by factors using varIdent 
summary(EC_GPP_LMM, corr = F)
anova(EC_GPP_LMM,EC_GPP_LMM_null)
Anova(EC_GPP_LMM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#p=0.157; not sig 



#testing the model without a random effect of plot_id
EC_GPP_LM<- lm(flux_value ~ landscape_position, data = df_EC_GPP )
summary(EC_GPP_LM, corr = F)
Anova(EC_GPP_LM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#no random effect, p = 0.0335

#sig without random effects, p = 0.0335
fit <- aov(flux_value ~ landscape_position, data = df_EC_GPP)
summary(fit)

#Testing with emmeans 
library(emmeans)
emmeans(EC_GPP_LMM, specs = "landscape_position") # put those means/CIs in a data frame
emmeans(EC_GPP_LMM, pairwise ~ landscape_position) #Tukeypairwise
#also shows contrast is not sig, p = 0.157

#testing with  a t-test for flux_value by landscape_position --> is sig at p=0.01381, but this does not include a random effect
t_test_result <- t.test(flux_value ~ landscape_position, data = df_EC_GPP)
# View the result
print(t_test_result)


```
```{r}
# Load lmerTest package for p-values - this does LMM with an added p-value 
library(lmerTest)

# Fit the model with lmerTest to obtain p-values
model <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_EC_GPP)
summary(model) #p=0.10254
Anova(model, type = "II", test.statistic = "F", ddf = "Kenward-Roger") #p=0.1574

```

#Tukey letters compact letter display  (just for practice, as there is no sig diff here)
```{r}
# Install multcomp if not already installed
install.packages("multcomp")
library(multcomp)
library(multcompView)
emGPP <- emmeans(EC_GPP_LMM, specs = "landscape_position")
cld(emGPP, Letter = "abcdefghijk")

#FOR TUKEY HSD 
# create compact letter display!
# using lme model, get est means with CIs
emmeans(EC_GPP_LMM, specs = "landscape_position") # put those means/CIs in a data frame
emmeans(EC_GPP_LMM, pairwise ~ landscape_position) #Tukeypairwise
emm_EC_GPP_LMM_df <- as.data.frame(emmeans(EC_GPP_LMM, specs = "landscape_position"))


# get the compact letter displays using multcomp package function cld() 
library(multcomp)
tuk_EC_GPP <- glht(EC_GPP_LMM, linfct = mcp(landscape_position = "Tukey"))
tuk_EC_GPP_cld <- cld(tuk_EC_GPP)
tuk_EC_GPP_cld

# add letters to data frame
emm_EC_GPP_LMM_df$letters <- tuk_EC_GPP_cld$mcletters$Letters

#plot with letters from tukey groupings 
library(ggplot2)
ggplot(df_EC_GPP, aes(x=landscape_position, y = flux_value))+
  geom_boxplot()+
  theme_minimal()+
   annotate("text", x = 1:2, y = 0.0002, label = tuk_EC_GPP_cld$mcletters$Letters)

#a more strict tukey, adheres to the p<0.05

library(multcomp)
library(multcompView)
emm <- emmeans(EC_GPP_LMM, specs = "landscape_position")
cld(emm, Letter = "abcdefghijk")

```
#LME of GPP, NEE, RECO among all plot types 

```{r}
# mixed effect models
library(lme4)
library(emmeans)
library(car)

library(nlme)
#model testing differences in landscape position among all plot types --> not sig, p = 0.3266
EC_GPP_lme <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_GPP, na.action=na.exclude) #lets R work around the 0 values in dataset )
summary(EC_GPP_lme, corr = F)
Anova(EC_GPP_lme,test.statistic = "F", type = "II", ddf = "Kenward-Roger") #does type 2 but only with chi-sqr in nlme package 

library(lme4)
# GPP among all plot types  --> not sig, p = 0.3459
model_GPP_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_GPP)
summary(model_GPP_plottype)
Anova(model_GPP_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# NEE among all plot types  --> not sig, p = 0.147
model_NEE_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_NEE)
summary(model_NEE_plottype)
Anova(model_NEE_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# RECO among all plot types  --> not sig, p = 0.89
model_RECO_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_RECO)
summary(model_RECO_plottype)
Anova(model_RECO_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")


```
#LME of GPP, NEE, RECO among plot types
```{r}

library(lme4)
# GPP among all plot types  --> not sig, p = 0.288
model_GPP_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_GPP)
summary(model_GPP_plottype2)
Anova(model_GPP_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# NEE among all plot types  --> not sig, p = 0.07
model_NEE_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_NEE)
summary(model_NEE_plottype2)
Anova(model_NEE_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# RECO among all plot types  --> not sig, p = 0.59
model_RECO_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_RECO)
summary(model_RECO_plottype2)
Anova(model_RECO_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")
```



```{r}
#Residuals and normality of mixed-effects model - ECC plots GPP

#checking normality of residuals distribution 
plot(model_NEE2) # check constant variance
lattice::qqmath(model_NEE2) # check normality of residuals
plot(model_NEE2 , plot_ID ~ resid(., scaled=TRUE)) # equal var within Plots
#---------------------------------------------------------------------------------------------------
#checking how this QQ plot compares to plots created with normally distributed residuals
#ASK CLAY - will this test work with my other datasets? just plug in diff dataset and test? 
#How to do a glmm / glmer model in case of transformations not working 
#how to transform / back transform (I do have negative values and 0 values in some cases)
op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(model_NEE2), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(model_NEE2))

# now create 24 qq plots using Normal data with sigma(PB19Eri.angCN_LMM
for(i in 1:24){
  # rnorm() samples from a Normal dist'n 
  d <- rnorm(length(residuals(model_NEE2)), 
             mean = 0, sd = sigma(model_NEE2))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#These residuals look pretty good 

#For further testing, if needed 
qqnorm(residuals(model_NEE2))
hist(residuals(model_NEE2)) #shows a bit of left skew

#brown forsythe test to test variance among groups, want p to be above 0.05 to show no sig diff
install.packages("onewaytests") #for a brown forscythe test
library(onewaytests)
##Testing for Homogeneity of variance
## but these test for normality in data, not residuals, right? 
# with Brown-Forsythe test
#bf.test(dependent variable ~ independent variable, data = dataset) 
bf.test(flux_value ~ landscape_position, data=df_NEE) #this is variance among groups, so diff not being stat sig is a good thing -> but p is 0.00843, so it is sig 
#variance within groups test

# with Levene's test
#leveneTest(dataset$dependent variable, dataset$independent variable)
leveneTest(df_NEE$flux_value, df_NEE$landscape_position) #p above 0.05 means there no sig variance within groups, so data is distr normally --> p = 0.57, so not sig 

#If I need to log transform:
#dataset$new_name of log dataset <- log(dataframe$dependent variable) 
#same for square rooting transformation, just use "sqrt"


#This not working, look into later 
#checking ratio of largest grp var to smallest group var, needs to be 3 or below 
grp_vars <- with(df_NEE, tapply(flux_type,landscape_position,var))
max(grp_vars)/min(grp_vars)



```

#EC RECO
```{r}
library(nlme)
library(lme4)
library(car)

df_RECO <- df_EClong %>% filter(flux_type == "RECO")


model_RECO <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_RECO)
summary(model_RECO)


df_GPP <- df_EClong %>% filter(flux_type == "GPP")

model_GPPtest <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_GPP)
summary(model_GPPtest)
anova(model_GPPtest)
```









