---
title: 'Council_Soil_Chamber_BGC.NEE' #breaking NEE apart by plot type for stats, adapted from NEE_statsmodels.Rmd 
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format, choose variables of interest 
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm, thawdepth) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
#Filter & make separate datasets for each plot_ID = "EC" "MW" and "BGC", and by flux type (GPP, NEE, RECO)

#BGC - eddy covar tower plot types 
df_BGClong <- df_long %>%
  filter(plot_type == "BGC")


#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")


#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_BGClong <- df_BGClong %>% arrange(flux_type)

```

#Filter for NEE 
```{r}
#make sure to use complete cases of all variables of interest 
df_BGClongNEE <- df_BGClong %>%
  filter(flux_type == "NEE") %>%
  filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC))


sum(is.na(df_BGClongNEE$VWC)) #check for 0's, in this one all the variables of interest have 0 NAs


```

# Categorical Predictors: Variance and normality of categorical predictor variables 
```{r}
#Flux and landscape pos
df_variance_landpos <- df_BGClongNEE %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland 


#Flux and inundated 
df_variance_inundated <- df_BGClongNEE %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)

# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in NOT* inundated plots 


# Shapiro-Wilk test for normality across entire dataset
shapiro_test <- shapiro.test(df_BGClongNEE$flux_value)
print(shapiro_test) #p<0.001, not normal 


# Test normality within each landscape position
df_BGClongNEE %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: lowland & upland = no; slope = yes


# Test normality based on inundated
df_BGClongNEE %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: not inun = no; inun = yes 


# Q-Q plot of full dataset 
qqnorm(df_BGClongNEE$flux_value)
qqline(df_BGClongNEE$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_BGClongNEE, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Landscape Position")


# histograms with density curves by inundated
ggplot(df_BGClongNEE, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for all main factors
leveneTest(flux_value ~ landscape_position, data = df_BGClongNEE) #p=0.87, homogeneity of var is ok
leveneTest(flux_value ~ inundated, data = df_BGClongNEE)#p = 0.81, ok


#for interactions
leveneTest(flux_value ~ landscape_position * inundated, data = df_BGClongNEE) #p=0.94, ok


```

#Continuous predictors: Variance and normality of continuous predictor variables 

###Testing to see which kind of correlation to use: Pearson for parametric, Spearman for non-parametric
```{r}
#test distribution normality of each numeric predictor variable 
shapiro.test(df_BGClongNEE$soil_temp_10_cm) #p=0.0039, NOT normal** 
shapiro.test(df_BGClongNEE$thawdepth) #p = <0.001, NOT normal **
shapiro.test(df_BGClongNEE$VWC) #p<0.001, NOT normal **

#visualize pairs to see if it looks linear
pairs(df_BGClongNEE[c("soil_temp_10_cm", "thawdepth", "VWC")]) #NOPE 

plot(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$thawdepth)

plot(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$VWC)

plot(df_BGClongNEE$thawdepth, df_BGClongNEE$VWC)

ggplot(df_BGClongNEE, aes(x = soil_temp_10_cm, y = thawdepth))+
  geom_point()+
    labs(
     x = "Soil Temperature (10 cm)", 
     y = "Thaw Depth")+
geom_abline(intercept = 0, slope = 7, color = "red") # 1:1 line in red, adj slope to account for diffs in scale


#So this means for correlations and looking at collinearity, I need to use non-parametric tests 


#Check homogeneity of variance: should see even spread of of points for homogeneity of variance 

# Fit a simple model
simple_model <- lm(flux_value ~ VWC, data = df_BGClongNEE)

# Extract fitted values and residuals
fitted_values <- fitted(simple_model)
residuals <- residuals(simple_model)

# Plot residuals vs. predictor
plot(df_BGClongNEE$VWC, residuals, 
     xlab = "Cont. Pred. Var", ylab = "Residuals",
     main = "Residuals vs. Predictor Var")
abline(h = 0, lty = 2)

# Create a scale-location plot --> should be randomly scattered with a relatively flat trend line (red) for homogeneity of variance 
plot(simple_model, which = 3)


#Breusch-Pagan test - specifically tests if variance of residuals is constant - want a p > 0.05 for homogeneity of var
library(lmtest)
bptest(simple_model)

#thawdepth = p=0.85, shows there is homogeneity of var 
#soil temp = p = 0.5, suggests homogeneity of var 
#VWC = p = 0.25,  suggests homogeneity of var 

```
```{r}
ggplot(df_BGClongNEE, aes(x = soil_temp_10_cm, y = thawdepth)) +
  geom_point() +
  labs(
    x = "Soil Temperature (10 cm)", 
    y = "Thaw Depth (cm)", 
    title = "Scatterplot of Soil Temperature vs. Thaw Depth") +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(intercept = 0, slope = 7, color = "blue", linetype = "dashed")

```


#Checking colinearity / correlations to determine which variables to use for models 

#Non-parametric tests of correlations - spearman's correlation 

#Corr of continuous variables 
```{r}
#Correlation between soil temp, thaw depth, and VWC

cor.test(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$thawdepth, method="spearman")#p<0.21, not sig, not correlated
cor.test(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$VWC, method="spearman") #p<0.001, *Sig*, correlated, one should be removed 
cor.test(df_BGClongNEE$thawdepth, df_BGClongNEE$VWC, method="spearman") #p<0.001, *sig*, correlated, one should be removed 

#Shows thawdepth and VWC are correlated, and VWC & soil temp are correlated 
```

#Checking correlation in categorical variables 
```{r}
# Check for separation in categorical variables
table(df_BGClongNEE$landscape_position, df_BGClongNEE$inundated) 


# Create contingency table to examine correlation quantitatively / confirm the separation test above 
cont_table <- table(df_BGClongNEE$landscape_position, df_BGClongNEE$inundated)
print(cont_table)

# Test for association
chisq.test(cont_table) #p<0.74 --> shows no correlation 
# Or for small sample sizes
fisher.test(cont_table) #p<0.70, not correlated 

#These are ok, don't show signs of correlation

```

#Determining model for BGC - NEE


#Step 1: Determine which of the correlated variables to keep
```{r}
# Create competing models with one variable at a time - use "ML" to compare models, use "REML" for fitting final model 

#thawdepth and VWC are correlated, and VWC & soil temp are correlated 
model_temp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 data = df_BGClongNEE, method = "ML")

model_thaw <- gls(flux_value ~ landscape_position + thawdepth, 
                 data = df_BGClongNEE, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 data = df_BGClongNEE, method = "ML")


# model_VWCsoiltemp <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
#                  data = df_BGClongNEE, method = "ML")
# 
# model_VWCthawdepth <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
#                  data = df_BGClongNEE, method = "ML")
# 
# model_soiltempthawdepth <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth, 
#                  data = df_BGClongNEE, method = "ML")

# Compare with AIC/BIC
AIC(model_temp, model_thaw, model_VWC) #model_VWC lowest AIC/BIC
#AIC(model_temp, model_thaw, model_VWC, model_VWCthawdepth, model_VWCsoiltemp, model_soiltempthawdepth) #model_VWC has lowest AIC, but not by much
BIC(model_temp, model_thaw, model_VWC)#model_VWC lowest AIC/BIC
#BIC(model_temp, model_thaw, model_VWC, model_VWCthawdepth, model_VWCsoiltemp, model_soiltempthawdepth) #model_VWC has lowest BIC 

#Of the single variables comparisons model_VWC is best predictor by a little, followed by model_thaw, suggests we should keep VWC...

```

#Step 2: Now, test whether adding soil temp or thaw depth improves model 
```{r}
model_VWC <- gls(flux_value ~ landscape_position + VWC,
                        data = df_BGClongNEE, method = "ML")

model_VWCsoiltemp <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm,
                        data = df_BGClongNEE, method = "ML")

model_VWCthawdepth <- gls(flux_value ~ landscape_position + VWC + thawdepth,
                        data = df_BGClongNEE, method = "ML")

anova(model_VWC, model_VWCsoiltemp, model_VWCthawdepth) #model_VWC has lowest AIC/BIC, but VWCsoiltemp has highest logLik, no p's are sig 
anova(model_VWC, model_VWCsoiltemp) #model VWC has lower AIC/BIC, higher logLik, p = 0.4, not sig 
anova(model_VWC, model_VWCthawdepth) #p=0.9, not sig, but model VWC has lower AIC/BIC
anova(model_VWCsoiltemp, model_VWCthawdepth) #VWCsoil temp has lower AIC but nearly the same BIC, higher LogLik (no p-value?)

#Appears neither thaw depth or soil temp improve model 



```

#Step 3: Test if random effect of plot_ID is appropriate
```{r}
# #If not already: Make plot_ID, inundated, landscape_position as factor so they'll work with gls
# df_MWlongNEE$plot_ID = factor(df_MWlongNEE$plot_ID)
# df_MWlongNEE$landscape_position = factor(df_MWlongNEE$landscape_position)
# df_MWlongNEE$inundated = factor(df_MWlongNEE$inundated)

#Use best predictor variable from Step 1, which was soil temp 
#use 'ML' for comparing models 

library(nlme)
#gls - no random effect
model_fixed <- gls(flux_value ~ landscape_position + VWC, 
                  data = df_BGClongNEE, method = "ML")

#lme - with random effect of plot_ID
model_random <- lme(flux_value ~ landscape_position + VWC, 
                   random = ~1|plot_ID, 
                   data = df_BGClongNEE, method = "ML")

# Compare
anova(model_fixed, model_random) #no sig diff and AIC/BIC marginally lower in model-fixed, p=0.99, so go for simpler model without random effect 
```

#Step 4: Test for variance structure 
```{r}
# test if adding a variance structure is appropriate 
model_homoscedastic <- gls(flux_value ~ landscape_position + VWC, 
                          data = df_BGClongNEE, method = "ML")

model_heteroscedastic <- gls(flux_value ~ landscape_position + VWC, 
                            weights = varIdent(form = ~1|landscape_position),
                            data = df_BGClongNEE, method = "ML")

# Compare
anova(model_homoscedastic, model_heteroscedastic) #p=<0.001 SIG*; hetero has lower AIC/BIC and higher LogLik so we do use a variance structure here * 
```
#**TO DO **
Step 5: Try adding any remaining non-correlated variables back in to see if it improves the model (use reduction and addition / reduced models to test if various un-correlated predictors improve the model or not, and favor the simpler model)
```{r}
# Assuming soil_temp was best and VWC is not strongly correlated with it
model_1var <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_MWlongNEE, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + soil_temp_10_cm + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_MWlongNEE, method = "ML")

# Compare
anova(model_1var, model_2var) #model 2var has lower AIc/BIC and higher LogLik, p <0.001 * add VWC back in* 
```


