---
title: 'Council_Soil_Chamber_EC.NEE' #breaking NEE apart by plot type for stats, adapted from NEE_statsmodels.Rmd 
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format, choose variables of interest 
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm, thawdepth) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
#Filter & make separate datasets for each plot_ID = "EC" "MW" and "BGC", and by flux type (GPP, NEE, RECO)

#EC - eddy covar tower plot types 
df_EClong <- df_long %>%
  filter(plot_type == "EC")


#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")


#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_EClong <- df_EClong %>% arrange(flux_type)

```

#Filter for NEE 
```{r}
# Calculate variance within groups
df_EClongNEE <- df_EClong %>%
  filter(flux_type == "NEE")

sum(is.na(df_EClongNEE$flux_value)) #check for 0's, in this one all the variables of interest have 0 NAs
# Look for missing values
any(is.na(df_EClongNEE[, c("flux_value", "landscape_position", "inundated", 
                           "soil_temp_10_cm", "thawdepth", "VWC")]))
```

# Categorical Predictors: Variance and normality of categorical predictor variables 
```{r}
#Flux and landscape pos
df_variance_landpos <- df_EClongNEE %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland landscape position than lowland* --> no slope here 


#Flux and inundated - dropped 2 NAs in inundated
df_variance_inundated <- df_EClongNEE %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)

# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance NOT* inundated plots 


# Shapiro-Wilk test for normality across entire dataset
shapiro_test <- shapiro.test(df_EClongNEE$flux_value)
print(shapiro_test) #p=0.067, above 0.05, so looks like it conforms to normality assumptions 


# Test normality within each landscape position
df_EClongNEE %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: all yes 


# Test normality based on inundated
df_EClongNEE %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: all yes 


# Q-Q plot of full dataset 
qqnorm(df_EClongNEE$flux_value)
qqline(df_EClongNEE$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_EClongNEE, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Landscape Position")


# histograms with density curves by inundated
ggplot(df_EClongNEE, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for all main factors
leveneTest(flux_value ~ landscape_position, data = df_EClongNEE) #p=0.57, homogeneity of var is ok
leveneTest(flux_value ~ inundated, data = df_EClongNEE)#p = 0.23, ok


#for interactions
leveneTest(flux_value ~ landscape_position * inundated, data = df_EClongNEE) #p=0.54, ok


```

#Continuous predictors: Variance and normality of continuous predictor variables 

###Testing to see which kind of correlation to use: Pearson for parametric, Spearman for non-parametric
```{r}
#test distribution normality of each numeric predictor variable 
shapiro.test(df_EClongNEE$soil_temp_10_cm) #p=0.41, normal 
shapiro.test(df_EClongNEE$thawdepth) #p = <0.001, NOT normal **
shapiro.test(df_EClongNEE$VWC) #p=0.35, normal 

#visualize pairs to see if it looks linear
pairs(df_EClongNEE[c("soil_temp_10_cm", "thawdepth", "VWC")]) #NOPE 

#So this means for correlations and looking at collinearity, I need to use non-parametric tests 


#Check homogeneity of variance: should see even spread of of points for homogeneity of variance 

# Fit a simple model
simple_model <- lm(flux_value ~ soil_temp_10_cm, data = df_EClongNEE)

# Extract fitted values and residuals
fitted_values <- fitted(simple_model)
residuals <- residuals(simple_model)

# Plot residuals vs. predictor
plot(df_EClongNEE$soil_temp_10_cm, residuals, 
     xlab = "Cont. Pred. Var", ylab = "Residuals",
     main = "Residuals vs. Thaw Depth")
abline(h = 0, lty = 2)

# Create a scale-location plot --> should be randomly scattered with a relatively flat trend line (red) for homogeneity of variance 
plot(simple_model, which = 3)

#thawdepth = spread is fine 
#soil temp = looks ok with spread, trendline a little bowed but relatively flat 
#VWC = spread looks ok, trendlind has downward trend

#Breusch-Pagan test - specifically tests if variance of residuals is constant - want a p > 0.05 for homogeneity of var
library(lmtest)
bptest(simple_model)

#thawdepth = p=0.4, shows there is homogeneity of var 
#soil temp = p = 0.9, suggests homogeneity of var 
#VWC = p = 0.03, suggest not homogeneous ** 

```
#Checking colinearity / correlations to determine which variables to use for models 


####I used Pearson corr for parametric distributions first, mistkanely - left code/results in for reference, but moved on to use Spearman, the non-parametric corr test, and need to rebuild the models based on these results**

####Pearson's Corr - parametric (incorrect, but leaving for when there is a parametric dataset)
```{r}
#Couldn't run the full gls model due to singularity - checked for collinearity and it seems all upland plots are not inundated and all lowland plots are inundated, creating a perfect correlation. Need to simplify the model to remove inundated* 

# Check for correlations between numeric predictors - this uses Pearson correlation, which assumes linear relationships and normal distribution among variables
cor_matrix <- cor(df_EClongNEE[c("soil_temp_10_cm", "thawdepth", "VWC")], 
                 use = "complete.obs")
print(cor_matrix) #this showed no correlations, BUT it was the incorretc test to use, as some of my data was not normally distirbuted*** that's why the spearman correlation tests below did highlight correlations missed in the cor_matrix 
```

#Non-parametric tests of correlations - spearman's correlation 
```{r}
#Correlation between soil temp and thaw depth 

cor.test(df_EClongNEE$soil_temp_10_cm, df_EClongNEE$thawdepth, method="spearman")#p<0.04, SIG* --> shows soil temp and thaw depth are capturing most of the same information / are correlated and one should be removed 
cor.test(df_EClongNEE$soil_temp_10_cm, df_EClongNEE$VWC, method="spearman") #p=0.2, not sig, not correlated 
cor.test(df_EClongNEE$thawdepth, df_EClongNEE$VWC, method="spearman") #p=0.11, not sig, not correlated 

```


#Checking correlation in categorical variables 
```{r}
# Check for separation in categorical variables
table(df_EClongNEE$landscape_position, df_EClongNEE$inundated) #all upland is N, all lowland = Y; highly correlated, should remove one of these variables 


# Create contingency table to examine correlation quantitatively / confirm the separation test above 
cont_table <- table(df_EClongNEE$landscape_position, df_EClongNEE$inundated)
print(cont_table)

# Test for association
chisq.test(cont_table) #p<0.001 --> shows correlation
# Or for small sample sizes
fisher.test(cont_table) #p<0.001, shows correlation 

# # Measure strength (if you have the vcd package)
# library(vcd)
# assocstats(cont_table)

#Shows that there is correlation as only lowland plots are inundated, but that's to be expected and not ALL lowland plots are inundated, so since the final model's VIF is <5, I can soundly keep both landscape position and inundated** 
```

####Testing models for EC NEE -- dataset already in complete cases for EC 
####Testing whether random effect plot_ID should be included or not 

#Important note: anova() with type="marginal" (or type="III") tests each main effect while accounting for all other terms, including interactions, while anova() without specifying the type defaults to Type I (sequential) testing, which considers terms in the order they appear in the model
```{r}
#Make plot_ID, inundated, landscape_position as factor so they'll work with gls
df_EClongNEE$plot_ID = factor(df_EClongNEE$plot_ID)
df_EClongNEE$landscape_position = factor(df_EClongNEE$landscape_position)
df_EClongNEE$inundated = factor(df_EClongNEE$inundated)


#use 'ML' for comparing models 
library(nlme)
#gls - no random effect
gls.EC.NEE = gls(flux_value ~  landscape_position + inundated + soil_temp_10_cm + thawdepth + VWC, data = df_EClongNEE, method = 'ML') #, na.action=na.exclude) --> don't need, already made dataset with complete cases

anova(gls.EC.NEE) #sig diff in landpos, p = 0.02
# Using the standard anova function with F tests
anova(gls.EC.NEE, type = "marginal", test = TRUE) #none sig


#remove thawdepth since it's correlated with soil temp 
gls.EC.NEE2 = gls(flux_value ~  landscape_position + inundated + soil_temp_10_cm + VWC, data = df_EClongNEE, method = 'ML') #, na.action=na.exclude) --> don't need, already made dataset with complete cases

anova(gls.EC.NEE2) #sig diff in landpos, p = 0.016
# Using the standard anova function with F tests
anova(gls.EC.NEE2, type = "marginal", test = TRUE) #none sig 




#lme - with random effect of plot_ID
lme.EC.NEE <- lme(flux_value ~ landscape_position + inundated + soil_temp_10_cm + thawdepth + VWC, 
               random = ~1 | plot_ID, 
               data = df_EClongNEE, # na.action=na.exclude, --> don't need, already made dataset with complete cases
               method = 'ML')
anova(lme.EC.NEE) #shows no sig diff in NEE among these 
# Using the standard anova function with F tests
anova(lme.EC.NEE, type = "marginal", test = TRUE) #none sig 


#lme - removing thaw depth due to correlation with soil temp 
lme.EC.NEE2 <- lme(flux_value ~ landscape_position + inundated + soil_temp_10_cm + VWC, 
               random = ~1 | plot_ID, 
               data = df_EClongNEE, 
               method = 'ML')

anova(lme.EC.NEE2) #shows no sig diff in NEE among these 
# Using the standard anova function with F tests
anova(lme.EC.NEE2, type = "marginal", test = TRUE) #none sig 





#Comparing the models 
anova(gls.EC.NEE, lme.EC.NEE) #p = 0.3, no improvement to model with random effect - so plot_ID random effect *NOT needed here 
anova(gls.EC.NEE2, lme.EC.NEE2) #same as above 
anova(gls.EC.NEE, gls.EC.NEE2) #no sig diff but without thaw depth has bit of a low AIC/BIC 
anova(lme.EC.NEE, lme.EC.NEE2) #no sig diff

```

####Multicollinearity in model- EC.NEE
```{r}
#variance inflation factor (vif) - vif = 1 = variables are not correlated; between 1 & 5 = moderately correlated,  5 or higher indicates highly correlated & multicollinearity among variables. GVIF is a generalized VIF for categorical variables with more than two levels; the adj VIF gives values for comparing across variables with diff degrees of freedom* --> look at both when assessing collinearity in a model 
library(car)
vif(gls.EC.NEE) # all variables ~2 or less, no impactful multicollinearity 
```

#### Variance Structre
```{r}
#testing variance structure - using only categorical variables here, not numeric (soil temp, thawdepth, VWC)

#Check sample size by group
table(df_EClongNEE$landscape_position, df_EClongNEE$inundated)

#based on inundation --> can't use this var structure due to small sample size in levels 
# var.EC.NEE.inundated = update(gls.EC.NEE, weights = varIdent(form = ~1|inundated))
# anova(var.EC.NEE.inundated) 
# anova(gls.EC.NEE, var.EC.NEE.inundated) 


#based on landscape position
var.EC.NEE.landpos = update(gls.EC.NEE, weights = varIdent(form = ~1|landscape_position))
anova(var.EC.NEE.landpos) #landpos and soil temp sig (sequential / heirarchical anova test)
anova(var.EC.NEE.landpos, type = "marginal", test = TRUE) #none sig --> trust this one for individual variable effects on NEE* 


#now testing which models are best fit 
anova(gls.EC.NEE, var.EC.NEE.landpos) #p=0.033, so var structure = sig improvement, but the differences in AIC/BIC/LogLik are very small. will test both models and see how they do...

```

#testing models - reducing variables, using landpos for variance  
```{r}
#Using the variance structure already determined - landpos 

library(nlme)

# full model.EC - all variables 
model.EC_full <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm + thawdepth + VWC, 
                  weights = varComb(varIdent(form = ~ 1| landscape_position)),
                 data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove VWC
model.EC_noVWC <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm + thawdepth, 
                        weights = varComb(varIdent(form = ~ 1| landscape_position)),
                        data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove thawdepth
model.EC_nothaw <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm + VWC, 
                          weights = varComb(varIdent(form = ~ 1| landscape_position)),
                         data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove thawdepth and VWC
model.EC_nothawVWC <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove soil temp
model.EC_nosoiltemp <- gls(flux_value ~ landscape_position + inundated + thawdepth + VWC, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove soil temp & VWC
model.EC_nosoiltempVWC <- gls(flux_value ~ landscape_position + inundated + thawdepth, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove soil temp & thawdepth
model.EC_nosoiltempthawdepth <- gls(flux_value ~ landscape_position + inundated + VWC, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "ML", na.action = na.omit)

# remove soil temp & thawdepth & VWC
model.EC_nosoiltempVWCthawdepth <- gls(flux_value ~ landscape_position + inundated, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "ML", na.action = na.omit)

# interaction of inundated and landpos, removing VWC as suggested by prev model.EC results 
#--> can't do interaction here as there are 0 upland plots that are inundated 



# Compare all variance structure model.ECs using AIC/BIC
model.ECs_list <- list(Full = model.EC_full, 
                   Nothaw = model.EC_nothaw,
                   NoVWC = model.EC_noVWC,
                   NothawVWC = model.EC_nothawVWC,
                   Nosoiltemp = model.EC_nosoiltemp,
                   NosoiltempVWC = model.EC_nosoiltempVWC,
                   Nosoiltempthawdepth = model.EC_nosoiltempthawdepth,
                   NosoiltempVWCthawdepth = model.EC_nosoiltempVWCthawdepth)

# Extract AIC and BIC
model.EC_comparison <- data.frame(
  model.EC = names(model.ECs_list),
  AIC = sapply(model.ECs_list, AIC),
  BIC = sapply(model.ECs_list, BIC)
)

# Sort by AIC
model.EC_comparison <- model.EC_comparison[order(model.EC_comparison$AIC),]
print(model.EC_comparison) #nothawVWC performs best 

# testing model.ECs 
anova(model.EC_full, model.EC_nothaw) #ML: no sig diff
anova(model.EC_full, model.EC_noVWC) #ML: no sig diff
anova(model.EC_full, model.EC_nothawVWC) #ML: no sig diff
anova(model.EC_full, model.EC_nosoiltemp) #ML: no sig diff
anova(model.EC_full, model.EC_nosoiltempthawdepth)#ML: no sig diff
anova(model.EC_full, model.EC_nosoiltempVWC)#no sig diff
anova(model.EC_full, model.EC_nosoiltempVWCthawdepth) #no sig diff

anova(model.EC_full, model.EC_nothaw, model.EC_noVWC, model.EC_nothawVWC, model.EC_nosoiltemp, model.EC_nosoiltempVWC, model.EC_nosoiltempthawdepth, model.EC_nosoiltempVWCthawdepth) 
#noVWC seems to be the best model.EC, but nothawVWC very close 
anova(model.EC_nothaw, model.EC_nothawVWC) #ML: no sig diff
anova(model.EC_noVWC, model.EC_nothawVWC) #ML: no sig diff

#Result: no model is sig...the best model appears to be no thawdepth and no VWC, which puts us back at original with landpos, plot type, inundated, and soil temp 
```

#Final model

```{r}
# Final EC NEE model: remove thawdepth and VWC - refit with REML 
EC.NEE.final <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm, 
                           weights = varComb(varIdent(form = ~ 1| landscape_position)),
                          data = df_EClongNEE, method = "REML", na.action = na.omit)
anova(EC.NEE.final)# sequential anova --> sig diff in lanpost (p<0.01) and soil temp marginal (p=0.048)
anova(EC.NEE.final, type = "marginal", test = TRUE) #landpos p=0.026; soil p = 0.048
```

#### Final model: EC.NEE.final model - Plot model residuals and qqplot
```{r}
#plot model residuals (homogeneity of variance)
plot(EC.NEE.final)

#qqplot to verify normality
qqnorm(EC.NEE.final)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(EC.NEE.final), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(EC.NEE.final))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(EC.NEE.final)), 
             mean = 0, sd = sigma(EC.NEE.final))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#doesn't look awful, doesn't look great....have Kyle take a look 

#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(EC.NEE.final))
hist(residuals(EC.NEE.final)) 
car::qqPlot(x = residuals(EC.NEE.final)) #shows where residuals breach normal distr

```

#Normality of Residuals 
```{r}
# Extract standardized/normalized residuals
std_resid <- residuals(EC.NEE.final, type = "normalized")

# Graphical assessment
par(mfrow = c(1, 2))
# Histogram of residuals
hist(std_resid, main = "Histogram of Standardized Residuals", 
     xlab = "Standardized Residuals", freq = FALSE)
curve(dnorm, add = TRUE, col = "red")

# QQ plot
qqnorm(std_resid, main = "Normal Q-Q Plot")
qqline(std_resid, col = "red")

# Formal test
shapiro_test <- shapiro.test(std_resid)
print(shapiro_test) #for gls final model: p=0.46 --> normality supported

```

#Homogeneity of variance 
```{r}
# Plot residuals vs fitted values
par(mfrow = c(1, 1))
plot(fitted(EC.NEE.final), std_resid,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# Check residuals by predictor variables
par(mfrow = c(2, 2))
boxplot(std_resid ~ df_EClongNEE$landscape_position, 
        main = "Residuals by Landscape Position")
abline(h = 0, lty = 2)

boxplot(std_resid ~ df_EClongNEE$inundated, 
        main = "Residuals by Inundation")
abline(h = 0, lty = 2)

plot(df_EClongNEE$soil_temp_10_cm, std_resid,
     xlab = "Soil Temperature (10 cm)", ylab = "Standardized Residuals")
abline(h = 0, lty = 2)

#Result: pretty good even spread here, indicates the variance structure captures the variance well * 
```


#Stats with final model
```{r}
# Final EC NEE model: remove thawdepth and VWC - refit with REML - used complete cases df_EClongNEE
EC.NEE.final <- gls(flux_value ~ landscape_position + inundated + soil_temp_10_cm, 
                           weights = varIdent(form = ~ 1| landscape_position),
                          data = df_EClongNEE, method = "REML", na.action = na.omit)
anova(EC.NEE.final)#sequential-->sig diff in lanpost (p<0.01) and soil temp marginal (p=0.048)
anova(EC.NEE.final, type = "marginal", test = TRUE)#landpops and soil temp sig 

library(emmeans)
emmeans(EC.NEE.final, adjust = "Tukey", pairwise ~ landscape_position) #p=0.027 - only upland and lowland  
emmeans(EC.NEE.final, adjust = "Tukey", pairwise ~ inundated) #p=0.64, not sig 

```






