---
title: "Council_Soil_Chamber_Landscape.CH4"  #FCH4 in gC/m2/s, flux_CH4 is in nano mol/m2/s
output: html_document
date: "2024-12-03"

---
#Note that for comparison purposes, both instruments 
were used to measure chamber fluxes on July 18, 2018 --> remove potential measurement duplicates from this date?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')
```


#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')

# head(df_NEE_RECO2_GPP)
# str(df_NEE_RECO2_GPP)

```

#Kyle recommended having units be umol, so convert from gC/m2/s to umolC/m2/s
```{r}
# #umol to g
# # Net CO2 Flux - convert from umol/m2/s to gC/m2/s
#  flux_CO2 * (1/1000000) * 12))
# 
# 
# #Net CH4 flux --> convert from nmol/m2/s to gC/m2/s
#  flux_CH4 * (1/1000000000)*12))

#flux-CO2 and flux-CH4 are in this df and are already in umol so below is not necessary but leaving for reference
#FCH4 and FCO2 are in g/m2/s 

#g to umol 
# Net CO2 Flux - convert from umol/m2/s to gC/m2/s
# df_NEE_RECO2_GPP <- df_NEE_RECO2_GPP%>%
#   mutate(FCO2 = ifelse(is.na(flux_CO2), 0, flux_CO2 * 1/12 * 1000000)) 
# 
# 
# #Net CH4 flux --> convert from nmol/m2/s to gC/m2/s
# df_NEE_RECO2_GPP < - df_NEE_RECO2_GPP %>%
#   mutate(FCH4 = ifelse(is.na(flux_CH4), 0, flux_CH4 * 1/12 * 1000000000)*12))
```


#Re-shape df 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format - leaving out soil temp 15cm because of a lot of missing values
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, FCH4, flux_CO2, flux_CH4, inundated, air_temp, soil_temp_10_cm, thawdepth, VWC) %>%
  pivot_longer(cols = c(flux_CO2, flux_CH4), 
               names_to = "flux_type", 
               values_to = "flux_value")



```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
# library(dplyr)

#Filter the dataframe for plot_ID = "EC" "MW" and "BGC", and by flux type to create df for diff analysis options

#EC - eddy covar tower plot types 
df_EC <- df_long%>%
  filter(plot_type == "EC")
#use complete cases 
df_EC <- df_EC %>%
   filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC, air_temp))

#MW - moisture warming plot types 
df_MW <- df_long %>%
  filter(plot_type == "MW")

#use complete cases 
df_MW<- df_MW %>%
   filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC, air_temp))

#BGC - biogeochem plot types - most representative of all the landscape positions 
df_BGC <- df_long %>%
  filter(plot_type == "BGC")

#use complete cases 
df_BGC <- df_BGC %>%
   filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC, air_temp)) #went from 204 to 204 obs 


#methane across all plot types 
df_FCH4 <-df_long %>%
  filter(flux_type == "flux_CH4")

sum(is.na(df_FCH4$flux_CH4)) #0
sum(is.na(df_FCH4$thawdepth)) #8
sum(is.na(df_FCH4$soil_temp_10_cm)) #10
sum(is.na(df_FCH4$VWC)) #28
sum(is.na(df_FCH4$air_temp)) #0 

df_FCH4 <- df_FCH4 %>%
   filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position,inundated, VWC, air_temp)) #went from 185 to 158 obs 


```


#checking data normality and homog of var 
```{r}
# Test normality within each landscape position
df_FCH4 %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: all no 


# Test normality based on inundated
df_FCH4 %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: all no  


# Q-Q plot of full dataset 
qqnorm(df_FCH4$flux_value)
qqline(df_FCH4$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_FCH4, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of FCH4 by Landscape Position")


# histograms with density curves by inundated
ggplot(df_FCH4, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of FCH4 by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for all main factors
leveneTest(flux_value ~ landscape_position, data = df_FCH4) #p=0.18, ok 
leveneTest(flux_value ~ inundated, data = df_FCH4)#p <0.001, NOT* homog of var


#for interactions
leveneTest(flux_value~ landscape_position * inundated, data = df_FCH4) #p<0.001, not homog*


#test distribution normality of each numeric predictor variable 
shapiro.test(df_FCH4$soil_temp_10_cm) #p=0.004, NOT normal** 
shapiro.test(df_FCH4$thawdepth) #p = <0.001, NOT normal **
shapiro.test(df_FCH4$VWC) #p<0.001, NOT normal **
shapiro.test(df_FCH4$air_temp) #p<0.001, NOT normal **
```
#Checking colinearity / correlations to determine which variables to use for models 

#Non-parametric tests of correlations - spearman's correlation 

#Corr of continuous variables 
```{r}
#checking for Correlation between soil temp, thaw depth, air temp, and VWC

#soil temp v thaw depth
cor.test(df_FCH4$soil_temp_10_cm, df_FCH4$thawdepth, method="spearman")#p=0.02, *SIG*, correlated **
#soil temp v VWC
cor.test(df_FCH4$soil_temp_10_cm, df_FCH4$VWC, method="spearman") #p=0.02, *Sig*, correlated **
#soil temp v air temp 
cor.test(df_FCH4$soil_temp_10_cm, df_FCH4$air_temp, method="spearman") #p<0.001, correlated **
#thaw depth v VWC
cor.test(df_FCH4$thawdepth, df_FCH4$VWC, method="spearman") #p<0.001, *sig*, correlated **
#thaw depth v air temp 
cor.test(df_FCH4$thawdepth, df_FCH4$air_temp, method="spearman") #p=0.59, not corr 
#air temp v VWC
cor.test(df_FCH4$air_temp, df_FCH4$VWC, method="spearman") #p=0.15, not corr 

#Shows soil temp is correlated with thaw depth, VWC, and air temp, & thawdepth and VWC are correlated
```

#Checking correlation in categorical variables 
```{r}
# Check for separation in categorical variables
table(df_FCH4$landscape_position, df_FCH4$inundated) 
  #          N  Y
  # lowland 33 14
  # slope   22  2
  # upland  81  6

# Create contingency table to examine correlation quantitatively / confirm the separation test above 
cont_table <- table(df_FCH4$landscape_position, df_FCH4$inundated)
print(cont_table)

# Test for association
chisq.test(cont_table) #p<0.001--> correlated
# Or for small sample sizes
fisher.test(cont_table) #p=0.001, correlated 

#landpos and inun are correlated, will remove inun --> but will tests models below just to be sure 


```

#Alt ways to test corr between continuous and categorical variables - help from Claude

```{r}
# Point-Biserial Correlation: For binary variables like inundated (Yes/No)
# The point-biserial correlation is essentially a Pearson correlation where one variable is continuous (VWC) and the other is dichotomous (inundated Yes/No). It ranges from -1 to +1, with the interpretation similar to other correlation coefficients:

library(ltm)
biserial.cor(df_FCH4$VWC, as.numeric(df_FCH4$inundated == "Y")) #means "Y" is coded as 1, and "N" is coded as 0
#Result: -0.4412363 --> shows moderate neg correlation, which seems to imply that the relationship is inverse which is odd......not sure about this approach 



#Kruskall-Wallis test - if sig, shows there is some kind of association between the variables 
kruskal.test(VWC ~ landscape_position, data = df_FCH4)#p<0.001, SIG
kruskal.test(inundated ~ landscape_position, data = df_FCH4)#p<0.001, SIG*
```















#Testing models for CH4 using complete cases 

#Step 1: Determine which of the correlated variables to keep
```{r}
# Create competing models with one variable at a time - use "ML" to compare models, use "REML" for fitting final model 

#Shows soil temp is correlated with thaw depth, VWC, and air temp, & thawdepth and VWC are correlated

model_temp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 data = df_FCH4, method = "ML")

model_thaw <- gls(flux_value ~ landscape_position + thawdepth, 
                 data = df_FCH4, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 data = df_FCH4, method = "ML")

model_airtemp <- gls(flux_value ~ landscape_position + air_temp, 
                 data = df_FCH4, method = "ML")

# Compare with AIC/BIC - lowest values denote better fit 
AIC(model_temp, model_thaw, model_VWC, model_airtemp) #thaw, then VWC have lowest AIC
BIC(model_temp, model_thaw, model_VWC, model_airtemp) #model_thaw, then VWC, have lowest BIC 

#air temp has highest AIC/BIC so removing airtemp since it's correlated with soil temp 
#model thaw seems best...followed by VWC, and they're correlated, so maybe remove VWC 



```

#Step 2: Now, test whether VWC and soil temp improve model or not 
```{r}
# using thawdepth as it seemed to be best predictor var
model_thaw <- gls(flux_value ~ landscape_position + thawdepth, 
                        data = df_FCH4, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                        data = df_FCH4, method = "ML")
#compare
anova(model_thaw, model_VWC) #model thaw better, lower AIC/BIC and higher logLik, suggests using thaw instead of VWC

#add VWC
model_with_VWC <- gls(flux_value ~ landscape_position + thawdepth + VWC, 
                     data = df_FCH4, method = "ML")

# Compare
anova(model_thaw, model_with_VWC) #significant improvement of model with VWC...p<0.001


#add soil temp 
model_with_soiltemp <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm, 
                     data = df_FCH4, method = "ML")

# Compare
anova(model_thaw, model_with_soiltemp) #p=0.001...= sig improvement, but marginally lower AIC/BIC including soil temp 

#add both soil temp and VWC 
model_with_all <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm + VWC, 
                     data = df_FCH4, method = "ML")
#compare
anova(model_with_all, model_thaw)#model with all lower AIC/BIC, p<0.001 so sig improvement...

#add inundated to with_all model 
model_with_allinun <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm + VWC + inundated, 
                     data = df_FCH4, method = "ML")
#compare
anova(model_with_all, model_with_allinun)#model including inun lower AIC/BIC, p<0.001 so sig improvement...

#compare all 
anova(model_thaw, model_with_soiltemp, model_with_VWC, model_with_all, model_with_allinun)

#despite correlations, it seems having all the variables, including inun, significantly improves the model...should check VIF for multicollinearity

#variance inflation factor (vif) - vif = 1 = variables are not correlated; between 1 & 5 = moderately correlated,  5 or higher indicates highly correlated & multicollinearity among variables. GVIF is a generalized VIF for categorical variables with more than two levels; the adj VIF gives values for comparing across variables with diff degrees of freedom* --> look at both when assessing collinearity in a model 
library(car)
vif(model_with_allinun) # all <2, so looks like it could be ok to keep all ?? *ask kyle 
```
#Step 3: Test if random effect of plot_ID or plot_type is appropriate
```{r}
#Use best predictor variable from Step 1, which was thawdepth
#use 'ML' for comparing models 

library(nlme)
#gls - no random effect
model_fixed <- gls(flux_value ~ landscape_position + thawdepth, 
                  data = df_FCH4, method = "ML")

#lme - with random effect of plot_ID
model_random <- lme(flux_value ~ landscape_position + thawdepth, 
                   random = ~1|plot_ID, 
                   data = df_FCH4, method = "ML")

# Compare
anova(model_fixed, model_random) #very marginally lower AIC/BIC in random, but p = 0.016, so sig improvement with random effect of plot ID

#lme - with random effect of plot type 
model_random2 <- lme(flux_value ~ landscape_position + thawdepth, 
                   random = ~1|plot_type, 
                   data = df_FCH4, method = "ML")

# Compare
anova(model_fixed, model_random2) #p=0.99, random effect of plot type does not improve model 
anova(model_random2, model_random)#plot_ID as random effect better, lower AIC/BIC  


#try will all variables 
model_fixed_all <- gls(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated, 
                  data = df_FCH4, method = "ML")

#lme - with random effect of plot_ID
model_random_all <- lme(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated, 
                   random = ~1|plot_ID, 
                   data = df_FCH4, method = "ML")

#lme - with random effect of plot_type
model_random_all2 <- lme(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated, 
                   random = ~1|plot_type, 
                   data = df_FCH4, method = "ML")

anova(model_fixed_all, model_random_all)#no sig improvement when adding plot ID, p = 0.4
anova(model_fixed_all, model_random_all2)#p=0.99, no sig improvement when adding plot type 
anova(model_random_all, model_random_all2) #no p, essentially the same 

anova(model_random, model_random_all) #model_random_all sig better, p<0.001, marginally lower AIC/BIC
anova(model_random, model_fixed_all) #p<0.001, model_fixed_all sig better 


anova(model_random, model_random2, model_random_all, model_random_all2, model_fixed, model_fixed_all)
#model_fixed_all has lowest AIC/BIC

vif(model_fixed_all) # all <2, so looks like it could be ok to keep all ? *Ask Kyle 

```

#Step 4: Test for variance structure 
```{r}
#using best model from code chunk before: 

# test if adding a variance structure is appropriate 

#var structure by landpos 
model_homoscedastic <- gls(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated,  
                          data = df_FCH4, method = "ML")

model_heteroscedastic <- gls(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated,  
                            weights = varIdent(form = ~1|landscape_position),
                            data = df_FCH4, method = "ML")

# Compare
anova(model_homoscedastic, model_heteroscedastic) #p=<0.001 SIG*; var structure of landpos improves model 


#variance structure by plot type 
model_heteroscedastic2 <- gls(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated,   
                            weights = varIdent(form = ~1|plot_type),
                            data = df_FCH4, method = "ML")

# Compare
anova(model_homoscedastic, model_heteroscedastic2) #p<0.001, model improves with plot type var structure 
anova(model_heteroscedastic, model_heteroscedastic2)#no p, appears using plot_type var structure may be better fit, lower AIC/BIC, higher logLik


#combined variance structure of landpos and plot type 
model_heteroscedastic3 <- gls(flux_value ~ landscape_position + thawdepth + VWC + soil_temp_10_cm + inundated,  
                            weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                            data = df_FCH4, method = "ML")
anova(model_heteroscedastic3, type = "marginal", test = TRUE)#landpos p=0.02, SIG*
                            
                            #varComb(varIdent(form = ~ 1| inundated), varIdent(form = ~ 1 | landscape_position)))

anova(model_homoscedastic, model_heteroscedastic3) # hetero3 p<0.001, hetero3 better model 
anova(model_heteroscedastic, model_heteroscedastic3) #p<0.001, hetero3 better 
anova(model_heteroscedastic2, model_heteroscedastic3)#p<0.001, hetero3 better model 

#suggests using both landpos and plot_type as var structures 


#Maybe try adding and reducing here?? check VIF again *** TO DO 

vif(model_heteroscedastic3) # all <2, so looks like it could be ok to keep all ? *Ask Kyle 
```
#Step 5: Double check best correlated var was chosen, with and without extra var, now that variance structure was added 

```{r}
#starting with thawdepth as that appeared to be strongest predictor variable per previous tests - re-testing for best correlated predictor variable now that var structure is added 
model_1varthawdepth <- gls(flux_value ~ landscape_position + thawdepth, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")


model_1varsoiltemp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")


model_1varVWC <- gls(flux_value ~ landscape_position + VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

# Compare
anova(model_1varthawdepth, model_1varsoiltemp) #no p, thawdepth had marginally lower AIC/BIC
anova(model_1varthawdepth, model_1varVWC) #no p, VWC had marginally lower AIC/BIC
anova(model_1varsoiltemp, model_1varVWC) #no p, VWC had marginally lower AIC/BIC
```


#Step 6: Try adding any remaining non-correlated variables back in to see if it improves the model (use reduction and addition / reduced models to test if various un-correlated predictors improve the model or not, and favor the simpler model)

#*Note: in this the variables are correlated but according to VIF not heavily, and adding them back in improved the model significantly 

#multiple var structure 
```{r}
#so now starting with model_1varVWC as base for comparison as that seems to maybe be best (marginally) predictor variable 

#add soil temp 
model_2var <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

# Compare
anova(model_1varVWC, model_2var) #p=0.53, model_1varVWC had marginally lower AIC/BIC, adding soil temp did not improve model  


#add thawdepth 
model_2var <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

# Compare
anova(model_1varVWC, model_2var) #p=0.33, model_1varVWC had marginally lower AIC/BIC, adding thaw does not significantly improve model 


model_3var <- gls(flux_value ~ landscape_position + VWC + thawdepth + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

anova(model_1varVWC, model_3var) #p=0.57, marginally lower AIC/BIC for model_1varVWC, adding extra variables does not improve the model 

#all variables 
model_4var <- gls(flux_value ~ landscape_position + VWC + thawdepth + soil_temp_10_cm + inundated, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

anova(model_1varVWC, model_4var) #p=0.001, SIG* - significantly improves the model - may be linked to inundation though....


#all variables 
model_VWCinun <- gls(flux_value ~ landscape_position + VWC + inundated, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

anova(model_VWCinun, model_4var) #p=0.64, no sig difference between these two, model_VWCinun slightly lower AIC/BIC 

anova(model_1varVWC, model_VWCinun, model_4var) #p=0.001, model_VWCinun sig better than model_1varVWC, but no sig improvement with model_4var, suggests inundation is important, but likely correlated ....how to handle this? 

vif(model_VWCinun) #all below 2, suggesting no significant multicollinearity 


#test interaction between land post and VWC
model_interact <- gls(flux_value ~ landscape_position * VWC, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

#interaction between VWC and inundated 
model_interact2 <- gls(flux_value ~ landscape_position + VWC * inundated, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")

#interaction between landpos, VWC and inundated 
model_interact3 <- gls(flux_value ~ landscape_position * VWC * inundated, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "ML")


# Compare
anova(model_1varVWC, model_interact)#p=0.008, *SIG*, interaction significantly improves model 
anova(model_1varVWC, model_interact2)#p<0.001, interact 2 significantly improves model 
anova(model_1varVWC, model_interact3)#p<0.001, interact3 significantly improves model 

anova(model_interact, model_interact2)#no p, model_interact2 marginally lower AIC/BIC, higher LogLik
anova(model_interact, model_interact3)#p<0.001 model_interact3 significantly improves model 
anova(model_interact3, model_interact2)#<0.001, model_interact3 significantly improves model 



anova(model_VWCinun, model_interact) #p=0.3,model_VWCinun has slightly lower AIC/BIC, suggests interaction effect is not important to model 
anova(model_VWCinun, model_interact2) #p=0.02, *SIG*, model_interact2 has very slightly lower AIC, but basically the same BIC, suggests interaction effect of VWC and inun may be important but that perhaps the complexity of the interaction is negating the explanatory power....*Ask Kyle*  
anova(model_VWCinun, model_interact3)#p<0.001, interact3 significantly improves model 
```

#Step 7: Test for multicollinearity in model 

####VIF - Checking multicollinearity among the different types of models with VIF 
```{r}
#Since I'm using continuous and categorical variables here, the adj VIF is more appropriate to follow as it specifically developed to address the limitation of standard VIF when dealing with factors or categorical variables. *Allows for comparison between continuous and categorical variables. For interpretation guidelines with adjusted VIF:
# Values below 2: Low multicollinearity
# Values between 2-4: Moderate multicollinearity  -- above 4 = high multicollinearity 

vif(model_VWCinun) #all below 2 so should be ok 
#                        GVIF Df GVIF^(1/(2*Df))
# landscape_position 1.514199  2        1.109292
# VWC                1.061871  1        1.030471
# inundated          1.478085  1        1.215765

vif(model_interact)
#                            GVIF Df GVIF^(1/(2*Df))
# landscape_position     18.153841  2        2.064154
# VWC                     9.805461  1        3.131367
# landscape_position:VWC 32.913386  2        2.395207


vif(model_interact2) #shows inundated and VWC are likely moderately correlated, but below the 5 threshold, and adj VIF are all at ~2 or less, so this might be acceptable....*Ask Kyle? 

#                        GVIF Df GVIF^(1/(2*Df))
# landscape_position 1.523141  2        1.110926
# VWC                1.155879  1        1.075118
# inundated          4.956730  1        2.226372
# VWC:inundated      4.626829  1        2.151006


vif(model_interact3) #HIGH correlations here, therefore we should not use this model * 
#                                       GVIF Df GVIF^(1/(2*Df))
# landscape_position                64.28515  2        2.831572
# VWC                               36.30070  1        6.025007
# inundated                         12.73873  1        3.569135
# landscape_position:VWC           142.67782  2        3.456122
# landscape_position:inundated     390.43981  2        4.445172
# VWC:inundated                     18.03519  1        4.246786
# landscape_position:VWC:inundated 404.86772  2        4.485680

```


#Step 8: Autocorrelation, model residuals (normality and homogeneity of variance)


#Test autocorrelation
from web: Autocorrelation analysis measures the relationship of the observations between the different points in time. Testing for autocorrelation is important for time series analysis and regression models to ensure the validity of statistical inferences and the reliability of model predictions. Autocorrelation refers to the correlation between values of the same variable at different time points, and its presence can lead to biased results
```{r}
# Extract normalized residuals from GLS model
residuals_gls <- residuals(model_interact3, type = "normalized")

# Plot the autocorrelation function
acf(residuals_gls, main = "Autocorrelation of Residuals") #want bars of Lag1 and higher to be below the blue-dotted lines of confidence -- ignore Lag0 ** Lag0 represents correlation of residuals with themselves and is always ~1, so this isn't useful and isn't a problem. Lag 1 shows corr between each residual and the prev residual, lag 2 = correlation between residuals 2 steps apart, etc. **Focus on Lag1 and higher* 

#Result: no apparent autocorrelation -- 
#All bars below the confidence interval dotted line = ok

# Compute lag-1 correlation in residuals - values close to 0 = little/no autocorrelation; value near 1 or -1 suggests sig correlation
cor(residuals_gls[-1], residuals_gls[-length(residuals_gls)]) 
#Result: -0.001, suggests no autocorrelation


#model_VWCinun: lag1 bar exceeds dotted blue lines, but corr = 0.19 = close to 0 so little to no autocorrelation
#model_interact: 0.159, ok 
#model_interact2: 0.17, ok 
#model_interact3: 0.095, ok 


```


#Model Residuals and QQ-plot
Going to do this for various models as unsure which one is actually best *
```{r}
#plot model residuals (homogeneity of variance)
plot(model_interact3)

#qqplot to verify normality - this plots the actual model, not the residuals 
qqnorm(model_interact3)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(model_interact3), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(model_interact3))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(model_interact3)), 
             mean = 0, sd = sigma(model_interact3))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}


#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(model_interact3)) #normality of residuals 
hist(residuals(model_interact3)) 
car::qqPlot(x = residuals(model_interact3)) #shows where residuals breach normal distr
qqPlot(residuals(model_interact3))


#Results
#model_VWCinun: not good at all 
#model_interact: not good at all 
#model_interact2: not good 
#model_interact3: not good 



```


#Normality of Residuals 
```{r}
# Extract standardized/normalized residuals
std_resid <- residuals(model_interact3, type = "normalized")

# Graphical assessment
par(mfrow = c(1, 2))
# Histogram of residuals
hist(std_resid, main = "Histogram of Standardized Residuals", 
     xlab = "Standardized Residuals", freq = FALSE)
curve(dnorm, add = TRUE, col = "red")

# QQ plot
qqnorm(std_resid, main = "Normal Q-Q Plot")
qqline(std_resid, col = "red")

# Formal test
shapiro_test <- shapiro.test(std_resid)
print(shapiro_test) 

#Results
#model_VWCinun: p<0.001, NOT supported 
#model_interact: p<0.001, NOT supported 
#model_interact2: p<0.001, NOT supported 
#model_interact3: p<0.001, NOT supported 

```




#Homogeneity of variance 
```{r}
# Plot residuals vs fitted values
par(mfrow = c(1, 1))
plot(fitted(model_interact3), std_resid,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# Check residuals by predictor variables
par(mfrow = c(2, 2))
boxplot(std_resid ~ df_FCH4$landscape_position, 
        main = "Residuals by Landscape Position")
abline(h = 0, lty = 2)

boxplot(std_resid ~ df_FCH4$inundated, 
        main = "Residuals by Inundation")
abline(h = 0, lty = 2)

plot(df_FCH4$soil_temp_10_cm, std_resid,
     xlab = "Soil Temperature (10 cm)", ylab = "Standardized Residuals")
abline(h = 0, lty = 2)

# Residuals vs. Fitted plot
plot(fitted(model_interact3), std_resid, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lty = 2)

# Levene’s Test for Homogeneity of Variance, need p greater than 0.05 
library(car)
leveneTest(std_resid ~ df_FCH4$landscape_position) 


#Results
#model_VWCinun: p=0.41, ok 
#model_interact: p=0.54, ok 
#model_interact2: p=0.51, ok 
#model_interact3: p=0.16, ok 

```
#just seeing what the gls model says in terms of variable significance
```{r}
library(car)
gls_VWC <- gls(flux_value ~ landscape_position + VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "REML")
summary(gls_VWC)
anova(gls_VWC) #landpos p<0.001, SIG*; VWC p = 0.044, *SIG marginal 
Anova(gls_VWC, type = "II") #landpos p <0.001; VWC p = 0.04
# Using the standard anova function with F tests
anova(gls_VWC, type = "marginal", test = TRUE) #landpos p<0.001, VWC p = 0.04

library(lsmeans)
lsmeans(gls_VWC, adjust = "Tukey", pairwise ~ landscape_position) #lowland - upland p<0.001, SIG*

#~~~~~~~~~~~~~~ VWC + inun ~~~~~~~~~~~~~~~~~~~~~~~~~

gls_VWCinun <- gls(flux_value ~ landscape_position + VWC + inundated, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_FCH4, method = "REML")
summary(gls_VWCinun) 
anova(gls_VWCinun) #landpos p<0.001, SIG*; VWC p = 0.038, SIG*, inundated p<0.001 SIG*
Anova(gls_VWCinun, type = "II") #landpos p <0.001; inundated p <0.001, VWC p = 0.06, not sig 
# Using the standard anova function with F tests
anova(gls_VWCinun, type = "marginal", test = TRUE) #landpos p=0.001, VWC p = 0.06, inundated p<0.001


lsmeans(gls_VWCinun, adjust = "Tukey", pairwise ~ landscape_position) #lowland - upland p=0.01, SIG*
```


#Non-parametric approach 

#### Important to note KW test is the non-parametric equivalent of a one-way anova, so it can only test one variable at a time, and can't test continuous/numeric variables. It also can't incorporate a variance structure. Spearman Correlation can be used for continuous variables*
#testing each var independently isn't ideal because it can't capture what might happen when other variables are included...therefore it's better to find a test that can account for multiple variables, like the robust regression in code chunk following this one 
```{r}
# Test landscape position effect on GPP
kruskal.test(flux_value ~ landscape_position, data = df_FCH4) #p=0.00031, *SIG* 

#wilcoxon test
# For comparing two landscape positions (unpaired)
wilcox.test(flux_value ~ landscape_position, 
            data = df_FCH4,
            subset = landscape_position %in% c("upland", "lowland")) #p<0.001, *SIG*

wilcox.test(flux_value ~ landscape_position, 
            data = df_FCH4,
            subset = landscape_position %in% c("upland", "slope"))#p=0.97, not sig 

wilcox.test(flux_value ~ landscape_position, 
            data = df_FCH4,
            subset = landscape_position %in% c("slope", "lowland"))#p=0.027, *SIG*

#just testing inundated since prev models indicate it may be really important 
wilcox.test(flux_value ~ inundated, 
            data = df_FCH4,
            subset = inundated %in% c("Y", "N")) #p<0.001, *SIG*


# For continuous variables, use Spearman correlation
cor.test(df_FCH4$flux_value, df_FCH4$soil_temp_10_cm, method = "spearman") #p=0.08, not sig 
cor.test(df_FCH4$flux_value, df_FCH4$VWC, method = "spearman") #p<0.001, *SIG*
cor.test(df_FCH4$flux_value, df_FCH4$thawdepth, method = "spearman") #p<0.001, *SIG*


```

#Robust Regression - less sensitive to violations of normality but cannot incorporate variance structure**
```{r}
library(MASS)
library(car)
#just testing reduced and full models - cross-validating 

robust_modelVWCinun <- rlm(flux_value ~ landscape_position + VWC + inundated, 
                    data = df_FCH4)
summary(robust_modelVWCinun) 
Anova(robust_modelVWCinun, type = "II") #landpos p = 0.29; VWC and inundated p<0.001, SIG* 



robust_model1 <- rlm(flux_value  ~ landscape_position + VWC, 
                    data = df_FCH4)
summary(robust_model1) 
Anova(robust_model1, type = "II") #landpos p=0.07, VWC <0.001 *SIG*


robust_model2 <- rlm(flux_value  ~ landscape_position + soil_temp_10_cm, 
                    data = df_FCH4)
summary(robust_model2) 
Anova(robust_model2, type = "II") #landpos p<0.001, SIG, and soil temp p=0.03, *SIG*, 


robust_model3 <- rlm(flux_value  ~ landscape_position + thawdepth, 
                    data = df_FCH4)
summary(robust_model3) 
Anova(robust_model3, type = "II") #lanpos p = 0.08; thawdepth <0.001 SIG*



robust_model4 <- rlm(flux_value  ~ landscape_position + VWC + thawdepth, 
                    data = df_FCH4)
summary(robust_model4) 
Anova(robust_model4, type = "II") #landpos p = 0.8; VWC and thawdepth p<0.001 *SIG



robust_model5 <- rlm(flux_value  ~ landscape_position + VWC + soil_temp_10_cm + thawdepth,
                    data = df_FCH4)
summary(robust_model5) 
Anova(robust_model5, type = "II") #landpos p = 0.7; soil temp p = 0.05, VWC and thawdepth p<0.001 *SIG


robust_model6 <- rlm(flux_value  ~ landscape_position + VWC + soil_temp_10_cm + thawdepth + inundated,
                    data = df_FCH4)
summary(robust_model6) 
Anova(robust_model6, type = "II") #landpos p = 0.8; soil temp p = 0.13; VWC, thawdepth, and inundated p <0.001 *SIG 


robust_interact <- rlm(flux_value  ~ landscape_position * VWC,
                    data = df_FCH4)
summary(robust_interact) 
Anova(robust_interact, type = "II") #all p<0.001, SIG*


robust_interact2 <- rlm(flux_value  ~ landscape_position + VWC * inundated,
                    data = df_FCH4)
summary(robust_interact2) 
Anova(robust_interact2, type = "II") #all p <0.001, SIG


robust_interact3 <- rlm(flux_value  ~ landscape_position * VWC * inundated,
                    data = df_FCH4)
summary(robust_interact3) 
Anova(robust_interact3, type = "II") #all p <0.001, SIG*


robust_interact4 <- rlm(flux_value  ~ landscape_position + inundated * VWC + thawdepth,
                    data = df_FCH4)
summary(robust_interact4) 
Anova(robust_interact4, type = "II") #all p <0.001, SIG*


robust_interact5 <- rlm(flux_value  ~ landscape_position + inundated * VWC * thawdepth,
                    data = df_FCH4)
summary(robust_interact5) 
Anova(robust_interact5, type = "II") 
#                          Df         F    Pr(>F)    
# landscape_position        2    0.3203    0.7264  --> not sig   
# inundated                 1 5969.7389 < 2.2e-16 ***
# VWC                       1   61.3760 8.445e-13 ***
# thawdepth                 1   61.8662 7.084e-13 ***
# inundated:VWC             1   26.9201 6.871e-07 ***
# inundated:thawdepth       1   23.1297 3.690e-06 ***
# VWC:thawdepth             1    0.0257    0.8728 --> not sig  
# inundated:VWC:thawdepth   1  382.3708 < 2.2e-16 ***
# Residuals               148 


robust_interact6 <- rlm(flux_value  ~ landscape_position * inundated * VWC + thawdepth,
                    data = df_FCH4)
summary(robust_interact6) 
Anova(robust_interact6, type = "II")#p all sig 

```

#Attempting to test robust regression models - coding help from Claude 
```{r}
# Define a function to calculate AIC for robust regression
rlm_AIC <- function(model) {
  n <- length(model$residuals)
  RSS <- sum(model$residuals^2)
  k <- length(coef(model)) + 1  # +1 for the error variance
  AIC <- n * log(RSS/n) + 2*k
  return(AIC)
}

# Compare models
rlm_AIC(robust_modelVWCinun)
rlm_AIC(robust_model1)
rlm_AIC(robust_model2)
rlm_AIC(robust_model3)
rlm_AIC(robust_model4)
rlm_AIC(robust_model5)
rlm_AIC(robust_model6)
rlm_AIC(robust_interact)
rlm_AIC(robust_interact2)
rlm_AIC(robust_interact3)
rlm_AIC(robust_interact4)
rlm_AIC(robust_interact5)
rlm_AIC(robust_interact6)

#Results:
# [1] 1473.265
# [1] 1499.908
# [1] 1507.356
# [1] 1495.145
# [1] 1488.723
# [1] 1487.508
# [1] 1487.508
# [1] 1475.603
# [1] 1473.41
# [1] 1355.546 --> interact 3 
# [1] 1473.168
# [1] 1478.262
# [1] 1354.918 * --> lowest, interact 6 
```
#IncMSE for soil chambers (?)
```{r}
library(randomForest)
rf_soilCH4_full <- randomForest(flux_value  ~ landscape_position + VWC + soil_temp_10_cm + thawdepth + inundated + air_temp,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_full) #shows thaw depth is most important, followed by inundated, VWC - air temp is least powerful predictor 



rf_soilCH4_fullinteract <- randomForest(flux_value  ~ landscape_position +  soil_temp_10_cm + thawdepth + VWC * inundated + air_temp,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_fullinteract) #shows thaw depth is most important, followed by inundated, VWC - air temp is least powerful predictor 

#removing air temp and inundated 
rf_soilCH4_reduced1 <- randomForest(flux_value  ~ landscape_position +  soil_temp_10_cm + thawdepth + VWC,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_reduced1) #shows thaw depth is most important, followed VWC - air temp is least powerful predictor 


#landpos and swapping out soil temp, thaw depth, VWC, inundated, and air temp   
rf_soilCH4_reduced2 <- randomForest(flux_value  ~ landscape_position + air_temp,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_reduced2) 
# soil temp more important than landpos, but incMSE was around 0.6
#VWC more important than landpos, incMSE was around 8
#thawdepth importance around 20 
#inundatd incMSE around 19
#airtemp incMSE around 3 


#taking out landpos   
rf_soilCH4_reduced3 <- randomForest(flux_value ~ soil_temp_10_cm + thawdepth + VWC + inundated + air_temp,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_reduced3) #thawdepth most important at 10, followed by inundated at 4, soil temp and VWC ~2


#taking out landpos & removing variables from full model    
rf_soilCH4_reduced4 <- randomForest(flux_value ~ thawdepth + VWC,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_reduced4)
#removed soil and air temp, so just thawdepth, VWC, and inundated: thawdepth most important at 12, then inundated at 8, last VWC at 4 
#removed VWC, just thawdepth and inundated: now inundated is most important at 17.5, thawdepth at 14.5 
#thaw depth and VWC: thawdepth is most important at 14, VWC at 8 

#interaction terms 
df_FCH4$VWC_thawdepth_interact <- df_FCH4$VWC * df_FCH4$thawdepth
#The interaction() function is designed for categorical variables and creates a new factor variable representing all possible combinations of levels.e.g., df_FCH4$inundated_VWC <- interaction(df_FCH4$inundated, df_FCH4$VWC)

rf_soilCH4_interact <- randomForest(flux_value ~ thawdepth + VWC + inundated + VWC_thawdepth_interact,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)


# Look at variable importance
varImpPlot(rf_soilCH4_interact) #thawdepth = 10, VWC_thawdepth interact = 8, inundated = 6, VWC = ~5 



rf_soilCH4_robustinteract6 <- randomForest(flux_value ~ landscape_position + inundated + VWC + thawdepth,
                             data = df_FCH4,
                             ntree = 500,
                             importance = TRUE)

varImpPlot(rf_soilCH4_robustinteract6)

```


#Partial regressions, code from Claude 
```{r}
#partial regressions looks at contribution of predictor variable while holding the other variables constant
# The partial regression plots show the relationship between each predictor and the response after removing the linear effects of all other predictors. The slope of the line in each plot corresponds to the regression coefficient for that variable in the full model.

# Method 1: Using car package
library(car)
avPlots(multiple_reg)  # Creates all partial regression plots
# Creating a plot for a specific variable:
avPlot(multiple_reg, "soil_temp_10_cm")

# Method 2: Manual calculation for one variable (e.g., soil_temp_10_cm)
# Step 1: Regress the response on all predictors except the focal one
model_without_soiltemp <- lm(flux_value ~ landscape_position + VWC + thawdepth + inundated, 
                            data = df_FCH4)
resid_y <- residuals(model_without_soiltemp)

# Step 2: Regress the focal predictor on all other predictors
model_soiltemp <- lm(soil_temp_10_cm ~ landscape_position + VWC + thawdepth + inundated, 
                    data = df_FCH4)
resid_x <- residuals(model_soiltemp)

# Step 3: Plot the residuals against each other
plot(resid_x, resid_y, 
     xlab = "soil_temp_10_cm | others", 
     ylab = "flux_value | others",
     main = "Partial Regression Plot for Soil Temperature")
abline(lm(resid_y ~ resid_x), col = "red")


#Results:
#landpos upland and slope both have relatively flat line, meaning there's a minimal relationship with methane flux after accounting for other variables. --> lowland was used as a reference point, so it tests upland vs lowland and slope vs lowland. If I want to see slope vs upland, I can change the reference point like this:

#df_FCH4$landscape_position <- relevel(factor(df_FCH4$landscape_position), ref = "upland")

#VWC shows a slight positive trend, but line also mostly flat, so effect seems minimal
#soil temp line is basically flat, so very minimal explanatory power after other variables are controlled for 
#thawdepth has most clear pos trend, showing it has most explanatory power for methane fluxes - followed by slight pos trend of inundated  


df_FCH4$landscape_position <- relevel(factor(df_FCH4$landscape_position), ref = "upland")
avPlots(multiple_reg) 
```


```{r}
library(ppcor) #Pearson corr - which assumes normality - so consider these carefully and cross-ref with non-parametric tests 
pcor.test(df_FCH4$flux_value, df_FCH4$soil_temp_10_cm, 
         df_FCH4[, c("VWC", "thawdepth")]) # Omitting categorical variables; uses Pearson corr
#partial correlation coeff is 0.18, p value is 0.018 *SIG* - so while relationship with methane fluxes is minimal, it appears significant 

pcor.test(df_FCH4$flux_value, df_FCH4$VWC,
         df_FCH4[, c("soil_temp_10_cm", "thawdepth")])#coeff = 0.19, p = 0.014 *SIG*, so minimal but significant effect of VWC on fluxes 

pcor.test(df_FCH4$flux_value, df_FCH4$thawdepth,
         df_FCH4[, c("soil_temp_10_cm", "VWC")]) #coeff = 0.33, p<0.001, SIG* 
```



#Singular var robust regression 
```{r}
#landpos
robust_landpos <- rlm(flux_value  ~ landscape_position, 
                    data = df_FCH4)
summary(robust_landpos) 
Anova(robust_landpos, type = "II") #landpos p<0.001, SIG*

#VWC
robust_VWC <- rlm(flux_value  ~ VWC, 
                    data = df_FCH4)
summary(robust_VWC) 
Anova(robust_VWC, type = "II") #p<0.001, SIG*


#soil temp 
robust_soiltemp <- rlm(flux_value  ~ soil_temp_10_cm, 
                    data = df_FCH4)
summary(robust_soiltemp) 
Anova(robust_soiltemp, type = "II") #p<0.02, SIG*

#thaw depth
robust_thawdepth <- rlm(flux_value  ~ thawdepth, 
                    data = df_FCH4)
summary(robust_thawdepth) 
Anova(robust_thawdepth, type = "II") #p<0.001, SIG*


#inundated
robust_inundated <- rlm(flux_value  ~ inundated, 
                    data = df_FCH4)
summary(robust_inundated) 
Anova(robust_inundated, type = "II") #p<0.001, SIG*


#air temp
robust_airtemp <- rlm(flux_value  ~ air_temp, 
                    data = df_FCH4)
summary(robust_airtemp) 
Anova(robust_airtemp, type = "II") #p<0.7, not sig 


#landpos * inundated
robust_interact<- rlm(flux_value  ~ landscape_position * inundated,  
                    data = df_FCH4)
summary(robust_interact) 
Anova(robust_interact, type = "II") #all p < 0.001, SIG* 


#landpos * VWC
robust_interact2<- rlm(flux_value  ~ landscape_position * VWC,  
                    data = df_FCH4)
summary(robust_interact2) 
Anova(robust_interact2, type = "II") #all p < 0.001, SIG* 

#inun * VWC
robust_interact3<- rlm(flux_value  ~ inundated * VWC,  
                    data = df_FCH4)
summary(robust_interact3) 
Anova(robust_interact3, type = "II") #all p < 0.001, SIG* 


#inun * VWC
robust_interact4<- rlm(flux_value  ~ thawdepth * VWC,  
                    data = df_FCH4)
summary(robust_interact4) 
Anova(robust_interact4, type = "II") #all p < 0.001, SIG* 


#inun * landpos
robust_interact5<- rlm(flux_value  ~ inundated * landscape_position,  
                    data = df_FCH4)
summary(robust_interact5) 
Anova(robust_interact5, type = "II") #all p < 0.001, SIG* 

```


#Best models (??? seemingly?)
```{r}
robust_model1 <- rlm(flux_value  ~ landscape_position + VWC, 
                    data = df_FCH4)
summary(robust_model1) 
Anova(robust_model1, type = "II") 
#                     Df      F    Pr(>F)    
# landscape_position   2  2.665   0.07281 .  
# VWC                  1 46.617 1.872e-10 ***
# Residuals          154 


robust_model3 <- rlm(flux_value  ~ landscape_position + thawdepth, 
                    data = df_FCH4)
summary(robust_model3) 
Anova(robust_model3, type = "II") 
#                     Df       F    Pr(>F)    
# landscape_position   2  2.5313   0.08286 .  
# thawdepth            1 60.7760 8.933e-13 ***
# Residuals          154 

#second-lowest AIC 
robust_interact3 <- rlm(flux_value  ~ landscape_position * VWC * inundated,
                    data = df_FCH4)
summary(robust_interact3) 
Anova(robust_interact3, type = "II") 
# Response: flux_value
#                                   Df        F    Pr(>F)    
# landscape_position                 2  2432.54 < 2.2e-16 ***
# VWC                                1   525.68 < 2.2e-16 ***
# inundated                          1 19333.36 < 2.2e-16 ***
# landscape_position:VWC             2   384.56 < 2.2e-16 ***
# landscape_position:inundated       2 12279.43 < 2.2e-16 ***
# VWC:inundated                      1  2149.55 < 2.2e-16 ***
# landscape_position:VWC:inundated   2 12518.89 < 2.2e-16 ***
# Residuals                        146  


#*lowest AIC 
robust_interact6 <- rlm(flux_value  ~ landscape_position * inundated * VWC + thawdepth,
                    data = df_FCH4)
summary(robust_interact6) 
Anova(robust_interact6, type = "II")
# Response: flux_value
#                                   Df        F    Pr(>F)    
# landscape_position                 2 1345.564 < 2.2e-16 ***
# inundated                          1 7977.205 < 2.2e-16 ***
# VWC                                1   95.134 < 2.2e-16 ***
# thawdepth                          1    7.972  0.005419 ** 
# landscape_position:inundated       2 5906.508 < 2.2e-16 ***
# landscape_position:VWC             2  225.144 < 2.2e-16 ***
# inundated:VWC                      1 1030.584 < 2.2e-16 ***
# landscape_position:inundated:VWC   2 5982.930 < 2.2e-16 ***
# Residuals                        145 



robust_m4 <- rlm(flux_value  ~ landscape_position + thawdepth + inundated + VWC, 
                    data = df_FCH4)
summary(robust_m4) 
Anova(robust_m4, type = "II") 
#                     Df         F    Pr(>F)    
# landscape_position   2    0.2249 0.7988563    
# thawdepth            1   15.7315 0.0001121 ***
# inundated            1 3456.9916 < 2.2e-16 ***
# VWC                  1   12.8491 0.0004543 ***
# Residuals          152 


robust_m5 <- rlm(flux_value  ~ landscape_position + thawdepth + inundated + VWC + soil_temp_10_cm, 
                    data = df_FCH4)
summary(robust_m5) 
Anova(robust_m5, type = "II") 

#                     Df         F    Pr(>F)    
# landscape_position   2    0.2127 0.8086155    
# thawdepth            1   14.9436 0.0001641 ***
# inundated            1 3195.8791 < 2.2e-16 ***
# VWC                  1   12.7624 0.0004750 ***
# soil_temp_10_cm      1    2.2654 0.1343804  
```




#Temp relationships 
```{r}

#FCH4 and soil temp 
ggplot(df_NEE_RECO2_GPP, aes(x=soil_temp_10_cm, y = FCH4*1000)) + geom_point()

FCH4.temp <- lm(FCH4 ~ soil_temp_10_cm, data = df_NEE_RECO2_GPP)
summary(FCH4.temp) #R2 = 0.05, p = 0.002
#extract R2 and p value for slope 
r2 <- summary(FCH4.temp)$r.squared
p_value <- summary(FCH4.temp)$coefficients[2, 4]
#plot 
ggplot(df_NEE_RECO2_GPP, aes(x = soil_temp_10_cm, y = FCH4)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add regression line
  annotate("text", x = 0, 
           y = 0, 
           label = paste("R² = ", round(r2, 2), "\n", 
                         "p = ", signif(p_value, 3)), 
           hjust = 0, size = 5) +
  labs(x = "Soil Temperature (10 cm)", y = "FCH4",
       title = "Relationship between FCH4 and Soil Temperature") +
  theme_minimal()


```

#AGU figs
```{r}
#looking at FCH4 (in g/m2/s) by landscape position and plot type, facet wrapped --> revert to umolCO2 and nmolCH4*
ggplot(df_long,
       aes(x=landscape_position, y = (FCH4 *1/12 * 1000000), fill = plot_type))+
  geom_boxplot()+
  labs( title = "Methane Fluxes within Plot Type vs Landscape Position ",
        x = "Landscape Position",
          #y = expression(Methane~Flux~(gCH[4]-C/m^2/s)),
       # y = expression(FCH[4]~"(" * mu * "mol" ~ CH[4]-C~/~m^-2~/~s)~")",
        y = expression(FCH[4]~(CH[4]-C~μmol~m^{-2}~s^{-1})),
          fill = "Plot Type")+
  geom_hline(yintercept=0)+
  #geom_hline(yintercept = 0.007)+
  #theme_minimal() +
  theme_bw()+
 scale_y_continuous(limits = c(-0.15,0.15), #cut out a few bit outliers here to show trend better, 0.25 looks good but still mutes trends 
  breaks = seq(-0.15, 0.15, 0.1))+
                      #labels = function(x) x / 1000)+# Scale y-axis labels back since FCH4 was scaled by 1000
    theme(
   axis.title.x = element_text(size = 14),  
    axis.text.x = element_text(size = 12, face = "bold"), 
   axis.title.y = element_text(size = 14),
   axis.text.y = element_text(size = 12, face = "bold")) +
   #facet_wrap(~plot_type, labeller = "label_both")
  facet_wrap(vars(plot_type), scales = "free_x")

ggplot(df_long,
       aes(x=landscape_position, y = (FCH4 *1/12 * 1000000), fill = plot_type))+
  geom_boxplot()+
  labs( title = "Methane Fluxes vs Landscape Position ",
        x = "Landscape Position",
          #y = expression(Methane~Flux~(gCH[4]-C/m^2/s)),
        #y = expression(FCH[4]~"(" * mu * "mol" ~ CH[4]-C~m^-2~s^-1),
         y = expression(FCH[4]~(CH[4]-C~μmol~m^{-2}~s^{-1})),
          fill = "Plot Type")+
  geom_hline(yintercept=0)+
  #geom_hline(yintercept = 0.007)+
  #theme_minimal() +
  theme_bw()+
 scale_y_continuous(limits = c(-0.15,0.15), #cut out a few bit outliers here to show trend better, 0.25 looks good but still mutes trends 
  breaks = seq(-0.15, 0.15, 0.1))+
                      #labels = function(x) x / 1000)+# Scale y-axis labels back since FCH4 was scaled by 1000
    theme(
   axis.title.x = element_text(size = 14),  
    axis.text.x = element_text(size = 12, face = "bold"), 
   axis.title.y = element_text(size = 14),
   axis.text.y = element_text(size = 12, face = "bold"))# +
   #facet_wrap(~plot_type, labeller = "label_both")
  facet_wrap(vars(plot_type), scales = "free_x")


# all plot type - FCH4 vs landscape position & inundation 
ggplot(df_long, aes(x = landscape_position, y = (FCH4 *1/12 * 1000000), fill = inundated)) +
  geom_boxplot() +
  labs(title = "Methane Fluxes within Plot Type vs Landscape Position and Inundation",
       x = "Landscape Position",
         #y = expression(Methane~Flux~(gCH[4]-C/m^2/s)),
        #y = expression(FCH[4]~"(" * mu * "mol" ~ CH[4]-C~m^-2~s^-1),
        y = expression(FCH[4]~(CH[4]-C~μmol~m^{-2}~s^{-1})),
       fill = "Inundated")+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-0.15,0.46), #cut out a few bit outliers here to show trend better, -0.25 and 0.5 looks good but trying others 
  breaks = seq(-0.15, 0.46, 0.15))+
                     #labels = function(x) x / 1000)+# Scale y-axis labels back since FCH4 was scaled by 1000
  theme_bw()+ 
  theme(
   axis.title.x = element_text(size = 14),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 12, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 14),
   axis.text.y = element_text(size = 12, face = "bold"))+
  facet_wrap(vars(plot_type), scales = "free_x")

# all plot type - FCH4 vs landscape position & inundation 
landscape_CH4_fluxes <- ggplot(df_long, aes(x = landscape_position, y = (FCH4 *1/12 * 1000000), fill = inundated)) +
  geom_boxplot() +
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Landscape Position",
         #y = expression(Methane~Flux~(gCH[4]-C/m^2/s)),
       #y = expression(FCH[4]~"(" * mu * "mol" ~ CH[4]-C~m^-2~s^-1),
       y = expression(FCH[4]~(CH[4]-C~μmol~m^{-2}~s^{-1})),
       fill = "Inundated")+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-0.15,0.46), #cut out a few bit outliers here to show trend better, -0.25 and 0.5 looks good but trying others 
  breaks = seq(-0.15, 0.46, 0.15))+
                     #labels = function(x) x / 1000)+# Scale y-axis labels back since FCH4 was scaled by 1000
  theme_bw()+ 
  theme(
   axis.title.x = element_text(size = 14),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 12, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 14),
   axis.text.y = element_text(size = 12, face = "bold"))#+
  facet_wrap(vars(plot_type), scales = "free_x")
  
  # ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/seasonal_budgets_CO2.png", seasonal_budgets_CO2, 
  #      width = 10, height = 7, dpi = 600)
```

#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes.png",  # File name and extension
  plot = landscape_CH4_fluxes,        # The plot to save (default is the last plot created)
  width = 15,                 # Width in inches
  height = 10,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```

