---
title: 'Council_Soil_Chamber_Landscape.GPP' 
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_GPP_RECO2 and df_GPP_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify GPP and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (GPP - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format, choose variables of interest 
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4,NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm, thawdepth) %>%
  pivot_longer(cols = c(GPP, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (NEE, GPP, RECO)

####CFilter for GPP and make dataset for just GPP fluxes 
```{r}
#Filter & make separate datasets for each plot_ID = "EC" "MW" and "BGC", and by flux type (GPP, GPP, RECO)

#GPP
df_GPP <-df_long %>%
  filter(flux_type == "GPP")

```

#make sure dataset has complete cases for variables of interest 
```{r}
#make sure to use complete cases of all variables of interest 
df_GPP <- df_GPP %>%
   filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC, air_temp))


sum(is.na(df_GPP$VWC)) 
sum(is.na(df_GPP$thawdepth)) 
sum(is.na(df_GPP$soil_temp_10_cm)) 
sum(is.na(df_GPP$inundated)) 
sum(is.na(df_GPP$air_temp))

```

# Categorical Predictors: Variance and normality of categorical predictor variables 
```{r}
#Flux and landscape pos
df_variance_landpos <- df_GPP %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland 


#Flux and inundated 
df_variance_inundated <- df_GPP %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)

# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in NOT* inundated plots 


# Shapiro-Wilk test for normality across entire dataset
shapiro_test <- shapiro.test(df_GPP$flux_value)
print(shapiro_test) #p<0.001, not normal 


# Test normality within each landscape position
df_GPP %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: lowland & upland = no; slope = yes


# Test normality based on inundated
df_GPP %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: N = no; Y = yes 


# Q-Q plot of full dataset 
qqnorm(df_GPP$flux_value)
qqline(df_GPP$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_GPP, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of GPP by Landscape Position")


# histograms with density curves by inundated
ggplot(df_GPP, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of GPP by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for categorical factors (see code chunk below for testing continuous/numerical factors)
leveneTest(flux_value ~ landscape_position, data = df_GPP) #p=0.83, homogeneity of var is ok
leveneTest(flux_value ~ inundated, data = df_GPP)#p = 0.54, ok


#for interactions
leveneTest(flux_value ~ landscape_position * inundated, data = df_GPP) #p=0.7, ok

```

#Continuous predictors: Variance and normality of continuous predictor variables 

###Testing to see which kind of correlation to use: Pearson for parametric, Spearman for non-parametric
```{r}
#test distribution normality of each numeric predictor variable (continuous factors)
shapiro.test(df_GPP$soil_temp_10_cm) #p=0.004, NOT normal** 
shapiro.test(df_GPP$thawdepth) #p = <0.001, NOT normal **
shapiro.test(df_GPP$VWC) #p<0.001, NOT normal **

#visualize pairs to see if it looks linear
pairs(df_GPP[c("soil_temp_10_cm", "thawdepth", "VWC")]) #NOPE 

plot(df_GPP$soil_temp_10_cm, df_GPP$thawdepth)

plot(df_GPP$soil_temp_10_cm, df_GPP$VWC)

plot(df_GPP$thawdepth, df_GPP$VWC)

ggplot(df_GPP, aes(x = soil_temp_10_cm, y = thawdepth))+
  geom_point()+
    labs(
     x = "Soil Temperature (10 cm)", 
     y = "Thaw Depth")+
geom_abline(intercept = 0, slope = 7, color = "red") # 1:1 line in red, adj slope to account for diffs in scale


#So this means for correlations and looking at collinearity, I need to use non-parametric tests 


#Check homogeneity of variance: should see even spread of of points for homogeneity of variance 

# Fit a simple model
simple_model <- lm(flux_value ~ VWC, data = df_GPP)

# Extract fitted values and residuals
fitted_values <- fitted(simple_model)
residuals <- residuals(simple_model)

# Plot residuals vs. predictor
plot(df_GPP$VWC, residuals, 
     xlab = "Cont. Pred. Var", ylab = "Residuals",
     main = "Residuals vs. Predictor Var")
abline(h = 0, lty = 2)

# Create a scale-location plot --> should be randomly scattered with a relatively flat trend line (red) for homogeneity of variance 
plot(simple_model, which = 3)


#Breusch-Pagan test - specifically tests if variance of residuals is constant - want a p > 0.05 for homogeneity of var
library(lmtest)
bptest(simple_model)

#thawdepth = p=0.49, shows there is homogeneity of var 
#soil temp = p = 0.7, suggests homogeneity of var 
#VWC = p = 0.22,  suggests homogeneity of var 

```
#Just looking at variables 
```{r}
ggplot(df_GPP, aes(x = soil_temp_10_cm, y = thawdepth)) +
  geom_point() +
  labs(
    x = "Soil Temperature (10 cm)", 
    y = "Thaw Depth (cm)", 
    title = "Scatterplot of Soil Temperature vs. Thaw Depth") +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(intercept = 0, slope = 7, color = "blue", linetype = "dashed")

```


#Checking colinearity / correlations to determine which variables to use for models 

#Non-parametric tests of correlations - spearman's correlation - because predictor variables are not* normally distributed

#Corr of continuous variables 
```{r}
#Correlation between soil temp, thaw depth, and VWC

cor.test(df_GPP$soil_temp_10_cm, df_GPP$thawdepth, method="spearman")#p=0.02, *SIG*, correlated
cor.test(df_GPP$soil_temp_10_cm, df_GPP$VWC, method="spearman") #p<0.02, *Sig*, correlated, one should be removed 
cor.test(df_GPP$thawdepth, df_GPP$VWC, method="spearman") #p<0.001, *sig*, correlated, one should be removed 

#Shows thawdepth and soil temp are correlated, thawdepth and VWC are correlated, and VWC & soil temp are correlated 
```

#Checking correlation in categorical variables 
```{r}
# Check for separation in categorical variables
table(df_GPP$landscape_position, df_GPP$inundated) 


# Create contingency table to examine correlation quantitatively / confirm the separation test above 
cont_table <- table(df_GPP$landscape_position, df_GPP$inundated)
print(cont_table)

# Test for association
chisq.test(cont_table) #p<0.001--> correlated
# Or for small sample sizes
fisher.test(cont_table) #p=0.001, correlated 

#landpos and inun are correlated, will remove inun 

```

#Determining model for landscape GPP


#Step 1: Determine which of the correlated variables to keep
```{r}
# Create competing models with one variable at a time - use "ML" to compare models, use "REML" for fitting final model 

#thawdepth and VWC are correlated, and VWC & soil temp are correlated 
model_temp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 data = df_GPP, method = "ML")

model_thaw <- gls(flux_value ~ landscape_position + thawdepth, 
                 data = df_GPP, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 data = df_GPP, method = "ML")

model_soiltempthawdepth1 <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm, 
                 data = df_GPP, method = "ML")

# Compare with AIC/BIC - lowest AIC and lowest BIC = better model 
AIC(model_temp, model_thaw, model_VWC, model_soiltempthawdepth1) #model_VWC lowest AIC/BIC

BIC(model_temp, model_thaw, model_VWC, model_soiltempthawdepth1)#model_VWC lowest AIC/BIC


#model_VWC has marginally lowest AIC/BIC 

```

#Step 2: Now, test whether adding soil temp or thaw depth improves model 
```{r}
model_VWC <- gls(flux_value ~ landscape_position + VWC,
                        data = df_GPP, method = "ML")

model_VWCsoiltemp <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm,
                        data = df_GPP, method = "ML")

model_VWCthawdepth <- gls(flux_value ~ landscape_position + VWC + thawdepth,
                        data = df_GPP, method = "ML")

anova(model_VWC, model_VWCsoiltemp, model_VWCthawdepth)
anova(model_VWC, model_VWCsoiltemp) #p=0.08, adding temp to VWC does not improve model 
anova(model_VWC, model_VWCthawdepth) #p=0.5, adding temp to VWC does not improve model 
anova(model_VWCsoiltemp, model_VWCthawdepth) #VWCsoiltemp lower AIc/BIC, no p 
anova(model_VWC, model_VWCsoiltemp)#p=0.08, no sig diff 

#Appears neither thaw depth or soil temp improve model, VWC model is best



```

#Step 3: Test if random effect of plot_ID or plot_type is appropriate
```{r}
#Use best predictor variable from Step 1, which was VWC
#use 'ML' for comparing models 

library(nlme)
#gls - no random effect
model_fixed <- gls(flux_value ~ landscape_position + VWC, 
                  data = df_GPP, method = "ML")

#lme - with random effect of plot_ID
model_random <- lme(flux_value ~ landscape_position + VWC, 
                   random = ~1|plot_ID, 
                   data = df_GPP, method = "ML")

#lme - with random effect of plot_type
model_random2 <- lme(flux_value ~ landscape_position + VWC, 
                   random = ~1|plot_type, 
                   data = df_GPP, method = "ML")

# Compare
anova(model_fixed, model_random) #p=0.9, random effect of plot ID does not improve model 
anova(model_fixed, model_random2) #p=0.99, random effect of plot type does not improve model 

#suggests having a random effect does not improve model, so exclude random effect 
```

#Step 4: Test for variance structure 
```{r}
# test if adding a variance structure is appropriate 

#landscape position as a variance structure 
#no var structure 
model_homoscedastic <- gls(flux_value ~ landscape_position + VWC, 
                          data = df_GPP, method = "ML")
#with var structure 
model_heteroscedastic <- gls(flux_value ~ landscape_position + VWC, 
                            weights = varIdent(form = ~1|landscape_position),
                            data = df_GPP, method = "ML")

# Compare
anova(model_homoscedastic, model_heteroscedastic) #p=<0.015 SIG*; var structure of landpos improves model 

#~~~~~~~~~~~~~~~~~~~~testing whether plot_type var structure is appropriate~~~~~~~~~~~~~~~~~~~~~
model_heteroscedastic2 <- gls(flux_value ~ landscape_position + VWC, 
                            weights = varIdent(form = ~1|plot_type),
                            data = df_GPP, method = "ML")
anova(model_heteroscedastic2, type = "marginal", test = TRUE)#landpos p = 0.05, marginally sig*

# Compare
anova(model_homoscedastic, model_heteroscedastic2) #p<0.0049, model improves with plot type var structure 
anova(model_heteroscedastic, model_heteroscedastic2)#no p, appears using plot type var structure may be better fit compared to using landscape position as var structure, lower AIC/BIC, higher logLik


#~~~~~~~~~~looking at multiple var structure, landscape position & plot type~~~~~~~~~~~~~~~~~~ 

model_heteroscedastic3 <- gls(flux_value ~ landscape_position + VWC, 
                            weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                            data = df_GPP, method = "ML")
anova(model_heteroscedastic3, type = "marginal", test = TRUE)#landpos p=0.02, SIG*
                            
                         

anova(model_homoscedastic, model_heteroscedastic3) # hetero3 p<0.001, hetero3 better model 
anova(model_heteroscedastic2, model_heteroscedastic3)#p<0.001, hetero3 better model 

#seems having both landpos and plot type as var structure significantly improves the model 
```

#Step 5: Try adding any remaining non-correlated variables back in to see if it improves the model (use reduction and addition / reduced models to test if various un-correlated predictors improve the model or not, and favor the simpler model)

####using single var structure for reference, see multiple var structure in code chunk below 
```{r}
# # test variable soil temp (but this was correlated with VWC, just running to check if var structure changes this)
# model_1var <- gls(flux_value ~ landscape_position + VWC, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# model_2var <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# # Compare
# anova(model_1var, model_2var) #p=0.048, adding soil temp maay marginally improve model - will have to cross validate 
# 
# #try variable thawdepth 
# model_1var <- gls(flux_value ~ landscape_position + VWC, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# model_2var <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# # Compare
# anova(model_1var, model_2var) #p=0.5, does not improve model 
# 
# #swap VWC for soil temp to check 
# model_1var2 <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# model_2var2 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# # Compare
# anova(model_1var2, model_2var2) #p=0.2, no sig improvement 
# anova(model_1var, model_1var2) #no p but soil temp model (1var2) may be better 
# 
# 
# model_3var <- gls(flux_value ~ landscape_position + VWC + thawdepth + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# anova(model_1var, model_3var) #p-0.6, lower AIC/BIC in 1var 
# 
# 
# #test interaction between land post and VWC
# model_1var <- gls(flux_value ~ landscape_position + VWC, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# 
# model_interact <- gls(flux_value ~ landscape_position * VWC, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# model_interact2 <- gls(flux_value ~ landscape_position * soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# 
# 
# # Compare
# anova(model_1var, model_interact) #p=0.3, model1var has slightly lower AIC/BIC, suggests interaction effect is not important to model 
# anova(model_1var2, model_interact2)#p=0.7, model without interaction better, not sig 


```
#multiple var structure 
```{r}
# test variable soil temp (but this was correlated with VWC, just running to check if var structure changes this)
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1var, model_2var) #p=0.048, adding soil temp may marginally improve model - will have to cross validate 

#try variable thawdepth 
model_1var <- gls(flux_value ~ landscape_position + VWC, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1var, model_2var) #p=0.5, does not improve model 

#swap VWC for soil temp to check 
model_1var2 <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var2 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1var2, model_2var2) #p=0.2, no sig improvement 
anova(model_1var, model_1var2) #no p but soil temp model (1var2) may be better 


model_3var <- gls(flux_value ~ landscape_position + VWC + thawdepth + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

anova(model_1var, model_3var) #p=0.6, no sig improvement, lower AIC/BIC in 1var 


#test interaction between land post and VWC
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")


model_interact <- gls(flux_value ~ landscape_position * VWC, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_interact2 <- gls(flux_value ~ landscape_position * soil_temp_10_cm, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")


# Compare
anova(model_1var, model_interact) #p=0.3, model1var has slightly lower AIC/BIC, suggests interaction effect is not important to model 
anova(model_1var2, model_interact2)#p=0.7, model without interaction better, not sig 


```

#*multiple var structure copied over 
```{r}
#combined var structure as suggested may improve the model 
# test variable soil temp (but this was correlated with VWC, just running to check if var structure changes this)
model_1varVWC <- gls(flux_value ~ landscape_position + VWC, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1varVWC, model_2var) #p=0.08, no improvement on model adding soil temp to VWC

#try variable thawdepth (but this was correlated with VWC)
model_2var <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1varVWC, model_2var) #p=0.1, thawdepth doesn't improve model 

#swap VWC for soil temp to check 
model_1varSoiltemp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + soil_temp_10_cm + VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_2var2 <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

# Compare
anova(model_1varSoiltemp, model_2var) #p=0.3, no sig improvement from adding VWC to soil temp  
anova(model_1varSoiltemp, model_2var2)#p=0.24, no improvement from adding thaw depth to soil temp 


model_3var <- gls(flux_value ~ landscape_position + VWC + thawdepth + soil_temp_10_cm, 
               weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

anova(model_1varVWC, model_3var) #p=0.055, adding thaw depth and soil temp may marginally improve model 
anova(model_1varSoiltemp, model_3var) #p=0.14, adding VWC and thaw depth does not improve model 


#test interaction between landpod and VWC, landpos and soil temp 
model_interactVWC <- gls(flux_value ~ landscape_position * VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_interactVWC2 <- gls(flux_value ~ landscape_position * VWC + thawdepth, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_interactSoiltemp <- gls(flux_value ~ landscape_position * soil_temp_10_cm, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_interactSoiltemp2 <- gls(flux_value ~ landscape_position * soil_temp_10_cm + thawdepth, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")


# Compare
anova(model_1varVWC, model_interactVWC) #p=0.6, no improvement 
anova(model_1varSoiltemp, model_interactSoiltemp) #p=0.9, no improvement 
anova(model_interactSoiltemp2, model_1varSoiltemp)#p=0.6, no improvement from adding thawdepth
anova(model_interactVWC2, model_1varVWC)#p=0.49, no improvement from adding thawdepth 


#comparing the leading models 
m_soiltempVWC<- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

m_soiltemp<- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

m_VWC<- gls(flux_value ~ landscape_position + VWC, 
                weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

 anova(m_soiltemp, m_VWC)#no p, but soiltemp marginally better
 anova(m_soiltemp, m_soiltempVWC) #p=0.29, no major improvement for adding VWC to soil temp 

```

#Step 6: Double check best correlated var was chosen now that variance structure was added 
```{r}
#test soil temp and thawdepth, and with and without VWC

model_soiltemp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                       weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

model_thawdepth <- gls(flux_value ~ landscape_position + thawdepth, 
                 weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "ML")

anova(model_VWC, model_soiltemp) #soiltemp marginally better, no p 
anova(model_VWC, model_thawdepth) #VWC slightly better, no p 
anova(model_soiltemp, model_thawdepth) #soiltemp better, no p

plot(df_GPP$soil_temp_10_cm, df_GPP$flux_value) #all show a very similar trend 
```


#Final model, refit with REML (trying with VWC and soil temp to cross ref )
```{r}
library(car)
GPP.final <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "REML")
summary(GPP.final) #shows no var are sig 
anova(GPP.final) #landpos p = 0.026 *SIG
Anova(GPP.final, type = "II") #landpos p = 0.03 SIG*
# Using the standard anova function with F tests
anova(GPP.final, type = "marginal", test = TRUE) #landpos p=0.02, soil marginal at p=0.06

library(lsmeans)
lsmeans(GPP.final, adjust = "Tukey", pairwise ~ landscape_position) #slope - upland marginally sig at p=0.051 


GPP.final2 <- gls(flux_value ~ landscape_position + VWC, 
                  weights = varComb(varIdent(form = ~1|plot_type),varIdent(form = ~ 1 |                                             landscape_position)),
                 data = df_GPP, method = "REML")
summary(GPP.final2) #landpos p = 0.02 *SIG
anova(GPP.final2) #landpos p = 0.039*
Anova(GPP.final2, type = "II") #none sig 
# Using the standard anova function with F tests
anova(GPP.final2, type = "marginal", test = TRUE) #landpos = 0.022, SIG*

lsmeans(GPP.final2, adjust = "Tukey", pairwise ~ landscape_position) #slope - upland marginally sig at p=0.051 
#lowland-upland marginally sig at p=0.05, slope - upland marginally sig at p=0.06

boxplot(flux_value ~ landscape_position, data = df_GPP)

anova(GPP.final, GPP.final2)#no p, but GPP.final better 

```


#Some extra exploratory validation tests (based on single var structure, keeping for reference but ignore for now)
```{r}
# m1 <- gls(flux_value ~ landscape_position + VWC, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# summary(m1) #shows no var are sig 
# Anova(m1) #none sig 
# Anova(m1, type = "II", test.statistic = "F") #none sig 
# # Using the standard anova function with F tests
# anova(m1, type = "marginal", test = TRUE) #none sig 
# 
# 
# m1.1 <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# summary(m1.1) #shows no var are sig 
# Anova(m1.1) #none sig 
# Anova(m1.1, type = "II", test.statistic = "F") #none sig 
# # Using the standard anova function with F tests
# anova(m1.1, type = "marginal", test = TRUE) #none sig 
# 
# 
# 
# m2 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# summary(m2) #shows no var are sig 
# Anova(m2) #none sig 
# Anova(m2, type = "II", test.statistic = "F") #none sig 
# # Using the standard anova function with F tests
# anova(m2, type = "marginal", test = TRUE) #none sig 
# 
# 
# 
# 
# m3 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm + thawdepth, 
#                  weights = varIdent(form = ~1|landscape_position),
#                  data = df_GPP, method = "ML")
# summary(m3) #shows no var are sig 
# Anova(m3) #none sig 
# Anova(m3, type = "II", test.statistic = "F") #none sig 
# # Using the standard anova function with F tests
# anova(m3, type = "marginal", test = TRUE) #none sig 
# 
# anova(m1, m1.1) #m1.1 with soil temp better, no p 
# anova(m1, m2) #p=0.048, m1 marginally better, adding soil temp may improve model
# anova(m1.1,m2)#p=0.22, adding VWC to soil temp does not improve model 
# anova(m1, m3)#p=0.09
# anova(m1.1, m3)#p=0.3
# 
# #soil temp model may be the best option for final model 
```




#Step 7: Test for multicollinearity in model 

####For testing this in models, use VIF test 
```{r}
#variance inflation factor (vif) - vif = 1 = variables are not correlated; between 1 & 5 = moderately correlated,  5 or higher indicates highly correlated & multicollinearity among variables. GVIF is a generalized VIF for categorical variables with more than two levels; the adj VIF gives values for comparing across variables with diff degrees of freedom* --> look at both when assessing collinearity in a model 
library(car)
vif(GPP.final) # all <2, so should be ok! 
```

#Plot model residuals and QQplot 
```{r}
#plot model residuals (homogeneity of variance)
plot(GPP.final)

#qqplot to verify normality - this plots the actual model, not the residuals 
qqnorm(GPP.final)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(GPP.final), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(GPP.final))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(GPP.final)), 
             mean = 0, sd = sigma(GPP.final))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}


#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(GPP.final)) #normality of residuals 
hist(residuals(GPP.final)) 
car::qqPlot(x = residuals(GPP.final)) #shows where residuals breach normal distr
qqPlot(residuals(GPP.final))

```
#Normality of Residuals 
```{r}
# Extract standardized/normalized residuals
std_resid <- residuals(GPP.final, type = "normalized")

# Graphical assessment
par(mfrow = c(1, 2))
# Histogram of residuals
hist(std_resid, main = "Histogram of Standardized Residuals", 
     xlab = "Standardized Residuals", freq = FALSE)
curve(dnorm, add = TRUE, col = "red")

# QQ plot
qqnorm(std_resid, main = "Normal Q-Q Plot")
qqline(std_resid, col = "red")

# Formal test
shapiro_test <- shapiro.test(std_resid)
print(shapiro_test) #for gls final model: p<0.001; normality NOT supported 

```
#Test autocorrelation
```{r}
# Extract normalized residuals from GLS model
residuals_gls <- residuals(GPP.final, type = "normalized")

# Plot the autocorrelation function
acf(residuals_gls, main = "Autocorrelation of Residuals") #want bars of Lag1 and higher to be below the blue-dotted lines of confidence -- ignore Lag0 ** Lag0 represents correlation of residuals with themselves and is always ~1, so this isn't useful and isn't a problem. Lag 1 shows corr between each residual and the prev residual, lag 2 = correlation between residuals 2 steps apart, etc. **Focus on Lag1 and higher* 

#Result: no apparent autocorrelation -- 
#All bars below the confidence interval dotted line = ok

# Compute lag-1 correlation in residuals - values close to 0 = little/no autocorrelation; value near 1 or -1 suggests sig correlation
cor(residuals_gls[-1], residuals_gls[-length(residuals_gls)]) 
#Result: -0.001, suggests no autocorrelation


```
#Homogeneity of variance 
```{r}
# Plot residuals vs fitted values
par(mfrow = c(1, 1))
plot(fitted(GPP.final), std_resid,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# Check residuals by predictor variables
par(mfrow = c(2, 2))
boxplot(std_resid ~ df_GPP$landscape_position, 
        main = "Residuals by Landscape Position")
abline(h = 0, lty = 2)

boxplot(std_resid ~ df_GPP$inundated, 
        main = "Residuals by Inundation")
abline(h = 0, lty = 2)

plot(df_GPP$soil_temp_10_cm, std_resid,
     xlab = "Soil Temperature (10 cm)", ylab = "Standardized Residuals")
abline(h = 0, lty = 2)

# Residuals vs. Fitted plot
plot(fitted(GPP.final), std_resid, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lty = 2)

# Levene’s Test for Homogeneity of Variance
library(car)
leveneTest(std_resid ~ df_GPP$landscape_position) #p=0.47; this says it's ok 

```
#Non-parametric approach 

#### Important to note KW test is the non-parametric equivalent of a one-way anova, so it can only test one variable at a time, and can't test continuous/numeric variables. It also can't incorporate a variance structure. Spearman Correlation can be used for continuous variables*
#testing each var independently isn't ideal because it can't capture what might happen when other variables are included...therefore it's better to find a test that can account for multiple variables, like the robust regression in code chunk following this one 
```{r}
# Test landscape position effect on GPP
kruskal.test(flux_value ~ landscape_position, data = df_GPP) #p=0.18, not sig 

#wilcoxon test
# For comparing two landscape positions (unpaired)
wilcox.test(flux_value ~ landscape_position, 
            data = df_GPP,
            subset = landscape_position %in% c("upland", "lowland")) #p=0.1, not sig 

wilcox.test(flux_value ~ landscape_position, 
            data = df_GPP,
            subset = landscape_position %in% c("upland", "slope"))#p=0.2, not sig 

wilcox.test(flux_value ~ landscape_position, 
            data = df_GPP,
            subset = landscape_position %in% c("slope", "lowland"))#p=0.89, not sig 


# For continuous variables, use Spearman correlation
cor.test(df_GPP$flux_value, df_GPP$soil_temp_10_cm, method = "spearman") #p=0.011, *SIG*
cor.test(df_GPP$flux_value, df_GPP$VWC, method = "spearman") #p=0.17, not sig 
cor.test(df_GPP$flux_value, df_GPP$thawdepth, method = "spearman") #p=0.09, not sig 


```
#Robust Regression - less sensitive to violations of normality but cannot incorporate variance structure**
```{r}
library(MASS)
library(car)
#just testing reduced and full models to cross-ref and cross-validate 
robust_model1 <- rlm(flux_value ~ landscape_position + soil_temp_10_cm , 
                    data = df_GPP)
summary(robust_model1) 
Anova(robust_model1, type = "II") #soil temp sig, p = 0.008


robust_model1.1 <- rlm(flux_value ~ landscape_position + VWC, 
                    data = df_GPP)
summary(robust_model1.1) 
Anova(robust_model1.1, type = "II") #VWC p = 0.07, not sig 


robust_model2 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                    data = df_GPP)
summary(robust_model2) 
Anova(robust_model2, type = "II") #soil temp sig, p = 0.01


robust_model3 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm + thawdepth,
                    data = df_GPP)
summary(robust_model3) 
Anova(robust_model3, type = "II") #soil temp sig, p = 0.01; VWC p = 0.037, but they're correlated so we ignore this 


```


#Attempting to test "goodness of fit" of the various robust regression models - coding help from Claude 
```{r}
# Define a function to calculate AIC for robust regression
rlm_AIC <- function(model) {
  n <- length(model$residuals)
  RSS <- sum(model$residuals^2)
  k <- length(coef(model)) + 1  # +1 for the error variance
  AIC <- n * log(RSS/n) + 2*k
  return(AIC)
}

# Compare models
rlm_AIC(robust_model1)
rlm_AIC(robust_model1.1)
rlm_AIC(robust_model2)
rlm_AIC(robust_model3)

#Results:
# [1] -3321.921
# [1] -3321.558
# [1] -3322.725 --> marginally lowest AIC, marginally best, soil temp + VWC model 
# [1] -3321.024
```

#Correlation matrix 

####Flux v Environmental correlation matrix - soil chambers GPP
```{r}
# Environmental relationships
# Create a correlation plot between fluxes and environmental variables

#variables of interest / used in model (except air temp), for reference 
#(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC))

#can only use numeric variables, so "landscape_position and inundated can't be used 
chamber_GPP_corr_matrix <- df_GPP %>%
  select(flux_value, thawdepth, soil_temp_10_cm, VWC, air_temp) %>%
  rename(
    GPP = flux_value,
    Soil_temp_10cm = soil_temp_10_cm,
    Soil_VWC = VWC,
    Air_temp = air_temp,
    Thaw_depth = thawdepth,
  )

cor_matrix2 <- cor(chamber_GPP_corr_matrix, use = "complete.obs")

# Convert correlation matrix to long format for plotting
cor_long2 <- as.data.frame(as.table(cor_matrix2))
names(cor_long2) <- c("Var1", "Var2", "Correlation")

ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,
                        limits = c(-1, 1)) +  # Add this to set the full range to be 1 to -1
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text = element_text(size = 12, face = "bold"),
       plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber GPP Fluxes and Environmental Variables")
   



# option 2: Fewer breaks for a cleaner look, only labels corr values above 0.3 --> I like this one best 
ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation values only for strong correlations
  geom_text(data = subset(cor_long2, abs(Correlation) > 0.2),  # Only show stronger correlations
            aes(label = sprintf("%.2f", Correlation),
                 color = abs(Correlation) < 0.5), # Changes text color based on correlation strength
            size = 3,
            fontface = "bold") + #to make numbers in tiles bold
  scale_fill_gradient2(low = "red", 
                      mid = "white", 
                      high = "blue", 
                      midpoint = 0,
                      limits = c(-1, 1),
                      breaks = c(-1, -0.5, 0, 0.5, 1)) +
    scale_color_manual(values = c("white", "black"), guide = "none") +  # For text visibility
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber GPP Fluxes and Environmental Variables")

 #saves the figure directly above
ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soil_chamber_GPP_corr_matrix.png",
       width = 8, height = 5, dpi = 600)

```

