---
title: 'Council_Soil_Chamber_BGC.NEE' #breaking NEE apart by plot type for stats, adapted from NEE_statsmodels.Rmd 
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')


```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs from inundation 
library(dplyr)
df_NEE_RECO2_GPP<- df_NEE_RECO2_GPP %>%
   filter(!is.na(inundated))


# Reshape the dataframe to long format, choose variables of interest 
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm, thawdepth) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


```


#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
#Filter & make separate datasets for each plot_ID = "EC" "MW" and "BGC", and by flux type (GPP, NEE, RECO)

#BGC - eddy covar tower plot types 
df_BGClong <- df_long %>%
  filter(plot_type == "BGC")


#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")


#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_BGClong <- df_BGClong %>% arrange(flux_type)

```

#Filter for NEE 
```{r}
#make sure to use complete cases of all variables of interest 
df_BGClongNEE <- df_BGClong %>%
  filter(flux_type == "NEE") %>%
  filter(complete.cases(flux_value, thawdepth, soil_temp_10_cm,landscape_position, inundated, VWC))


sum(is.na(df_BGClongNEE$VWC)) #check for 0's, in this one all the variables of interest have 0 NAs
sum(is.na(df_BGClongNEE$thawdepth)) #check for 0's, in this one all the variables of interest have 0 NAs
sum(is.na(df_BGClongNEE$soil_temp_10_cm)) #check for 0's, in this one all the variables of interest have 0 NAs

```

# Categorical Predictors: Variance and normality of categorical predictor variables 
```{r}
#Flux and landscape pos
df_variance_landpos <- df_BGClongNEE %>%
  group_by(landscape_position) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_landpos)

# Variance plot - flux and landscape pos
ggplot(df_variance_landpos, aes(x = landscape_position, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Landscape Position",
       y = "Variance")
#higher variance in upland 


#Flux and inundated 
df_variance_inundated <- df_BGClongNEE %>%
  group_by(inundated) %>%
  summarize(variance = var(flux_value, na.rm = TRUE))

print(df_variance_inundated)

# Variance plot - flux and inundated
ggplot(df_variance_inundated, aes(x = inundated, y = variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Variance of Flux Value by Inundation",
       y = "Variance")
#higher variance in NOT* inundated plots 


# Shapiro-Wilk test for normality across entire dataset
shapiro_test <- shapiro.test(df_BGClongNEE$flux_value)
print(shapiro_test) #p<0.001, not normal 


# Test normality within each landscape position
df_BGClongNEE %>%
  group_by(landscape_position) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: lowland & upland = no; slope = yes


# Test normality based on inundated
df_BGClongNEE %>%
  group_by(inundated) %>%
  summarize(
    shapiro_p = shapiro.test(flux_value)$p.value,
    normal = ifelse(shapiro_p > 0.05, "Yes", "No")
      )
#Results: not inun = no; inun = yes 


# Q-Q plot of full dataset 
qqnorm(df_BGClongNEE$flux_value)
qqline(df_BGClongNEE$flux_value, col = "red")


#histograms with density curves by landscape position
ggplot(df_BGClongNEE, aes(x = flux_value, fill = landscape_position)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~landscape_position) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Landscape Position")


# histograms with density curves by inundated
ggplot(df_BGClongNEE, aes(x = flux_value, fill = inundated)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity", bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~inundated) +
  theme_minimal() +
  labs(title = "Distribution of NEE by Inundation")



#homogeneity of variance - levene's test
# p > 0.05: Variances are homogeneous (no significant difference between variances).
# p ≤ 0.05: Variances are not homogeneous.
library(car)

# Test homogeneity of variance for all main factors
leveneTest(flux_value ~ landscape_position, data = df_BGClongNEE) #p=0.87, homogeneity of var is ok
leveneTest(flux_value ~ inundated, data = df_BGClongNEE)#p = 0.81, ok


#for interactions
leveneTest(flux_value ~ landscape_position * inundated, data = df_BGClongNEE) #p=0.94, ok


```

#Continuous predictors: Variance and normality of continuous predictor variables 

###Testing to see which kind of correlation to use: Pearson for parametric, Spearman for non-parametric
```{r}
#test distribution normality of each numeric predictor variable 
shapiro.test(df_BGClongNEE$soil_temp_10_cm) #p=0.0065, NOT normal** 
shapiro.test(df_BGClongNEE$thawdepth) #p = <0.001, NOT normal **
shapiro.test(df_BGClongNEE$VWC) #p<0.001, NOT normal **

#visualize pairs to see if it looks linear
pairs(df_BGClongNEE[c("soil_temp_10_cm", "thawdepth", "VWC")]) #NOPE 

plot(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$thawdepth)

plot(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$VWC)

plot(df_BGClongNEE$thawdepth, df_BGClongNEE$VWC)

ggplot(df_BGClongNEE, aes(x = soil_temp_10_cm, y = thawdepth))+
  geom_point()+
    labs(
     x = "Soil Temperature (10 cm)", 
     y = "Thaw Depth")+
geom_abline(intercept = 0, slope = 7, color = "red") # 1:1 line in red, adj slope to account for diffs in scale


#So this means for correlations and looking at collinearity, I need to use non-parametric tests 


#Check homogeneity of variance: should see even spread of of points for homogeneity of variance 

# Fit a simple model
simple_model <- lm(flux_value ~ soil_temp_10_cm, data = df_BGClongNEE)

# Extract fitted values and residuals
fitted_values <- fitted(simple_model)
residuals <- residuals(simple_model)

# Plot residuals vs. predictor
plot(df_BGClongNEE$VWC, residuals, 
     xlab = "Cont. Pred. Var", ylab = "Residuals",
     main = "Residuals vs. Predictor Var")
abline(h = 0, lty = 2)

# Create a scale-location plot --> should be randomly scattered with a relatively flat trend line (red) for homogeneity of variance 
plot(simple_model, which = 3)


#Breusch-Pagan test - specifically tests if variance of residuals is constant - want a p > 0.05 for homogeneity of var
library(lmtest)
bptest(simple_model)

#thawdepth = p=0.77, shows there is homogeneity of var 
#soil temp = p = 0.5, suggests homogeneity of var 
#VWC = p = 0.25,  suggests homogeneity of var 

```
```{r}
ggplot(df_BGClongNEE, aes(x = soil_temp_10_cm, y = thawdepth)) +
  geom_point() +
  labs(
    x = "Soil Temperature (10 cm)", 
    y = "Thaw Depth (cm)", 
    title = "Scatterplot of Soil Temperature vs. Thaw Depth") +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(intercept = 0, slope = 7, color = "blue", linetype = "dashed")

```


#Checking colinearity / correlations to determine which variables to use for models 

#Non-parametric tests of correlations - spearman's correlation 

#Corr of continuous variables 
```{r}
#Correlation between soil temp, thaw depth, and VWC

cor.test(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$thawdepth, method="spearman")#p=0.58, not sig, not correlated
cor.test(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$VWC, method="spearman") #p<0.001, *Sig*, correlated, one should be removed 
cor.test(df_BGClongNEE$thawdepth, df_BGClongNEE$VWC, method="spearman") #p<0.001, *sig*, correlated, one should be removed 

#Shows thawdepth and VWC are correlated, and VWC & soil temp are correlated 
```

#Checking correlation in categorical variables 
```{r}
# Check for separation in categorical variables
table(df_BGClongNEE$landscape_position, df_BGClongNEE$inundated) 


# Create contingency table to examine correlation quantitatively / confirm the separation test above 
cont_table <- table(df_BGClongNEE$landscape_position, df_BGClongNEE$inundated)
print(cont_table)

# Test for association
chisq.test(cont_table) #p<0.64 --> shows no correlation 
# Or for small sample sizes
fisher.test(cont_table) #p<0.64, not correlated 

#These are ok, don't show signs of correlation

```

#Determining model for BGC - NEE


#Step 1: Determine which of the correlated variables to keep
```{r}
# Create competing models with one variable at a time - use "ML" to compare models, use "REML" for fitting final model 

#thawdepth and VWC are correlated, and VWC & soil temp are correlated 
model_temp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 data = df_BGClongNEE, method = "ML")

model_thaw <- gls(flux_value ~ landscape_position + thawdepth, 
                 data = df_BGClongNEE, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 data = df_BGClongNEE, method = "ML")

model_soiltempthawdepth1 <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm, 
                 data = df_BGClongNEE, method = "ML")


# model_VWCsoiltemp <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
#                  data = df_BGClongNEE, method = "ML")
# 
# model_VWCthawdepth <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
#                  data = df_BGClongNEE, method = "ML")
# 
# model_soiltempthawdepth <- gls(flux_value ~ landscape_position + soil_temp_10_cm + thawdepth, 
#                  data = df_BGClongNEE, method = "ML")

# Compare with AIC/BIC
AIC(model_temp, model_thaw, model_VWC, model_soiltempthawdepth1) #model_VWC lowest AIC/BIC
#AIC(model_temp, model_thaw, model_VWC, model_VWCthawdepth, model_VWCsoiltemp, model_soiltempthawdepth) #model_VWC has lowest AIC, but not by much
BIC(model_temp, model_thaw, model_VWC, model_soiltempthawdepth1)#model_VWC lowest AIC/BIC
#BIC(model_temp, model_thaw, model_VWC, model_VWCthawdepth, model_VWCsoiltemp, model_soiltempthawdepth) #model_VWC has lowest BIC 

#model_VWC has lowest AIC/BIC 

```

#Step 2: Now, test whether adding soil temp or thaw depth improves model 
```{r}
model_VWC <- gls(flux_value ~ landscape_position + VWC,
                        data = df_BGClongNEE, method = "ML")

model_VWCsoiltemp <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm,
                        data = df_BGClongNEE, method = "ML")

model_VWCthawdepth <- gls(flux_value ~ landscape_position + VWC + thawdepth,
                        data = df_BGClongNEE, method = "ML")

anova(model_VWC, model_VWCsoiltemp, model_VWCthawdepth) #model_VWC has lowest AIC/BIC, but VWCsoiltemp has highest logLik, no p's are sig 
anova(model_VWC, model_VWCsoiltemp) #model VWC has lower AIC/BIC, higher logLik, p = 0.4, not sig 
anova(model_VWC, model_VWCthawdepth) #p=0.9, not sig, but model VWC has lower AIC/BIC
anova(model_VWCsoiltemp, model_VWCthawdepth) #VWCsoil temp has lower AIC but nearly the same BIC, higher LogLik, no p 
anova(model_VWC, model_VWCsoiltemp)#p<0.001, model VWC has lower AIC/BIC, is sig better 

#Appears neither thaw depth or soil temp improve model 



```

#Step 3: Test if random effect of plot_ID is appropriate
```{r}
# #If not already: Make plot_ID, inundated, landscape_position as factor so they'll work with gls
# df_MWlongNEE$plot_ID = factor(df_MWlongNEE$plot_ID)
# df_MWlongNEE$landscape_position = factor(df_MWlongNEE$landscape_position)
# df_MWlongNEE$inundated = factor(df_MWlongNEE$inundated)

#Use best predictor variable from Step 1, which was soil temp 
#use 'ML' for comparing models 

library(nlme)
#gls - no random effect
model_fixed <- gls(flux_value ~ landscape_position + VWC, 
                  data = df_BGClongNEE, method = "ML")

#lme - with random effect of plot_ID
model_random <- lme(flux_value ~ landscape_position + VWC, 
                   random = ~1|plot_ID, 
                   data = df_BGClongNEE, method = "ML")

# Compare
anova(model_fixed, model_random) #no sig diff and AIC/BIC marginally lower in model-fixed, p=0.99, so go for simpler model without random effect 
```

#Step 4: Test for variance structure 
```{r}
# test if adding a variance structure is appropriate 
model_homoscedastic <- gls(flux_value ~ landscape_position + VWC, 
                          data = df_BGClongNEE, method = "ML")

model_heteroscedastic <- gls(flux_value ~ landscape_position + VWC, 
                            weights = varIdent(form = ~1|landscape_position),
                            data = df_BGClongNEE, method = "ML")

# Compare
anova(model_homoscedastic, model_heteroscedastic) #p=<0.001 SIG*; hetero has lower AIC/BIC and higher LogLik so we do use a variance structure here * 
```

#Step 5: Try adding any remaining non-correlated variables back in to see if it improves the model (use reduction and addition / reduced models to test if various un-correlated predictors improve the model or not, and favor the simpler model)
```{r}
# test variable soil temp (but this was correlated with VWC, just running to check if var structure changes this)
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

# Compare
anova(model_1var, model_2var) #model 1 var has slightly lower AIC/BIC, p = 0.8, suggests soil temp doesn't improve model  

#try variable thawdepth (but this was correlated with VWC)
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + thawdepth, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

# Compare
anova(model_1var, model_2var) #p=0.7, model 1 var has slightly lower AIC/BIC, suggests thawdepth doesn't improve model 

#try variable "inundated"
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_2var <- gls(flux_value ~ landscape_position + VWC + inundated, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

# Compare
anova(model_1var, model_2var) #p=0.2, model 1 var has slightly lower AIC/BIC - suggests inundated doesn't improve model 


#test interaction between land post and VWC
model_1var <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")


model_interact <- gls(flux_value ~ landscape_position * VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")


# Compare
anova(model_1var, model_interact) #p=0.2, model1var has slightly lower AIC/BIC, suggests interaction effect is not important to model 



```
#Step 6: Double check best correlated var was chosen, with and without extra var, now that variance structure was added 
```{r}
#test soil temp and thawdepth, and with and without VWC

model_soiltemp <- gls(flux_value ~ landscape_position + soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_VWC <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_thawdepth <- gls(flux_value ~ landscape_position + thawdepth, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

anova(model_VWC, model_soiltemp) #the same BIC/AIC, no p 
anova(model_VWC, model_thawdepth) #same AIC/BIC, no p 
anova(model_soiltemp, model_thawdepth) #same AIC/BIC, no p 

plot(df_BGClongNEE$soil_temp_10_cm, df_BGClongNEE$flux_value) #all show a very similar trend 

#Add in inundated 
model_soiltempinun <- gls(flux_value ~ landscape_position + soil_temp_10_cm + inundated, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_VWCinun <- gls(flux_value ~ landscape_position + VWC + inundated, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_thawdepthinun <- gls(flux_value ~ landscape_position + thawdepth + inundated, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

anova(model_VWC, model_VWCinun) #p=0.2, VWC slightly better 
anova(model_soiltemp, model_soiltempinun) #p=0.3, soiltemp slightly better 
anova(model_thawdepth, model_thawdepthinun) #p=0.2, thawdepth slightly better 

#suggests adding inundated does not significantly improve the model 


model_soiltempVWC <- gls(flux_value ~ landscape_position + soil_temp_10_cm + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_thawdepthVWC <- gls(flux_value ~ landscape_position + thawdepth + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_soiltempthawdepth <- gls(flux_value ~ landscape_position + thawdepth + soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")



anova(model_soiltempVWC, model_thawdepthVWC) #same AIC/BIC, no p 
anova(model_soiltempVWC, model_soiltempthawdepth) #same
anova(model_thawdepthVWC, model_soiltempthawdepth) #same 
anova(model_thawdepthVWC, model_thawdepth) #thawdepth slightly better, p=0.7
anova(model_soiltempthawdepth, model_thawdepth) #p=0.8, thawdepth slightly better
anova(model_soiltempthawdepth, model_soiltemp)#p=0.6, soil temp slightly better
anova(model_soiltempVWC, model_soiltemp)#p=0.6, soil temp slightly better 
anova(model_thawdepthVWC, model_VWC)#p=0.8, VWC slightly better 
anova(model_soiltempVWC, model_VWC) #p=0.8, VWC slightly better 
#suggests the two correlated variables together does not improve model 


model_interactVWC <- gls(flux_value ~ landscape_position * VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_interactthawdepth <- gls(flux_value ~ landscape_position * thawdepth, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")

model_interactsoiltemp <- gls(flux_value ~ landscape_position * soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")


anova(model_interactsoiltemp, model_interactthawdepth) #soiltemp interact slightly better, no p 
anova(model_interactsoiltemp, model_interactVWC) #interactVWC slightly better, no p
anova(model_interactVWC, model_interactthawdepth) #same interactVWC slightly better, no p
anova(model_VWC, model_interactVWC)#p=0.2, VWC slightly better 
anova(model_soiltemp, model_interactsoiltemp)#p=0.47, soil temp slightly better 
anova(model_thawdepth, model_interactthawdepth) #p=0.9, thawdepth slightly better
#suggests interaction effect does not sig improve model 



anova(model_soiltemp, model_VWC) #same
anova(model_thawdepth, model_VWC) #same
anova(model_VWC, model_interact) #model VWC better, lower AIC/BIC, p=0.2

```
#Final model, refit with REML
```{r}
BGC.NEE.final <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "REML")
summary(BGC.NEE.final) #shows no var are sig 
Anova(BGC.NEE.final) #none sig 
Anova(BGC.NEE.final, type = "II", test.statistic = "F") #none sig 
# Using the standard anova function with F tests
anova(BGC.NEE.final, type = "marginal", test = TRUE) #none sig 

library(lsmeans)
lsmeans(BGC.NEE.final, adjust = "Tukey", pairwise ~ landscape_position) #none sig 

boxplot(flux_value ~ landscape_position, data = df_BGClongNEE)

```
#**TO DO REDO WITH ML
#Some extra exploratory validation tests
```{r}
m1 <- gls(flux_value ~ landscape_position + VWC, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")
summary(m1) #shows no var are sig 
Anova(m1) #none sig 
Anova(m1, type = "II", test.statistic = "F") #none sig 
# Using the standard anova function with F tests
anova(m1, type = "marginal", test = TRUE) #none sig 



m2 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")
summary(m2) #shows no var are sig 
Anova(m2) #none sig 
Anova(m2, type = "II", test.statistic = "F") #none sig 
# Using the standard anova function with F tests
anova(m2, type = "marginal", test = TRUE) #none sig 




m3 <- gls(flux_value ~ landscape_position + VWC + soil_temp_10_cm + inundated, 
                 weights = varIdent(form = ~1|landscape_position),
                 data = df_BGClongNEE, method = "ML")
summary(m3) #shows no var are sig 
Anova(m3) #none sig 
Anova(m3, type = "II", test.statistic = "F") #none sig 
# Using the standard anova function with F tests
anova(m3, type = "marginal", test = TRUE) #none sig 


anova(m1, m2) #p=0.8, m1 better, no sig diff
anova(m1, m3)#p<0.4, m1 better, no sig diff
```




#Step 7: Test for multicollinearity in model 

####For testing this in models, use VIF test 
```{r}
#variance inflation factor (vif) - vif = 1 = variables are not correlated; between 1 & 5 = moderately correlated,  5 or higher indicates highly correlated & multicollinearity among variables. GVIF is a generalized VIF for categorical variables with more than two levels; the adj VIF gives values for comparing across variables with diff degrees of freedom* --> look at both when assessing collinearity in a model 
library(car)
vif(BGC.NEE.final) # all <2, so should be ok! 
```

Plot model residuals and QQplot 
```{r}
#plot model residuals (homogeneity of variance)
plot(BGC.NEE.final)

#qqplot to verify normality - this plots the actual model, not the residuals 
qqnorm(BGC.NEE.final)

#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(BGC.NEE.final), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(BGC.NEE.final))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(BGC.NEE.final)), 
             mean = 0, sd = sigma(BGC.NEE.final))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}


#comparing data QQplot to a normal QQplot and hist 
qqnorm(residuals(BGC.NEE.final)) #normality of residuals 
hist(residuals(BGC.NEE.final)) 
car::qqPlot(x = residuals(BGC.NEE.final)) #shows where residuals breach normal distr
qqPlot(residuals(BGC.NEE.final))

```
#Normality of Residuals 
```{r}
# Extract standardized/normalized residuals
std_resid <- residuals(BGC.NEE.final, type = "normalized")

# Graphical assessment
par(mfrow = c(1, 2))
# Histogram of residuals
hist(std_resid, main = "Histogram of Standardized Residuals", 
     xlab = "Standardized Residuals", freq = FALSE)
curve(dnorm, add = TRUE, col = "red")

# QQ plot
qqnorm(std_resid, main = "Normal Q-Q Plot")
qqline(std_resid, col = "red")

# Formal test
shapiro_test <- shapiro.test(std_resid)
print(shapiro_test) #for gls final model: p<0.001; normality NOT supported 

```
#Test autocorrelation
```{r}
# Extract normalized residuals from GLS model
residuals_gls <- residuals(BGC.NEE.final, type = "normalized")

# Plot the autocorrelation function
acf(residuals_gls, main = "Autocorrelation of Residuals") #want bars of Lag1 and higher to be below the blue-dotted lines of confidence -- ignore Lag0 ** Lag0 represents correlation of residuals with themselves and is always ~1, so this isn't useful and isn't a problem. Lag 1 shows corr between each residual and the prev residual, lag 2 = correlation between resisuals 2 steps apart, etc. **Focus on Lag1 and higher* 

#Result: no apparent autocorrelation -- 
#All bars below the confidence interval dotted line = ok

# Compute lag-1 correlation in residuals - values close to 0 = little/no autocorrelation; value near 1 or -1 suggests sig correlation
cor(residuals_gls[-1], residuals_gls[-length(residuals_gls)]) 
#Result: 0.05, suggests no autocorrelation


```
#Homogeneity of variance 
```{r}
# Plot residuals vs fitted values
par(mfrow = c(1, 1))
plot(fitted(BGC.NEE.final), std_resid,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

# Check residuals by predictor variables
par(mfrow = c(2, 2))
boxplot(std_resid ~ df_BGClongNEE$landscape_position, 
        main = "Residuals by Landscape Position")
abline(h = 0, lty = 2)

boxplot(std_resid ~ df_BGClongNEE$inundated, 
        main = "Residuals by Inundation")
abline(h = 0, lty = 2)

plot(df_BGClongNEE$soil_temp_10_cm, std_resid,
     xlab = "Soil Temperature (10 cm)", ylab = "Standardized Residuals")
abline(h = 0, lty = 2)

#Result: residuals vs fitted: spread is split, clumping on either end 


# Residuals vs. Fitted plot
plot(fitted(BGC.NEE.final), std_resid, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lty = 2)

# Levene’s Test for Homogeneity of Variance
library(car)
leveneTest(std_resid ~ df_BGClongNEE$landscape_position) #p=0.47; this says it's ok 

```
#Non-parametric approach 

#### Important to note KW test is the non-parametric equivalent of a one-way anova, so it can only test one variable at a time, and can't test continuous/numeric variables. It also can't incorporate a variance structure. Spearman Correlation can be used for continuous variables*
#testing each var independently isn't ideal because it can't capture what might happen when other variables are included...
```{r}
# Test landscape position effect on NEE
kruskal.test(flux_value ~ landscape_position, data = df_BGClongNEE) #p=0.5, not sig 

#wilcoxon test
# For comparing two landscape positions (unpaired)
wilcox.test(flux_value ~ landscape_position, 
            data = df_BGClongNEE,
            subset = landscape_position %in% c("upland", "lowland")) #p=0.7, not sig 

wilcox.test(flux_value ~ landscape_position, 
            data = df_BGClongNEE,
            subset = landscape_position %in% c("upland", "slope"))#p=0.4, not sig 

wilcox.test(flux_value ~ landscape_position, 
            data = df_BGClongNEE,
            subset = landscape_position %in% c("slope", "lowland"))#p=0.2, not sig 


# For continuous variables, use Spearman correlation
cor.test(df_BGClongNEE$flux_value, df_BGClongNEE$soil_temp_10_cm, method = "spearman") #p=0.9, not sig
cor.test(df_BGClongNEE$flux_value, df_BGClongNEE$VWC, method = "spearman") #p=0.12, not sig 
cor.test(df_BGClongNEE$flux_value, df_BGClongNEE$thawdepth, method = "spearman") #p=0.94, not sig 
#The spearman results match the findings from the final gls model 

```
#Robust Regression - less sensitive to violations of normality but cannot incorporate variance structure**
```{r}
library(MASS)
library(car)
#just testing reduced and full models 
robust_model1 <- rlm(flux_value ~ landscape_position + VWC, 
                    data = df_BGClongNEE)
summary(robust_model1) 
Anova(robust_model1, type = "II") #none sig 


robust_model2 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm, 
                    data = df_BGClongNEE)
summary(robust_model2) 
Anova(robust_model2, type = "II") #none sig 


robust_model3 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm + inundated, 
                    data = df_BGClongNEE)
summary(robust_model3) 
Anova(robust_model3, type = "II") #none sig 


```
#Attempting to test robust regression models - coding help from Claude 
```{r}
# Define a function to calculate AIC for robust regression
rlm_AIC <- function(model) {
  n <- length(model$residuals)
  RSS <- sum(model$residuals^2)
  k <- length(coef(model)) + 1  # +1 for the error variance
  AIC <- n * log(RSS/n) + 2*k
  return(AIC)
}

# Compare models
rlm_AIC(robust_model1)
rlm_AIC(robust_model2)
rlm_AIC(robust_model3)

#Results:
# [1] -2173.55 #--> model with just landpos and VWC has lowest AIC
# [1] -2171.623
# [1] -2169.813
```
#Prediction error
```{r}
# MAE (model absolute error) values, interpretation is straightforward:
# 
# The model with the lower MAE provides better predictions on average
# The difference between MAE values indicates the average improvement in prediction accuracy from adding soil temperature to your model

#models 
model1 <- rlm(flux_value ~ landscape_position + VWC, data = df_BGClongNEE)
model2 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm, data = df_BGClongNEE)
model3 <- rlm(flux_value ~ landscape_position + VWC + soil_temp_10_cm + inundated, data = df_BGClongNEE)

# Calculate MAE for each model
MAE_model1 <- mean(abs(residuals(model1)))
MAE_model2 <- mean(abs(residuals(model2)))
MAE_model3 <- mean(abs(residuals(model3)))

# Compare
print(paste("Model 1 MAE:", MAE_model1))
print(paste("Model 2 MAE:", MAE_model2))
print(paste("Model 3 MAE:", MAE_model3))

#Results:
# [1] "Model 1 MAE: 1.18936072155054e-05"
# [1] "Model 2 MAE: 1.18966188202511e-05"
# [1] "Model 3 MAE: 1.18781833706506e-05"

```
#Cross-validation using RMSE
```{r}

# RMSE measures the average prediction error (lower is better)
# Each model shows 6 values because the caret package runs multiple tuning parameters by default for robust regression

library(caret)

# Create training control for 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Compare models using cross-validation
cv_full <- train(flux_value ~ landscape_position + soil_temp_10_cm + VWC + inundated, 
                data = df_BGClongNEE,
                method = "rlm",
                trControl = train_control)

cv_reduced1 <- train(flux_value ~ landscape_position + soil_temp_10_cm + VWC, 
                   data = df_BGClongNEE,
                   method = "rlm",
                   trControl = train_control)

cv_reduced2 <- train(flux_value ~ landscape_position + VWC, 
                   data = df_BGClongNEE,
                   method = "rlm",
                   trControl = train_control)

cv_reduced3 <- train(flux_value ~ landscape_position + soil_temp_10_cm, 
                   data = df_BGClongNEE,
                   method = "rlm",
                   trControl = train_control)

# Compare RMSE
cv_full$results$RMSE
cv_reduced1$results$RMSE
cv_reduced2$results$RMSE
cv_reduced3$results$RMSE #has lowest RMSE, suggesting this is best model 

#the AIC suggests landpos + VWC is best, RMSE suggests landpos + soil temp is best, but both are very close, which makes sense given the correlation. 
#Ran stats using both versions as a final model to cross validate 
```
#testing both versions of a final model 
```{r}
library(MASS)
library(car)
#VWC
final_VWC <- rlm(flux_value ~ landscape_position + VWC, 
                    data = df_BGClongNEE)
summary(final_VWC) 
Anova(final_VWC, type = "II") #none sig 


#soil temp
final_soiltemp <- rlm(flux_value ~ landscape_position + soil_temp_10_cm, 
                    data = df_BGClongNEE)
summary(final_soiltemp) 
Anova(final_soiltemp, type = "II") #none sig 

#the two models agree 
```


