---
title: "Council Soil Chamber Analysis" #time is in AK daylight time** NEE/RECO/GPP are in gC/m2/s; FCO2 and FCH4 are in gC/m2/s, flux_CO2 is in umol/m2/s; flux_CH4 is in nano mol/m2/s
output: html_document
date: "2024-10-22"
---

#Working code - very far from finalized / still a mess 

#Note that for comparison purposes, both instruments 
were used to measure chamber fluxes on July 18, 2018 --> remove potential measurement duplicates from this date?

#measure the Net Ecosystem Exchange (NEE) with the transparent chamber during the day (when photosynthesis is occurring) and the Ecosystem Respiration (Reco) with the opaque chamber during the night (when only respiration is happening), then subtract the Reco value from the NEE value to get GPP: GPP = NEE (transparent chamber) - Reco (opaque chamber)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
#filtered for p<0.05; units umol/m2/s or nmol/m2/s
df_soilchambers_filtered = fread('./council_filtered_soil_chamber_fluxes_2017to2019.csv')

#fluxes and moisture/temp df merged; FCO2 in units g/m2/s
df_fulljoin = fread('./council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************
#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('./council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('./council_NEE_RECO2_GPP_2017to2019.csv')


```


#Boxplots exploring some data 
```{r}
library(tidyr)


# Reshape the dataframe to long format
df_long <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>%
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")

# Create the boxplot: NEE and RECO flux vs landscape position 
ggplot(df_long, aes(x = landscape_position, y = flux_value, fill = flux_type)) +
  geom_boxplot() +
  labs(title = "NEE, GPP, and RECO vs Landscape Position",
       x = "Landscape Position",
       y = "Flux Value (gCO2/m²/s)",
       fill = "Flux Type") +
   geom_hline(yintercept = 0)+
  theme_minimal()


# EC plot type NEE and RECO flux vs landscape position 
ggplot(df_long %>% filter(plot_type == "EC"),
       aes(x = landscape_position, y = flux_value, fill = flux_type)) +
  geom_boxplot() +
  labs(title = "EC plots: NEE, GPP, and RECO vs Landscape Position",
       x = "Landscape Position",
       y = "Flux Value (gCO2/m²/s)",
       fill = "Flux Type") +
   geom_hline(yintercept = 0)+
  theme_minimal()

# MW plots by landscape position 
ggplot(df_long %>% filter(plot_type == "MW"),
       aes(x = landscape_position, y = flux_value, fill = flux_type)) +
  geom_boxplot() +
  labs(title = "MW plots: NEE, GPP, and RECO vs Landscape Position",
       x = "Landscape Position",
       y = "Flux Value (gCO2/m²/s)",
       fill = "Flux Type") +
   geom_hline(yintercept = 0)+
  theme_minimal()

# BGC plots by landscape position 
ggplot(df_long %>% filter(plot_type == "BGC"),
       aes(x = landscape_position, y = flux_value, fill = flux_type)) +
  geom_boxplot() +
  labs(title = "BGC plots: NEE, GPP, and RECO vs Landscape Position",
       x = "Landscape Position",
       y = "Flux Value (gCO2/m²/s)",
       fill = "Flux Type") +
   geom_hline(yintercept = 0)+
  theme_minimal()

#looking at FCH4 (in g/m2/s) by landscape position and plot type - scaled 
ggplot(df_long,
       aes(x=landscape_position, y = (FCH4 * 1000), fill = plot_type))+
  geom_boxplot()+
  labs( title = "FCH4 by plot type vs inundation",
        x = "Landscape position",
        y = "FCH4 (gCH4/m2/s)",
        fill = "Plot Type")+
  geom_hline(yintercept=0)+
  theme_minimal() +
  scale_y_continuous(limits = c(-0.0005, 0.0007))


#looking at FCH4 (in g/m2/s) by inundation - scaled
ggplot(df_long,
       aes(x=inundated, y = (FCH4 * 1000), fill = plot_type))+
  geom_boxplot()+
  labs( title = "FCH4 vs inundation",
        x = "Inundation",
        y = "FCH4 (gCH4/m2/s)",
        fill = "plot type")+
  geom_hline(yintercept=0)+
  theme_minimal() +
  scale_y_continuous(limits = c(-0.0005, 0.0007))

```

#Boxplots of GPP flux by plot_type and landscape position / inundation
```{r}
#GPP of each plot type by landscape position 
ggplot(df_long %>% filter(flux_type == "GPP"),
       aes(x = landscape_position, y = flux_value, fill = plot_type)) +
  geom_boxplot() +
  labs(title = "GPP vs Landscape Position",
       x = "landscape position",
       y = "GPP(gCO2/m²/s)",
       fill = "Plot Type") +
   geom_hline(yintercept = 0)+
  theme_minimal()



#landscape position, plot type, and inundation 
ggplot(df_long %>% filter(flux_type == "GPP"),
       aes(x = plot_ID, y = flux_value, color = plot_type, shape = inundated )) +
  geom_point() +
  labs(title = "GPP vs Landscape Position & inundation",
       x = "plot_ID",
       y = "GPP(gCO2/m²/s)",
       color = "Plot Type",
       shape = 'inundated') +
   geom_hline(yintercept = 0)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#closer look at inundation 
ggplot(df_long %>% filter(flux_type == "GPP") %>%
         drop_na(inundated), #drop_na () drops NAs, but here I can specify which NAs in which relevant rows to drop -- here I specify only to drop a row if "inundated = NA" (and leave other rows in NA alone)
       aes(x = inundated, y = flux_value, color = plot_type )) +
  geom_boxplot() +
  labs(title = "GPP vs Landscape Position & inundation",
       x = "plot_ID",
       y = "GPP(gCO2/m²/s)",
       color = "Inundation") +
   geom_hline(yintercept = 0)+
  theme_minimal() + 
  scale_y_continuous(limits = c(-0.0001, 0.0002))

#NEE based on inundation and landscape position --> no real differences seen 
ggplot(df_long %>% filter(flux_type == "NEE") %>%
         drop_na(inundated), 
       aes(x=landscape_position, y = flux_value, color = inundated))+
  geom_jitter() +
  theme_minimal()

#RECO based on inundation and landscape position --> no real differences seen 
ggplot(df_long %>% filter(flux_type == "RECO") %>%
         drop_na(inundated), 
       aes(x=landscape_position, y = flux_value, color = inundated))+
  geom_jitter() +
  theme_minimal()
  

#GPP based on inundation and landscape position --> no real differences seen 
ggplot(df_long %>% filter(flux_type == "GPP") %>%
         drop_na(inundated), 
       aes(x=landscape_position, y = flux_value, color = inundated))+
  geom_jitter() +
  theme_minimal()

#Checking to see which plots are inundated 


# Filter for inundated plots and create a bar plot to see which plots have inundation 
ggplot(df_long %>% filter(inundated == "Y"), aes(x = plot_ID, fill = inundated)) +
  geom_bar() +
  labs(title = "Inundated Plots", x = "Plot ID", y = "Count") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))

#seeing inundated vs not 
ggplot(df_long %>% drop_na(inundated))+
       aes(x=plot_ID, fill = inundated)+
  geom_bar(position = "dodge")+
  theme_minimal() 


ggplot(df_long %>% drop_na(inundated))+
       aes(x=plot_ID, fill = inundated)+
  geom_bar()+
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust =1))


# looking at measurement date counts for each plot ID 
ggplot(df_long, aes(x = plot_ID, fill = measurement_date)) +
  geom_bar() +
  labs(title = "Inundated Plots", x = "Plot ID", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

#Filter df by landscape position and flux type (GPP, NEE, RECO)

####Create new df for each plot type for analysis 
```{r}
library(nlme)
library(lme4)
# library(dplyr)
# library(tidyr)

model_GPPtest <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_GPP)
summary(model_GPPtest)



# Step 1: Filter the dataframe for plot_ID = "EC" "MW" and "BGC" to create df for analysis 

#EC - eddy covar tower plot types 
df_EC <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "EC")

#EC - eddy covar tower plot types; just another way to subset - doing it this way saves you from having to make df_EC into d_long format, but below way allows you to choose variables of interest and simplify the df
df_EC2 <- df_long %>%
  filter(plot_type == "EC")

#MW - moisture warming plot types 
df_MW <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "MW")

#BGC - biogeochem plot types 
df_BGC <- df_NEE_RECO2_GPP %>%
  filter(plot_type == "BGC")

#GPP
df_GPP <-df_long %>%
  filter(flux_type == "GPP")

#NEE
df_NEE <-df_long %>%
  filter(flux_type == "NEE")
#RECO
df_RECO <-df_long %>%
  filter(flux_type == "RECO")


# Step 2: Reshape the dataframe to long format and choose variables of interest 

#EC
df_EClong <- df_EC %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>% *choosing variables of interest 
  select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")

#MW
df_MWlong <- df_MW %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>%
   select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")

#BGC
df_BGClong <- df_BGC %>%
  #select(plot_ID, plot_type, landscape_position, measurement_date, NEE, RECO, GPP) %>%
   select(plot_ID, plot_type, landscape_position, measurement_date, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, soil_temp_15_cm) %>% #choosing all variables or just subset from the d_long data 
  pivot_longer(cols = c(NEE, RECO, GPP), 
               names_to = "flux_type", 
               values_to = "flux_value")


#Re-arrange by flux type (NEE, GPP, RECO) so you can analyze more easily 

# Sort the dataframe by the flux_type column 
df_EClong <- df_EClong %>% arrange(flux_type)
df_MWlong <- df_MWlong %>% arrange(flux_type)
df_BGClong <- df_BGClong %>% arrange(flux_type)

```

#Exploratory analysis 

#### EC plots - NEE 
```{r}
# Load nlme package
library(nlme)
library(lme4)
library(agricolae)
library(car)
library(emmeans)

df_EC_NEE <- df_EClong %>% filter(flux_type == "NEE")


#histogram of distr of data in df_NEE
hist(df_EC_NEE$flux_value)
#skewed a bit to the left

library(nlme)
model_NEE <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_NEE)
summary(model_NEE)
anova(model_NEE) #p = 0.07
Anova(model_NEE, test.statistic = "F", type = "II", ddf = "Kenward-Roger") # p = 0.0361?! ask why this is diff 


library(nlme)
EC_NEE_NLME <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_NEE)
summary(EC_NEE_NLME)
anova(EC_NEE_NLME)
Anova(EC_NEE_NLME, test.statistic = "F", type = "II", ddf = "Kenward-Roger") #can't customize with nlme, only uses chi-sqr
#this is still chi-sq....p = 0.07  --> now says sig at 0.036?! **Need to ask about this 

library(lme4)
EC_NEE_LMM<- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_NEE )
summary(EC_NEE_LMM, corr = F)
Anova(EC_NEE_LMM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#p=0.087; not sig 

#testing the model without a random effect of plot_id
model_NEE_LM<- lm(flux_value ~ landscape_position, data = df_EC_NEE )
summary(model_NEE_LM, corr = F)
Anova(model_NEE_LM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#without plot_ID as a random effect, this is sig with p = 0.015

 #Testing which model is a beter fit: NEE_LMM = with random effect of plot_ID; NEE_LM = without random effect of plot_ID 
anova(EC_NEE_LMM, model_NEE_LM)
#Results show lower AIC and BIC for model_NEE2, with random effect of plot_ID, so we will proceed with this version 

#Testing with emmeans 
library(emmeans)
emmeans(EC_NEE_LMM, specs = "landscape_position") # 
emmeans(EC_NEE_LMM, pairwise ~ landscape_position) #Tukeypairwise --> not sig, p = 0.0872

```


#Residuals and normality of mixed-effects model - ECC plots NEE2
```{r}
#checking normality of residuals distribution 
plot(model_NEE2) # check constant variance
lattice::qqmath(model_NEE2) # check normality of residuals
plot(model_NEE2 , plot_ID ~ resid(., scaled=TRUE)) # equal var within Plots
#---------------------------------------------------------------------------------------------------
#checking how this QQ plot compares to plots created with normally distributed residuals

op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(model_NEE2), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(model_NEE2))

# now create 24 qq plots using Normal data with sigma(dataset)
for(i in 1:24){
  # rnorm() samples from a Normal distribution  
  d <- rnorm(length(residuals(model_NEE2)), 
             mean = 0, sd = sigma(model_NEE2))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#these look pretty good

#For further testing, if needed 
qqnorm(residuals(model_NEE2))
hist(residuals(model_NEE2)) #shows a bit of left skew

#brown forsythe test to test variance among groups, want p to be above 0.05 to show no sig diff
install.packages("onewaytests") #for a brown forscythe test
library(onewaytests)
##Testing for Homogeneity of variance with Brown-Forsythe test

#code syntax: #bf.test(dependent variable ~ independent variable, data = dataset) 
bf.test(flux_value ~ landscape_position, data=df_NEE) #this is variance among groups, so diff not being stat sig is a good thing -> but p is 0.00843, so it is sig 

#variance within groups test with Levene's test
#code syntax: leveneTest(dataset$dependent variable, dataset$independent variable)
leveneTest(df_NEE$flux_value, df_NEE$landscape_position) #p above 0.05 means there no sig variance within groups, so data is distr normally --> p = 0.57, so not sig 

#If I need to log transform:
#dataset$new_name of log dataset <- log(dataframe$dependent variable) 
#same for square rooting transformation, just use "sqrt"


#This not working, look into later 
#checking ratio of largest grp var to smallest group var, should be 3 or below 
grp_vars <- with(df_NEE, tapply(flux_type,landscape_position,var))
max(grp_vars)/min(grp_vars)


```

####EC plots - GPP

```{r}
library(nlme)
library(lme4)
library(agricolae)
library(car)
library(emmeans)

df_EC_GPP <- df_EClong %>% filter(flux_type == "GPP")

ggplot(df_EC_GPP, aes (x=landscape_position, y = flux_value))+geom_boxplot()

hist(df_EC_GPP$flux_value)
#skewed right 

library(nlme)
EC_GPP_NLME <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_EC_GPP)
summary(EC_GPP_NLME)
anova(EC_GPP_NLME) #not sig, p = 0.1462; only works with chi-sqr


library(lme4)
EC_GPP_LMM <- lmer(flux_value ~ landscape_position + (1|plot_ID), data = df_EC_GPP )
summary(EC_GPP_LMM, corr = F)
Anova(EC_GPP_LMM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#p=0.157; not sig 



#testing the model without a random effect of plot_id
EC_GPP_LM<- lm(flux_value ~ landscape_position, data = df_EC_GPP )
summary(EC_GPP_LM, corr = F)
Anova(EC_GPP_LM, type = "II", test.statistic = "F", ddf = "Kenward-Roger")
#no random effect, p = 0.0335

#sig without random effects, p = 0.0335
fit <- aov(flux_value ~ landscape_position, data = df_EC_GPP)
summary(fit)

#Testing with emmeans 
library(emmeans)
emmeans(EC_GPP_LMM, specs = "landscape_position") # put those means/CIs in a data frame
emmeans(EC_GPP_LMM, pairwise ~ landscape_position) #Tukeypairwise
#also shows contrast is not sig, p = 0.157

#testing with  a t-test for flux_value by landscape_position --> is sig at p=0.01381, but this does not include a random effect
t_test_result <- t.test(flux_value ~ landscape_position, data = df_EC_GPP)
# View the result
print(t_test_result)


```
```{r}
# Load lmerTest package for p-values - this does LMM with an added p-value 
library(lmerTest)

# Fit the model with lmerTest to obtain p-values
model <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_EC_GPP)
summary(model) #p=0.10254
Anova(model, type = "II", test.statistic = "F", ddf = "Kenward-Roger") #p=0.1574

```

#Tukey letters compact letter display  (just for practice, as there is no sig diff here)
```{r}
# Install multcomp if not already installed
install.packages("multcomp")
library(multcomp)
library(multcompView)
emGPP <- emmeans(EC_GPP_LMM, specs = "landscape_position")
cld(emGPP, Letter = "abcdefghijk")

#FOR TUKEY HSD 
# create compact letter display!
# using lme model, get est means with CIs
emmeans(EC_GPP_LMM, specs = "landscape_position") # put those means/CIs in a data frame
emmeans(EC_GPP_LMM, pairwise ~ landscape_position) #Tukeypairwise
emm_EC_GPP_LMM_df <- as.data.frame(emmeans(EC_GPP_LMM, specs = "landscape_position"))


# get the compact letter displays using multcomp package function cld() 
library(multcomp)
tuk_EC_GPP <- glht(EC_GPP_LMM, linfct = mcp(landscape_position = "Tukey"))
tuk_EC_GPP_cld <- cld(tuk_EC_GPP)
tuk_EC_GPP_cld

# add letters to data frame
emm_EC_GPP_LMM_df$letters <- tuk_EC_GPP_cld$mcletters$Letters

#plot with letters from tukey groupings 
library(ggplot2)
ggplot(df_EC_GPP, aes(x=landscape_position, y = flux_value))+
  geom_boxplot()+
  theme_minimal()+
   annotate("text", x = 1:2, y = 0.0002, label = tuk_EC_GPP_cld$mcletters$Letters)

#a more strict tukey, adheres to the p<0.05

library(multcomp)
library(multcompView)
emm <- emmeans(EC_GPP_LMM, specs = "landscape_position")
cld(emm, Letter = "abcdefghijk")

```
#LME of GPP, NEE, RECO among all plot types 

```{r}
# mixed effect models
library(lme4)
library(emmeans)
library(car)

library(nlme)
#model testing differences in landscape position among all plot types --> not sig, p = 0.3266
EC_GPP_lme <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_GPP, na.action=na.exclude) #lets R work around the 0 values in dataset )
summary(EC_GPP_lme, corr = F)
Anova(EC_GPP_lme,test.statistic = "F", type = "II", ddf = "Kenward-Roger") #does type 2 but only with chi-sqr in nlme package 

library(lme4)
# GPP among all plot types  --> not sig, p = 0.3459
model_GPP_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_GPP)
summary(model_GPP_plottype)
Anova(model_GPP_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# NEE among all plot types  --> not sig, p = 0.147
model_NEE_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_NEE)
summary(model_NEE_plottype)
Anova(model_NEE_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# RECO among all plot types  --> not sig, p = 0.89
model_RECO_plottype <- lmer(flux_value ~ landscape_position + (1 | plot_ID), data = df_RECO)
summary(model_RECO_plottype)
Anova(model_RECO_plottype,test.statistic = "F", type = "II", ddf = "Kenward-Roger")


```
#LME of GPP, NEE, RECO among plot types
```{r}

library(lme4)
# GPP among all plot types  --> not sig, p = 0.288
model_GPP_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_GPP)
summary(model_GPP_plottype2)
Anova(model_GPP_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# NEE among all plot types  --> not sig, p = 0.07
model_NEE_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_NEE)
summary(model_NEE_plottype2)
Anova(model_NEE_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")

# RECO among all plot types  --> not sig, p = 0.59
model_RECO_plottype2 <- lmer(flux_value ~ plot_type + (1 | plot_ID), data = df_RECO)
summary(model_RECO_plottype2)
Anova(model_RECO_plottype2,test.statistic = "F", type = "II", ddf = "Kenward-Roger")
```



```{r}
#Residuals and normality of mixed-effects model - ECC plots GPP

#checking normality of residuals distribution 
plot(model_NEE2) # check constant variance
lattice::qqmath(model_NEE2) # check normality of residuals
plot(model_NEE2 , plot_ID ~ resid(., scaled=TRUE)) # equal var within Plots
#---------------------------------------------------------------------------------------------------
#checking how this QQ plot compares to plots created with normally distributed residuals
#ASK CLAY - will this test work with my other datasets? just plug in diff dataset and test? 
#How to do a glmm / glmer model in case of transformations not working 
#how to transform / back transform (I do have negative values and 0 values in some cases)
op <- par(mar = c(2,2,1,1), mfrow = c(5,5))

# create first qq plot using model residuals
# color it red
qqnorm(residuals(model_NEE2), xlab = "", ylab = "", main = "", 
       col = "red")
qqline(residuals(model_NEE2))

# now create 24 qq plots using Normal data with sigma(PB19Eri.angCN_LMM
for(i in 1:24){
  # rnorm() samples from a Normal dist'n 
  d <- rnorm(length(residuals(model_NEE2)), 
             mean = 0, sd = sigma(model_NEE2))
  qqnorm(d, xlab = "", ylab = "", main = "")
  qqline(d)
}

#These residuals look pretty good 

#For further testing, if needed 
qqnorm(residuals(model_NEE2))
hist(residuals(model_NEE2)) #shows a bit of left skew

#brown forsythe test to test variance among groups, want p to be above 0.05 to show no sig diff
install.packages("onewaytests") #for a brown forscythe test
library(onewaytests)
##Testing for Homogeneity of variance
## but these test for normality in data, not residuals, right? 
# with Brown-Forsythe test
#bf.test(dependent variable ~ independent variable, data = dataset) 
bf.test(flux_value ~ landscape_position, data=df_NEE) #this is variance among groups, so diff not being stat sig is a good thing -> but p is 0.00843, so it is sig 
#variance within groups test

# with Levene's test
#leveneTest(dataset$dependent variable, dataset$independent variable)
leveneTest(df_NEE$flux_value, df_NEE$landscape_position) #p above 0.05 means there no sig variance within groups, so data is distr normally --> p = 0.57, so not sig 

#If I need to log transform:
#dataset$new_name of log dataset <- log(dataframe$dependent variable) 
#same for square rooting transformation, just use "sqrt"


#This not working, look into later 
#checking ratio of largest grp var to smallest group var, needs to be 3 or below 
grp_vars <- with(df_NEE, tapply(flux_type,landscape_position,var))
max(grp_vars)/min(grp_vars)



```

#EC RECO
```{r}
library(nlme)
library(lme4)
library(car)

df_RECO <- df_EClong %>% filter(flux_type == "RECO")


model_RECO <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_RECO)
summary(model_RECO)


df_GPP <- df_EClong %>% filter(flux_type == "GPP")

model_GPPtest <- lme(flux_value ~ landscape_position, random = ~1 | plot_ID, data = df_GPP)
summary(model_GPPtest)
anova(model_GPPtest)
```


#Checking out the overall data, filtered, but not subset 
```{r}
library(ggplot2)

#plot - filtered df but not converted or subset, units are umolC/m2/s

# Create the scatter plot
ggplot(df_soilchambers_filtered, aes(x = measurement_date, y = flux_CO2, 
                                      color = landscape_position, shape = chamber_type)) +
  geom_point(size = 3, position = position_jitter(width = 0.8, height = 0)) +  # Adjust point size for better visibility
  labs(x = "Measurement Date", y = "CO2 Flux (umol/m2/s)", 
       title = "CO2 Flux vs Measurement Date by Landscape Position and Chamber Type") +
  scale_color_manual(values = c("upland" = "green", "lowland" = "blue", "slope" = "red")) +  # Adjust color palette as needed
  theme_minimal() +  # Apply minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability



```

#Below this are working notes, ignore!


#can subset by year
#can subset by landscape popsition 
#def need to subset by chamber 

#find a way to add the temp and inundation data into here** 





#Need to make new df in the d_long format so I can run analyses 
#Mixed effects model - NLME - daytime chambers (tr)
```{r}
library(nlme)
# landscape_NLME <- lme(flux_CO2 ~ landscape_position, random = ~1|PlotNumber, data = percentC) 
# #na.action=na.exclude) #lets R work around the 0 values in dataset 
# summary(percentC_NLME, corr = F)
# anova(percentC_NLME)
# # each plot gets a fitted intercept
# coef(percentC_NLME)


#Anova(StDead_LMM, test.statistic = "F", type = "II", ddf = "Kenward-Roger")
library(nlme)

landscape_NLME <- lme(flux_CO2 ~ landscape_position * plot_type, data = df_trns )
summary(landscape_NLME, corr = F)
anova(Slandscape_NLME)
# each plot gets a fitted intercept
coef(landscape_NLME)


library(emmeans)
emmeans(landscape_NLME, revpairwise ~ Location | SiteType) |>
  confint() 
#ASK CLAY: can we go over these results one more time, just to make sure I understand what's useful to report --> these results show up the average data value for a site type at each location, and the df and CL, right? What are the contrast estimates again? simply the difference in means for each site type between locations? just shows us which location has higher nutrient data averages and by how much? --> yes**

#might not need to add in siteId as random, as the StDead_LMM model already incorporated it...?
emmeans(landscape_NLME, pairwise ~ landscape_position | plot_type, random = ~1|plot_ID, adjust="tukey")


# mixed effect models
library(lme4)
library(emmeans)
library(car)
# lme model, pretty much an ANOVA
PB19_PRSprobesNO3N_LMM <- lmer(data ~ SiteType + (1|PlotNumber), data = NO3N )
summary(PB19_PRSprobesNO3N_LMM, corr = F)
Anova(PB19_PRSprobesNO3N_LMM,test.statistic = "F", type = "II", ddf = "Kenward-Roger")
# each plot gets a fitted intercept
coef(PB19_PRSprobesNO3N_LMM)
```






