---
title: "Soil_Chamber_Enviro_Figs_Stats"
output: html_document
date: "2025-10-26"
---
#"Soil_Chamber_EnviroVariable_Figs" ##time is in AK daylight time** NEE/RECO/GPP are in gC/m2/s; FCO2 and FCH4 are in gC/m2/s, flux_CO2 is in umol/m2/s; flux_CH4 is in nano mol/m2/s

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')



```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs 
library(dplyr)

# Reshape the dataframe to long format, choose variables of interest 
df_wide <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) #exclude soil temp 15 cm since most of it is NA 

#make NEE, GPP, RECO, and FCH4 all g/m2/s -> umol/m2/s
df_wide$NEE_umol <- df_wide$NEE * 1/12 * 1000000
df_wide$GPP_umol <- df_wide$GPP * 1/12 * 1000000
df_wide$RECO_umol <- df_wide$RECO * 1/12 * 1000000
df_wide$FCH4_umol <- df_wide$FCH4 * 1/12 * 1000000

#Filter out inundated slope with only n=2
# Filter out combinations with n < 5
# This removes slope-inundated (slope-Y)
df_wide_filtered <- df_wide %>%
  group_by(landscape_position, inundated) %>%
 filter(flux_CH4 <= 400) %>% #filter out the outlier of flux CH4 = 600 nmol, occurs only one time in 2019 and rest are < 400 
  filter(n() >= 5) %>%
  ungroup()


df_long2 <- df_wide_filtered  %>%
  # select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) %>% #exclude soil temp 15 cm since most of it is NA 
  pivot_longer(cols = c(NEE, RECO, GPP, FCH4),
               names_to = "flux_type",
               values_to = "flux_value")

#df_long2$FCH4_umol <- df_long2$FCH4 * 1/12 * 1000000
df_long2$flux_umol <- df_long2$flux_value * 1/12 * 1000000

colnames(df_wide)
colnames(df_long2)
```


#looking at variables  across landscape positions & inundation
```{r}

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, color = flux_type)) +
  geom_boxplot(position = "dodge") +
  stat_summary(fun = mean, geom = "point", position = position_dodge(0.8), shape = 23, size = 4)+
    facet_wrap(~inundated) +
  labs(title = "Observed NEE by Land pos",
       x = "Landscape Position",
       y = "NEE (flux_value)") +
  theme_bw()

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, fill = flux_type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 4,
    position = position_dodge(0.8)  # same dodge width as boxplot
  ) +
  labs(
    title = "Observed NEE by Landscape Position",
    x = "Landscape Position",
    y = "Flux Value (umol/m2.s)",
    fill = "Flux Type"
  ) +
  theme_bw()



#thawdepth by landpos
thaw_landpos <- ggplot(df_long2, aes(x = landscape_position, y = thawdepth)) +
  geom_boxplot() +
  labs(title = "Observed Thaw Depth by Landscape Position",
       x = "Landscape Position",
       y = "Thaw Depth (cm)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

thaw_landpos

#thawdepth by landpos and inundated 
thaw_landposinun <- ggplot(df_long2, aes(x = landscape_position, y = thawdepth)) +
  geom_boxplot() +
   facet_wrap(~inundated) +
  labs(title = "Observed thawdepth by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Thaw Depth (cm)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

thaw_landposinun

#VWC by landpos 
VWC_landpos <- ggplot(df_long2, aes(x = landscape_position, y = VWC)) +
  geom_boxplot() +
    labs(title = "Observed VWC by Landscape Position",
       x = "Landscape Position",
       y = "VWC (%)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

VWC_landpos

#VWC by landpos and inundated
VWC_landposinun <- ggplot(df_long2, aes(x = landscape_position, y = VWC)) +
  geom_boxplot() +
  facet_wrap(~inundated)+
    labs(title = "Observed VWC by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "VWC (%)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

VWC_landposinun


#soil temp by landpos 
soiltemp_landpos <- ggplot(df_long2, aes(x = landscape_position, y = soil_temp_10_cm)) +
  geom_boxplot() +
    labs(title = "Observed Soil Temp by Landscape Position",
       x = "Landscape Position",
       y = "Soil Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

soiltemp_landpos


#soiltemp by landpos and inundated 
soiltemp_landposinun <- ggplot(df_long2, aes(x = landscape_position, y = soil_temp_10_cm)) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    labs(title = "Observed Soil Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Soil Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

soiltemp_landposinun

#air temp by landpos 
airtemp_landpos <- ggplot(df_long2, aes(x = landscape_position, y = air_temp)) +
  geom_boxplot() +
    labs(title = "Observed Air Temp by Landscape Position",
       x = "Landscape Position",
       y = "Air Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

airtemp_landpos

#airtemp by landpos and inundated 
airtemp_landposinun <- ggplot(df_long2, aes(x = landscape_position, y = air_temp)) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    labs(title = "Observed Air Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Air Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

airtemp_landposinun

#Number of inundated plots by landpos 
library(dplyr)
library(ggplot2)

df_counts <- df_long2 %>%
  distinct(plot_ID, landscape_position, inundated) %>%  # keep unique plots only
  count(landscape_position, inundated)                  # count per group

ggplot(df_counts, aes(x = landscape_position, y = n, fill = inundated)) +
  geom_col(position = "dodge") +
  labs(
    title = "Unique Plot Counts by Landscape Position and Inundation",
    x = "Landscape Position",
    y = "Count of Unique Plots",
    fill = "Inundated"
  ) +
  theme_bw() 
  


n_distinct(df_long2$plot_ID) #35 total plots

```

#Arrange timeseries figures into a grid to save as single figure
```{r}
# Install if you don't have it
 install.packages("patchwork")
library(patchwork)

# Arrange plots in two columns (3 rows x 2 columns) 
combined_plot <- (title1 + title2) /
                 (soiltemp_landpos + soiltemp_landposinun) /
                 (airtemp_landpos + airtemp_landposinun) /
                 (thaw_landpos + thaw_landposinun)/
                 (VWC_landpos + VWC_landposinun) 
                # plot_layout(guides = "collect") # Collects all legends into one
combined_plot



#remove titles from each individual plot 
library(patchwork)

combined_plot <- (
  
  soiltemp_landpos + labs(title = NULL) + 
  soiltemp_landposinun + labs(title = NULL)
) /
(
  airtemp_landpos + labs(title = NULL) + 
  airtemp_landposinun + labs(title = NULL)
) /
(
  thaw_landpos + labs(title = NULL) + 
  thaw_landposinun + labs(title = NULL)
) /
(
  VWC_landpos + labs(title = NULL) + 
  VWC_landposinun + labs(title = NULL)
) 


combined_plot


```
#patchwork not working, trying GridExtra
```{r}
# ===================================================================
# STEP 5: Combine Plots Using gridExtra (Alternative to patchwork)
# ===================================================================

library(gridExtra)
library(grid)  # For textGrob

#  individual plots:
# soiltemp_landpos, soiltemp_landposinun
# airtemp_landpos, airtemp_landposinun  
# thaw_landpos, thaw_landposinun
# VWC_landpos, VWC_landposinun

# Method 1: Simple grid arrangement
combined_plot <- grid.arrange(
  soiltemp_landpos, soiltemp_landposinun,
  airtemp_landpos, airtemp_landposinun,
  thaw_landpos, thaw_landposinun,
  VWC_landpos, VWC_landposinun,
  ncol = 2,  # Two columns
  nrow = 4   # Four rows
)

# Method 2: With custom column headers
# Create text grobs for headers
header_left <- textGrob("Landscape Position", 
                        gp = gpar(fontsize = 14, fontface = "bold"))
header_right <- textGrob("Landscape Position and Inundation", 
                         gp = gpar(fontsize = 14, fontface = "bold"))

# Arrange with headers
combined_plot_with_headers <- grid.arrange(
  header_left, header_right,
  soiltemp_landpos + labs(title = NULL), 
  soiltemp_landposinun + labs(title = NULL),
  airtemp_landpos + labs(title = NULL), 
  airtemp_landposinun + labs(title = NULL),
  thaw_landpos + labs(title = NULL), 
  thaw_landposinun + labs(title = NULL),
  VWC_landpos + labs(title = NULL), 
  VWC_landposinun + labs(title = NULL),
  ncol = 2,
  nrow = 5,  # 5 rows now (1 for headers, 4 for plots)
  heights = c(0.5, 2, 2, 2, 2)  # Make header row shorter
)



# Alternative using cowplot (another option)
# install.packages("cowplot")
library(cowplot)

combined_cowplot <- plot_grid(
  soiltemp_landpos + labs(title = NULL),
  soiltemp_landposinun + labs(title = NULL),
  airtemp_landpos + labs(title = NULL),
  airtemp_landposinun + labs(title = NULL),
  thaw_landpos + labs(title = NULL),
  thaw_landposinun + labs(title = NULL),
  VWC_landpos + labs(title = NULL),
  VWC_landposinun + labs(title = NULL),
  ncol = 2,
  nrow = 4,
  labels = "AUTO",  # Adds A, B, C, D... labels
  align = "hv"  # Aligns plots horizontally and vertically
)

print(combined_cowplot)
```



#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun.png",  # File name and extension
  plot = combined_plot,        # The plot to save (default is the last plot created)
  width = 20,                 # Width in inches
  height = 20,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```

#Adding column titles to the combined fig 
```{r}
library(patchwork)
library(ggplot2)

# First, remove individual titles from each plot
soiltemp_landpos <- soiltemp_landpos + theme(plot.title = element_blank())
soiltemp_landposinun <- soiltemp_landposinun + theme(plot.title = element_blank())
airtemp_landpos <- airtemp_landpos + theme(plot.title = element_blank())
airtemp_landposinun <- airtemp_landposinun + theme(plot.title = element_blank())
thaw_landpos <- thaw_landpos + theme(plot.title = element_blank())
thaw_landposinun <- thaw_landposinun + theme(plot.title = element_blank())
VWC_landpos <- VWC_landpos + theme(plot.title = element_blank())
VWC_landposinun <- VWC_landposinun + theme(plot.title = element_blank())

# Create empty plots for column titles
title1 <- ggplot() + 
  annotate("text", x = 0.5, y = 0.5, label = "Landscape Position", size = 5, fontface = "bold") + 
  theme_void()

title2 <- ggplot() + 
  annotate("text", x = 0.5, y = 0.5, label = "Landscape Position and Inundation", size = 5, fontface = "bold") + 
  theme_void()

# Arrange plots with column titles
combined_plot2 <- (title1 + title2) /
                 (soiltemp_landpos + soiltemp_landposinun) /
                 (airtemp_landpos + airtemp_landposinun) /
                 (thaw_landpos + thaw_landposinun) /
                 (VWC_landpos + VWC_landposinun) +
                 plot_layout(heights = c(0.5, 1, 1, 1, 1)) #, widths = c(0.3, 3)) #widths not working, may adjust later 


# # Add overall bottom title if needed
# combined_plot2 <- combined_plot + 
#   plot_annotation(
#     caption = "Landscape Position",
#     theme = theme(plot.caption = element_text(hjust = 0.5, size = 12, face = "bold"))
#   )

combined_plot2
```
#Patchwork not working, try gridExtra
```{r}
# ===================================================================
# COMBINING PLOTS WITH GRIDEXTRA (Alternative to patchwork)
# ===================================================================

library(gridExtra)
library(grid)
library(ggplot2)

# First, remove individual titles from each plot
soiltemp_landpos <- soiltemp_landpos + theme(plot.title = element_blank())
soiltemp_landposinun <- soiltemp_landposinun + theme(plot.title = element_blank())
airtemp_landpos <- airtemp_landpos + theme(plot.title = element_blank())
airtemp_landposinun <- airtemp_landposinun + theme(plot.title = element_blank())
thaw_landpos <- thaw_landpos + theme(plot.title = element_blank())
thaw_landposinun <- thaw_landposinun + theme(plot.title = element_blank())
VWC_landpos <- VWC_landpos + theme(plot.title = element_blank())
VWC_landposinun <- VWC_landposinun + theme(plot.title = element_blank())

# Create text grobs for column titles
# textGrob creates text objects that can be placed in the grid
title1 <- textGrob("Landscape Position", 
                   gp = gpar(fontsize = 14, fontface = "bold"))

title2 <- textGrob("Landscape Position and Inundation", 
                   gp = gpar(fontsize = 14, fontface = "bold"))

# Arrange all plots in a grid with titles
# The layout is: 2 columns, 5 rows (1 title row + 4 plot rows)
combined_plot2 <- grid.arrange(
  title1, title2,                              # Row 1: Column headers
  soiltemp_landpos, soiltemp_landposinun,     # Row 2: Soil temp plots
  airtemp_landpos, airtemp_landposinun,       # Row 3: Air temp plots
  thaw_landpos, thaw_landposinun,             # Row 4: Thaw depth plots
  VWC_landpos, VWC_landposinun,               # Row 5: VWC plots
  ncol = 2,                                    # 2 columns
  nrow = 5,                                    # 5 rows total
  heights = c(0.5, 2, 2, 2, 2)                # Make title row shorter (0.5 units)
                                               # and each plot row taller (2 units each)
)

# Display the combined plot
combined_plot2


```

#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun2.png",  # File name and extension
  plot = combined_plot2,        # The plot to save (default is the last plot created)
  width = 20,                 # Width in inches
  height = 20,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```



#Stats
```{r}
# ===================================================================
# STEP 1: Sample Size Check and Data Filtering
# ===================================================================

# Load required packages
library(dplyr)
library(ggplot2)

# Count observations by landscape position (ignoring plot_type and inundation)
sample_sizes_landscape <- df_wide %>%
  group_by(landscape_position) %>%
  summarise(n = n()) %>%
  arrange(landscape_position)

print("Sample sizes by landscape position:")
print(sample_sizes_landscape)

# Check sample sizes by landscape position AND inundation status
sample_sizes_combo <- df_wide %>%
  group_by(landscape_position, inundated) %>%
  summarise(n = n(), .groups = 'drop') %>%
  arrange(landscape_position, inundated)

print("Sample sizes by landscape position × inundation:")
print(sample_sizes_combo)

# Filter out any combinations with n < 5 for the inundation analysis
# This should remove "slope + inundated=Y" if n < 5
df_filtered <- df_wide %>%
  group_by(landscape_position, inundated) %>%
  filter(n() >= 5) %>%
  ungroup()

# Verify what was removed
print("Combinations excluded (n<5):")
excluded <- anti_join(
  df_wide %>% distinct(landscape_position, inundated),
  df_filtered %>% distinct(landscape_position, inundated)
)
print(excluded)

#so need to exclude slope inundated with n = 2 
```

#Testing normality / homogeneity of var - by landscape position 
```{r}
#  test for LANDSCAPE POSITION ONLY (3 groups: lowland, slope, upland)

library(car)  # For Levene's test

# Create a list of environmental variables to test
env_vars <- c("soil_temp_10_cm", "air_temp", "thawdepth", "VWC")
var_names <- c("Soil Temperature (10cm)", "Air Temperature", 
               "Thaw Depth", "Volumetric Water Content")

# Initialize a dataframe to store assumption test results
assumption_results <- data.frame(
  Variable = var_names,
  Shapiro_p = NA,
  Levene_p = NA,
  Normal = NA,
  Homogeneous = NA
)

# Loop through each environmental variable and test assumptions
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]
  
  cat("\n==========================================\n")
  cat("Testing assumptions for:", var_names[i], "\n")
  cat("==========================================\n")
  
  # Extract the variable as a vector
  var_data <- df_wide[[var]]
  
  # NORMALITY TEST: Shapiro-Wilk
  # Note: Shapiro-Wilk can only handle n <= 5000
  # If you have more data, take a random sample
  if(length(var_data) > 5000) {
    shapiro_result <- shapiro.test(sample(var_data, 5000))
  } else {
    shapiro_result <- shapiro.test(var_data)
  }
  
  cat("Shapiro-Wilk test p-value:", shapiro_result$p.value, "\n")
  cat("  Interpretation:", ifelse(shapiro_result$p.value > 0.05, 
                                  "Data appear normal (p > 0.05)", 
                                  "Data are NOT normal (p <= 0.05)"), "\n")
  
  # HOMOGENEITY OF VARIANCE TEST: Levene's Test
  # Tests if variance is equal across groups
  formula_str <- paste(var, "~ landscape_position")
  levene_result <- leveneTest(as.formula(formula_str), data = df_wide)
  levene_p <- levene_result$`Pr(>F)`[1]
  
  cat("Levene's test p-value:", levene_p, "\n")
  cat("  Interpretation:", ifelse(levene_p > 0.05, 
                                  "Variances are homogeneous (p > 0.05)", 
                                  "Variances are NOT homogeneous (p <= 0.05)"), "\n")
  
  # Store results
  assumption_results$Shapiro_p[i] <- shapiro_result$p.value
  assumption_results$Levene_p[i] <- levene_p
  assumption_results$Normal[i] <- shapiro_result$p.value > 0.05
  assumption_results$Homogeneous[i] <- levene_p > 0.05
}

# Display summary table

cat("SUMMARY OF ASSUMPTION TESTS\n")

print(assumption_results)

# Decision: Which test to use?
cat("\nRECOMMENDATION:\n")
if(all(assumption_results$Normal) & all(assumption_results$Homogeneous)) {
  cat("✓ All assumptions met: Use ANOVA\n")
} else {
  cat("✗ Assumptions violated: Use Kruskal-Wallis (non-parametric)\n")
}

#all have homogeneity of var, but all not normal 
```

#check assumptions, with QQ-plots, landscape level only 
```{r}

# tests whether data meet assumptions for parametric tests
# (normality and homogeneity of variance) for each environmental variable
# across landscape positions (lowland, slope, upland)
# ===================================================================

# Load required packages
library(dplyr)      # For data manipulation
library(ggplot2)    # For creating plots
library(car)        # For Levene's test (homogeneity of variance)
library(gridExtra)  # For combining multiple plots
library(grid)       # For textGrob

# ===================================================================
# PART A: SET UP VARIABLES TO TEST
# ===================================================================

# Create a list of environmental variables we want to test
env_vars <- c("soil_temp_10_cm", "air_temp", "thawdepth", "VWC")

# Create readable names for plots and tables
var_names <- c("Soil Temperature (10cm)", 
               "Air Temperature", 
               "Thaw Depth", 
               "Volumetric Water Content")

# Create a dataframe to store all assumption test results
assumption_results <- data.frame(
  Variable = var_names,              # Variable name
  Shapiro_p = NA,                    # p-value from Shapiro-Wilk test
  Levene_p = NA,                     # p-value from Levene's test
  Normal = NA,                       # TRUE/FALSE: Is data normal?
  Homogeneous = NA,                  # TRUE/FALSE: Are variances equal?
  Recommended_Test = NA              # Which test should we use?
)

# ===================================================================
# PART B: CREATE QQ PLOTS FOR VISUAL ASSESSMENT OF NORMALITY
# ===================================================================

# Create an empty list to store all QQ plots
qq_plot_list <- list()

# Loop through each environmental variable to create QQ plots
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]        # Current variable name (e.g., "soil_temp_10_cm")
  var_label <- var_names[i] # Readable name (e.g., "Soil Temperature (10cm)")
  
  # Create QQ plot for this variable
  # geom_qq() plots theoretical quantiles vs sample quantiles
  # geom_qq_line() adds the reference line (what we'd expect if normal)
  qq_plot <- ggplot(df_wide, aes(sample = .data[[var]])) +
    geom_qq(alpha = 0.5, color = "steelblue") +        # Plot the points
    geom_qq_line(color = "red", linewidth = 1) +       # Add reference line
    labs(
      title = var_label,
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    ) +
    theme_bw() +                                        # Clean black/white theme
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      axis.title = element_text(size = 9)
    )
  
  # Store this plot in our list
  qq_plot_list[[i]] <- qq_plot
}

# Combine all QQ plots into a single figure
# This makes it easy to compare normality across all variables
combined_qq <- grid.arrange(
  grobs = qq_plot_list,    # Our list of QQ plots
  ncol = 2,                # Arrange in 2 columns
  nrow = 2,                # And 2 rows (4 plots total)
  top = textGrob("QQ Plots: Visual Assessment of Normality", 
                 gp = gpar(fontsize = 14, fontface = "bold"))
)

# Display the combined QQ plot
combined_qq


# ===================================================================
# PART C: FORMAL STATISTICAL TESTS OF ASSUMPTIONS
# ===================================================================

# Loop through each environmental variable and run formal tests
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("\n==========================================\n")
  cat("Testing assumptions for:", var_label, "\n")
  cat("==========================================\n")
  
  # Extract the variable as a vector (removing any NAs)
  var_data <- df_wide[[var]]
  var_data <- var_data[!is.na(var_data)]  # Remove missing values
  
  # -------------------------------------------------------------------
  # TEST 1: NORMALITY - Shapiro-Wilk Test
  # -------------------------------------------------------------------
  # If p > 0.05, we fail to reject H0 (data appear normal)
  # If p <= 0.05, we reject H0 (data are NOT normal)
  
  # Note: Shapiro-Wilk can only handle sample sizes <= 5000
  # For larger datasets, we test a random sample
  if(length(var_data) > 5000) {
    cat("  Note: Sample size > 5000, testing random sample of 5000\n")
    shapiro_result <- shapiro.test(sample(var_data, 5000))
  } else {
    shapiro_result <- shapiro.test(var_data)
  }
  
  # Display results
  cat("\n1. NORMALITY TEST (Shapiro-Wilk):\n")
  cat("   W statistic:", round(shapiro_result$statistic, 4), "\n")
  cat("   p-value:", format.pval(shapiro_result$p.value, digits = 3), "\n")
  
  # Interpret the result
  is_normal <- shapiro_result$p.value > 0.05
  if(is_normal) {
    cat("   ✓ Data appear NORMAL (p > 0.05)\n")
    cat("   → Points should fall along line in QQ plot\n")
  } else {
    cat("   ✗ Data are NOT NORMAL (p <= 0.05)\n")
    cat("   → Points deviate from line in QQ plot\n")
  }
  
  # -------------------------------------------------------------------
  # TEST 2: HOMOGENEITY OF VARIANCE - Levene's Test
  # -------------------------------------------------------------------
  # If p > 0.05, we fail to reject H0 (variances are homogeneous)
  # If p <= 0.05, we reject H0 (variances are NOT equal)
  
  # Create the formula for testing (e.g., "soil_temp_10_cm ~ landscape_position")
  formula_str <- paste(var, "~ landscape_position")
  
  # Run Levene's test
  levene_result <- leveneTest(as.formula(formula_str), data = df_wide)
  levene_p <- levene_result$`Pr(>F)`[1]  # Extract p-value
  levene_f <- levene_result$`F value`[1] # Extract F-statistic
  
  # Display results
  cat("\n2. HOMOGENEITY OF VARIANCE (Levene's Test):\n")
  cat("   F statistic:", round(levene_f, 4), "\n")
  cat("   p-value:", format.pval(levene_p, digits = 3), "\n")
  
  # Interpret the result
  is_homogeneous <- levene_p > 0.05
  if(is_homogeneous) {
    cat("   ✓ Variances are HOMOGENEOUS (p > 0.05)\n")
    cat("   → Groups have similar spread in boxplots\n")
  } else {
    cat("   ✗ Variances are NOT HOMOGENEOUS (p <= 0.05)\n")
    cat("   → Groups have different spread in boxplots\n")
  }
  
  # -------------------------------------------------------------------
  # DECISION: Which statistical test should we use?
  # -------------------------------------------------------------------
  cat("\n3. RECOMMENDATION:\n")
  if(is_normal & is_homogeneous) {
    cat("   → Use PARAMETRIC test (ANOVA with Tukey post-hoc)\n")
    recommended <- "ANOVA"
  } else {
    cat("   → Use NON-PARAMETRIC test (Kruskal-Wallis with Dunn post-hoc)\n")
    if(!is_normal) cat("      Reason: Data not normally distributed\n")
    if(!is_homogeneous) cat("      Reason: Variances not homogeneous\n")
    recommended <- "Kruskal-Wallis"
  }
  
  # Store all results in our summary table
  assumption_results$Shapiro_p[i] <- shapiro_result$p.value
  assumption_results$Levene_p[i] <- levene_p
  assumption_results$Normal[i] <- is_normal
  assumption_results$Homogeneous[i] <- is_homogeneous
  assumption_results$Recommended_Test[i] <- recommended
}

# ===================================================================
# PART D: CREATE SUMMARY TABLE OF ASSUMPTION TESTS
# ===================================================================

cat("SUMMARY TABLE: ASSUMPTION TEST RESULTS\n")

# Print the results table
print(assumption_results)

# Create a nicely formatted table for saving/reporting
library(knitr)  # For nice table formatting

# Format p-values to show more clearly
assumption_results_formatted <- assumption_results %>%
  mutate(
    Shapiro_p = format.pval(Shapiro_p, digits = 3, eps = 0.001),
    Levene_p = format.pval(Levene_p, digits = 3, eps = 0.001),
    Normal = ifelse(Normal, "Yes ✓", "No ✗"),
    Homogeneous = ifelse(Homogeneous, "Yes ✓", "No ✗")
  )

# Display formatted table
kable(assumption_results_formatted, 
      caption = "Statistical Assumption Tests for Environmental Variables")


# ===================================================================
# PART E: CREATE ADDITIONAL QQ PLOTS BY GROUP (OPTIONAL)
# ===================================================================

# Sometimes it's useful to look at QQ plots for each landscape position
# separately to see if normality differs by group

# Example: QQ plots for soil temperature by landscape position
qq_by_group <- ggplot(df_wide, aes(sample = soil_temp_10_cm)) +
  geom_qq(alpha = 0.5, color = "steelblue") +
  geom_qq_line(color = "red", linewidth = 1) +
  facet_wrap(~landscape_position, ncol = 3) +  # Separate panel for each group
  labs(
    title = "QQ Plots: Soil Temperature by Landscape Position",
    subtitle = "Check if normality holds within each group",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    strip.text = element_text(size = 10, face = "bold")
  )

# Display the grouped QQ plot
qq_by_group


#All not normally distributed, but have homogeneity of var --> use non-parametric 

```


# Landscape position only - Kruskall wallis with post hoc Dunn
```{r}

library(FSA)  # For Dunn's post-hoc test

# Create a dataframe to store all results
kw_results <- data.frame(
  Variable = var_names,
  Chi_squared = NA,
  df = NA,
  p_value = NA,
  Significant = NA
)

# Store post-hoc results separately as list 
dunn_results_list <- list()

# Loop through each environmental variable
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]
  
  cat("Kruskal-Wallis Test:", var_names[i], "\n")

  
  # Run Kruskal-Wallis test

  formula_str <- paste(var, "~ landscape_position")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_wide_filtered)
  
  # Extract results
  chi_sq <- kw_test$statistic
  df_val <- kw_test$parameter
  p_val <- kw_test$p.value
  
  # Store in results table
  kw_results$Chi_squared[i] <- chi_sq
  kw_results$df[i] <- df_val
  kw_results$p_value[i] <- p_val
  kw_results$Significant[i] <- ifelse(p_val < 0.05, "YES", "NO")
  
  # Print results
  cat("Chi-squared =", round(chi_sq, 3), "\n")
  cat("df =", df_val, "\n")
  cat("p-value =", format.pval(p_val, digits = 3), "\n")
  cat("Significant?", ifelse(p_val < 0.05, "YES (p < 0.05)", "NO (p >= 0.05)"), "\n")
  
  # If significant, run Dunn's post-hoc test
  if(p_val < 0.05) {
    cat("\n--- Running Dunn's post-hoc test ---\n")
    
    # Dunn's test for pairwise comparisons
    # method="bh" uses Benjamini-Hochberg correction for multiple comparisons
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_wide_filtered, 
                          method = "bh")
    
    # Store results
    dunn_results_list[[var_names[i]]] <- dunn_test$res
    
    # Print pairwise comparisons
    print(dunn_test$res)
    
    # Identify which groups are different
    sig_comparisons <- dunn_test$res %>%
      filter(P.adj < 0.05)
    
    if(nrow(sig_comparisons) > 0) {
      cat("\nSignificant pairwise differences (p < 0.05):\n")
      for(j in 1:nrow(sig_comparisons)) {
        cat("  ", sig_comparisons$Comparison[j], 
            " (adjusted p =", round(sig_comparisons$P.adj[j], 4), ")\n")
      }
    } else {
      cat("\nNo significant pairwise differences after correction\n")
    }
  } else {
    cat("\n(Skipping post-hoc test - overall test not significant)\n")
  }
}

# Display final summary table

cat("SUMMARY: KRUSKAL-WALLIS RESULTS\n")

print(kw_results)
```

#Landscape position only - Compact letter display and means 
```{r}

# Calculate means and standard deviations for each group
summary_stats_landpos <- df_wide_filtered %>%
  group_by(landscape_position) %>%
  summarise(
    soil_temp_mean = mean(soil_temp_10_cm, na.rm = TRUE),
    soil_temp_sd = sd(soil_temp_10_cm, na.rm = TRUE),
    soil_temp_se = sd(soil_temp_10_cm, na.rm = TRUE)/sqrt(sum(!is.na(soil_temp_10_cm))),
    air_temp_mean = mean(air_temp, na.rm = TRUE),
    air_temp_sd = sd(air_temp, na.rm = TRUE),
    air_temp_se = sd(air_temp, na.rm = TRUE)/sqrt(sum(!is.na(air_temp))),
    thaw_mean = mean(thawdepth, na.rm = TRUE),
    thaw_sd = sd(thawdepth, na.rm = TRUE),
    thaw_se = sd(thawdepth, na.rm = TRUE)/sqrt(sum(!is.na(thawdepth))),
    VWC_mean = mean(VWC, na.rm = TRUE),
    VWC_sd = sd(VWC, na.rm = TRUE),
    VWC_se = sd(VWC, na.rm = TRUE)/sqrt(sum(!is.na(VWC))),
    n = n()
  )

print("Summary Statistics by Landscape Position:")
print(summary_stats_landpos)


library(writexl)
write_xlsx(summary_stats_landpos, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos2_outlierremoved.xlsx")

# Create compact letter display for each variable
# This tells us which groups are significantly different
# Groups that share a letter are NOT significantly different

# Helper function to assign letters based on Dunn's test results
assign_letters <- function(dunn_results, var_name) {
  
  if(is.null(dunn_results[[var_name]])) {
    # No significant differences - all groups get same letter
    return(data.frame(
      landscape_position = c("lowland", "slope", "upland"),
      letters = c("a", "a", "a")
    ))
  }
  
  # Extract pairwise comparison results
  comparisons <- dunn_results[[var_name]]
  
  # Simple letter assignment logic
  # (For complex cases, you might want to use multcompView package)
  groups <- data.frame(
    landscape_position = c("lowland", "slope", "upland"),
    letters = c("", "", "")
  )

 }

# create a manual approach to assigning letters based on Dunn's test results
# Run this AFTER you see the Dunn's test results above
# manually assign letters based on which groups differ

# Example (adjust based on your actual results):
letter_assignments <- list(
  soil_temp = data.frame(
    landscape_position = c("lowland", "slope", "upland"),
    letters = c("a", "a", "a")  # Replace with actual groupings
  ),
  air_temp = data.frame(
    landscape_position = c("lowland", "slope", "upland"),
    letters = c("a", "a", "a")  # Replace with actual groupings
  ),
  thaw = data.frame(
    landscape_position = c("lowland", "slope", "upland"),
    letters = c("a", "ab", "b")  # Replace with actual groupings
  ),
  VWC = data.frame(
    landscape_position = c("lowland", "slope", "upland"),
    letters = c("a", "bc", "c")  # Replace with actual groupings
  )
)



#print("Letter assignments for plotting:")
print(letter_assignments)
```

#pub table w/o slope-inun & FCH4 outlier removed 
```{r}
#Table with FCH4 outlier & slope-inundated removed 
pub_table_landpos <- summary_stats_landpos %>%
  group_by(landscape_position) %>%
  summarise(
    N = n(),
    `Soil Temp (°C)` = paste0(sprintf("%.2f", soil_temp_mean), 
                                 " ± ", 
                                 sprintf("%.2f", soil_temp_se)),
    `Air Temp (°C)` = paste0(sprintf("%.2f", air_temp_mean), 
                                " ± ", 
                                sprintf("%.2f", air_temp_se)),
    `Thaw (cm)` = paste0(sprintf("%.2f", thaw_mean), 
                                " ± ", 
                                sprintf("%.2f", thaw_se)),
    `VWC (%)` = paste0(sprintf("%.2f", VWC_mean), 
                                 " ± ", 
                                 sprintf("%.2f", VWC_se)),
    .groups = 'drop'
  ) %>%
select(landscape_position, N,
         `Soil Temp (°C)`, `Air Temp (°C)`, `Thaw (cm)`, `VWC (%)`)

print(pub_table_landpos)

#Save
 library(writexl)
 write_xlsx(pub_table_landpos, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_pubtable2.xlsx")
 

#Rounding variable means to match # of digits reported in dataset 
 
pub_table_landpos2 <- summary_stats_landpos %>%
  group_by(landscape_position) %>%
  summarise(
    N = n(),
    `Soil Temp (°C)` = paste0(sprintf("%.1f", soil_temp_mean), 
                                 " ± ", 
                                 sprintf("%.1f", soil_temp_se)),
    `Air Temp (°C)` = paste0(sprintf("%.2f", air_temp_mean), 
                                " ± ", 
                                sprintf("%.2f", air_temp_se)),
    `Thaw (cm)` = paste0(sprintf("%.0f", thaw_mean), 
                                " ± ", 
                                sprintf("%.0f", thaw_se)),
    `VWC (%)` = paste0(sprintf("%.1f", VWC_mean), 
                                 " ± ", 
                                 sprintf("%.1f", VWC_se)),
    .groups = 'drop'
  ) %>%
select(landscape_position, N,
         `Soil Temp (°C)`, `Air Temp (°C)`, `Thaw (cm)`, `VWC (%)`)

print(pub_table_landpos2)

#Save
 library(writexl)
 write_xlsx(pub_table_landpos2, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_pubtable2_rounded.xlsx")

```
#********************
#Figures with stats assigned

#### denoting exclusion of inundated-slope from analyses by chaning transparency of the box
```{r}
#dataset with slope-inundated included for SI figure 

#Add transparency to inundated slope boxplot to denote it being left out of analyses
# Create a variable to control transparency
df_wide2 <- df_wide %>%
  filter(flux_CH4 <= 400) %>% #remove outlier 
  mutate(excluded = ifelse(landscape_position == "slope" & inundated == "Y", 
                           "excluded", 
                           "included"))
#include FCH outlier row 
df_wide3 <- df_wide %>%
  mutate(excluded = ifelse(landscape_position == "slope" & inundated == "Y", 
                           "excluded", 
                           "included"))

```


##### Var vs landscape position only - use df_wide_filtered here to exclude the inundated slope and FCH4 outlier from the means * 
```{r}

#thawdepth by landpos
thaw_landpos <- ggplot(df_wide_filtered, aes(x = landscape_position, y = thawdepth, fill = landscape_position)) +
  geom_boxplot() +
  labs(title = "Observed Thaw Depth by Landscape Position",
       x = "Landscape Position",
       y = "Thaw Depth (cm)") +
  theme_bw(base_size = 15) +
    labs(fill = "Landscape Position") +
    stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black") +
  annotate("text", x = "lowland", y = 140, label = "a", fontface = "bold", size = 5) +
  annotate("text", x = "slope", y = 140, label = "ab", fontface = "bold", size = 5) +
    annotate("text", x = "upland", y = 140, label = "b", fontface = "bold", size = 5) 
  
  
thaw_landpos


#VWC by landpos 
VWC_landpos <- ggplot(df_wide_filtered, aes(x = landscape_position, y = VWC, fill = landscape_position)) +
  geom_boxplot() +
    labs(title = "Observed VWC by Landscape Position",
       x = "Landscape Position",
       y = "SWC (%)") +
  theme_bw(base_size = 15) +
    labs(fill = "Landscape Position") +
   stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black") +
  annotate("text", x = "lowland", y = 130, label = "a", fontface = "bold", size = 5) +
  annotate("text", x = "slope", y = 130, label = "b", fontface = "bold", size = 5) +
    annotate("text", x = "upland", y = 130, label = "b", fontface = "bold", size = 5) 

VWC_landpos



#soil temp by landpos 
soiltemp_landpos <- ggplot(df_wide_filtered, aes(x = landscape_position, y = soil_temp_10_cm, fill = landscape_position)) +
  geom_boxplot() +
    labs(title = "Observed Soil Temp by Landscape Position",
       x = "Landscape Position",
       y = "Soil Temp (C)") +
  theme_bw(base_size = 15) +
    labs(fill = "Landscape Position") +
  stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black")

soiltemp_landpos



#air temp by landpos 
airtemp_landpos <- ggplot(df_wide_filtered, aes(x = landscape_position, y = air_temp, fill = landscape_position)) +
  geom_boxplot() +
    labs(title = "Observed Air Temp by Landscape Position",
       x = "Landscape Position",
       y = "Air Temp (C)") +
  theme_bw(base_size = 15) +
  labs(fill = "Landscape Position") +
  stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black")

airtemp_landpos



  
```
#stats - landpos-inun

#### summary - landpos-inun

```{r}
#slope-inun and FCH4 outlier removed 
summary_inundation <- df_wide_filtered %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    soil_temp_mean = mean(soil_temp_10_cm, na.rm = TRUE),
    soil_temp_sd = sd(soil_temp_10_cm, na.rm = TRUE),
    soil_temp_se = sd(soil_temp_10_cm, na.rm = TRUE)/sqrt(sum(!is.na(soil_temp_10_cm))),
    air_temp_mean = mean(air_temp, na.rm = TRUE),
    air_temp_sd = sd(air_temp, na.rm = TRUE),
    air_temp_se = sd(air_temp, na.rm = TRUE)/sqrt(sum(!is.na(air_temp))),
    thaw_mean = mean(thawdepth, na.rm = TRUE),
    thaw_sd = sd(thawdepth, na.rm = TRUE),
    thaw_se = sd(thawdepth, na.rm = TRUE)/sqrt(sum(!is.na(thawdepth))),
    VWC_mean = mean(VWC, na.rm = TRUE),
    VWC_sd = sd(VWC, na.rm = TRUE),
    VWC_se = sd(VWC, na.rm = TRUE)/sqrt(sum(!is.na(VWC))),
    n = n()
  )
 
print(summary_inundation)

#Save
 library(writexl)
 write_xlsx(summary_inundation, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_inun2.xlsx")
```
#pub table w/o slope-inun & FCH4 outlier removed 
```{r}
#Table with FCH4 outlier & slope-inundated removed 
pub_table_landposinun <- summary_inundation %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    N = n(),
    `Soil Temp (°C)` = paste0(sprintf("%.2f", soil_temp_mean), 
                                 " ± ", 
                                 sprintf("%.2f", soil_temp_se)),
    `Air Temp (°C)` = paste0(sprintf("%.2f", air_temp_mean), 
                                " ± ", 
                                sprintf("%.2f", air_temp_se)),
    `Thaw (cm)` = paste0(sprintf("%.2f", thaw_mean), 
                                " ± ", 
                                sprintf("%.2f", thaw_se)),
    `VWC (%)` = paste0(sprintf("%.2f", VWC_mean), 
                                 " ± ", 
                                 sprintf("%.2f", VWC_se)),
    .groups = 'drop'
  ) %>%
select(landscape_position, inundated, N,
         `Soil Temp (°C)`, `Air Temp (°C)`, `Thaw (cm)`, `VWC (%)`)

print(pub_table_landposinun)

#Save
 library(writexl)
 write_xlsx(pub_table_landposinun, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_inun2_outlierremoved.xlsx")

 #------------- variables rounded to match formatting reported in dataset ----------------
 
#Table with FCH4 outlier & slope-inundated removed 
pub_table_landposinun2 <- summary_inundation %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    N = n(),
    `Soil Temp (°C)` = paste0(sprintf("%.1f", soil_temp_mean), 
                                 " ± ", 
                                 sprintf("%.1f", soil_temp_se)),
    `Air Temp (°C)` = paste0(sprintf("%.2f", air_temp_mean), 
                                " ± ", 
                                sprintf("%.2f", air_temp_se)),
    `Thaw (cm)` = paste0(sprintf("%.0f", thaw_mean), 
                                " ± ", 
                                sprintf("%.0f", thaw_se)),
    `VWC (%)` = paste0(sprintf("%.1f", VWC_mean), 
                                 " ± ", 
                                 sprintf("%.1f", VWC_se)),
    .groups = 'drop'
  ) %>%
select(landscape_position, inundated, N,
         `Soil Temp (°C)`, `Air Temp (°C)`, `Thaw (cm)`, `VWC (%)`)

print(pub_table_landposinun2)

#Save
 library(writexl)
 write_xlsx(pub_table_landposinun2, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_inun2_outlierremoved_rounded.xlsx")
```


#pub table - WITH inundated slope and FCH4 outlier - ignore for now
```{r}
#Table with the inundated slope included & FCH4 outlier 
pub_table2 <- df_wide3 %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    N = n(),
    `Soil Temp (°C)` = paste0(sprintf("%.2f", soil_temp_mean), 
                                 " ± ", 
                                 sprintf("%.2f", soil_temp_se)),
    `Air Temp (°C)` = paste0(sprintf("%.2f", air_temp_mean), 
                                " ± ", 
                                sprintf("%.2f", air_temp_se)),
    `Thaw (cm)` = paste0(sprintf("%.2f", thaw_mean), 
                                " ± ", 
                                sprintf("%.2f", thaw_se)),
    `VWC (%)` = paste0(sprintf("%.2f", VWC_mean), 
                                 " ± ", 
                                 sprintf("%.2f", VWC_se)),
    .groups = 'drop'
  ) %>%
select(landscape_position, inundated, N,
         `Soil Temp (°C)`, `Air Temp (°C)`, `Thaw (cm)`, `VWC (%)`)

print(pub_table2)

#Save
 library(writexl)
 write_xlsx(pub_table, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos_inun2.xlsx")
```


#Breaking up the tests 

inundated: differences in var among lowland, slope, upland; not-inundated: differences in var among lowland, slope, upland; 
then 
upland: differences in var by inundated, no inundated; lowland: differences in var by inundated and not inundated (excluding slope from here since it only has a non-inundated category)
 
#TO DO: change "median" to mean in script below ** 

streamlined code with help from claude 
```{r}
# ===================================================================
# BREAKDOWN ANALYSIS: LANDSCAPE × INUNDATION INTERACTIONS
# ===================================================================
# This script breaks down the landscape × inundation interaction into
# specific comparisons to understand WHERE differences occur:
# 1) Within each inundation status: do landscape positions differ?
# 2) Within each landscape position: does inundation matter?
# ===================================================================

library(dplyr)
library(FSA)        # For Dunn's test
library(ggplot2)

# Define variables
env_vars <- c("soil_temp_10_cm", "air_temp", "thawdepth", "VWC")
var_names <- c("Soil Temperature (10cm)", 
               "Air Temperature", 
               "Thaw Depth", 
               "Volumetric Water Content")
var_short <- c("Soil Temp", "Air Temp", "Thaw Depth", "VWC")

# ===================================================================
# PART A: SUBSET 1 - INUNDATED SITES ONLY
# ===================================================================
# Question: Among INUNDATED sites, do landscape positions differ?




cat("║  ANALYSIS 1: INUNDATED SITES ONLY                            ║\n")
cat("║  Testing: Do landscape positions differ in inundated areas?  ║\n")


# Filter to only inundated sites
df_inundated <- df_wide_filtered %>%
  filter(inundated == "Y") %>%
  group_by(landscape_position) %>%
  filter(n() >= 5) %>%  # Exclude groups with n < 5
  ungroup()

# Check sample sizes
cat("Sample sizes for INUNDATED sites:\n")
inundated_n <- df_inundated %>%
  count(landscape_position) %>%
  arrange(landscape_position)
print(inundated_n)
cat("\n")

# Test each environmental variable
for(i in 1:length(env_vars)) {
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("──────────────────────────────────────────────────────────────\n")
  cat("Variable:", var_label, "\n")
  cat("──────────────────────────────────────────────────────────────\n")
  
  # Only run test if we have more than 2 groups
  n_groups <- df_inundated %>% 
    distinct(landscape_position) %>% 
    nrow()
  
  if(n_groups < 2) {
    cat("⚠ Not enough groups to test (need at least 2)\n\n")
    next
  }
  
  # Kruskal-Wallis test
  formula_str <- paste(var, "~ landscape_position")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_inundated)
  
  cat("Kruskal-Wallis: χ² =", round(kw_test$statistic, 3), 
      ", p =", format.pval(kw_test$p.value, digits = 3), "\n")
  
  # If significant and more than 2 groups, run Dunn's test
  if(kw_test$p.value < 0.05 && n_groups > 2) {
    cat("✓ Significant! Running Dunn's post-hoc...\n\n")
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_inundated, 
                          method = "bh")
    print(dunn_test$res)
  } else if(kw_test$p.value < 0.05) {
    cat("✓ Significant difference between the two groups\n")
  } else {
    cat("✗ No significant differences\n")
  }
  cat("\n")
}

# ===================================================================
# PART B: SUBSET 2 - NON-INUNDATED SITES ONLY
# ===================================================================
# Question: Among NON-INUNDATED sites, do landscape positions differ?



cat("║  ANALYSIS 2: NON-INUNDATED SITES ONLY                        ║\n")
cat("║  Testing: Do landscape positions differ in dry areas?        ║\n")


# Filter to only non-inundated sites
df_not_inundated <- df_wide_filtered %>%
  filter(inundated == "N")

# Check sample sizes
cat("Sample sizes for NON-INUNDATED sites:\n")
not_inundated_n <- df_not_inundated %>%
  count(landscape_position) %>%
  arrange(landscape_position)
print(not_inundated_n)
cat("\n")

# Test each environmental variable
for(i in 1:length(env_vars)) {
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("──────────────────────────────────────────────────────────────\n")
  cat("Variable:", var_label, "\n")
  cat("──────────────────────────────────────────────────────────────\n")
  
  # Kruskal-Wallis test
  formula_str <- paste(var, "~ landscape_position")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_not_inundated)
  
  cat("Kruskal-Wallis: χ² =", round(kw_test$statistic, 3), 
      ", p =", format.pval(kw_test$p.value, digits = 3), "\n")
  
  # If significant, run Dunn's test
  if(kw_test$p.value < 0.05) {
    cat("✓ Significant! Running Dunn's post-hoc...\n\n")
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_not_inundated, 
                          method = "bh")
    print(dunn_test$res)
  } else {
    cat("✗ No significant differences\n")
  }
  cat("\n")
}

# ===================================================================
# PART C: SUBSET 3 - LOWLAND SITES ONLY
# ===================================================================
# Question: Within LOWLAND, does inundation status matter?
# Test: Mann-Whitney U (Wilcoxon rank-sum test) -- kruskall-wallis needs at least 3 levels


cat("║  ANALYSIS 3: LOWLAND SITES ONLY                              ║\n")
cat("║  Testing: Does inundation affect lowland environments?       ║\n")


# Filter to only lowland sites
df_lowland <- df_wide_filtered %>%
  filter(landscape_position == "lowland")

# Check sample sizes
cat("Sample sizes for LOWLAND sites:\n")
lowland_n <- df_lowland %>%
  count(inundated) %>%
  arrange(inundated)
print(lowland_n)
cat("\n")

# Test each environmental variable
# Use Wilcoxon test (Mann-Whitney U) for 2-group comparison
for(i in 1:length(env_vars)) {
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("──────────────────────────────────────────────────────────────\n")
  cat("Variable:", var_label, "\n")
  cat("──────────────────────────────────────────────────────────────\n")
  
  # Wilcoxon rank-sum test (Mann-Whitney U)
  # This is the 2-group version of Kruskal-Wallis
  formula_str <- paste(var, "~ inundated")
  wilcox_test <- wilcox.test(as.formula(formula_str), data = df_lowland)
  
  cat("Wilcoxon rank-sum test:\n")
  cat("  W =", wilcox_test$statistic, "\n")
  cat("  p =", format.pval(wilcox_test$p.value, digits = 3), "\n")
  
  # Calculate medians for interpretation
  medians <- df_lowland %>%
    group_by(inundated) %>%
    summarise(median = median(.data[[var]], na.rm = TRUE))
  
  cat("\n  Medians:\n")
  print(medians)
  
  if(wilcox_test$p.value < 0.05) {
    cat("\n  ✓ SIGNIFICANT difference between inundated and non-inundated lowlands\n")
    
    # Determine direction of effect
    inundated_median <- medians$median[medians$inundated == "Y"]
    not_inundated_median <- medians$median[medians$inundated == "N"]
    
    if(inundated_median > not_inundated_median) {
      cat("  → Inundated lowlands have HIGHER values\n")
    } else {
      cat("  → Inundated lowlands have LOWER values\n")
    }
  } else {
    cat("\n  ✗ No significant difference\n")
  }
  cat("\n")
}

# ===================================================================
# PART D: SUBSET 4 - UPLAND SITES ONLY
# ===================================================================
# Question: Within UPLAND, does inundation status matter?
# Test: Mann-Whitney U (Wilcoxon rank-sum test)


cat("║  ANALYSIS 4: UPLAND SITES ONLY                               ║\n")
cat("║  Testing: Does inundation affect upland environments?        ║\n")


# Filter to only upland sites
df_upland <- df_wide_filtered %>%
  filter(landscape_position == "upland")

# Check sample sizes
cat("Sample sizes for UPLAND sites:\n")
upland_n <- df_upland %>%
  count(inundated) %>%
  arrange(inundated)
print(upland_n)
cat("\n")

# Test each environmental variable
for(i in 1:length(env_vars)) {
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("──────────────────────────────────────────────────────────────\n")
  cat("Variable:", var_label, "\n")
  cat("──────────────────────────────────────────────────────────────\n")
  
  # Wilcoxon rank-sum test
  formula_str <- paste(var, "~ inundated")
  wilcox_test <- wilcox.test(as.formula(formula_str), data = df_upland)
  
  cat("Wilcoxon rank-sum test:\n")
  cat("  W =", wilcox_test$statistic, "\n")
  cat("  p =", format.pval(wilcox_test$p.value, digits = 3), "\n")
  
  # Calculate medians
  medians <- df_upland %>%
    group_by(inundated) %>%
    summarise(median = median(.data[[var]], na.rm = TRUE))
  
  cat("\n  Medians:\n")
  print(medians)
  
  if(wilcox_test$p.value < 0.05) {
    cat("\n  ✓ SIGNIFICANT difference between inundated and non-inundated uplands\n")
    
    # Determine direction of effect
    inundated_median <- medians$median[medians$inundated == "Y"]
    not_inundated_median <- medians$median[medians$inundated == "N"]
    
    if(inundated_median > not_inundated_median) {
      cat("  → Inundated uplands have HIGHER values\n")
    } else {
      cat("  → Inundated uplands have LOWER values\n")
    }
  } else {
    cat("\n  ✗ No significant difference\n")
  }
  cat("\n")
}


```




#Figure - adding stats 

#looking at variables  across landscape positions & inundation
```{r}
#use df_wide3 which includes both the outlier and slope-inun for SI figure 

#Here is where you use the alpha level 

#To add stat labels: 

# Create letter labels for thaw depth
# N panel: lowland = a, slope = ab, upland = b
# Y panel: no letters (no significant differences)
# thaw_letters <- data.frame(
#   landscape_position = c("lowland", "slope", "upland"),  # Only N panel
#   inundated = c("N", "N", "N"),                          # Only N panel
#   label = c("a", "ab", "b"),                             # Your statistical results
#   y_position = c(140, 140, 140)                          # Position at top
# )


 #Letters for N panel
thaw_letters_only <- data.frame(
  landscape_position = c("lowland", "slope", "upland"),
  inundated = c("N", "N", "N"),
  label = c("a", "ab", "b"),
  y_position = c(140, 140, 140)
)

# Asterisks for Y panel - BOTH lowland and upland
thaw_asterisk <- data.frame(
  landscape_position = c("lowland", "upland"),  # Both positions
  inundated = c("Y", "Y"),                      # Both in Y panel
  label = c("*", "*"),                          # Both get asterisks
  y_position = c(135, 135)                      # Same height
)

thaw_landposinun <- ggplot(df_wide3, aes(x = landscape_position, y = thawdepth, fill = landscape_position, alpha = excluded)) +
  geom_boxplot() +
  facet_wrap(~inundated) +
  # Hide the alpha legend but keep the fill legend
  scale_alpha_manual(values = c("excluded" = 0.3, "included" = 1.0), 
                     guide = "none") +  # This hides the include/excluded legend
  theme_bw(base_size = 15) +
    stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black", alpha = 1) +
  # Add letters (regular size)
  geom_text(data = thaw_letters_only, 
            aes(x = landscape_position, y = y_position, label = label),
            fontface = "bold", 
            size = 5,
            alpha = 1, #explicitly set to solid 
            inherti.aes = FALSE)+ #don't inherit alpha transparency from main aes 
  # Add asterisk (bigger size)
  geom_text(data = thaw_asterisk, 
            aes(x = landscape_position, y = y_position, label = label),
            fontface = "bold", 
            size = 8,  # Bigger! Adjust this number as needed
            alpha = 1,  # Explicitly set to solid
            inherit.aes = FALSE) +  # Don't inherit alpha from main aes
  labs(title = "Observed thawdepth by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Thaw Depth (cm)",
       fill = "Landscape Position")


thaw_landposinun



# ===================================================================
# VWC FIGURE WITH SIGNIFICANCE LETTERS
# ===================================================================

# Create letter labels for VWC
# N panel: lowland = a, slope = b, upland = b
# Y panel: no letters (no significant differences)
# vwc_letters <- data.frame(
#   landscape_position = c("lowland", "slope", "upland"),  # Only N panel
#   inundated = c("N", "N", "N"),                          # Only N panel
#   label = c("a", "b", "b"),                              # Your statistical results
#   y_position = c(90, 90, 90)                             # Adjust if needed
# )


#Letters for N panel
vwc_letters_only <- data.frame(
  landscape_position = c("lowland", "slope", "upland"),
  inundated = c("N", "N", "N"),
  label = c("a", "b", "b"),
  y_position = c(99, 99, 99) 
)

# Asterisks for Y panel - BOTH lowland and upland
vwc_asterisk <- data.frame(
  landscape_position = c("lowland", "upland"),  # Both positions
  inundated = c("Y", "Y"),                      # Both in Y panel
  label = c("*", "*"),                          # Both get asterisks
  y_position = c(97, 97)                      # Same height
)


VWC_landposinun <- ggplot(df_wide3, aes(x = landscape_position, y = VWC, fill = landscape_position, alpha = excluded)) +
  geom_boxplot() +
  facet_wrap(~inundated) +
    # Hide the alpha legend but keep the fill legend
  scale_alpha_manual(values = c("excluded" = 0.3, "included" = 1.0), 
                     guide = "none") +  # This hides the include/excluded legend
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "black", alpha = 1) +
  # Add letters only to N panel
  geom_text(data = vwc_letters_only, 
            aes(x = landscape_position, y = y_position, label = label),
            fontface = "bold", 
            size = 5,
            alpha = 1,  # Explicitly set to solid
            inherit.aes = FALSE) +  # Don't inherit alpha from main aes
  # Add asterisk (bigger size)
  geom_text(data = vwc_asterisk, 
            aes(x = landscape_position, y = y_position, label = label),
            fontface = "bold", 
            size = 8,  # Bigger! Adjust this number as needed
            alpha = 1,  # Explicitly set to solid
            inherit.aes = FALSE) +  # Don't inherit alpha from main aes
  labs(title = "Observed VWC by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "SWC (%)",
       fill = "Landscape Position") +
  theme_bw(base_size = 15)+
  scale_y_continuous(limits =c(0,100))

VWC_landposinun


#========= soil temp ================

# Asterisks for Y panel - BOTH lowland and upland
soilT_asterisk <- data.frame(
  landscape_position = c("lowland"),  # Both positions
  inundated = c("Y"),                      # Both in Y panel
  label = c("*"),                          # Both get asterisks
  y_position = c(15)                      # Same height
)


#soiltemp by landpos and inundated 
soiltemp_landposinun <- ggplot(df_wide3, aes(x = landscape_position, y = soil_temp_10_cm, fill = landscape_position, alpha = excluded)) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    # Hide the alpha legend but keep the fill legend
  scale_alpha_manual(values = c("excluded" = 0.3, "included" = 1.0), 
                     guide = "none") +  # This hides the include/excluded legend
  # Add asterisk (bigger size)
  geom_text(data = soilT_asterisk, 
            aes(x = landscape_position, y = y_position, label = label),
            fontface = "bold", 
            size = 8,  # Bigger! Adjust this number as needed
            alpha = 1,  # Explicitly set to solid
            inherit.aes = FALSE) +  # Don't inherit alpha from main aes
    labs(title = "Observed Soil Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Soil Temp (C)",
       fill = "Landscape Position") +
  theme_bw(base_size = 15) +
  stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black", alpha = 1) 
soiltemp_landposinun



#airtemp by landpos and inundated 
airtemp_landposinun <- ggplot(df_wide3, aes(x = landscape_position, y = air_temp, fill = landscape_position, alpha = excluded )) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    # Hide the alpha legend but keep the fill legend
  scale_alpha_manual(values = c("excluded" = 0.3, "included" = 1.0), 
                     guide = "none") +  # This hides the include/excluded legend
    labs(title = "Observed Air Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Air Temp (C)",
       fill = "Landscape Position") +
  theme_bw() +
  theme_bw(base_size = 15) +
    labs(fill = "Landscape Position") +
    stat_summary(fun=mean, geom="point", shape=23, size=4, fill = "black", alpha = 1)

airtemp_landposinun


```
 

#Final image for paper 

#Full combined plot with landpos, and landpos inun 

```{r}
library(cowplot)

# Remove legends and titles from all plots
soiltemp_landpos <- soiltemp_landpos + 
  theme(plot.title = element_blank()) #, legend.position = "none")
soiltemp_landposinun <- soiltemp_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
airtemp_landpos <- airtemp_landpos + 
  theme(plot.title = element_blank()) #, legend.position = "none")
airtemp_landposinun <- airtemp_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
thaw_landpos <- thaw_landpos + 
  theme(plot.title = element_blank()) #, legend.position = "none")
thaw_landposinun <- thaw_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
VWC_landpos <- VWC_landpos + 
  theme(plot.title = element_blank()) #, legend.position = "none")
VWC_landposinun <- VWC_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")

# Extract legend
# soiltemp_with_legend <- soiltemp_landpos + 
#   theme(legend.position = "right")
# shared_legend <- get_legend(soiltemp_with_legend)

# Create column titles
title1 <- ggdraw() + 
  draw_label("Landscape Position", fontface = "bold", size = 14)
title2 <- ggdraw() + 
  draw_label("Landscape Position and Inundation", fontface = "bold", size = 14)

# Arrange plots in grid
plot_grid <- plot_grid(  
  title1, title2,
  soiltemp_landpos, soiltemp_landposinun,
  airtemp_landpos, airtemp_landposinun,
  thaw_landpos, thaw_landposinun,
  VWC_landpos, VWC_landposinun,
  ncol = 2,
  rel_heights = c(0.3, 1, 1, 1, 1)
)

# Add legend to the right
# combined_plot2 <- plot_grid(
#   plot_grid, 
#   shared_legend,
#   ncol = 2,
#   rel_widths = c(10, 1)
# )

#combined_plot2

plot_grid

```



```{r}
library(cowplot)

# Remove legends and titles from ALL plots
soiltemp_landpos <- soiltemp_landpos + 
  theme(plot.title = element_blank(), legend.position = "none")
soiltemp_landposinun <- soiltemp_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
airtemp_landpos <- airtemp_landpos + 
  theme(plot.title = element_blank() , legend.position = "none")
airtemp_landposinun <- airtemp_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
thaw_landpos <- thaw_landpos + 
  theme(plot.title = element_blank() , legend.position = "none")
thaw_landposinun <- thaw_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")
VWC_landpos <- VWC_landpos + 
  theme(plot.title = element_blank() , legend.position = "none")
VWC_landposinun <- VWC_landposinun + 
  theme(plot.title = element_blank(), legend.position = "none")

# Create column titles using ggdraw
title1 <- ggdraw() + 
  draw_label("Landscape Position", fontface = "bold", size = 14)

title2 <- ggdraw() + 
  draw_label("Landscape Position and Inundation", fontface = "bold", size = 14)

# Arrange all plots in grid - no legend needed!
combined_plot2 <- plot_grid(
  title1, title2,
  soiltemp_landpos, soiltemp_landposinun,
  airtemp_landpos, airtemp_landposinun,
  thaw_landpos, thaw_landposinun,
  VWC_landpos, VWC_landposinun,
  ncol = 2,
  nrow = 5,
  rel_heights = c(0.3, 1, 1, 1, 1)  # Titles shorter, plots equal height
)

combined_plot2

# Save (commented out)
# ggsave("environmental_conditions_combined.png",
#        combined_plot2,
#        width = 12,
#        height = 16,
#        dpi = 300)

```


#save 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun3.png",  # File name and extension
  plot = combined_plot2,        # The plot to save (default is the last plot created)
  width = 20,                 # Width in inches
  height = 20,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)




```


#Combined plot final 

```{r}

# ===================================================================
# COMBINING PLOTS WITH GRIDEXTRA (Alternative to patchwork)
# ===================================================================

library(gridExtra)
library(grid)
library(ggplot2)

# First, remove individual titles from each plot
soiltemp_landpos <- soiltemp_landpos + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))             

soiltemp_landposinun <- soiltemp_landposinun + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

airtemp_landpos <- airtemp_landpos + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

airtemp_landposinun <- airtemp_landposinun + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

thaw_landpos <- thaw_landpos + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

thaw_landposinun <- thaw_landposinun + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

VWC_landpos <- VWC_landpos + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

VWC_landposinun <- VWC_landposinun + theme(plot.title = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(face = "bold"),text = element_text(color = "black"), axis.text = element_text(color = "black"), strip.text = element_text(face = "bold"))

# Create text grobs for column titles
# textGrob creates text objects that can be placed in the grid
title1 <- textGrob("Landscape Position", 
                   gp = gpar(fontsize = 20, fontface = "bold"))

title2 <- textGrob("Landscape Position and Inundation", 
                   gp = gpar(fontsize = 20, fontface = "bold"))

# Arrange all plots in a grid with titles
#The layout is: 2 columns, 5 rows (1 title row + 4 plot rows)
combined_plot2 <- grid.arrange(
  title1, title2,                              # Row 1: Column headers
  soiltemp_landpos, soiltemp_landposinun,     # Row 2: Soil temp plots
  airtemp_landpos, airtemp_landposinun,       # Row 3: Air temp plots
  thaw_landpos, thaw_landposinun,             # Row 4: Thaw depth plots
  VWC_landpos, VWC_landposinun,               # Row 5: VWC plots
  ncol = 2,                                    # 2 columns
  nrow = 5,                                    # 5 rows total
  heights = c(0.5, 2, 2, 2, 2),                # Make title row shorter (0.5 units) # and each plot row taller (2 units each)
   widths = c(0.7, 1)  #ratio of right column to left column                                             
)

# Display the combined plot
combined_plot2

```

#save 
```{r}
#nolegend2 = redone with the adjusted landscape means to remove outlier and slope inun for SI fig 
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun_nolegend2.png",  # File name and extension
  plot = combined_plot2,        # The plot to save (default is the last plot created)
  width = 35,                 # Width in inches
  height = 35,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm",
  background = "white"# Units for width and height (can be "in", "cm", or "mm")
)
```

#END


#================================
#comprehensive stats streamlined from claude based on code above -- have not checked this over for correctness yet* 

```{r}
# ===================================================================
# STATISTICAL TESTS FOR LANDSCAPE POSITION AND INUNDATION
# ===================================================================
# This script runs statistical tests to determine if environmental
# variables differ among:
# 1) Landscape positions (lowland, slope, upland)
# 2) Landscape position × inundation status combinations
# ===================================================================

# Load required packages
library(dplyr)      # For data manipulation
library(ggplot2)    # For creating plots
library(FSA)        # For Dunn's post-hoc test
library(knitr)      # For formatted tables

# ===================================================================
# PART A: PREPARE DATA AND VARIABLES
# ===================================================================

# Define environmental variables to test
env_vars <- c("soil_temp_10_cm", "air_temp", "thawdepth", "VWC")

# Create readable names for output
var_names <- c("Soil Temperature (10cm)", 
               "Air Temperature", 
               "Thaw Depth", 
               "Volumetric Water Content")

# Short names for tables
var_short <- c("Soil Temp", "Air Temp", "Thaw Depth", "VWC")

# ===================================================================
# PART B: FILTER DATA TO EXCLUDE GROUPS WITH n < 5
# ===================================================================

cat("==========================================\n")
cat("STEP B: CHECKING SAMPLE SIZES\n")
cat("==========================================\n\n")

# First, check sample sizes for landscape position only
sample_sizes_landscape <- df_wide %>%
  group_by(landscape_position) %>%
  summarise(n = n()) %>%
  arrange(landscape_position)

cat("Sample sizes by landscape position:\n")
print(sample_sizes_landscape)

# Check sample sizes for landscape × inundation combinations
sample_sizes_combo <- df_wide %>%
  group_by(landscape_position, inundated) %>%
  summarise(n = n(), .groups = 'drop') %>%
  arrange(landscape_position, inundated)

cat("\nSample sizes by landscape position × inundation:\n")
print(sample_sizes_combo)

# Filter out combinations with n < 5
# This creates our dataset for the inundation analysis
df_filtered <- df_wide %>%
  group_by(landscape_position, inundated) %>%
  filter(n() >= 5) %>%
  ungroup()

# Check what was excluded
excluded_combos <- sample_sizes_combo %>%
  filter(n < 5)

if(nrow(excluded_combos) > 0) {
  cat("\n⚠ EXCLUDED combinations (n < 5):\n")
  print(excluded_combos)
} else {
  cat("\n✓ No combinations excluded (all n >= 5)\n")
}

cat("\nFinal sample size for inundation analysis:", nrow(df_filtered), "\n")

# ===================================================================
# PART C: ANALYSIS 1 - LANDSCAPE POSITION ONLY
# ===================================================================

cat("\n\n==========================================\n")
cat("ANALYSIS 1: LANDSCAPE POSITION ONLY\n")
cat("==========================================\n")
cat("Testing: Do environmental variables differ among\n")
cat("         lowland, slope, and upland positions?\n")
cat("Method:  Kruskal-Wallis test (non-parametric ANOVA)\n")
cat("==========================================\n")

# Create dataframe to store Kruskal-Wallis results
kw_results <- data.frame(
  Variable = var_short,
  Chi_squared = NA,
  df = NA,
  p_value = NA,
  Significant = NA
)

# Create a list to store Dunn's test results for each variable
dunn_results_landscape <- list()

# Loop through each environmental variable
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("\n------------------------------------------\n")
  cat("Variable:", var_label, "\n")
  cat("------------------------------------------\n")
  
  # -------------------------------------------------------------------
  # Run Kruskal-Wallis Test
  # -------------------------------------------------------------------

  
  formula_str <- paste(var, "~ landscape_position")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_wide)
  
  # Extract results
  chi_sq <- kw_test$statistic
  df_val <- kw_test$parameter
  p_val <- kw_test$p.value
  
  # Store results
  kw_results$Chi_squared[i] <- chi_sq
  kw_results$df[i] <- df_val
  kw_results$p_value[i] <- p_val
  kw_results$Significant[i] <- ifelse(p_val < 0.05, "YES", "NO")
  
  # Display results
  cat("Kruskal-Wallis Test:\n")
  cat("  χ² =", round(chi_sq, 3), "\n")
  cat("  df =", df_val, "\n")
  cat("  p-value =", format.pval(p_val, digits = 3), "\n")
  
  if(p_val < 0.05) {
    cat("  ✓ SIGNIFICANT (p < 0.05)\n")
    cat("  → At least one landscape position differs\n")
  } else {
    cat("  ✗ NOT SIGNIFICANT (p >= 0.05)\n")
    cat("  → No evidence of differences among landscape positions\n")
  }
  
  # -------------------------------------------------------------------
  # If significant, run Dunn's Post-hoc Test
  # -------------------------------------------------------------------
  if(p_val < 0.05) {
    cat("\n  ═══════════════════════════════════════\n")
    cat("  DUNN'S POST-HOC TEST FOR:", var_label, "\n")
    cat("  ═══════════════════════════════════════\n")
    
    # Dunn's test performs pairwise comparisons
    # method="bh" = Benjamini-Hochberg correction for multiple comparisons
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_wide, 
                          method = "bh")
    
    # Store results
    dunn_results_landscape[[var_short[i]]] <- dunn_test$res
    
    # Display pairwise comparisons
    cat("\n  Pairwise Comparisons (Dunn's test with BH correction):\n\n")
    dunn_table <- dunn_test$res %>%
      mutate(
        Z = round(Z, 3),
        P.unadj = format.pval(P.unadj, digits = 3),
        P.adj = format.pval(P.adj, digits = 3),
        Sig = ifelse(P.adj < 0.05, "***", "ns")
      )
    print(dunn_table)
    cat("\n  ═══════════════════════════════════════\n")
    
    # Summarize significant differences
    sig_pairs <- dunn_test$res %>% filter(P.adj < 0.05)
    
    if(nrow(sig_pairs) > 0) {
      cat("\n  ✓ Significant pairwise differences (p < 0.05):\n")
      for(j in 1:nrow(sig_pairs)) {
        cat("     •", sig_pairs$Comparison[j], 
            "(p =", round(sig_pairs$P.adj[j], 4), ")\n")
      }
    } else {
      cat("\n  Note: Overall test significant but no pairwise differences\n")
      cat("        after correction for multiple comparisons\n")
    }
  } else {
    cat("\n  (Skipping post-hoc test - overall test not significant)\n")
    dunn_results_landscape[[var_short[i]]] <- NULL
  }
}

# Display summary table of Kruskal-Wallis results
cat("\n\n==========================================\n")
cat("SUMMARY: LANDSCAPE POSITION ANALYSIS\n")
cat("==========================================\n\n")
print(kw_results)



# ===================================================================
# PART D: ANALYSIS 2 - LANDSCAPE POSITION × INUNDATION
# ===================================================================

cat("\n\n==========================================\n")
cat("ANALYSIS 2: LANDSCAPE POSITION × INUNDATION\n")
cat("==========================================\n")
cat("Testing: Do environmental variables differ among\n")
cat("         landscape-inundation combinations?\n")
cat("Method:  Kruskal-Wallis test on combined factor\n")
cat("Note:    Excluding slope-inundated (n < 5)\n")
cat("==========================================\n")

# Create a combined factor for landscape × inundation
# This allows us to test all combinations at once
df_filtered <- df_filtered %>%
  mutate(landscape_inundation = paste(landscape_position, 
                                      inundated, 
                                      sep = "-"))

# Show the combinations we're testing
cat("\nCombinations being tested:\n")
unique_combos <- df_filtered %>%
  count(landscape_inundation) %>%
  arrange(landscape_inundation)
print(unique_combos)

# Create dataframe to store results
kw_results_inundation <- data.frame(
  Variable = var_short,
  Chi_squared = NA,
  df = NA,
  p_value = NA,
  Significant = NA
)

# Create a list to store Dunn's test results
dunn_results_inundation <- list()

# Loop through each environmental variable
for(i in 1:length(env_vars)) {
  
  var <- env_vars[i]
  var_label <- var_names[i]
  
  cat("\n------------------------------------------\n")
  cat("Variable:", var_label, "\n")
  cat("------------------------------------------\n")
  
  # -------------------------------------------------------------------
  # Run Kruskal-Wallis Test on combined factor
  # -------------------------------------------------------------------
  formula_str <- paste(var, "~ landscape_inundation")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_filtered)
  
  # Extract results
  chi_sq <- kw_test$statistic
  df_val <- kw_test$parameter
  p_val <- kw_test$p.value
  
  # Store results
  kw_results_inundation$Chi_squared[i] <- chi_sq
  kw_results_inundation$df[i] <- df_val
  kw_results_inundation$p_value[i] <- p_val
  kw_results_inundation$Significant[i] <- ifelse(p_val < 0.05, "YES", "NO")
  
  # Display results
  cat("Kruskal-Wallis Test:\n")
  cat("  χ² =", round(chi_sq, 3), "\n")
  cat("  df =", df_val, "\n")
  cat("  p-value =", format.pval(p_val, digits = 3), "\n")
  
  if(p_val < 0.05) {
    cat("  ✓ SIGNIFICANT (p < 0.05)\n")
    cat("  → At least one landscape-inundation combination differs\n")
  } else {
    cat("  ✗ NOT SIGNIFICANT (p >= 0.05)\n")
    cat("  → No evidence of differences among combinations\n")
  }
  
  # -------------------------------------------------------------------
  # If significant, run Dunn's Post-hoc Test
  # -------------------------------------------------------------------
  if(p_val < 0.05) {
    cat("\n  ═══════════════════════════════════════\n")
    cat("  DUNN'S POST-HOC TEST FOR:", var_label, "\n")
    cat("  ═══════════════════════════════════════\n")
    cat("  (This may take a moment - many pairwise comparisons)\n")
    
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_filtered, 
                          method = "bh")
    
    # Store results
    dunn_results_inundation[[var_short[i]]] <- dunn_test$res
    
    # Display only significant pairwise comparisons
    sig_pairs <- dunn_test$res %>% 
      filter(P.adj < 0.05) %>%
      arrange(P.adj) %>%
      mutate(
        Z = round(Z, 3),
        P.adj = round(P.adj, 4)
      )
    
    if(nrow(sig_pairs) > 0) {
      cat("\n  ✓ Significant pairwise differences (p < 0.05):\n\n")
      print(sig_pairs %>% select(Comparison, Z, P.adj))
      
      cat("\n  Total significant pairs:", nrow(sig_pairs), 
          "out of", nrow(dunn_test$res), "\n")
    } else {
      cat("\n  Note: Overall test significant but no pairwise differences\n")
      cat("        after correction for multiple comparisons\n")
    }
    
    cat("  ═══════════════════════════════════════\n")
    
    # Save full Dunn's test results for this variable (commented out)
    # write.csv(dunn_test$res, 
    #           paste0("dunn_test_", var, "_inundation.csv"), 
    #           row.names = FALSE)
    
  } else {
    cat("\n  (Skipping post-hoc test - overall test not significant)\n")
    dunn_results_inundation[[var_short[i]]] <- NULL
  }
}

# Display summary table
cat("\n\n==========================================\n")
cat("SUMMARY: LANDSCAPE × INUNDATION ANALYSIS\n")
cat("==========================================\n\n")
print(kw_results_inundation)

# Save results (commented out)
# write.csv(kw_results_inundation, 
#           "kruskal_wallis_inundation_results.csv", 
#           row.names = FALSE)

# ===================================================================
# PART E: CALCULATE DESCRIPTIVE STATISTICS
# ===================================================================

cat("\n\n==========================================\n")
cat("DESCRIPTIVE STATISTICS\n")
cat("==========================================\n")

# Calculate means and SDs by landscape position
cat("\nBy Landscape Position:\n")
summary_landscape <- df_wide %>%
  group_by(landscape_position) %>%
  summarise(
    across(all_of(env_vars), 
           list(mean = ~mean(., na.rm = TRUE),
                sd = ~sd(., na.rm = TRUE)),
           .names = "{.col}_{.fn}"),
    n = n()
  )

print(summary_landscape)

# Save summary statistics (commented out)
# write.csv(summary_landscape, 
#           "summary_stats_landscape.csv", 
#           row.names = FALSE)

# Calculate means and SDs by landscape × inundation
cat("\n\nBy Landscape Position × Inundation:\n")
summary_inundation <- df_filtered %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    across(all_of(env_vars), 
           list(mean = ~mean(., na.rm = TRUE),
                sd = ~sd(., na.rm = TRUE)),
           .names = "{.col}_{.fn}"),
    n = n(),
    .groups = 'drop'
  ) %>%
  arrange(landscape_position, inundated)

print(summary_inundation)

# Save summary statistics (commented out)
# write.csv(summary_inundation, 
#           "summary_stats_inundation.csv", 
#           row.names = FALSE)

# ===================================================================
# PART F: CREATE LETTER GROUPINGS FOR PLOTS
# ===================================================================

cat("\n\n==========================================\n")
cat("CREATING LETTER GROUPINGS FOR PLOTS\n")
cat("==========================================\n")

# This function assigns letters (a, b, c, etc.) based on Dunn's test
# Groups that share a letter are NOT significantly different
assign_letters_from_dunn <- function(dunn_results, groups) {
  
  # If no Dunn results (test not significant), all groups get same letter
  if(is.null(dunn_results)) {
    letters_df <- data.frame(
      group = groups,
      letters = rep("a", length(groups))
    )
    return(letters_df)
  }
  
  # Initialize all groups with empty letters
  letter_assignment <- data.frame(
    group = groups,
    letters = "",
    stringsAsFactors = FALSE
  )
  
  # Get significant pairs (p < 0.05)
  sig_pairs <- dunn_results %>% filter(P.adj < 0.05)
  
  # Simple algorithm: assign letters based on pairwise differences
  # Groups that are NOT significantly different share letters
  
  current_letter <- 1  # Start with 'a'
  
  for(i in 1:length(groups)) {
    current_group <- groups[i]
    
    if(letter_assignment$letters[i] == "") {
      # Assign next available letter
      letter_assignment$letters[i] <- letters[current_letter]
      
      # Find all groups NOT significantly different from current group
      for(j in (i+1):length(groups)) {
        if(j > length(groups)) break
        
        compare_group <- groups[j]
        
        # Check if these two groups are significantly different
        comparison1 <- paste(current_group, compare_group, sep = " - ")
        comparison2 <- paste(compare_group, current_group, sep = " - ")
        
        is_different <- any(sig_pairs$Comparison %in% c(comparison1, comparison2))
        
        # If NOT different, assign same letter
        if(!is_different && letter_assignment$letters[j] == "") {
          letter_assignment$letters[j] <- letters[current_letter]
        }
      }
      
      current_letter <- current_letter + 1
    }
  }
  
  return(letter_assignment)
}

# Create letter assignments for landscape position
cat("\nLetter assignments for LANDSCAPE POSITION:\n")
cat("(Groups with same letter are not significantly different)\n\n")

landscape_letters_list <- list()

for(i in 1:length(env_vars)) {
  var_label <- var_short[i]
  
  cat(var_label, ":\n")
  
  groups <- c("lowland", "slope", "upland")
  letters_df <- assign_letters_from_dunn(dunn_results_landscape[[var_label]], 
                                         groups)
  
  colnames(letters_df) <- c("landscape_position", "letters")
  landscape_letters_list[[var_label]] <- letters_df
  
  print(letters_df)
  cat("\n")
}

# Create letter assignments for landscape × inundation
cat("\nLetter assignments for LANDSCAPE × INUNDATION:\n")
cat("(Groups with same letter are not significantly different)\n\n")

inundation_letters_list <- list()

for(i in 1:length(env_vars)) {
  var_label <- var_short[i]
  
  cat(var_label, ":\n")
  
  groups <- unique(df_filtered$landscape_inundation)
  letters_df <- assign_letters_from_dunn(dunn_results_inundation[[var_label]], 
                                         groups)
  
  colnames(letters_df) <- c("landscape_inundation", "letters")
  inundation_letters_list[[var_label]] <- letters_df
  
  print(letters_df)
  cat("\n")
}

# ===================================================================
# PART G: SAVE LETTER ASSIGNMENTS FOR PLOTTING
# ===================================================================

# These letter assignments can be merged with your plotting data
# to add significance labels to your boxplots

# Example: Save soil temperature letters
# write.csv(landscape_letters_list[["Soil Temp"]], 
#           "letters_soil_temp_landscape.csv", 
#           row.names = FALSE)


```

