---
title: "Soil_Chambers_Comprehensive_Flux_stats_landpos"
output: html_document
date: "2025-10-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#"Soil_Chamber_EnviroVariable_Figs" ##time is in AK daylight time** NEE/RECO/GPP are in gC/m2/s; FCO2 and FCH4 are in gC/m2/s, flux_CO2 is in umol/m2/s; flux_CH4 is in nano mol/m2/s

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')



```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs 
library(dplyr)

# Reshape the dataframe to long format, choose variables of interest 
df_wide <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) #exclude soil temp 15 cm since most of it is NA 

#make NEE, GPP, RECO, and FCH4 all g/m2/s -> umol/m2/s
df_wide$NEE_umol <- df_wide$NEE * 1/12 * 1000000
df_wide$GPP_umol <- df_wide$GPP * 1/12 * 1000000
df_wide$RECO_umol <- df_wide$RECO * 1/12 * 1000000
df_wide$FCH4_umol <- df_wide$FCH4 * 1/12 * 1000000


#Filter out inundated slope with only n=2
# Filter out combinations with n < 5
# This removes slope-inundated (slope-Y)
df_wide_filtered <- df_wide %>%
  group_by(landscape_position, inundated) %>%
 filter(flux_CH4 <= 400) %>% #filter out the outlier of flux CH4 = 600 nmol, occurs only one time in 2019 and rest are < 400 
  filter(n() >= 5) %>%
  ungroup()


#================ DF LONG ============================================
df_long2 <- df_wide_filtered %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) %>% #exclude soil temp 15 cm since most of it is NA 
  pivot_longer(cols = c(NEE, RECO, GPP, FCH4),
               names_to = "flux_type",
               values_to = "flux_value")

#df_long2$FCH4_umol <- df_long2$FCH4 * 1/12 * 1000000
df_long2$flux_umol <- df_long2$flux_value * 1/12 * 1000000


```


#looking at variables  across landscape positions & inundation
```{r}

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, color = flux_type)) +
  geom_boxplot(position = "dodge") +
  stat_summary(fun = mean, geom = "point", position = position_dodge(0.8), shape = 23, size = 4)+
    facet_wrap(~inundated) +
  labs(title = "Observed NEE by Land pos",
       x = "Landscape Position",
       y = "NEE (flux_value)") +
  theme_bw()

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, fill = flux_type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 4,
    position = position_dodge(0.8)  # same dodge width as boxplot
  ) +
  labs(
    title = "Observed NEE by Landscape Position",
    x = "Landscape Position",
    y = "Flux Value (umol/m2.s)",
    fill = "Flux Type"
  ) +
  theme_bw()

#filter(flux_value >= -6.5 & flux_value <= 6.5)


ggplot(df_long2 %>%
         filter(flux_umol >= -6.5 & flux_umol <= 6.5), 
       aes(x = landscape_position, y = flux_umol, fill = flux_type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 4,
    position = position_dodge(0.8)  # same dodge width as boxplot
  ) +
  labs(
    title = "Observed NEE by Landscape Position",
    x = "Landscape Position",
    y = "Flux Value (umol/m2.s)",
    fill = "Flux Type"
  ) +
  theme_bw()
```
#NEE, GPP, RECO

#Comprehensive stats (streamlined code with help from Claude)
```{r}

# FLUX ANALYSIS: LANDSCAPE POSITION DIFFERENCES

# This script tests whether CO2 and CH4 fluxes differ among
# landscape positions (lowland, slope, upland)
# 
# Fluxes analyzed:
# - RECO (Ecosystem Respiration) in ┬╡mol/m┬▓/s
# - GPP (Gross Primary Production) in ┬╡mol/m┬▓/s
# - NEE (Net Ecosystem Exchange) in ┬╡mol/m┬▓/s
# - FCH4 (Methane Flux) in ┬╡mol/m┬▓/s


library(dplyr)      # For data manipulation
library(ggplot2)    # For creating plots
library(car)        # For Levene's test
library(FSA)        # For Dunn's test
library(gridExtra)  # For combining plots
library(grid)       # For textGrob
library(knitr)      # For formatted tables

# ===================================================================
# STEP 1: CHECK SAMPLE SIZES AND FILTER DATA
# ===================================================================

# STEP 1: SAMPLE SIZE CHECK AND FILTERING 


# Check sample sizes by landscape position ├Ч inundation
cat("Sample sizes by landscape position ├Ч inundation:\n")
sample_sizes_combo_flux <- df_wide_filtered %>%
  group_by(landscape_position, inundated) %>%
  summarise(n = n(), .groups = 'drop') %>%
  arrange(landscape_position, inundated)

print(sample_sizes_combo_flux)
cat("\n")

# Filter out combinations with n < 5
# This removes slope-inundated (slope-Y)
# df_wide_filtered <- df_wide %>%
#   group_by(landscape_position, inundated) %>%
#   filter(flux_CH4 <= 400) %>%
#   filter(n() >= 5) %>%
#   ungroup()

# Check what was excluded
excluded_combos_flux <- sample_sizes_combo_flux %>%
  filter(n < 5)

if(nrow(excluded_combos_flux) > 0) {
  cat("тЪа EXCLUDED combinations (n < 5):\n")
  print(excluded_combos_flux)
  cat("\n")
} else {
  cat("тЬУ No combinations excluded (all n >= 5)\n\n")
}

cat("Original sample size:", nrow(df_wide), "\n")
cat("Filtered sample size:", nrow(df_wide_filtered), "\n")
cat("Observations removed:", nrow(df_wide) - nrow(df_wide_filtered), "\n\n")

# Sample sizes by landscape position after filtering
sample_sizes_flux <- df_wide_filtered %>%
  group_by(landscape_position) %>%
  summarise(n = n()) %>%
  arrange(landscape_position)

cat("Sample sizes by landscape position (after filtering):\n")
print(sample_sizes_flux)
cat("\n")

# ===================================================================
# STEP 2: TEST ASSUMPTIONS WITH QQ PLOTS OF RESIDUALS
# ===================================================================


# STEP 2: TEST STATISTICAL ASSUMPTIONS (RESIDUALS)   


# Define flux variables (using umol versions)
flux_vars <- c("RECO_umol", "GPP_umol", "NEE_umol", "FCH4_umol")

# Create readable names
flux_names <- c("Ecosystem Respiration (RECO)", 
                "Gross Primary Production (GPP)", 
                "Net Ecosystem Exchange (NEE)", 
                "Methane Flux (FCH4)")

# Short names for tables
flux_short <- c("RECO", "GPP", "NEE", "FCH4")

# Create assumption results table
assumption_results_flux <- data.frame(
  Variable = flux_names,
  Shapiro_p = NA,
  Levene_p = NA,
  Normal = NA,
  Homogeneous = NA,
  Recommended_Test = NA
)

# -------------------------------------------------------------------
# Create QQ plots for RESIDUALS (correct approach)
# -------------------------------------------------------------------

qq_flux_list <- list()

for(i in 1:length(flux_vars)) {
  
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  # Fit ANOVA model
  formula_str <- paste(var, "~ landscape_position")
  model <- aov(as.formula(formula_str), data = df_wide_filtered)
  
  # Extract residuals from the model
  residuals_data <- data.frame(residuals = residuals(model))
  
  # Create QQ plot of RESIDUALS (not raw data)
  qq_plot <- ggplot(residuals_data, aes(sample = residuals)) +
    geom_qq(alpha = 0.5, color = "darkgreen") +
    geom_qq_line(color = "red", linewidth = 1) +
    labs(
      title = var_label,
      subtitle = "Model Residuals",
      x = "Theoretical Quantiles",
      y = "Residual Quantiles"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      plot.subtitle = element_text(size = 9),
      axis.title = element_text(size = 9)
    )
  
  qq_flux_list[[i]] <- qq_plot
}

# Combine QQ plots
combined_qq_flux <- grid.arrange(
  grobs = qq_flux_list,
  ncol = 2,
  nrow = 2,
  top = textGrob("QQ Plots: Normality of Model Residuals (Correct for ANOVA)", 
                 gp = gpar(fontsize = 14, fontface = "bold"))
)

combined_qq_flux

# Save QQ plots (commented out)
# ggsave("qq_plots_flux_residuals.png", 
#        combined_qq_flux,
#        width = 10, 
#        height = 8, 
#        dpi = 300)

# -------------------------------------------------------------------
# Run formal statistical tests on RESIDUALS
# -------------------------------------------------------------------

for(i in 1:length(flux_vars)) {
  
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("\nтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Testing assumptions for:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  # STEP 1: Fit ANOVA model to get residuals
  formula_str <- paste(var, "~ landscape_position")
  model <- aov(as.formula(formula_str), data = df_wide_filtered)
  
  # Extract residuals from the model
  residuals_vals <- residuals(model)
  
  # NORMALITY TEST: Shapiro-Wilk on RESIDUALS (correct approach)
  if(length(residuals_vals) > 5000) {
    cat("  Note: Sample size > 5000, testing random sample of 5000\n")
    shapiro_result <- shapiro.test(sample(residuals_vals, 5000))
  } else {
    shapiro_result <- shapiro.test(residuals_vals)
  }
  
  cat("\n1. NORMALITY TEST on RESIDUALS (Shapiro-Wilk):\n")
  cat("   W statistic:", round(shapiro_result$statistic, 4), "\n")
  cat("   p-value:", format.pval(shapiro_result$p.value, digits = 3), "\n")
  
  is_normal <- shapiro_result$p.value > 0.05
  if(is_normal) {
    cat("   тЬУ Residuals appear NORMAL (p > 0.05)\n")
  } else {
    cat("   тЬЧ Residuals are NOT NORMAL (p <= 0.05)\n")
  }
  
  # HOMOGENEITY OF VARIANCE TEST: Levene's Test
  levene_result <- leveneTest(as.formula(formula_str), data = df_wide_filtered)
  levene_p <- levene_result$`Pr(>F)`[1]
  levene_f <- levene_result$`F value`[1]
  
  cat("\n2. HOMOGENEITY OF VARIANCE (Levene's Test):\n")
  cat("   F statistic:", round(levene_f, 4), "\n")
  cat("   p-value:", format.pval(levene_p, digits = 3), "\n")
  
  is_homogeneous <- levene_p > 0.05
  if(is_homogeneous) {
    cat("   тЬУ Variances are HOMOGENEOUS (p > 0.05)\n")
  } else {
    cat("   тЬЧ Variances are NOT HOMOGENEOUS (p <= 0.05)\n")
  }
  
  # RECOMMENDATION
  cat("\n3. RECOMMENDATION:\n")
  if(is_normal & is_homogeneous) {
    cat("   тЖТ Use PARAMETRIC test (ANOVA with Tukey post-hoc)\n")
    recommended <- "ANOVA"
  } else {
    cat("   тЖТ Use NON-PARAMETRIC test (Kruskal-Wallis with Dunn post-hoc)\n")
    if(!is_normal) cat("      Reason: Residuals not normally distributed\n")
    if(!is_homogeneous) cat("      Reason: Variances not homogeneous\n")
    recommended <- "Kruskal-Wallis"
  }
  
  # Store results
  assumption_results_flux$Shapiro_p[i] <- shapiro_result$p.value
  assumption_results_flux$Levene_p[i] <- levene_p
  assumption_results_flux$Normal[i] <- is_normal
  assumption_results_flux$Homogeneous[i] <- is_homogeneous
  assumption_results_flux$Recommended_Test[i] <- recommended
}

# Display summary table

# SUMMARY: ASSUMPTION TEST RESULTS    


print(assumption_results_flux)

# Save assumption results (commented out)
# write.csv(assumption_results_flux, 
#           "assumption_test_results_fluxes.csv", 
#           row.names = FALSE)

# ===================================================================
# STEP 3: KRUSKAL-WALLIS TESTS
# ===================================================================


# STEP 3: KRUSKAL-WALLIS TESTS BY LANDSCAPE POSITION    

# Testing: Do fluxes differ among lowland, slope, and upland?\n")
# Method:  Kruskal-Wallis test (non-parametric ANOVA)\n")


# Create results table
kw_results_flux <- data.frame(
  Variable = flux_short,
  Chi_squared = NA,
  df = NA,
  p_value = NA,
  Significant = NA
)

# Store Dunn's test results
dunn_results_flux <- list()

# Loop through each flux variable
for(i in 1:length(flux_vars)) {
  
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("\nтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Variable:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  # Run Kruskal-Wallis test
  formula_str <- paste(var, "~ landscape_position")
  kw_test <- kruskal.test(as.formula(formula_str), data = df_wide_filtered)
  
  # Extract results
  chi_sq <- kw_test$statistic
  df_val <- kw_test$parameter
  p_val <- kw_test$p.value
  
  # Store results
  kw_results_flux$Chi_squared[i] <- chi_sq
  kw_results_flux$df[i] <- df_val
  kw_results_flux$p_value[i] <- p_val
  kw_results_flux$Significant[i] <- ifelse(p_val < 0.05, "YES", "NO")
  
  # Display results
  cat("Kruskal-Wallis Test:\n")
  cat("  ╧З┬▓ =", round(chi_sq, 3), "\n")
  cat("  df =", df_val, "\n")
  cat("  p-value =", format.pval(p_val, digits = 3), "\n")
  
  if(p_val < 0.05) {
    cat("  тЬУ SIGNIFICANT (p < 0.05)\n")
    cat("  тЖТ At least one landscape position differs\n")
  } else {
    cat("  тЬЧ NOT SIGNIFICANT (p >= 0.05)\n")
    cat("  тЖТ No evidence of differences among landscape positions\n")
  }
  
  # If significant, run Dunn's post-hoc test
  if(p_val < 0.05) {
    cat("\n  тХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХР\n")
    cat("  DUNN'S POST-HOC TEST FOR:", var_label, "\n")
    cat("  тХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХР\n")
    
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_wide_filtered, 
                          method = "bh")
    
    # Store results
    dunn_results_flux[[flux_short[i]]] <- dunn_test$res
    
    # Display pairwise comparisons
    cat("\n  Pairwise Comparisons (Dunn's test with BH correction):\n\n")
    dunn_table <- dunn_test$res %>%
      mutate(
        Z = round(Z, 3),
        P.unadj = format.pval(P.unadj, digits = 3),
        P.adj = format.pval(P.adj, digits = 3),
        Sig = ifelse(P.adj < 0.05, "***", "ns")
      )
    print(dunn_table)
    cat("\n  тХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХР\n")
    
    # Summarize significant differences
    sig_pairs <- dunn_test$res %>% filter(P.adj < 0.05)
    
    if(nrow(sig_pairs) > 0) {
      cat("\n  тЬУ Significant pairwise differences (p < 0.05):\n")
      for(j in 1:nrow(sig_pairs)) {
        cat("     тАв", sig_pairs$Comparison[j], 
            "(p =", round(sig_pairs$P.adj[j], 4), ")\n")
      }
    } else {
      cat("\n  Note: Overall test significant but no pairwise differences\n")
      cat("        after correction for multiple comparisons\n")
    }
    
  } else {
    cat("\n  (Skipping post-hoc test - overall test not significant)\n")
    dunn_results_flux[[flux_short[i]]] <- NULL
  }
}

# Display summary table

cat("тХС  SUMMARY: KRUSKAL-WALLIS RESULTS FOR FLUXES  тХС\n")


print(kw_results_flux)

# Save results (commented out)
# write.csv(kw_results_flux, 
#           "kruskal_wallis_flux_results.csv", 
#           row.names = FALSE)

# ===================================================================
# STEP 4: DESCRIPTIVE STATISTICS
# ===================================================================


# STEP 4: DESCRIPTIVE STATISTICS BY LANDSCAPE POSITION 


# Calculate means, SDs, and medians for each landscape position
summary_flux <- df_wide_filtered %>%
  group_by(landscape_position) %>%
  summarise(
    across(all_of(flux_vars), 
           list(mean = ~mean(., na.rm = TRUE),
                sd = ~sd(., na.rm = TRUE),
                median = ~median(., na.rm = TRUE)),
           .names = "{.col}_{.fn}"),
    n = n()
  )

cat("Summary Statistics (┬╡mol/m┬▓/s):\n")
print(summary_flux)

# Save summary statistics (commented out)
# write.csv(summary_flux, 
#           "summary_stats_fluxes_landscape.csv", 
#           row.names = FALSE)

# Create a more readable summary table
summary_flux_readable <- df_wide_filtered %>%
  group_by(landscape_position) %>%
  summarise(
    RECO_mean = mean(RECO_umol, na.rm = TRUE),
    RECO_sd = sd(RECO_umol, na.rm = TRUE),
    GPP_mean = mean(GPP_umol, na.rm = TRUE),
    GPP_sd = sd(GPP_umol, na.rm = TRUE),
    NEE_mean = mean(NEE_umol, na.rm = TRUE),
    NEE_sd = sd(NEE_umol, na.rm = TRUE),
    FCH4_mean = mean(FCH4_umol, na.rm = TRUE),
    FCH4_sd = sd(FCH4_umol, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

cat("\n\nReadable Summary (Mean ┬▒ SD in ┬╡mol/m┬▓/s):\n")
print(summary_flux_readable)

# ===================================================================
# STEP 5: CREATE LETTER GROUPINGS FOR PLOTS
# ===================================================================


cat("тХС  STEP 5: LETTER GROUPINGS FOR PLOTTING тХС\n")


# Function to assign letters based on Dunn's test
assign_flux_letters <- function(dunn_results, flux_name) {
  
  groups <- c("lowland", "slope", "upland")
  
  # If no significant results, all groups get same letter
  if(is.null(dunn_results)) {
    letters_df <- data.frame(
      landscape_position = groups,
      letters = rep("a", 3)
    )
    return(letters_df)
  }
  
  # Get significant pairs
  sig_pairs <- dunn_results %>% filter(P.adj < 0.05)
  
  # If no significant pairwise differences, all get same letter
  if(nrow(sig_pairs) == 0) {
    letters_df <- data.frame(
      landscape_position = groups,
      letters = rep("a", 3)
    )
    return(letters_df)
  }
  
  # Simple letter assignment logic
  letter_assignment <- data.frame(
    landscape_position = groups,
    letters = c("", "", ""),
    stringsAsFactors = FALSE
  )
  
  # Start with first group
  letter_assignment$letters[1] <- "a"
  current_letter <- 1
  
  # Assign letters based on pairwise comparisons
  for(i in 2:length(groups)) {
    # Check if current group differs from any previous group
    differs <- FALSE
    for(j in 1:(i-1)) {
      comp1 <- paste(groups[j], groups[i], sep = " - ")
      comp2 <- paste(groups[i], groups[j], sep = " - ")
      if(any(sig_pairs$Comparison %in% c(comp1, comp2))) {
        differs <- TRUE
        break
      }
    }
    
    if(differs) {
      current_letter <- current_letter + 1
      letter_assignment$letters[i] <- letters[current_letter]
    } else {
      letter_assignment$letters[i] <- letter_assignment$letters[i-1]
    }
  }
  
  return(letter_assignment)
}

# Create letter assignments for each flux
cat("Letter assignments for each flux:\n")
cat("(Groups with same letter are not significantly different)\n\n")

flux_letters_list <- list()

for(i in 1:length(flux_vars)) {
  var_name <- flux_short[i]
  
  cat(var_name, ":\n")
  
  letters_df <- assign_flux_letters(dunn_results_flux[[var_name]], var_name)
  flux_letters_list[[var_name]] <- letters_df
  
  print(letters_df)
  cat("\n")
}

# ===================================================================
# STEP 6: CREATE EXAMPLE BOXPLOTS WITH LETTERS
# ===================================================================


cat("тХС  STEP 6: EXAMPLE PLOTS WITH SIGNIFICANCE LETTERS  тХС\n")


# Function to create boxplot with letters
create_flux_plot <- function(data, var, var_label, letters_df, y_pos) {
  
  # Add y position to letters dataframe
  letters_df$y_position <- y_pos
  
  # Create plot
  p <- ggplot(data, aes(x = landscape_position, y = .data[[var]])) +
    geom_boxplot() +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 4) +
    geom_text(data = letters_df,
              aes(x = landscape_position, y = y_position, label = letters),
              fontface = "bold",
              size = 5) +
    labs(
      title = var_label,
      x = "Landscape Position",
      y = "Flux (┬╡mol/m┬▓/s)"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(face = "bold", size = 11)
    )
  
  return(p)
}

# Create plots for each flux

# RECO plot
reco_plot <- create_flux_plot(
  data = df_wide_filtered,
  var = "RECO_umol",
  var_label = "Ecosystem Respiration (RECO)",
  letters_df = flux_letters_list[["RECO"]],
  y_pos = max(df_wide_filtered$RECO_umol, na.rm = TRUE) * 1.1
)

# GPP plot
gpp_plot <- create_flux_plot(
  data = df_wide_filtered,
  var = "GPP_umol",
  var_label = "Gross Primary Production (GPP)",
  letters_df = flux_letters_list[["GPP"]],
  y_pos = max(df_wide_filtered$GPP_umol, na.rm = TRUE) * 1.1
)

# NEE plot
nee_plot <- create_flux_plot(
  data = df_wide_filtered,
  var = "NEE_umol",
  var_label = "Net Ecosystem Exchange (NEE)",
  letters_df = flux_letters_list[["NEE"]],
  y_pos = max(df_wide_filtered$NEE_umol, na.rm = TRUE) * 1.1
)

# FCH4 plot
fch4_plot <- create_flux_plot(
  data = df_wide_filtered,
  var = "FCH4_umol",
  var_label = "Methane Flux (FCH4)",
  letters_df = flux_letters_list[["FCH4"]],
  y_pos = max(df_wide_filtered$FCH4_umol, na.rm = TRUE) * 1.1
)

# Display individual plots
print(reco_plot)
print(gpp_plot)
print(nee_plot)
print(fch4_plot)

# Combine all flux plots
combined_flux_plots <- grid.arrange(
  reco_plot + labs(title = NULL),
  gpp_plot + labs(title = NULL),
  nee_plot + labs(title = NULL),
  fch4_plot + labs(title = NULL),
  ncol = 2,
  nrow = 2,
  top = textGrob("COтВВ and CHтВД Fluxes by Landscape Position", 
                 gp = gpar(fontsize = 14, fontface = "bold"))
)

combined_flux_plots

# Save plots (commented out)
# ggsave("flux_plots_landscape_combined.png",
#        combined_flux_plots,
#        width = 10,
#        height = 10,
#        dpi = 300)


```

#Save summary stats of C fluxes by landpos 
```{r}
summary_flux_readable

library(writexl)
write_xlsx(summary_flux_readable, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_Flux_means_landpos.xlsx")
```

#FCH4

# Comprehensive stats - landpos x inundation 
```{r}
# ===================================================================
# FLUX BREAKDOWN ANALYSIS: LANDSCAPE ├Ч INUNDATION INTERACTIONS
# ===================================================================
# This script breaks down the landscape ├Ч inundation interaction for fluxes:
# 1) Within each inundation status: do landscape positions differ?
# 2) Within each landscape position: does inundation matter?
# 
# Fluxes: RECO, GPP, NEE, FCH4 (all in ┬╡mol/m┬▓/s)
# Note: Slope-inundated (n=2) excluded from all analyses
# All analyses test assumptions on model RESIDUALS
# ===================================================================

library(dplyr)
library(FSA)        # For Dunn's test
library(ggplot2)
library(car)        # For Levene's test
library(gridExtra)
library(grid)

# Define variables
flux_vars <- c("RECO_umol", "GPP_umol", "NEE_umol", "FCH4_umol")
flux_names <- c("Ecosystem Respiration (RECO)", 
                "Gross Primary Production (GPP)", 
                "Net Ecosystem Exchange (NEE)", 
                "Methane Flux (FCH4)")
flux_short <- c("RECO", "GPP", "NEE", "FCH4")

# ===================================================================
# If not already done, FILTER OUT SLOPE-INUNDATED (n=2)
# ===================================================================


cat("тХС  FILTERING DATA                                              тХС\n")


# Remove slope-inundated combinations throughout
df_wide_filtered_flux <- df_wide_filtered %>%
  filter(!(landscape_position == "slope" & inundated == "Y"))

# cat("Original sample size:", nrow(df_wide), "\n")
# cat("Filtered sample size:", nrow(df_wide_filtered_flux), "\n")
# cat("Observations removed (slope-Y):", nrow(df_wide) - nrow(df_wide_filtered_flux), "\n\n")

# ===================================================================
# ANALYSIS 1: INUNDATED SITES ONLY
# ===================================================================


cat("тХС  ANALYSIS 1: INUNDATED SITES ONLY                            тХС\n")
cat("тХС  Testing: Do landscape positions differ in inundated areas?  тХС\n")


df_inundated_flux <- df_wide_filtered_flux %>%
  filter(inundated == "Y")

cat("Sample sizes for INUNDATED sites:\n")
inundated_n_flux <- df_inundated_flux %>%
  count(landscape_position) %>%
  arrange(landscape_position)
print(inundated_n_flux)
cat("\n")

for(i in 1:length(flux_vars)) {
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Variable:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  n_groups <- df_inundated_flux %>% 
    distinct(landscape_position) %>% 
    nrow()
  
  if(n_groups < 2) {
    cat("тЪа Not enough groups to test\n\n")
    next
  }
  
  # Test assumptions on RESIDUALS
  cat("\nTesting assumptions on model residuals:\n")
  formula_str <- paste(var, "~ landscape_position")
  model <- aov(as.formula(formula_str), data = df_inundated_flux)
  residuals_vals <- residuals(model)
  
  # Shapiro-Wilk on residuals
  if(length(residuals_vals) > 5000) {
    shapiro_result <- shapiro.test(sample(residuals_vals, 5000))
  } else {
    shapiro_result <- shapiro.test(residuals_vals)
  }
  
  cat("  Shapiro-Wilk (residuals): p =", format.pval(shapiro_result$p.value, digits = 3))
  cat(ifelse(shapiro_result$p.value > 0.05, " тЬУ Normal\n", " тЬЧ Not normal\n"))
  
  # Levene's test
  levene_result <- leveneTest(as.formula(formula_str), data = df_inundated_flux)
  levene_p <- levene_result$`Pr(>F)`[1]
  
  cat("  Levene's test: p =", format.pval(levene_p, digits = 3))
  cat(ifelse(levene_p > 0.05, " тЬУ Homogeneous\n", " тЬЧ Not homogeneous\n"))
  
  if(shapiro_result$p.value > 0.05 & levene_p > 0.05) {
    cat("  тЖТ Could use", ifelse(n_groups == 2, "t-test", "ANOVA"), 
        ", but using Kruskal-Wallis for consistency\n\n")
  } else {
    cat("  тЖТ Using Kruskal-Wallis (non-parametric)\n\n")
  }
  
  # Kruskal-Wallis test
  kw_test <- kruskal.test(as.formula(formula_str), data = df_inundated_flux)
  
  cat("Kruskal-Wallis: ╧З┬▓ =", round(kw_test$statistic, 3), 
      ", p =", format.pval(kw_test$p.value, digits = 3), "\n")
  
  if(kw_test$p.value < 0.05 && n_groups > 2) {
    cat("тЬУ Significant! Running Dunn's post-hoc...\n\n")
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_inundated_flux, 
                          method = "bh")
    print(dunn_test$res)
  } else if(kw_test$p.value < 0.05) {
    cat("тЬУ Significant difference between the two groups\n")
  } else {
    cat("тЬЧ No significant differences\n")
  }
  cat("\n")
}

# ===================================================================
# ANALYSIS 2: NON-INUNDATED SITES ONLY
# ===================================================================


cat("тХС  ANALYSIS 2: NON-INUNDATED SITES ONLY                        тХС\n")
cat("тХС  Testing: Do landscape positions differ in dry areas?        тХС\n")


df_not_inundated_flux <- df_wide_filtered_flux %>%
  filter(inundated == "N")

cat("Sample sizes for NON-INUNDATED sites:\n")
not_inundated_n_flux <- df_not_inundated_flux %>%
  count(landscape_position) %>%
  arrange(landscape_position)
print(not_inundated_n_flux)
cat("\n")

for(i in 1:length(flux_vars)) {
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Variable:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  # Test assumptions on RESIDUALS
  cat("\nTesting assumptions on model residuals:\n")
  formula_str <- paste(var, "~ landscape_position")
  model <- aov(as.formula(formula_str), data = df_not_inundated_flux)
  residuals_vals <- residuals(model)
  
  # Shapiro-Wilk on residuals
  if(length(residuals_vals) > 5000) {
    shapiro_result <- shapiro.test(sample(residuals_vals, 5000))
  } else {
    shapiro_result <- shapiro.test(residuals_vals)
  }
  
  cat("  Shapiro-Wilk (residuals): p =", format.pval(shapiro_result$p.value, digits = 3))
  cat(ifelse(shapiro_result$p.value > 0.05, " тЬУ Normal\n", " тЬЧ Not normal\n"))
  
  # Levene's test
  levene_result <- leveneTest(as.formula(formula_str), data = df_not_inundated_flux)
  levene_p <- levene_result$`Pr(>F)`[1]
  
  cat("  Levene's test: p =", format.pval(levene_p, digits = 3))
  cat(ifelse(levene_p > 0.05, " тЬУ Homogeneous\n", " тЬЧ Not homogeneous\n"))
  
  if(shapiro_result$p.value > 0.05 & levene_p > 0.05) {
    cat("  тЖТ Could use ANOVA, but using Kruskal-Wallis for consistency\n\n")
  } else {
    cat("  тЖТ Using Kruskal-Wallis (non-parametric)\n\n")
  }
  
  # Kruskal-Wallis test
  kw_test <- kruskal.test(as.formula(formula_str), data = df_not_inundated_flux)
  
  cat("Kruskal-Wallis: ╧З┬▓ =", round(kw_test$statistic, 3), 
      ", p =", format.pval(kw_test$p.value, digits = 3), "\n")
  
  if(kw_test$p.value < 0.05) {
    cat("тЬУ Significant! Running Dunn's post-hoc...\n\n")
    dunn_test <- dunnTest(as.formula(formula_str), 
                          data = df_not_inundated_flux, 
                          method = "bh")
    print(dunn_test$res)
  } else {
    cat("тЬЧ No significant differences\n")
  }
  cat("\n")
}

# ===================================================================
# ANALYSIS 3: LOWLAND SITES ONLY
# ===================================================================


cat("тХС  ANALYSIS 3: LOWLAND SITES ONLY                              тХС\n")
cat("тХС  Testing: Does inundation affect lowland fluxes?             тХС\n")


df_lowland_flux <- df_wide_filtered_flux %>%
  filter(landscape_position == "lowland")

cat("Sample sizes for LOWLAND sites:\n")
lowland_n_flux <- df_lowland_flux %>%
  count(inundated) %>%
  arrange(inundated)
print(lowland_n_flux)
cat("\n")

for(i in 1:length(flux_vars)) {
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Variable:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  # Test assumptions on RESIDUALS
  cat("\nTesting assumptions on model residuals:\n")
  formula_str <- paste(var, "~ inundated")
  model <- aov(as.formula(formula_str), data = df_lowland_flux)
  residuals_vals <- residuals(model)
  
  # Shapiro-Wilk on residuals
  if(length(residuals_vals) > 5000) {
    shapiro_result <- shapiro.test(sample(residuals_vals, 5000))
  } else {
    shapiro_result <- shapiro.test(residuals_vals)
  }
  
  cat("  Shapiro-Wilk (residuals): p =", format.pval(shapiro_result$p.value, digits = 3))
  cat(ifelse(shapiro_result$p.value > 0.05, " тЬУ Normal\n", " тЬЧ Not normal\n"))
  
  # Levene's test
  levene_result <- leveneTest(as.formula(formula_str), data = df_lowland_flux)
  levene_p <- levene_result$`Pr(>F)`[1]
  
  cat("  Levene's test: p =", format.pval(levene_p, digits = 3))
  cat(ifelse(levene_p > 0.05, " тЬУ Homogeneous\n", " тЬЧ Not homogeneous\n"))
  
  if(shapiro_result$p.value > 0.05 & levene_p > 0.05) {
    cat("  тЖТ Could use t-test, but using Wilcoxon for consistency\n\n")
  } else {
    cat("  тЖТ Using Wilcoxon (non-parametric)\n\n")
  }
  
  # Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(as.formula(formula_str), data = df_lowland_flux)
  
  cat("Wilcoxon rank-sum test:\n")
  cat("  W =", wilcox_test$statistic, "\n")
  cat("  p =", format.pval(wilcox_test$p.value, digits = 3), "\n")
  
  # Calculate medians
  medians <- df_lowland_flux %>%
    group_by(inundated) %>%
    summarise(median = median(.data[[var]], na.rm = TRUE))
  
  cat("\n  Medians:\n")
  print(medians)
  
  if(wilcox_test$p.value < 0.05) {
    cat("\n  тЬУ SIGNIFICANT difference between inundated and non-inundated lowlands\n")
    
    inundated_median <- medians$median[medians$inundated == "Y"]
    not_inundated_median <- medians$median[medians$inundated == "N"]
    
    if(inundated_median > not_inundated_median) {
      cat("  тЖТ Inundated lowlands have HIGHER flux values\n")
    } else {
      cat("  тЖТ Inundated lowlands have LOWER flux values\n")
    }
  } else {
    cat("\n  тЬЧ No significant difference\n")
  }
  cat("\n")
}

# ===================================================================
# ANALYSIS 4: UPLAND SITES ONLY
# ===================================================================


cat("тХС  ANALYSIS 4: UPLAND SITES ONLY                               тХС\n")
cat("тХС  Testing: Does inundation affect upland fluxes?              тХС\n")


df_upland_flux <- df_wide_filtered_flux %>%
  filter(landscape_position == "upland")

cat("Sample sizes for UPLAND sites:\n")
upland_n_flux <- df_upland_flux %>%
  count(inundated) %>%
  arrange(inundated)
print(upland_n_flux)
cat("\n")

for(i in 1:length(flux_vars)) {
  var <- flux_vars[i]
  var_label <- flux_names[i]
  
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  cat("Variable:", var_label, "\n")
  cat("тФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФАтФА\n")
  
  # Test assumptions on RESIDUALS
  cat("\nTesting assumptions on model residuals:\n")
  formula_str <- paste(var, "~ inundated")
  model <- aov(as.formula(formula_str), data = df_upland_flux)
  residuals_vals <- residuals(model)
  
  # Shapiro-Wilk on residuals
  if(length(residuals_vals) > 5000) {
    shapiro_result <- shapiro.test(sample(residuals_vals, 5000))
  } else {
    shapiro_result <- shapiro.test(residuals_vals)
  }
  
  cat("  Shapiro-Wilk (residuals): p =", format.pval(shapiro_result$p.value, digits = 3))
  cat(ifelse(shapiro_result$p.value > 0.05, " тЬУ Normal\n", " тЬЧ Not normal\n"))
  
  # Levene's test
  levene_result <- leveneTest(as.formula(formula_str), data = df_upland_flux)
  levene_p <- levene_result$`Pr(>F)`[1]
  
  cat("  Levene's test: p =", format.pval(levene_p, digits = 3))
  cat(ifelse(levene_p > 0.05, " тЬУ Homogeneous\n", " тЬЧ Not homogeneous\n"))
  
  if(shapiro_result$p.value > 0.05 & levene_p > 0.05) {
    cat("  тЖТ Could use t-test, but using Wilcoxon for consistency\n\n")
  } else {
    cat("  тЖТ Using Wilcoxon (non-parametric)\n\n")
  }
  
  # Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(as.formula(formula_str), data = df_upland_flux)
  
  cat("Wilcoxon rank-sum test:\n")
  cat("  W =", wilcox_test$statistic, "\n")
  cat("  p =", format.pval(wilcox_test$p.value, digits = 3), "\n")
  
  # Calculate medians
  medians <- df_upland_flux %>%
    group_by(inundated) %>%
    summarise(median = median(.data[[var]], na.rm = TRUE))
  
  cat("\n  Medians:\n")
  print(medians)
  
  if(wilcox_test$p.value < 0.05) {
    cat("\n  тЬУ SIGNIFICANT difference between inundated and non-inundated uplands\n")
    
    inundated_median <- medians$median[medians$inundated == "Y"]
    not_inundated_median <- medians$median[medians$inundated == "N"]
    
    if(inundated_median > not_inundated_median) {
      cat("  тЖТ Inundated uplands have HIGHER flux values\n")
    } else {
      cat("  тЖТ Inundated uplands have LOWER flux values\n")
    }
  } else {
    cat("\n  тЬЧ No significant difference\n")
  }
  cat("\n")
}

# ===================================================================
# SUMMARY
# ===================================================================


#SUMMARY: FLUX BREAKDOWN ANALYSIS RESULTS  


#"What was tested

# All analyses exclude slope-inundated (n=2)
#  Analyses 1 & 2: Kruskal-Wallis for landscape comparisons
# Analyses 3 & 4: Wilcoxon for inundation comparisons



```
#Summary by landpos only (regardless of inun)
```{r}
# Calculate detailed statistics for each landscape ├Ч inundation combination
summary_table_landpos <- df_flux_summary2 %>% 
  group_by(landscape_position) %>%
  summarise(
    n = n(),
    
    # RECO statistics
    RECO_mean = mean(RECO_umol, na.rm = TRUE),
    RECO_se = sd(RECO_umol, na.rm = TRUE) / sqrt(n()),

    
    # GPP statistics
    GPP_mean = mean(GPP_umol, na.rm = TRUE),
    GPP_se = sd(GPP_umol, na.rm = TRUE) / sqrt(n()),

    
    # NEE statistics
    NEE_mean = mean(NEE_umol, na.rm = TRUE),
    NEE_se = sd(NEE_umol, na.rm = TRUE) / sqrt(n()),

    
    # FCH4 statistics - umol 
    FCH4_mean = mean(FCH4_umol, na.rm = TRUE),
    FCH4_se = sd(FCH4_umol, na.rm = TRUE) / sqrt(n()),

    
    # flux_CH4 statistics - nmol 
    flux_CH4_mean = mean(flux_CH4, na.rm = TRUE),
    flux_CH4_se = sd(flux_CH4, na.rm = TRUE) / sqrt(n()),

    
    .groups = 'drop'
  ) %>%

  # Round all numeric columns to 2 decimal places
  mutate(across(where(is.numeric) & !n, ~round(., 2)))


print(summary_table_landpos)


```



#Summary stats for fluxes ~  lanpos x inun
```{r}
library(dplyr)
library(knitr)      # For kable() formatting
library(kableExtra) # For enhanced table styling (optional)

# Filter out slope-inundated (n<5)
df_flux_summary <- df_wide %>%
  filter(!(landscape_position == "slope" & inundated == "Y"))

# Filter out slope-inundated (n<5) & the outlier of flux_CO2 > 400 (outlier is at 600, single point in upland non-inun)
df_flux_summary2 <- df_wide %>%
  filter(!(landscape_position == "slope" & inundated == "Y")) %>%
  filter(flux_CH4 <= 400)

cat("Creating comprehensive summary statistics table...\n\n")

# Calculate detailed statistics for each landscape ├Ч inundation combination
summary_table <- df_flux_summary2 %>% 
  group_by(landscape_position, inundated) %>%
  summarise(
    n = n(),
    
    # RECO statistics
    RECO_mean = mean(RECO_umol, na.rm = TRUE),
    RECO_sd = sd(RECO_umol, na.rm = TRUE),
    RECO_se = sd(RECO_umol, na.rm = TRUE) / sqrt(n()),
    RECO_median = median(RECO_umol, na.rm = TRUE),
    RECO_min = min(RECO_umol, na.rm = TRUE),
    RECO_max = max(RECO_umol, na.rm = TRUE),
    
    # GPP statistics
    GPP_mean = mean(GPP_umol, na.rm = TRUE),
    GPP_sd = sd(GPP_umol, na.rm = TRUE),
    GPP_se = sd(GPP_umol, na.rm = TRUE) / sqrt(n()),
    GPP_median = median(GPP_umol, na.rm = TRUE),
    GPP_min = min(GPP_umol, na.rm = TRUE),
    GPP_max = max(GPP_umol, na.rm = TRUE),
    
    # NEE statistics
    NEE_mean = mean(NEE_umol, na.rm = TRUE),
    NEE_sd = sd(NEE_umol, na.rm = TRUE),
    NEE_se = sd(NEE_umol, na.rm = TRUE) / sqrt(n()),
    NEE_median = median(NEE_umol, na.rm = TRUE),
    NEE_min = min(NEE_umol, na.rm = TRUE),
    NEE_max = max(NEE_umol, na.rm = TRUE),
    
    # FCH4 statistics - umol 
    FCH4_mean = mean(FCH4_umol, na.rm = TRUE),
    FCH4_sd = sd(FCH4_umol, na.rm = TRUE),
    FCH4_se = sd(FCH4_umol, na.rm = TRUE) / sqrt(n()),
    FCH4_median = median(FCH4_umol, na.rm = TRUE),
    FCH4_min = min(FCH4_umol, na.rm = TRUE),
    FCH4_max = max(FCH4_umol, na.rm = TRUE),
    
    # flux_CH4 statistics - nmol 
    flux_CH4_mean = mean(flux_CH4, na.rm = TRUE),
    flux_CH4_sd = sd(flux_CH4, na.rm = TRUE),
    flux_CH4_se = sd(flux_CH4, na.rm = TRUE) / sqrt(n()),
    flux_CH4_median = median(flux_CH4, na.rm = TRUE),
    flux_CH4_min = min(flux_CH4, na.rm = TRUE),
    flux_CH4_max = max(flux_CH4, na.rm = TRUE),
    
    .groups = 'drop'
  ) %>%
  # Create a combined label for clarity
  mutate(
    Group = paste0(
      tools::toTitleCase(landscape_position), 
      " - ", 
      ifelse(inundated == "Y", "Inundated", "Not Inundated")
    )
  ) %>%
  # Reorder columns
  select(Group, landscape_position, inundated, n, everything(), 
         -landscape_position, -inundated) %>%
  # Round all numeric columns to 2 decimal places
  mutate(across(where(is.numeric) & !n, ~round(., 2)))

# Display the full table
cat("FULL SUMMARY STATISTICS TABLE\n")
print(summary_table)

# Save as CSV
# write.csv(summary_table, 
#           "flux_summary_stats_landscape_inundation.csv", 
#           row.names = FALSE)

# ===================================================================
# TABLE 2: COMPACT MEAN ┬▒ SD FORMAT (FOR MANUSCRIPT)
# ===================================================================

cat("\n\nCOMPACT TABLE: MEAN ┬▒ SD\n")

# Create compact table with mean ┬▒ SD format
compact_table <- df_flux_summary2 %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    n = n(),
    RECO = paste0(round(mean(RECO_umol, na.rm = TRUE), 2), " ┬▒ ", 
                  round(sd(RECO_umol, na.rm = TRUE), 2)),
    GPP = paste0(round(mean(GPP_umol, na.rm = TRUE), 2), " ┬▒ ", 
                 round(sd(GPP_umol, na.rm = TRUE), 2)),
    NEE = paste0(round(mean(NEE_umol, na.rm = TRUE), 2), " ┬▒ ", 
                 round(sd(NEE_umol, na.rm = TRUE), 2)),
    FCH4 = paste0(round(mean(FCH4_umol, na.rm = TRUE), 2), " ┬▒ ", 
                  round(sd(FCH4_umol, na.rm = TRUE), 2)),
    FCH4_nmol = paste0(round(mean(flux_CH4, na.rm = TRUE), 2), " ┬▒ ", 
                  round(sd(flux_CH4, na.rm = TRUE), 2)),
    .groups = 'drop'
  ) %>%
  mutate(
    Landscape = tools::toTitleCase(landscape_position),
    Inundated = ifelse(inundated == "Y", "Yes", "No")
  ) %>%
  select(Landscape, Inundated, n, RECO, GPP, NEE, FCH4)

print(compact_table)

# Create a nicely formatted kable table
compact_kable <- kable(compact_table, 
                       caption = "Table S1. Mean ┬▒ SD of CO2 and CH4 fluxes (┬╡mol C/m┬▓/s) by landscape position and inundation status",
                       col.names = c("Landscape Position", "Inundated", "n", 
                                   "RECO", "GPP", "NEE", "FCH4"),
                       align = c('l', 'l', 'c', 'c', 'c', 'c', 'c'))

print(compact_kable)

# Save as CSV
# write.csv(compact_table, 
#           "flux_summary_compact_landscape_inundation.csv", 
#           row.names = FALSE)

# ===================================================================
# TABLE 3: SEPARATE TABLES BY FLUX TYPE
# ===================================================================

# cat("\n\nSEPARATE TABLES BY FLUX TYPE\n")
# cat("тХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХРтХР\n\n")
# 
# # Function to create summary table for one flux
create_flux_table <- function(data, flux_var, flux_name) {
  data %>%
    group_by(landscape_position, inundated) %>%
    summarise(
      n = n(),
      Mean = mean(.data[[flux_var]], na.rm = TRUE),
      SD = sd(.data[[flux_var]], na.rm = TRUE),
      SE = sd(.data[[flux_var]], na.rm = TRUE) / sqrt(n()),
      Median = median(.data[[flux_var]], na.rm = TRUE),
      Min = min(.data[[flux_var]], na.rm = TRUE),
      Max = max(.data[[flux_var]], na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    mutate(
      Landscape = tools::toTitleCase(landscape_position),
      Inundated = ifelse(inundated == "Y", "Yes", "No"),
      `Mean ┬▒ SD` = paste0(round(Mean, 2), " ┬▒ ", round(SD, 2))
    ) %>%
    select(Landscape, Inundated, n, Mean, SD, SE, Median, Min, Max, `Mean ┬▒ SD`) %>%
    mutate(across(where(is.numeric) & !n, ~round(., 2)))
}

# # Create table for each flux
# reco_table <- create_flux_table(df_flux_summary, "RECO_umol", "RECO")
# gpp_table <- create_flux_table(df_flux_summary, "GPP_umol", "GPP")
# nee_table <- create_flux_table(df_flux_summary, "NEE_umol", "NEE")
# fch4_table <- create_flux_table(df_flux_summary, "FCH4_umol", "FCH4")
#fch4_table_nmol <-create_flux_table(df_flux_summary, "flux_CH4", "flux_CH4")
# 
# cat("\nRECO (Ecosystem Respiration) - ┬╡mol C/m┬▓/s:\n")
# print(reco_table)
# 
# cat("\n\nGPP (Gross Primary Production) - ┬╡mol C/m┬▓/s:\n")
# print(gpp_table)
# 
# cat("\n\nNEE (Net Ecosystem Exchange) - ┬╡mol C/m┬▓/s:\n")
# print(nee_table)
# 
# cat("\n\nFCH4 (Methane Flux) - ┬╡mol C/m┬▓/s:\n")
# print(fch4_table)

# Save individual tables
# write.csv(reco_table, "summary_RECO_landscape_inundation.csv", row.names = FALSE)
# write.csv(gpp_table, "summary_GPP_landscape_inundation.csv", row.names = FALSE)
# write.csv(nee_table, "summary_NEE_landscape_inundation.csv", row.names = FALSE)
# write.csv(fch4_table, "summary_FCH4_landscape_inundation.csv", row.names = FALSE)

# ===================================================================
# TABLE 4: PUBLICATION-READY FORMATTED TABLE
# ===================================================================

cat("\n\nPUBLICATION-READY TABLE\n")

# Standard error function
se <- function(x, na.rm = TRUE) {
  sd(x, na.rm = na.rm) / sqrt(sum(!is.na(x)))
}


# Create a publication-ready table with all fluxes with SE 
pub_table <- df_flux_summary2 %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    N = n(),
    `RECO (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(RECO_umol, na.rm = TRUE)), 
                                 " ┬▒ ", 
                                 sprintf("%.2f", se(RECO_umol, na.rm = TRUE))),
    `GPP (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(GPP_umol, na.rm = TRUE)), 
                                " ┬▒ ", 
                                sprintf("%.2f", se(GPP_umol, na.rm = TRUE))),
    `NEE (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(NEE_umol, na.rm = TRUE)), 
                                " ┬▒ ", 
                                sprintf("%.2f", se(NEE_umol, na.rm = TRUE))),
    `FCH4 (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(FCH4_umol, na.rm = TRUE)), 
                                 " ┬▒ ", 
                                 sprintf("%.2f", se(FCH4_umol, na.rm = TRUE))),
`FCH4_nmol (nmol/m┬▓/s)` = paste0(sprintf("%.2f", mean(flux_CH4, na.rm = TRUE)), 
                                 " ┬▒ ", 
                                 sprintf("%.2f", se(flux_CH4, na.rm = TRUE))),
    
    .groups = 'drop'
  ) %>%
  mutate(
    `Landscape Position` = tools::toTitleCase(landscape_position),
    Inundation = ifelse(inundated == "Y", "Inundated", "Non-inundated")
  ) %>%
  select(`Landscape Position`, Inundation, N, 
         `RECO (┬╡mol/m┬▓/s)`, `GPP (┬╡mol/m┬▓/s)`, 
         `NEE (┬╡mol/m┬▓/s)`, `FCH4 (┬╡mol/m┬▓/s)`,
         `FCH4_nmol (nmol/m┬▓/s)`)

print(pub_table)

# Create formatted kable
pub_kable <- kable(pub_table,
                   caption = "Table S1. Summary statistics (Mean ┬▒ SD) of CO2 and CH4 fluxes by landscape position and inundation status",
                   align = c('l', 'l', 'c', 'c', 'c', 'c', 'c'))

print(pub_kable)

# Save as CSV
# write.csv(pub_table, 
#           "flux_summary_publication_ready.csv", 
#           row.names = FALSE)

# ===================================================================
# TABLE 5: SUMMARY BY LANDSCAPE POSITION ONLY 
# ===================================================================

# cat("\n\nSUMMARY BY LANDSCAPE POSITION (ALL INUNDATION COMBINED)\n")

# 
# landscape_only_table <- df_flux_summary %>%
#   group_by(landscape_position) %>%
#   summarise(
#     N = n(),
#     `RECO` = paste0(sprintf("%.2f", mean(RECO_umol, na.rm = TRUE)), 
#                     " ┬▒ ", sprintf("%.2f", sd(RECO_umol, na.rm = TRUE))),
#     `GPP` = paste0(sprintf("%.2f", mean(GPP_umol, na.rm = TRUE)), 
#                    " ┬▒ ", sprintf("%.2f", sd(GPP_umol, na.rm = TRUE))),
#     `NEE` = paste0(sprintf("%.2f", mean(NEE_umol, na.rm = TRUE)), 
#                    " ┬▒ ", sprintf("%.2f", sd(NEE_umol, na.rm = TRUE))),
#     `FCH4` = paste0(sprintf("%.2f", mean(FCH4_umol, na.rm = TRUE)), 
#                     " ┬▒ ", sprintf("%.2f", sd(FCH4_umol, na.rm = TRUE))),
#     .groups = 'drop'
#   ) %>%
#   mutate(`Landscape Position` = tools::toTitleCase(landscape_position)) %>%
#   select(`Landscape Position`, N, RECO, GPP, NEE, FCH4)
# 
# print(landscape_only_table)

# ===================================================================
# TABLE 6: SUMMARY BY INUNDATION ONLY 
# ===================================================================
# 
# cat("\n\nSUMMARY BY INUNDATION STATUS (ALL LANDSCAPES COMBINED)\n")

# 
# inundation_only_table <- df_flux_summary %>%
#   group_by(inundated) %>%
#   summarise(
#     N = n(),
#     `RECO` = paste0(sprintf("%.2f", mean(RECO_umol, na.rm = TRUE)), 
#                     " ┬▒ ", sprintf("%.2f", sd(RECO_umol, na.rm = TRUE))),
#     `GPP` = paste0(sprintf("%.2f", mean(GPP_umol, na.rm = TRUE)), 
#                    " ┬▒ ", sprintf("%.2f", sd(GPP_umol, na.rm = TRUE))),
#     `NEE` = paste0(sprintf("%.2f", mean(NEE_umol, na.rm = TRUE)), 
#                    " ┬▒ ", sprintf("%.2f", sd(NEE_umol, na.rm = TRUE))),
#     `FCH4` = paste0(sprintf("%.2f", mean(FCH4_umol, na.rm = TRUE)), 
#                     " ┬▒ ", sprintf("%.2f", sd(FCH4_umol, na.rm = TRUE))),
#     .groups = 'drop'
#   ) %>%
#   mutate(`Inundation Status` = ifelse(inundated == "Y", "Inundated", "Non-inundated")) %>%
#   select(`Inundation Status`, N, RECO, GPP, NEE, FCH4)
# 
# print(inundation_only_table)


```

#save summary table 
```{r}

library(writexl)
write_xlsx(pub_table, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_Flux_landpos.inun_summary.xlsx")
```

#Table with the inundated slope included
```{r}
# Create a publication-ready table with all fluxes
pub_table2 <- df_wide %>%
  group_by(landscape_position, inundated) %>%
  summarise(
    N = n(),
    `RECO (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(RECO_umol, na.rm = TRUE)), 
                                 " ┬▒ ", 
                                 sprintf("%.2f", se(RECO_umol, na.rm = TRUE))),
    `GPP (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(GPP_umol, na.rm = TRUE)), 
                                " ┬▒ ", 
                                sprintf("%.2f", se(GPP_umol, na.rm = TRUE))),
    `NEE (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(NEE_umol, na.rm = TRUE)), 
                                " ┬▒ ", 
                                sprintf("%.2f", se(NEE_umol, na.rm = TRUE))),
    `FCH4 (┬╡mol/m┬▓/s)` = paste0(sprintf("%.2f", mean(FCH4_umol, na.rm = TRUE)), 
                                 " ┬▒ ", 
                                 sprintf("%.2f", se(FCH4_umol, na.rm = TRUE))),
    .groups = 'drop'
  ) %>%
  mutate(
    `Landscape Position` = tools::toTitleCase(landscape_position),
    Inundation = ifelse(inundated == "Y", "Inundated", "Non-inundated")
  ) %>%
  select(`Landscape Position`, Inundation, N, 
         `RECO (┬╡mol/m┬▓/s)`, `GPP (┬╡mol/m┬▓/s)`, 
         `NEE (┬╡mol/m┬▓/s)`, `FCH4 (┬╡mol/m┬▓/s)`)

print(pub_table2)
```

#save summary table with inundated slope 
```{r}

library(writexl)
write_xlsx(pub_table2, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_Flux_landpos.inun_summary2.xlsx")
```


#Save table 
```{r}
library(writexl)
write_xlsx(summary_stats_landpos, "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soilchamber_enviro_vars_means_landpos.xlsx")
```


#save figure 
```{r}

ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun3.png",  # File name and extension
  plot = combined_plot2,        # The plot to save (default is the last plot created)
  width = 27,                 # Width in inches
  height = 35,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)

```


#AGU figs

#NEE, RECO, GPP - landscape

```{r}
# Create the boxplot: NEE and RECO flux vs landscape position 

df_long_cc <- df_long2 %>%
  filter(!is.na(flux_value))

#filter out the inundated slope with n=2
df_long_cc<- df_long_cc %>%
  filter(!(landscape_position == "slope" & inundated == "Y"))

#filter out extremes 
 #filter(flux_umol >= -6.5 & flux_umol <= 6.5), 

unique(df_long_cc$flux_type)

# Asterisks for 
# vwc_asterisk <- data.frame(
#   landscape_position = c("lowland", "upland"),  # Both positions
#   inundated = c("Y", "Y"),                      # Both in Y panel
#   label = c("*", "*"),                          # Both get asterisks
#   y_position = c(97, 97)                      # Same height
# )


#=== Landscape fluxes without FCH4 ========================
landscape_C_fluxes <- ggplot(df_long_cc %>%
                               filter(flux_type !="FCH4"), 
                             aes(x = landscape_position, y = (flux_umol), fill = flux_type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 3, position = position_dodge(0.8))+
  labs(title = "Council Soil Chamber GPP, NEE, and RECO by Landscape Position",
       x = "Landscape Position",
       #y = expression(Flux~Value~(gCO[2]-C~m^-2~s^-1)),
        y = expression(atop("CO"[2]~"Flux", "("*mu* "mol" ~ CO[2]~m^-2~s^-1*")")),
        #y = expression("Flux Value (" * mu * "mol " ~ C ~ m^-2 * ~ s^-1 * ")"),
       fill = "Flux Type") +
  geom_hline(yintercept = 0) +
    scale_y_continuous(limits = c(-6, 6))+ 
                     # breaks = seq(-7, 7, 1.5)+  # Specify the tick positions you want)+
  theme_bw() +
  theme(
   axis.title.x = element_text(size = 14),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 12, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 14),
   axis.text.y = element_text(size = 12, face = "bold"))

landscape_C_fluxes 


#=== Landscape fluxes without FCH4 axis flipped ========================
landscape_C_fluxes2 <- ggplot(df_long_cc %>%
                               filter(flux_type !="FCH4"), 
                             aes(x = flux_type, y = (flux_umol), fill = landscape_position)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 3, position = position_dodge(0.8))+
  labs(#title = "Council Soil Chamber GPP, NEE, and RECO by Landscape Position",
      # x = "Landscape Position",
       x = "Flux Type",
       #y = expression(Flux~Value~(gCO[2]-C~m^-2~s^-1)),
        y = expression(bold(atop("CO"[2]~"Flux", "("*mu* "mol" ~ CO[2]~m^-2~s^-1*")"))),
        #y = expression("Flux Value (" * mu * "mol " ~ C ~ m^-2 * ~ s^-1 * ")"),
       #fill = "Flux Type") +
       fill = "Landscape Position") +
  geom_hline(yintercept = 0) +
    scale_x_discrete(labels = c("NEE" = "NEE",
                               "GPP" = "GPP",
                               "RECO" = expression(R[ECO]))) +
    scale_y_continuous(limits = c(-6, 6))+ 
                     # breaks = seq(-7, 7, 1.5)+  # Specify the tick positions you want)+
  theme_bw() +
  theme(
   axis.title.x = element_text(size = 14, face = "bold", color = "black"),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 12, face = "bold", color = "black"), #Adj size for x-axis text
   axis.title.y = element_text(size = 14, face = "bold", color = "black"),
   axis.text.y = element_text(size = 12, face = "bold", color = "black"),
   legend.text = element_text(size = 14, face = "bold"),
   legend.title = element_text(size = 14, face = "bold"),
   legend.position = "bottom") #moe legend to bottom for combined fig 

landscape_C_fluxes2 


```

#save image to maintain dimensions 
```{r}
#fig 1 
 ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_C_fluxes_final.png",  # File name and extension
  plot = landscape_C_fluxes,        # The plot to save (default is the last plot created)
  width = 18,                 # Width in inches
  height = 12,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)


#flipped axis - fig2
 ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_C_fluxes_final2.png",  # File name and extension
  plot = landscape_C_fluxes2,        # The plot to save (default is the last plot created)
  width = 18,                 # Width in inches
  height = 12,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
 
```
# *****************

#Paneled figure with CO2 fluxes and CH4 fluxes on landscape scale 


```{r}
library(patchwork)

#small dataframe for stats letters
ch4_letters <- data.frame(
  landscape_position = c("lowland", "slope", "upland"),
  label = c("a", "b", "b"),
  y = 330  # y position for all letters
)

#CH4 in nmol 
p_ch4 <- ggplot(df_wide_filtered,
                aes(x = "CH4", y = flux_CH4, fill = landscape_position)) +
  geom_boxplot(position = position_dodge(0.8)) +
  # stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 3,
  #              position = position_dodge(0.8)) +
  
  stat_summary(aes(fill = landscape_position), 
             fun = mean, geom = "point", shape = 23,
             color = "black", size = 2, stroke = 2,
             position = position_dodge(0.8)) +
scale_fill_manual(values = c("lowland" = "#F8766D",   # red 
                              "slope" = "#00BA38",        # green
                              "upland" = "#619CFF"),    # blue 
                  aesthetics = "fill") +

  labs(#title = "Council Soil Chamber CH4 by Landscape Position",
       x = "Flux Type",
       y = expression(bold(atop("CH"[4]~"Flux", "(nmol"~ CH[4]~m^-2~s^-1*")"))),
       fill = "Landscape Position") +
  geom_hline(yintercept = 0) +
  scale_x_discrete(labels = c("CH4" = expression(CH[4]))) +
  scale_y_continuous(limits = c(-100, 335),
                       breaks = c(-100, 0, 100, 200, 300),  # forces -70 to appear
                     position = "right") + #move to right for shared figure 
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold", color = "black"),
    axis.text.x = element_text(size = 12, face = "bold", color = "black"),
    axis.title.y = element_text(size = 14, face = "bold", color = "black"),
    axis.text.y = element_text(size = 12, face = "bold", color = "black"),
    legend.position = "none" ) +
  
#add stat letters
  geom_text(data = ch4_letters,
          aes(x = "CH4", y = y, label = label, group = landscape_position),
          position = position_dodge(0.8),  # matches boxplot dodging exactly
          fontface = "bold", size = 5,
          inherit.aes = FALSE)
      
 p_ch4

# Combine with patchwork
# (landscape_C_fluxes2 + p_ch4) +
#   plot_layout(guides = "collect", widths = c(3, 1)) &
#  theme(legend.position = "bottom")
  

# # Combine with shared legend at bottom
# (p_co2 + p_ch4) +
#   plot_layout(guides = "collect", widths = c(3, 1)) &  # wider CO2 panel since 3 flux types
#   theme(legend.position = "bottom")


```


```{r}
combined_fig <- ((landscape_C_fluxes2 + theme(legend.margin = margin(l = 100))) + (p_ch4 + theme(legend.position = "none"))) +
  #plot_layout(guides = "collect", widths = c(3, 1)) &
  plot_layout(widths = c(3, 1)) &
  theme(#legend.position = "bottom",
        axis.title.x = element_blank())  # remove individual x-axis labels

# Add shared title and x-axis label
combined_fig <- combined_fig +
  plot_annotation(
    title = "Chamber Fluxes by Landscape Position",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

# Add shared x-axis label using a bottom label
# combined_fig <- wrap_elements(combined_fig) +
#   labs(tag = "Flux Type") +
#   theme(plot.tag.position = "bottom",
#         plot.tag = element_text(size = 14, hjust = 0.5, vjust = 0))

print(combined_fig)

```

#Save combined fig of all fluxes ~ landscape position
```{r}
# #flipped axis - fig2
 ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_C_fluxes_all.png",  # File name and extension
  plot = combined_fig,        # The plot to save (default is the last plot created)
  width = 24,                 # Width in inches
  height = 12,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)

# ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_C_fluxes_all", combined_fig,
#         width = 7, height = 5, dpi = 600)


```


#CH4 figs 

#CH4 landscape fig with stats 

```{r}
# all plot type - FCH4 vs landscape position only- in umol/m2/s
landscape_CH4_fluxes <- ggplot(df_wide_filtered, aes(x = landscape_position, y = flux_CH4, fill = landscape_position)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8))+
  labs(title = "Methane Fluxes vs Landscape Position",
       x = "Landscape Position",
       y = expression("Flux ("*mu* "mol" ~ CH[4]/m^2/s*")"))+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-70, 315)) + #, #cut out a few bit outliers here to show trend 
  theme_bw(base_size = 15)+ 
  theme(legend.position = "none")+
  theme(
   axis.title.x = element_text(size = 17),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 14, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 17),
   axis.text.y = element_text(size = 14, face = "bold"))+
      annotate("text", x = "lowland", y = 315, label = "a", fontface = "bold", size = 7) +
      annotate("text", x = "slope", y = 315, label = "bc", fontface = "bold", size = 7) +
      annotate("text", x = "upland", y = 315, label = "c", fontface = "bold", size = 7) 
  
landscape_CH4_fluxes


#looking at upland outlier which is > 600 nmol FCH4
landscape_CH4_fluxes2 <- ggplot(df_wide_filtered, aes(x = landscape_position, y = flux_CH4, fill = landscape_position)) +
  geom_boxplot(position = position_dodge(0.8)) 

landscape_CH4_fluxes2

```


#FCH4 landpos vs inun fig with stats
```{r}
#=========== Landpos x Inundated ========================


# all plot type - FCH4 vs landscape position & inundation - in umol/m2/s
landscapeinun_CH4_fluxes <- ggplot(df_wide_filtered, aes(x = landscape_position, y = (FCH4_umol), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8))+
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Landscape Position",
       y = expression("Flux ("*mu* "mol" ~ CH[4]/m^2/s*")"),
       fill = "Inundated")+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-0.13,0.4)) + #, #cut out a few bit outliers here to show trend 
  theme_bw(14)+ 
  theme(
   axis.title.x = element_text(size = 17),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 14, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 17),
   axis.text.y = element_text(size = 14, face = "bold")) +

# With dodging, positions are roughly:
# lowland-N: 0.8, lowland-Y: 1.2
# slope-N: 1.8, slope-Y: 2.2
# upland-N: 2.8, upland-Y: 3.2
  
  # Letters on non-inundated boxes
  annotate("text", x = 0.8, y = 0.4, label = "a", fontface = "bold", size = 7) +  # lowland-N
  annotate("text", x = 1.8, y = 0.4, label = "b", fontface = "bold", size = 7) +  # slope-N
  annotate("text", x = 2.8, y = 0.4, label = "b", fontface = "bold", size = 7) +  # upland-N
  
  # Asterisks on inundated boxes
  annotate("text", x = 1.2, y = 0.38, label = "*", fontface = "bold", size = 10) +  # lowland-Y
  annotate("text", x = 3.2, y = 0.38, label = "*", fontface = "bold", size = 10)    # upland-Y
  
landscapeinun_CH4_fluxes
  

```


#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_stats.png",  # File name and extension
  plot = landscape_CH4_fluxes,        # The plot to save (default is the last plot created)
  width = 15,                 # Width in inches
  height = 10,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)

#with inundation 
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscapeinun_CH4_fluxes_stats.png",  # File name and extension
  plot = landscapeinun_CH4_fluxes,        # The plot to save (default is the last plot created)
  width = 18,                 # Width in inches
  height = 13,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```

#facetwrapped
```{r}
# all plot type - FCH4 vs landscape position & inundation - in umol/m2/s
landscapeinun_CH4_fluxes2 <- ggplot(df_wide_filtered, aes(x = landscape_position, y = (FCH4_umol), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~inundated)+
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8))+
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Landscape Position",
       y = expression(atop("CH"[4]~"Flux", "("*mu* "mol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated")+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-0.13,0.4)) + #, #cut out a few bit outliers here to show trend 
  theme_bw(12)+ 
  theme(
   axis.title.x = element_text(size = 17),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 10, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 17),
   axis.text.y = element_text(size = 14, face = "bold")) # +
 # annotate("text", x = 0.8, y = "lowland", label = "a", fontface = "bold", size = 7) + 
 #  annotate("text", x = 1.8, y = "slope", label = "b", fontface = "bold", size = 7)+
 #    annotate("text", x = 2.8, y = "upland", label = "b", fontface = "bold", size = 7)


landscapeinun_CH4_fluxes2



landscapeinun_CH4_fluxes3 <- ggplot(df_wide_filtered %>% filter(!landscape_position == "slope"), aes(x = inundated, y = (FCH4_umol), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~landscape_position)+
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8))+
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Inundation",
      y = expression(atop("CH"[4]~"Flux", "("*mu* "mol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated")+
   geom_hline(yintercept = 0)+
 scale_y_continuous(limits = c(-0.13,0.4)) + #, #cut out a few bit outliers here to show trend 
  theme_bw(base_size=12)+ 
  theme(
   axis.title.x = element_text(size = 17),  # Adjust size for x-axis label
    axis.text.x = element_text(size = 10, face = "bold"), # Adjust size for x-axis text
   axis.title.y = element_text(size = 17),
   axis.text.y = element_text(size = 14, face = "bold")) 

landscapeinun_CH4_fluxes3
```
```{r}
landscapeinun_CH4_fluxes3 <- ggplot(df_wide_filtered, aes(x = landscape_position, y = (flux_CH4), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~inundated)+
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8))

landscapeinun_CH4_fluxes3 
```


```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_inun.png",  # File name and extension
  plot = landscapeinun_CH4_fluxes2,        # The plot to save (default is the last plot created)
  width = 15,                 # Width in inches
  height = 10,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)


ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_landpos.png",  # File name and extension
  plot = landscapeinun_CH4_fluxes3,        # The plot to save (default is the last plot created)
  width = 15,                 # Width in inches
  height = 10,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```


#adding stats annotations, combining figures 
```{r}
library(patchwork)

# ============================================================================
# FIGURE 1: Add letters "a", "b", "b" to inundated = N panel
# ============================================================================

Fig1 <- ggplot(df_wide_filtered, aes(x = landscape_position, y = (FCH4_umol), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~inundated) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8)) +
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Landscape Position",
     y = expression(atop("CH"[4]~"Flux", "("*mu* "mol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated") +
  geom_hline(yintercept = 0) +
  scale_y_continuous(limits = c(-0.13, 0.4)) +
  theme_bw() + 
  theme(
    axis.title.x = element_text(size = 17),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 17, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold") ,
      strip.text = element_text(face = "bold") 
  ) +
  
  # Add letters ONLY to inundated = N panel
  # geom_text places them above the boxes
  geom_text(
    data = df_wide_filtered %>% filter(inundated == "N"),
    aes(x = landscape_position, y = 0.37, label = c("a", "b", "b")[as.numeric(factor(landscape_position))]),
    fontface = "bold", 
    size = 5,
    inherit.aes = FALSE
  )

print(Fig1)

# ============================================================================
# FIGURE 2: Add "*" above inundated = Y for lowland and upland
# ============================================================================

Fig2 <- ggplot(df_wide_filtered %>% filter(!landscape_position == "slope"), 
               aes(x = inundated, y = (FCH4_umol), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~landscape_position) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8)) +
  labs(
    #title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Inundation",
        y = expression(atop("CH"[4]~"Flux", "("*mu* "mol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated") +
  geom_hline(yintercept = 0) +
  scale_y_continuous(limits = c(-0.13, 0.4)) +
  theme_bw() + 
  theme(
    axis.title.x = element_text(size = 17),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 17, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold"),
       strip.text = element_text(face = "bold") 
  ) +
  
  # Add "*" above Y for both lowland and upland
  geom_text(
    data = df_wide_filtered %>% 
      filter(!landscape_position == "slope", inundated == "Y") %>%
      distinct(landscape_position, inundated),
    aes(x = inundated, y = 0.37, label = "*"),
    fontface = "bold",
    size = 10,
    inherit.aes = FALSE
  )

print(Fig2)

# ============================================================================
# COMBINE FIGURES (stacked vertically)
# ============================================================================

combined_figure <- Fig1 / Fig2 #+
  #plot_annotation(tag_levels = 'A')  # Optional: adds (A) and (B) labels

print(combined_figure)

# Save
# ggsave("combined_ch4_landscape_figures.png", combined_figure, 
#        width = 10, height = 12, dpi = 300)
```


#Save
```{r}
ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_landpos_inun_combined2.png", combined_figure, 
        width = 8, height = 8, dpi = 600)

```

#Save combined fig 
```{r}
# ggsave(
#   filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_landpos_inun_combined2.png",  # File name and extension
#   plot = combined_figure,        # The plot to save (default is the last plot created)
#   width = 10,                 # Width in inches
#   height = 10,                # Height in inches
#   dpi = 600,                 # Resolution in dots per inch
#   units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
# )
```

#FCH4 in nmol 

#adding stats annotations, combining figures 
```{r}
library(patchwork)

# ============================================================================
# FIGURE 1: Add letters "a", "b", "b" to inundated = N panel
# ============================================================================

CH4_nmol.1 <- ggplot(df_wide_filtered, aes(x = landscape_position, y = (flux_CH4), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~inundated) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8)) +
  labs(title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Landscape Position",
        y = expression(atop("CH"[4]~"Flux", "(nmol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated") +
  geom_hline(yintercept = 0) +
  #scale_y_continuous(limits = c(-0.13, 0.4)) +
      scale_y_continuous(limits = c(-100, 335),
                       breaks = c(-100, 0, 100, 200, 300)) +  # forces -70 to appear
  theme_bw() + 
  theme(
    axis.title.x = element_text(size = 17),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 17, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold") ,
      strip.text = element_text(face = "bold") 
  ) +
  
  # Add letters ONLY to inundated = N panel
  # geom_text places them above the boxes
  geom_text(
    data = df_wide_filtered %>% filter(inundated == "N"),
    aes(x = landscape_position, y = 330, label = c("a", "b", "b")[as.numeric(factor(landscape_position))]),
    fontface = "bold", 
    size = 5,
    inherit.aes = FALSE
  )

print(CH4_nmol.1)

# ============================================================================
# FIGURE 2: Add "*" above inundated = Y for lowland and upland
# ============================================================================

CH4_nmol.2 <- ggplot(df_wide_filtered %>% filter(!landscape_position == "slope"), 
               aes(x = inundated, y = (flux_CH4), fill = inundated)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_wrap(~landscape_position) +
  stat_summary(fun = mean, geom = "point", shape = 18, color = "black", size = 4, position = position_dodge(0.8)) +
  labs(
    #title = "Methane Fluxes vs Landscape Position and Inundation",
       x = "Inundation",
        y = expression(atop("CH"[4]~"Flux", "(nmol" ~ CH[4]~m^-2~s^-1*")")),
       fill = "Inundated") +
  geom_hline(yintercept = 0) +
  #scale_y_continuous(limits = c(-0.13, 0.4)) +
    scale_y_continuous(limits = c(-100, 335),
                       breaks = c(-100, 0, 100, 200, 300)) +  # forces -70 to appear
  theme_bw() + 
  theme(
    axis.title.x = element_text(size = 17),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 17, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold"),
       strip.text = element_text(face = "bold") 
  ) +
  
  # Add "*" above Y for both lowland and upland
  geom_text(
    data = df_wide_filtered %>% 
      filter(!landscape_position == "slope", inundated == "Y") %>%
      distinct(landscape_position, inundated),
    aes(x = inundated, y = 300, label = "*"),
    fontface = "bold",
    size = 10,
    inherit.aes = FALSE
  )

print(CH4_nmol.2)

# ============================================================================
# COMBINE FIGURES (stacked vertically)
# ============================================================================

combined_figure2 <- CH4_nmol.1 / CH4_nmol.2 #+
  #plot_annotation(tag_levels = 'A')  # Optional: adds (A) and (B) labels

print(combined_figure2)

# Save
# ggsave("combined_ch4_landscape_figures.png", combined_figure, 
#        width = 10, height = 12, dpi = 300)
```

#save combined fig in nmol 

#Save
```{r}
ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/landscape_CH4_fluxes_landpos_inun_combined_nmol.png", combined_figure2, 
        width = 8, height = 8, dpi = 600)

```

#END



#=== old, exploratory working code ==========================

```{r}



df_wide_FCH4 <- df_wide_filtered %>%
  group_by(landscape_position, inundated) %>%
  filter(n() >= 5) %>%
  ungroup()

#Kruskall-Wallis test - if sig, shows there is some kind of association between the variables 
kruskal.test(VWC ~ landscape_position, data = df_FCH4)#p<0.001, SIG
kruskal.test(inundated ~ landscape_position, data = df_FCH4)#
```


```{r}
#inun
df_inundated <- df_wide_filtered_flux %>%
  filter(inundated == "Y")

wilcox.test(flux_CH4 ~ landscape_position, 
            data = df_inundated,
            subset = landscape_position %in% c("upland", "lowland")) #p=0.07

#non-inun
df_notinundated <- df_wide_filtered_flux %>%
  filter(inundated == "N")

wilcox.test(flux_CH4 ~ landscape_position, 
            data = df_notinundated,
            subset = landscape_position %in% c("upland", "lowland")) #p = 0.00048 * SIG 


wilcox.test(flux_CH4 ~ landscape_position, 
            data = df_notinundated,
            subset = landscape_position %in% c("upland", "slope")) #p = 0.99

wilcox.test(flux_CH4 ~ landscape_position, 
            data = df_notinundated,
            subset = landscape_position %in% c("lowland", "slope")) #p = 0.0503 *borderline ***


```
#Playing around with permutations to account for small sample size and unequal sample sizes -- using to cocmpare with and verify the kruskall-wallis and wilcox tests 

#Permutation test тАФ makes no assumptions about distribution and handles unequal sample sizes well by randomly shuffling group labels many times to build a null distribution

```{r}
library(coin)
kruskal_test(flux_CH4 ~ factor(landscape_position), 
             #data = df_wide_filtered_flux %>% filter(inundated == "N"),
             data = df_notinundated,
             distribution = approximate(nresample = 9999)) #p-value = 0.0023
```
#Bootstrap Wilcoxon 
```{r}
wilcox_test(flux_CH4 ~ factor(landscape_position),
            #data = df_inundated %>% filter(landscape_position %in% c("upland", "lowland")),
            data = df_inundated,
            distribution = approximate(nresample = 9999)) #p-value = 0.06961
```
```{r}
library(coin)
library(rcompanion)

df_noninundated <- df_wide_filtered_flux %>%
  filter(inundated == "N") %>%
  mutate(landscape_position = factor(landscape_position))



# Pairwise permutation tests
pairs <- combn(levels(df_noninundated$landscape_position), 2, simplify = FALSE)

pairwise_results <- lapply(pairs, function(pair) {
  df_pair <- df_noninundated %>%
    filter(landscape_position %in% pair) %>%
    mutate(landscape_position = factor(landscape_position))
  
  test <- wilcox_test(flux_CH4 ~ landscape_position,
                      data = df_pair,
                      distribution = approximate(nresample = 9999))
  
  data.frame(
    group1 = pair[1],
    group2 = pair[2],
    p_value = round(pvalue(test), 4)
  )
})

# Combine results
pairwise_df <- do.call(rbind, pairwise_results)

# Apply Benjamini-Hochberg correction for multiple comparisons
pairwise_df$p_adjusted <- round(p.adjust(pairwise_df$p_value, method = "BH"), 4)

print(pairwise_df)
```
```{r}
library(dunn.test)

cat("=== Kruskal-Wallis Test ===\n")
kw_test <- kruskal.test(flux_CH4 ~ landscape_position, data = df_noninundated)
print(kw_test)

cat("\n=== Dunn's Post-hoc: NO correction ===\n")
dunn.test(df_noninundated$flux_CH4, 
          df_noninundated$landscape_position,
          method = "none",
          kw = FALSE,
          label = TRUE)

cat("\n=== Dunn's Post-hoc: Bonferroni correction ===\n")
dunn.test(df_noninundated$flux_CH4, 
          df_noninundated$landscape_position,
          method = "bonferroni",
          kw = FALSE,
          label = TRUE)

cat("\n=== Dunn's Post-hoc: BH correction ===\n")
dunn.test(df_noninundated$flux_CH4, 
          df_noninundated$landscape_position,
          method = "bh",
          kw = FALSE,
          label = TRUE)
```

