---
title: "Soil_Chamber_EnviroVariable_Figs" ##time is in AK daylight time** NEE/RECO/GPP are in gC/m2/s; FCO2 and FCH4 are in gC/m2/s, flux_CO2 is in umol/m2/s; flux_CH4 is in nano mol/m2/s
output: html_document
date: "2025-04-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r, include=FALSE}
rm(list= ls())

library(data.table)
library(ggplot2)
library(cowplot)
library(openair)
library(plotrix)
library(signal)
library(svMisc)
library(zoo)
library(stringr)
library(plyr)
library(viridis)
library(lubridate)
library(tidyverse)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(pracma)
library(dplyr)
library(openair)
library(nlme)
library(lme4)

Sys.setenv(TZ='UTC')

```

#Load filtered and merged df of soil chamber fluxes, moisture, temp (I upload multiples but only using df_NEE_RECO2 and df_NEE_RECO2_GPP for analysis below)
```{r}
# #filtered for p<0.05; units umol/m2/s or nmol/m2/s
# df_soilchambers_filtered = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_filtered_soil_chamber_fluxes_2017to2019.csv')
# 
# #fluxes and moisture/temp df merged; FCO2 in units g/m2/s
# df_fulljoin = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

# *****************Use these two, above are just extra if needed for looking at*********************

#used transparent and opaque chambers to identify NEE and RECO, then merged back together 
df_NEE_RECO2 = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_fulljoin_soilchamber_fluxes_moisttemp_2017to2019.csv')

#calculated GPP (NEE - Reco)
df_NEE_RECO2_GPP = fread('C:/Users/kkent/Documents/Council Data/Soil Chambers_Council/council_NEE_RECO2_GPP_2017to2019.csv')



```


#Re-shape df into long format 
```{r}
library(tidyr)

#Remove the NAs 
library(dplyr)

# Reshape the dataframe to long format, choose variables of interest 
df_wide <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) #exclude soil temp 15 cm since most of it is NA 

#make NEE, GPP, RECO, and FCH4 all g/m2/s -> umol/m2/s
df_wide$NEE_umol <- df_wide$NEE * 1/12 * 1000000
df_wide$GPP_umol <- df_wide$GPP * 1/12 * 1000000
df_wide$RECO_umol <- df_wide$RECO * 1/12 * 1000000
df_wide$FCH4_umol <- df_wide$FCH4 * 1/12 * 1000000




df_long2 <- df_NEE_RECO2_GPP %>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth) %>%
  #remove NAs
   filter(complete.cases( VWC, air_temp, flux_CO2, flux_CH4, FCH4, NEE, RECO, GPP, inundated, soil_temp_10_cm, thawdepth)) %>% #exclude soil temp 15 cm since most of it is NA 
  pivot_longer(cols = c(NEE, RECO, GPP, FCH4),
               names_to = "flux_type",
               values_to = "flux_value")

#df_long2$FCH4_umol <- df_long2$FCH4 * 1/12 * 1000000
df_long2$flux_umol <- df_long2$flux_value * 1/12 * 1000000

colnames(df_wide)
colnames(df_long2)
```


#looking at variables  across landscape positions & inundation
```{r}

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, color = flux_type)) +
  geom_boxplot(position = "dodge") +
  stat_summary(fun = mean, geom = "point", position = position_dodge(0.8), shape = 23, size = 4)+
    facet_wrap(~inundated) +
  labs(title = "Observed NEE by Land pos",
       x = "Landscape Position",
       y = "NEE (flux_value)") +
  theme_bw()

ggplot(df_long2, aes(x = landscape_position, y = flux_umol, fill = flux_type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 4,
    position = position_dodge(0.8)  # same dodge width as boxplot
  ) +
  labs(
    title = "Observed NEE by Landscape Position",
    x = "Landscape Position",
    y = "Flux Value (umol/m2.s)",
    fill = "Flux Type"
  ) +
  theme_bw()



#thawdepth by landpos
thaw_landpos <- ggplot(df_long, aes(x = landscape_position, y = thawdepth)) +
  geom_boxplot() +
  labs(title = "Observed Thaw Depth by Landscape Position",
       x = "Landscape Position",
       y = "Thaw Depth (cm)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

thaw_landpos

#thawdepth by landpos and inundated 
thaw_landposinun <- ggplot(df_long, aes(x = landscape_position, y = thawdepth)) +
  geom_boxplot() +
   facet_wrap(~inundated) +
  labs(title = "Observed thawdepth by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Thaw Depth (cm)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

thaw_landposinun

#VWC by landpos 
VWC_landpos <- ggplot(df_long, aes(x = landscape_position, y = VWC)) +
  geom_boxplot() +
    labs(title = "Observed VWC by Landscape Position",
       x = "Landscape Position",
       y = "VWC (%)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

VWC_landpos

#VWC by landpos and inundated
VWC_landposinun <- ggplot(df_long, aes(x = landscape_position, y = VWC)) +
  geom_boxplot() +
  facet_wrap(~inundated)+
    labs(title = "Observed VWC by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "VWC (%)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

VWC_landposinun


#soil temp by landpos 
soiltemp_landpos <- ggplot(df_long, aes(x = landscape_position, y = soil_temp_10_cm)) +
  geom_boxplot() +
    labs(title = "Observed Soil Temp by Landscape Position",
       x = "Landscape Position",
       y = "Soil Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

soiltemp_landpos


#soiltemp by landpos and inundated 
soiltemp_landposinun <- ggplot(df_long, aes(x = landscape_position, y = soil_temp_10_cm)) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    labs(title = "Observed Soil Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Soil Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

soiltemp_landposinun

#air temp by landpos 
airtemp_landpos <- ggplot(df_long, aes(x = landscape_position, y = air_temp)) +
  geom_boxplot() +
    labs(title = "Observed Air Temp by Landscape Position",
       x = "Landscape Position",
       y = "Air Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

airtemp_landpos

#airtemp by landpos and inundated 
airtemp_landposinun <- ggplot(df_long, aes(x = landscape_position, y = air_temp)) +
  geom_boxplot() +
    facet_wrap(~inundated)+
    labs(title = "Observed Air Temp by Landscape Position & Inundation",
       x = "Landscape Position",
       y = "Air Temp (C)") +
  theme_bw() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

airtemp_landposinun

#Number of inundated plots by landpos 
library(dplyr)
library(ggplot2)

df_counts <- df_long2 %>%
  distinct(plot_ID, landscape_position, inundated) %>%  # keep unique plots only
  count(landscape_position, inundated)                  # count per group

ggplot(df_counts, aes(x = landscape_position, y = n, fill = inundated)) +
  geom_col(position = "dodge") +
  labs(
    title = "Unique Plot Counts by Landscape Position and Inundation",
    x = "Landscape Position",
    y = "Count of Unique Plots",
    fill = "Inundated"
  ) +
  theme_bw() 
  


n_distinct(df_long$plot_ID) #35 total plots

```
#Arrange timeseries figures into a grid to save as single figure
```{r}
# Install if you don't have it
 install.packages("patchwork")
library(patchwork)

# Arrange plots in two columns (3 rows x 2 columns) 
combined_plot <- (title1 + title2) /
                 (soiltemp_landpos + soiltemp_landposinun) /
                 (airtemp_landpos + airtemp_landposinun) /
                 (thaw_landpos + thaw_landposinun)/
                 (VWC_landpos + VWC_landposinun) 
                # plot_layout(guides = "collect") # Collects all legends into one
combined_plot



#remove titles from each individual plot 
library(patchwork)

combined_plot <- (
  
  soiltemp_landpos + labs(title = NULL) + 
  soiltemp_landposinun + labs(title = NULL)
) /
(
  airtemp_landpos + labs(title = NULL) + 
  airtemp_landposinun + labs(title = NULL)
) /
(
  thaw_landpos + labs(title = NULL) + 
  thaw_landposinun + labs(title = NULL)
) /
(
  VWC_landpos + labs(title = NULL) + 
  VWC_landposinun + labs(title = NULL)
) 


combined_plot






```

#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun.png",  # File name and extension
  plot = combined_plot,        # The plot to save (default is the last plot created)
  width = 20,                 # Width in inches
  height = 20,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```

#Adding column titles to the combined fig 
```{r}
library(patchwork)
library(ggplot2)

# First, remove individual titles from each plot
soiltemp_landpos <- soiltemp_landpos + theme(plot.title = element_blank())
soiltemp_landposinun <- soiltemp_landposinun + theme(plot.title = element_blank())
airtemp_landpos <- airtemp_landpos + theme(plot.title = element_blank())
airtemp_landposinun <- airtemp_landposinun + theme(plot.title = element_blank())
thaw_landpos <- thaw_landpos + theme(plot.title = element_blank())
thaw_landposinun <- thaw_landposinun + theme(plot.title = element_blank())
VWC_landpos <- VWC_landpos + theme(plot.title = element_blank())
VWC_landposinun <- VWC_landposinun + theme(plot.title = element_blank())

# Create empty plots for column titles
title1 <- ggplot() + 
  annotate("text", x = 0.5, y = 0.5, label = "Landscape Position", size = 5, fontface = "bold") + 
  theme_void()

title2 <- ggplot() + 
  annotate("text", x = 0.5, y = 0.5, label = "Landscape Position and Inundation", size = 5, fontface = "bold") + 
  theme_void()

# Arrange plots with column titles
combined_plot2 <- (title1 + title2) /
                 (soiltemp_landpos + soiltemp_landposinun) /
                 (airtemp_landpos + airtemp_landposinun) /
                 (thaw_landpos + thaw_landposinun) /
                 (VWC_landpos + VWC_landposinun) +
                 plot_layout(heights = c(0.5, 1, 1, 1, 1)) #, widths = c(0.3, 3)) #widths not working, may adjust later 


# # Add overall bottom title if needed
# combined_plot2 <- combined_plot + 
#   plot_annotation(
#     caption = "Landscape Position",
#     theme = theme(plot.caption = element_text(hjust = 0.5, size = 12, face = "bold"))
#   )

combined_plot2
```
#save image to maintain dimensions 
```{r}
ggsave(
  filename = "C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/enviro_variables_landposinun2.png",  # File name and extension
  plot = combined_plot2,        # The plot to save (default is the last plot created)
  width = 20,                 # Width in inches
  height = 20,                # Height in inches
  dpi = 600,                 # Resolution in dots per inch
  units = "cm"               # Units for width and height (can be "in", "cm", or "mm")
)
```


#Testing differences in enviro features among landscapes - exclude inundated slope**




#Correlation matrix 

####Create correlation matrix dataset of variables of interest - can only use continuous variables* 

```{r}
# Environmental relationships
# Create a correlation plot between fluxes and environmental variables

#can only use numeric variables, so landscape_position and inundated can't be used 
chamber_corr_matrix <- df_long %>%
  select(NEE, GPP, RECO, thawdepth, soil_temp_10_cm, VWC, air_temp) %>%
  rename(
    Soil_Temp_10cm = soil_temp_10_cm,
    Soil_VWC = VWC,
    Air_Temperature = air_temp,
    Thaw_Depth = thawdepth
  )

#make sure you have complete cases, if not already 
#cor_matrix2 <- cor(chamber_corr_matrix, use = "complete.obs")
```


####Pearson Corr - Flux v Environmental correlation matrix - soil chambers 
*But this is for normally distributed data, which most of these fluxes/data are not, so see Spearman corr in code chunk below - keeping both for reference 
```{r}
# Calculate the correlation matrix
chamber_corr_matrix_result <- cor(chamber_corr_matrix , use = "pairwise.complete.obs")
print(chamber_corr_matrix_result)

# Convert correlation matrix to long format for plotting
cor_long2 <- reshape2::melt(chamber_corr_matrix_result)
names(cor_long2) <- c("Var1", "Var2", "Correlation")



ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,
                        limits = c(-1, 1)) +  # Add this to set the full range to be 1 to -1
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text = element_text(size = 12, face = "bold"),
       plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber GPP Fluxes and Environmental Variables")
   



# option 2: Fewer breaks for a cleaner look, only labels corr values above 0.2 --> I like this one best 
#this uses linear correlation / pearson correlation *** beware if your data is not normally distributed 
ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation values only for strong correlations
  geom_text(data = subset(cor_long2, abs(Correlation) > 0.2),  # Only show stronger correlations
            aes(label = sprintf("%.2f", Correlation),
                 color = abs(Correlation) < 0.5), # Changes text color based on correlation strength
            size = 3,
            fontface = "bold") + #to make numbers in tiles bold
  scale_fill_gradient2(low = "red", 
                      mid = "white", 
                      high = "blue", 
                      midpoint = 0,
                      limits = c(-1, 1),
                      breaks = c(-1, -0.5, 0, 0.5, 1)) +
    scale_color_manual(values = c("white", "black"), guide = "none") +  # For text visibility
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber C Fluxes and Environmental Variables")

 #saves the figure directly above
ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soil_chamber_Cflux_corr_matrix.png",
       width = 8, height = 5, dpi = 600)

```


#Correlation matrix - spearman for non-parametric 

####Spearman Corr
```{r}
#calc correlation matrix with spearman corr
chamber_SPcorr_result <- cor(chamber_corr_matrix, method = "spearman", use = "pairwise.complete.obs")
print(chamber_SPcorr_result)

#For p-values, use rcorr
library(Hmisc)
chamber_corr_with_p <- rcorr(as.matrix(chamber_corr_matrix), type = "spearman")
print(chamber_corr_with_p)

# View results
print(chamber_SPcorr_result)
print("P-values:")
print(chamber_corr_with_p$P)

#correlation coeffs
#                   NEE   GPP  RECO Thaw_Depth Soil_Temp_10cm Soil_VWC Air_Temperature
# NEE              1.00 -0.75  0.13       0.16           0.09    -0.11            0.18
# GPP             -0.75  1.00  0.48      -0.13           0.20     0.11            0.17
# RECO             0.13  0.48  1.00      -0.02           0.51     0.09            0.60
# Thaw_Depth       0.16 -0.13 -0.02       1.00           0.18     0.34           -0.06
# Soil_Temp_10cm   0.09  0.20  0.51       0.18           1.00     0.19            0.64
# Soil_VWC        -0.11  0.11  0.09       0.34           0.19     1.00            0.11
# Air_Temperature  0.18  0.17  0.60      -0.06           0.64     0.11            1.00
# 
# n= 155 
# 
# 
# P values 
#                 NEE    GPP    RECO   Thaw_Depth Soil_Temp_10cm Soil_VWC Air_Temperature
# NEE                    0.0000 0.0967 0.0463     0.2819         0.1796   0.0257         
# GPP             0.0000        0.0000 0.0947     0.0111         0.1713   0.0380         
# RECO            0.0967 0.0000        0.8328     0.0000         0.2771   0.0000         
# Thaw_Depth      0.0463 0.0947 0.8328            0.0235         0.0000   0.4957         
# Soil_Temp_10cm  0.2819 0.0111 0.0000 0.0235                    0.0199   0.0000         
# Soil_VWC        0.1796 0.1713 0.2771 0.0000     0.0199                  0.1590         
# Air_Temperature 0.0257 0.0380 0.0000 0.4957     0.0000         0.1590   
```

#Plot correlation matrix - tiled figure 
```{r}
# Convert correlation matrix to long format for plotting
cor_long2 <- reshape2::melt(chamber_SPcorr_result )
names(cor_long2) <- c("Var1", "Var2", "Correlation")


#correlation matrix using spearman's corr, for non-parametric / not normally distributed data 
ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation values only for strong correlations
  geom_text(data = subset(cor_long2, abs(Correlation) > 0.3),  # Only show stronger correlations
            aes(label = sprintf("%.2f", Correlation),
                 color = abs(Correlation) < 0.5), # Changes text color based on correlation strength
            size = 3,
            fontface = "bold") + #to make numbers in tiles bold
  scale_fill_gradient2(low = "red", 
                      mid = "white", 
                      high = "blue", 
                      midpoint = 0,
                      limits = c(-1, 1),
                      breaks = c(-1, -0.5, 0, 0.5, 1)) +
    scale_color_manual(values = c("white", "black"), guide = "none") +  # For text visibility
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber C Fluxes and Environmental Variables")


```





```{r}
#double checking 
cor.test(df_long$NEE, df_long$GPP, method="spearman")
cor.test(chamber_corr_matrix$NEE, chamber_corr_matrix$GPP, method="spearman")
cor.test(chamber_corr_matrix$RECO, chamber_corr_matrix$GPP, method="spearman")
```

#Adding significance codes to correlation matrix figure - coding help from Claude, keeping the redundant steps for reference 
```{r}
# Step 1: Calculate correlation and p-values
library(Hmisc)
corr_result <- rcorr(as.matrix(chamber_corr_matrix), type = "spearman")

# Step 2: Extract correlation coefficients and p-values
cor_coef <- corr_result$r
cor_p <- corr_result$P

# Step 3: Create a long format data frame for plotting
cor_long <- reshape2::melt(cor_coef)
names(cor_long) <- c("Var1", "Var2", "Correlation")

# Step 4: Add p-values to the long format data
p_long <- reshape2::melt(cor_p)
cor_long$p_value <- p_long$value

# Step 5: Add significance indicators
cor_long$significance <- ""
cor_long$significance[cor_long$p_value < 0.05] <- "*"    # p < 0.05
cor_long$significance[cor_long$p_value < 0.01] <- "**"   # p < 0.01
cor_long$significance[cor_long$p_value < 0.001] <- "***" # p < 0.001

# Step 6: Create the correlation plot with significance indicators
ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation coefficients with significance stars
  geom_text(aes(label = paste0(round(Correlation, 2), significance)), 
            color = "black", size = 3) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,
                      limits = c(-1, 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  labs(title = "Correlation Matrix of Soil Chamber C Fluxes and Environmental Variables")




#correlation matrix using spearman's corr, for non-parametric / not normally distributed data 
ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation values only for strong correlations
  geom_text(data = subset(cor_long, abs(Correlation) > 0.3),  # Only show stronger correlations
            aes(label = paste0(sprintf("%.2f", Correlation), significance),
                 color = abs(Correlation) < 0.5), # Changes text color based on correlation strength
            size = 4.5,
            fontface = "bold") + #to make numbers in tiles bold
  scale_fill_gradient2(low = "red", 
                      mid = "white", 
                      high = "blue", 
                      midpoint = 0,
                      limits = c(-1, 1),
                      breaks = c(-1, -0.5, 0, 0.5, 1)) +
    scale_color_manual(values = c("white", "black"), guide = "none") +  # For text visibility
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber C Fluxes and Environmental Variables")


#Display corr text on tiles if it's >0.3 OR if it has a sig p 

ggplot(cor_long2, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  # Add correlation values for EITHER strong correlations OR significant ones
  geom_text(data = subset(cor_long2, abs(Correlation) > 0.3 | p_value < 0.05), #the | represents "OR" so this displays correlations above 0.3 OR significant correlations; also use "abs" to use absolute value of the corr so it's printed whether it's positive or negative*  
            aes(label = paste0(sprintf("%.2f", Correlation), significance),
                color = abs(Correlation) < 0.5), # Changes text color based on correlation strength
            size = 4.5,
            fontface = "bold") + #to make numbers in tiles bold
  scale_fill_gradient2(low = "red", 
                      mid = "white", 
                      high = "blue", 
                      midpoint = 0,
                      limits = c(-1, 1),
                      breaks = c(-1, -0.5, 0, 0.5, 1)) +
    scale_color_manual(values = c("white", "black"), guide = "none") +  # For text visibility
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +                 
  labs(title = "Correlation Matrix of Soil Chamber C Fluxes and Environmental Variables")

 #saves the figure directly above
ggsave("C:/Users/kkent/Documents/Github Flux Network/Council_Flux_Analysis_Paper/Council Figures/Soil Chambers/soil_chamber_Cflux_corr_matrix.png",
       width = 9, height = 5, dpi = 600)
```

```{r}
ggplot(df_long, aes(x=thawdepth, y = NEE))+
  geom_point()
```
# ** TO DO************
```{r}
# Create visualization functions to reuse
plot_flux_by_position <- function(flux_var, df_long) {
  ggplot(df_long, aes_string(x = "landscape_position", y = flux_var, fill = "landscape_position")) +
    geom_boxplot() +
    facet_wrap(~inundated, labeller = labeller(inundated = c("N" = "Not Inundated", "Y" = "Inundated"))) +
    labs(title = paste(flux_var, "by Landscape Position and Inundation Status"),
         x = "Landscape Position", y = flux_var) +
    theme_minimal() +
    theme(legend.position = "none")
}

# Create scatter plots with smooth lines for continuous variables
plot_flux_by_continuous <- function(flux_var, cont_var, df_long) {
  ggplot(df_long, aes_string(x = cont_var, y = flux_var, color = "inundated")) + #color = landpos 
    geom_point(alpha = 0.6) +
    geom_smooth(method = "loess", se = TRUE) +
    facet_wrap(~landscape_position) +
    labs(title = paste(flux_var, "vs", cont_var, "by Landscape Position"),
         x = cont_var, y = flux_var) +
    theme_minimal()
}

# Generate these plots for each flux
plot_flux_by_position("GPP_umol", df_long)
plot_flux_by_position("NEE_umol", df_long)
plot_flux_by_position("RECO_umol", df_long)
plot_flux_by_position("GPP_umol", df_long)
plot_flux_by_position("FCH4_umol", df_long)
plot_flux_by_continuous("GPP_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous("GPP_umol", "air_temp", df_long)
plot_flux_by_continuous("GPP_umol", "VWC", df_long)
plot_flux_by_continuous("GPP_umol", "thawdepth", df_long)
plot_flux_by_continuous("NEE_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous("NEE_umol", "air_temp", df_long)
plot_flux_by_continuous("NEE_umol", "VWC", df_long)
plot_flux_by_continuous("NEE_umol", "thawdepth", df_long)
plot_flux_by_continuous("RECO_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous("RECO_umol", "air_temp", df_long)
plot_flux_by_continuous("RECO_umol", "VWC", df_long)
plot_flux_by_continuous("RECO_umol", "thawdepth", df_long)
plot_flux_by_continuous("FCH4_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous("FCH4_umol", "air_temp", df_long)
plot_flux_by_continuous("FCH4_umol", "VWC", df_long)
plot_flux_by_continuous("FCH4_umol", "thawdepth", df_long)
```

```{r}
# Create scatter plots with smooth lines for continuous variables, faceted by both landscape position and inundation # loess = smooth line, lm = linear line
plot_flux_by_position2 <- function(flux_var, df_long) {
  ggplot(df_long, aes_string(x = "landscape_position", y = flux_var, fill = "landscape_position")) +
    geom_boxplot() +
  facet_wrap(~inundated, 
               labeller = labeller(inundated = c("N" = "Not Inundated", "Y" = "Inundated"))) +
    labs(title = paste(flux_var, "by Landscape Position and Inundation Status"),
         x = "Landscape Position", y = flux_var) +
    theme_minimal() +
    theme(legend.position = "none")
}


plot_flux_by_continuous2 <- function(flux_var, cont_var, df_long) {
  ggplot(df_long, aes_string(x = cont_var, y = flux_var, color = "inundated")) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE) + #method = "loess"
    facet_wrap(~landscape_position) +
    labs(title = paste(flux_var, "vs", cont_var, "by Landscape Position and Inundation Status"),
         x = cont_var, y = flux_var) +
    theme_minimal()
}

plot_flux_by_position2("GPP_umol", df_long)
plot_flux_by_position2("NEE_umol", df_long)
plot_flux_by_position2("RECO_umol", df_long)
plot_flux_by_position2("GPP_umol", df_long)
plot_flux_by_position2("FCH4_umol", df_long)
plot_flux_by_continuous2("GPP_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous2("GPP_umol", "air_temp", df_long)
plot_flux_by_continuous2("GPP_umol", "VWC", df_long)
plot_flux_by_continuous2("GPP_umol", "thawdepth", df_long)
plot_flux_by_continuous2("NEE_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous2("NEE_umol", "air_temp", df_long)
plot_flux_by_continuous2("NEE_umol", "VWC", df_long)
plot_flux_by_continuous2("NEE_umol", "thawdepth", df_long)
plot_flux_by_continuous2("RECO_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous2("RECO_umol", "air_temp", df_long)
plot_flux_by_continuous2("RECO_umol", "VWC", df_long)
plot_flux_by_continuous2("RECO_umol", "thawdepth", df_long)
plot_flux_by_continuous2("FCH4_umol", "soil_temp_10_cm", df_long)
plot_flux_by_continuous2("FCH4_umol", "air_temp", df_long)
plot_flux_by_continuous2("FCH4_umol", "VWC", df_long)
plot_flux_by_continuous2("FCH4_umol", "thawdepth", df_long)
```



#Looking at fluxes with interacting enviro variables 
```{r}
# NEE and soil + VWC
ggplot(df_long, aes(x = soil_temp_10_cm, y = NEE_umol, color = VWC)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(title = "NEE Flux vs Soil Temp",
       subtitle = "Colored by Soil Water Content",
       x = "Temperature",
       y = "NEE Flux")


#This code breaks SWC into 3 even levels (low, intermediate, high) to show the change in temp relationship depending on SWC
ggplot(df_long, aes(x = soil_temp_10_cm, y = NEE_umol)) +
  geom_point(aes(color = VWC), alpha = 0.3) +
  geom_smooth(aes(group = cut(VWC, breaks = 3)), method = "lm") +
  theme_minimal() +
  labs(title = "NEE Flux vs Temperature",
       subtitle = "Colored by Soil Water Content",
       x = "Temperature (°C)",
       y = "NEE Flux",
       color = "Soil Water Content")


```
#Bins - for soil temp, air temp, VWC, and thawdepth  
```{r}
#check ranges of variables 
summary(df_long$VWC) #min = 0, max = 90.3
summary(df_long$soil_temp_10_cm) #min = 0.6, max = 16.5
summary(df_long$thawdepth) #min= 10, max = 118cm
summary(df_long$air_temp) #min = 3.7, max = 27.10 

# Create VWC bins
df_long$VWC_bin <- cut(df_long$VWC, 
                       breaks = c(-0.1, 1, 20, 40, 60, 95), #max VWC is 90.30
                       labels = c("Dry (0-1%)",
                                "Very Dry (1-20%)",
                                "Dry (20-40%)",
                                "Wet (40-60%)",
                                "Very Wet (60-95%)"))
                          #include.lowest = TRUE)



# Create soil temperature bins - min is 6, max is 16.4 
df_long$soiltemp_bin <- cut(df_long$soil_temp_10_cm, 
                       breaks = c(0, 4, 8, 12, 16.5),  
                       labels = c("Cold (0 to 4°C)",
                                "Cool (4 to 8°C)",
                                "Warm (8 to 12°C)",
                                "Warmest (12 to 16.5°C)"))

# Create soil temperature bins - min is 6, max is 16.4 
df_long$thawdepth_bin <- cut(df_long$thawdepth, 
                       breaks = c(0, 25, 50, 75, 120),  
                       labels = c("Shallowest (0 to 25cm)",
                                "Shallow (25 to 50cm)",
                                "Deep (50 to 75cm)",
                                "Deepest (75 to 120cm)"))


# Create soil temperature bins - min is 6, max is 16.4 
df_long$airtemp_bin <- cut(df_long$air_temp, 
                       breaks = c(0, 5, 10, 15, 20, 30),  
                       labels = c("Cold (0 to 5°C)",
                                "Cool (5 to 10°C)",
                                "Warm (10 to 15°C)",
                                "Warmer (15 to 20°C)",
                                "Hot (20 - 30°C"))
```

#to test whether a linear or quadratic or polynomial line fits data best 
```{r}
# For a specific bin
bin_data <- df_long[df_long$VWC_bin == "Wet (40-60%)",]

# Fit linear model
linear_model <- lm(NEE_umol ~ soil_temp_10_cm, data = bin_data)

# Fit quadratic model
quadratic_model <- lm(NEE_umol ~ soil_temp_10_cm + I(soil_temp_10_cm^2), data = bin_data)

# Compare models
anova(linear_model, quadratic_model)

# Get R-squared and p-values for the better model
summary(quadratic_model)
```


#Interaction of Enviro Vars
```{r}
# soiltemp and VWC 
ggplot(df_long, aes(x = soil_temp_10_cm, y = VWC, color = VWC)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~VWC_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "VWC vs Temperature by Moisture Range",
       x = "Soil Temperature (°C)",
       y = "VWC",
       color = "VWC (%)") +
  theme_minimal()

#NEE x soiltemp and thawdepth
ggplot(df_long, aes(x = soil_temp_10_cm, y = NEE_umol, color = thawdepth)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~thawdepth_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "NEE vs Temperature by Thaw Depth Range",
       x = "Soil Temperature (°C)",
       y = "NEE Flux",
       color = "Thaw Depth (cm)") +
  theme_minimal()


#NEE x airtemp and VWC
ggplot(df_long, aes(x = air_temp, y = NEE_umol, color = VWC)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~VWC_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "NEE vs Air Temperature by VWC Range",
       x = "AirTemperature (°C)",
       y = "NEE Flux",
       color = "VWC (%)") +
  theme_minimal()
```







#NEE x interacting enviro vars
```{r}

# NEE x soiltemp and VWC 
ggplot(df_long, aes(x = soil_temp_10_cm, y = NEE_umol, color = VWC)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~VWC_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "NEE vs Temperature by Moisture Range",
       x = "Soil Temperature (°C)",
       y = "NEE Flux",
       color = "VWC (%)") +
  theme_minimal()

#NEE x soiltemp and thawdepth
ggplot(df_long, aes(x = soil_temp_10_cm, y = NEE_umol, color = thawdepth)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~thawdepth_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "NEE vs Temperature by Thaw Depth Range",
       x = "Soil Temperature (°C)",
       y = "NEE Flux",
       color = "Thaw Depth (cm)") +
  theme_minimal()


#NEE x airtemp and VWC
ggplot(df_long, aes(x = air_temp, y = NEE_umol, color = VWC)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~VWC_bin, scales = "free") +
  scale_color_viridis_c() +
  labs(title = "NEE vs Air Temperature by VWC Range",
       x = "AirTemperature (°C)",
       y = "NEE Flux",
       color = "VWC (%)") +
  theme_minimal()




```




```{r}
# Function to test differences by landscape position
test_by_position <- function(flux_var, df_long) {
  # Formula creation
  formula <- as.formula(paste(flux_var, "~ landscape_position"))
  
  # Check normality first
  shapiro_test <- shapiro.test(df_long[[flux_var]])
  
  if (shapiro_test$p.value < 0.05) {
    # Non-parametric test if not normal
    kruskal_result <- kruskal.test(formula, data = df_long)
    return(list(test = "Kruskal-Wallis", result = kruskal_result))
  } else {
    # ANOVA if normal
    anova_result <- aov(formula, data = df_long)
    return(list(test = "ANOVA", result = summary(anova_result), 
                posthoc = TukeyHSD(anova_result)))
  }
}

# Test interaction with inundation
test_interaction <- function(flux_var, df_long) {
  formula <- as.formula(paste(flux_var, "~ landscape_position * inundated"))
  model <- aov(formula, data = df_long)
  return(summary(model))
}

# Apply to each flux
 gpp_position_test <- test_by_position("GPP", df)
 gpp_interaction <- test_interaction("GPP", df)
```


#RLM and KW tests of continuous vars by landscape position 
```{r}
rlm1 <- rlm(VWC ~ landscape_position, data = df_NEE)
Anova(rlm1, type = "II") 

#soil temp x landpos p = 0.7, not sig 
#VWC p < 0.001 *SIG* 
#air temp p=0.8, not sig 
#thawdepth p < 0.001 *SIG*

rlm2 <- rlm(air_temp ~ landscape_position, data = df_NEE)
Anova(rlm2, type = "II") 

#soil temp x landpos p = 0.7, not sig 
#VWC p < 0.001 *SIG* 
#air temp p=0.8, not sig 
#thawdepth p < 0.001 *SIG*

# Test landscape position on continuous variable 
kruskal.test(thawdepth ~ landscape_position, data = df_NEE)

#soil temp x landpos p = 0.6, not sig 
#VWC p < 0.001 *SIG* 
#air temp p=0.8, not sig 
#thawdepth p < 0.001 *SIG*

# Test inundation on continuous variable 
kruskal.test(air_temp ~ inundated, data = df_NEE)

#soil temp x inundated p = 0.008 *SIG 
#VWC p < 0.001 *SIG* 
#air temp p=0.4, not sig 
#thawdepth p < 0.001 *SIG*

#suggests VWC and thawdepth vary significantly among landpos; and all variables except for air temp vary significantly by inundation --> maybe run PCA on this?


```
#Variance partitioning - coding help from Claude 
```{r}
# Venn diagram showing how much variance in your environmental variables (thaw depth, soil temperature, and VWC) is explained by landscape position and inundation. Shows the unique variance explained by each factor, plus their shared variance.


library(vegan)
# Prepare data
env_vars <- df_long %>% select(thawdepth, soil_temp_10_cm, VWC)
cat_vars <- df_long %>% select(landscape_position, inundated)

# Convert categorical variables to dummy variables
cat_dummy <- model.matrix(~ landscape_position + inundated - 1, data = cat_vars)

# Variance partitioning
vp <- varpart(env_vars, 
              cat_dummy[,grep("landscape", colnames(cat_dummy))], 
              cat_dummy[,grep("inundated", colnames(cat_dummy))])

# Plot results
plot(vp, digits = 2, Xnames = c("Landscape Position", "Inundation"))

#Shows landpos along explains ~3% of variance in the numeric enviro factors, inundation explains about 11% of variance, and their shared variance = ~ 5%. Residuals show 80% of variance in the enviro factors is explained by something else 



library(vegan)
# Prepare data
env_vars2 <- df_long %>% select(thawdepth, soil_temp_10_cm, VWC)
flux_varFCH4 <- df_long %>% select(FCH4)
flux_varNEE <- df_long %>% select(NEE)
flux_vars <- df_long %>% select(NEE, GPP, RECO)
cat_vars2 <- df_long %>% select(landscape_position, inundated)

# Convert categorical variables to dummy variables
cat_dummy <- model.matrix(~ landscape_position + inundated - 1, data = cat_vars2)

# Variance partitioning
vp2 <- varpart(flux_vars, 
              cat_dummy[,grep("landscape", colnames(cat_dummy))], 
              cat_dummy[,grep("inundated", colnames(cat_dummy))])

# Plot results
plot(vp2, digits = 2, Xnames = c("Landscape Position", "Inundation"))
```




#PCA - only works with continuous/numeric data 

```{r}
# Modified PCA code
library(FactoMineR)

env_vars <- select(df_long, soil_temp_10_cm, air_temp, VWC, thawdepth)
pca_result <- PCA(env_vars, graph = FALSE)

# Extract the coordinates and add back the categorical variables
pca_df <- data.frame(pca_result$ind$coord)
pca_df$landscape_position <- df_long$landscape_position
pca_df$inundated <- df_long$inundated

# Now create the plot
ggplot(pca_df, aes(x = Dim.1, y = Dim.2, color = landscape_position, shape = inundated)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(title = "PCA of Environmental Variables")
```


#PCA for each landscape position 
```{r}
# Function to create PCA for a specific landscape position
create_position_pca <- function(position_name, data) {
  # Filter data for the specific landscape position
  position_data <- data[data$landscape_position == position_name, ]
  
  # Select environmental variables
  env_vars <- position_data[, c("soil_temp_10_cm", "air_temp", "VWC", "thawdepth")]
  
  # Run PCA if we have enough data
  if(nrow(env_vars) > 3) { # Need at least 4 rows for PCA with 4 variables
    pca_result <- prcomp(env_vars, scale. = TRUE)
    
    # Create plot data
    pca_plot_data <- data.frame(
      PC1 = pca_result$x[,1],
      PC2 = pca_result$x[,2],
      inundated = position_data$inundated
    )
    
    # Create the plot
    plot <- ggplot(pca_plot_data, aes(x = PC1, y = PC2, shape = inundated, color = inundated)) +
      geom_point(size = 3) +
      stat_ellipse() +
      labs(title = paste("PCA of Environmental Variables -", position_name),
           x = "PC1", y = "PC2") +
      theme_minimal()
    
    # Additional info to return
    var_explained <- summary(pca_result)$importance[2, 1:2] * 100
    loadings <- pca_result$rotation
    
    return(list(plot = plot, 
                variance_explained = var_explained, 
                loadings = loadings,
                pca_result = pca_result))
  } else {
    return(NULL)
  }
}

# Apply to each landscape position
lowland_pca <- create_position_pca("lowland", df_long)
slope_pca <- create_position_pca("slope", df_long)
upland_pca <- create_position_pca("upland", df_long)

# View the plots
lowland_pca$plot
slope_pca$plot
upland_pca$plot

# Check which environmental variables drive variation within each position
lowland_pca$loadings
slope_pca$loadings
upland_pca$loadings

#LOWLAND
#                       PC1        PC2        PC3        PC4
# soil_temp_10_cm 0.6024493  0.2616177 -0.4371061 -0.6144504
# air_temp        0.4259289  0.6731998  0.1778673  0.5777108
# VWC             0.5030634 -0.3491144  0.7649874 -0.1996011
# thawdepth       0.4500741 -0.5970572 -0.4382873  0.4988589

#SLOPE
#                        PC1        PC2        PC3        PC4
# soil_temp_10_cm -0.6003161  0.2636395  0.6747952  0.3387716
# air_temp        -0.3520804  0.6710585 -0.2793839 -0.5896308
# VWC             -0.6086657 -0.2416271 -0.6448954  0.3940207
# thawdepth       -0.3810328 -0.6494544  0.2251798 -0.6183179

#UPLAND
#                        PC1        PC2        PC3         PC4
# soil_temp_10_cm -0.6500843 -0.3233936  0.1376162 -0.67369777
# air_temp        -0.6969717 -0.1066950 -0.1408980  0.69497793
# VWC              0.2081485 -0.6482817 -0.7313549 -0.03905338
# thawdepth        0.2197611 -0.6810019  0.6529392  0.24821732

# To add carbon fluxes as supplementary variables
position_data <- df_long[df_long$landscape_position == "lowland", ]
env_vars <- position_data[, c("soil_temp_10_cm", "air_temp", "VWC", "thawdepth")]
flux_vars <- position_data[, c("GPP", "NEE", "RECO", "FCH4")]

pca_result <- PCA(env_vars, graph = FALSE, 
                  quanti.sup = 1:ncol(flux_vars), # Add flux variables as supplementary
                  row.w = rep(1, nrow(env_vars)))
```


#PCA of continuous variables and C fluxes 
```{r}
# Step 1: Run PCA on your environmental variables
env_vars <- df_long %>% 
  select(soil_temp_10_cm, air_temp, VWC, thawdepth)

# Run PCA
pca_result <- prcomp(env_vars, scale = TRUE)

# Summary of PCA
summary(pca_result)

# Look at variable contributions to principal components
pca_loadings <- pca_result$rotation
print(pca_loadings)

# Step 2: Plot the PCA biplot
biplot(pca_result, scale = 0)

# Step 3: Enhanced visualization with ggplot2
library(ggplot2)
library(ggfortify)

# Create PCA biplot with ggplot2

#NEE
autoplot(pca_result, data = df_long, colour = 'NEE_umol', shape = 'inundated',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 4) +
  theme_minimal() +
  labs(title = "PCA of Environmental Variables",
       subtitle = "Colored by NEE flux value")

#Gpp
autoplot(pca_result, data = df_long, colour = 'GPP_umol', shape = 'inundated',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 4) +
  theme_minimal() +
  labs(title = "PCA of Environmental Variables",
       subtitle = "Colored by GPP flux value")
#RECO
autoplot(pca_result, data = df_long, colour = 'RECO_umol', shape = 'inundated',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 4) +
  theme_minimal() +
  labs(title = "PCA of Environmental Variables",
       subtitle = "Colored by RECO flux value")

#FCH4
autoplot(pca_result, data = df_long, colour = 'FCH4_umol', shape = 'inundated',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 4) +
  theme_minimal() +
  labs(title = "PCA of Environmental Variables",
       subtitle = "Colored by FCH4 flux value")



# Step 4: Correlation between PC scores and carbon fluxes
df_with_pcs <- cbind(df_long, pca_result$x)

# Correlation between principal components and carbon fluxes
cor_pcs_fluxes <- cor(df_with_pcs[, c("PC1", "PC2", "PC3", "PC4")], 
                      df_with_pcs[, c("NEE_umol", "GPP_umol", "RECO_umol", "FCH4_umol")],
                      method = "spearman")
print(cor_pcs_fluxes)

#suggests PC1 is a thermal gradient (mainly contributed to by air temp and soil temp), and PC2 is a moisture & thaw gradient, driven by VWC and thawdepth
#PC1 - pos values = cooler temps (due to negative relationship in loadings, so inverse relationship), neg values = warmer 
#PC2 - pos values = wetter soils and deeper thaw depths, neg values = drier soils and shallow thaw depths 
#Pc1 = ~42% of variance, so thermal plays a large role 
#PC2 = ~35% of variance
```

#Piece-wise SEM

#### using glm 
```{r}
library(piecewiseSEM)

#dataset without any NAs
df_long_cc <- df_long%>%
  select(plot_ID, plot_type, landscape_position, measurement_date, time, date, VWC, air_temp, FCH4_umol, NEE_umol, RECO_umol, GPP_umol, inundated, soil_temp_10_cm, thawdepth)


any(is.na(df_long_cc))

# Create dummy variables for landscape position since it can only handle numeric data 
df_long_cc$lp_slope <- ifelse(df_long_cc$landscape_position == "slope", 1, 0)
df_long_cc$lp_lowland <- ifelse(df_long_cc$landscape_position == "lowland", 1, 0)
#df_long_cc$lp_upland <- ifelse(df_long_cc$landscape_position == "upland", 1, 0) #swapped out for lowland to get lowland values, using upland as reference point
df_long_cc$inundated_numeric <- ifelse(df_long_cc$inundated == "Y", 1, 0)


# Create individual models - make sure these are all lm or glm objects
model_soil <- glm(soil_temp_10_cm ~ lp_slope + lp_lowland + inundated_numeric, data = df_long_cc)
model_air <- glm(air_temp ~ lp_slope + lp_lowland + inundated_numeric, data = df_long_cc)
model_vwc <- glm(VWC ~ lp_slope + lp_lowland + inundated_numeric, data = df_long_cc)
model_thaw <- glm(thawdepth ~ lp_slope + lp_lowland + inundated_numeric, data = df_long_cc)
model_nee <- glm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + lp_slope + lp_lowland + inundated_numeric, data = df_long_cc)

# Specify the model structure correctly
psem_model <- psem(
  model_soil,
  model_air,
  model_vwc, 
  model_thaw,
  model_nee
)

# Get summary
summary(psem_model)

plot(psem_model)
```

#Visualizing piecewise SEM - pSEM

#TO DO: include lowland landpos 

```{r}
#from glm pSEM - pSEM does not work with rlm 

#have to add coefficients in by hand ** based on pSEM results 

library(DiagrammeR)

# Create a basic path diagram
grViz("
  digraph path_model {
    rankdir=LR;
    
    # Nodes
    node [shape=box]
    lp_slope [label='Slope Position']
    lp_upland [label='Upland Position']
    inundated [label='Inundation']
    
    node [shape=ellipse]
    soil_temp [label='Soil Temp']
    air_temp [label='Air Temp']  
    vwc [label='VWC']
    thaw [label='Thaw Depth']
    nee [label='NEE']
    
    # Edges with significant paths
    lp_slope -> vwc [label='-0.18*']
    lp_upland -> vwc [label='-0.24**'] 
    lp_upland -> thaw [label='-0.22*']
    inundated -> soil_temp [label='0.20*']
    inundated -> vwc [label='0.39***']
    inundated -> thaw [label='0.31***']
    
    soil_temp -> air_temp [style=dashed, label='']
    air_temp -> nee [label='0.33**']
    vwc -> nee [label='-0.19*']
    vwc -> thaw [style=dashed, label=''] 
    air_temp -> thaw [style=dashed, label='']
    thaw -> nee [label='0.16.']
  }
")
```
#TO DO
#Need to add lowland in....? maybe run psem with diff reference point to get the lowland effect **
```{r}
library(DiagrammeR)

# Create a complete path diagram that includes lowland position
grViz("
  digraph path_model {
    rankdir=LR;
    
    # Nodes
    node [shape=box]
    lp_lowland [label='Lowland Position']
    lp_slope [label='Slope Position']
    lp_upland [label='Upland Position']
    inundated [label='Inundation']
    
    node [shape=ellipse]
    soil_temp [label='Soil Temp']
    air_temp [label='Air Temp']  
    vwc [label='VWC']
    thaw [label='Thaw Depth']
    nee [label='NEE']
    
    # Edges for lowland (as reference category)
    lp_lowland -> vwc [label='+', color='green']
    lp_lowland -> thaw [label='+', color='green']
    
    # Edges for slope and upland (using actual coefficients)
    lp_slope -> vwc [label='-0.18*', color='red']
    lp_upland -> vwc [label='-0.24**', color='red'] 
    lp_upland -> thaw [label='-0.22*', color='red']
    
    # Other significant paths
    inundated -> soil_temp [label='0.20*', color='green']
    inundated -> vwc [label='0.39***', color='green']
    inundated -> thaw [label='0.31***', color='green']
    
    # Relationships between environmental variables  
    soil_temp -> air_temp [style=dashed]
    air_temp -> nee [label='0.33**', color='green']
    vwc -> nee [label='-0.19*', color='red']
    vwc -> thaw [style=dashed] 
    air_temp -> thaw [style=dashed]
    thaw -> nee [label='0.16', color='green']
    
    # Visual enhancements
    {rank=same; lp_lowland; lp_slope; lp_upland; inundated}
    {rank=same; soil_temp; air_temp; vwc; thaw}
  }
")
```


```{r}
# For each component model in your piecewise SEM
# Extract and check the residuals

# Access your individual models in the psem directly
model_list <- list(
  soil_temp = model_soil,
  air_temp = model_air,
  vwc = model_vwc,
  thaw = model_thaw,
  nee = model_nee
)

# Then check residuals as shown above

# Check residuals for each model
par(mfrow=c(2,2))
for (i in seq_along(model_list)) {
  model_name <- names(model_list)[i]
  if (is.null(model_name)) model_name <- paste("Model", i)
  
  resids <- residuals(model_list[[i]])
  
  # Histogram of residuals
  hist(resids, main=paste("Residuals -", model_name), 
       xlab="Residuals", freq=FALSE)
  curve(dnorm(x, mean=mean(resids), sd=sd(resids)), add=TRUE, col="red")
  
  # QQ plot
  qqnorm(resids, main=paste("QQ Plot -", model_name))
  qqline(resids, col="red")
}



# Shapiro-Wilk test for normality on each model's residuals
residual_tests <- list()
for (i in seq_along(model_list)) {
  model_name <- ifelse(is.null(names(model_list)[i]), paste("Model", i), names(model_list)[i])
  resids <- residuals(model_list[[i]])
  sw_test <- shapiro.test(resids)
  residual_tests[[model_name]] <- list(
    W = sw_test$statistic,
    p_value = sw_test$p.value,
    normal = sw_test$p.value > 0.05
  )
}

# Print results
print(residual_tests)





```


#checking linearity among flux and enviro variables 
####NEE
```{r}
# Check each relationship for linearity
# Example for NEE model
par(mfrow=c(2,2))

# Soil temp vs NEE
plot(df_long_cc$soil_temp_10_cm, df_long_cc$NEE_umol, 
     main="NEE vs Soil Temp", xlab="Soil Temp", ylab="NEE")
lines(lowess(df_long_cc$soil_temp_10_cm, df_long_cc$NEE_umol), col="red")
abline(lm(NEE_umol ~ soil_temp_10_cm, data=df_long_cc), col="blue")

# Air temp vs NEE
plot(df_long_cc$air_temp, df_long_cc$NEE_umol, 
     main="NEE vs Air Temp", xlab="Air Temp", ylab="NEE")
lines(lowess(df_long_cc$air_temp, df_long_cc$NEE_umol), col="red")
abline(lm(NEE_umol ~ air_temp, data=df_long_cc), col="blue")

# VWC vs NEE
plot(df_long_cc$VWC, df_long_cc$NEE_umol, 
     main="NEE vs VWC", xlab="VWC", ylab="NEE")
lines(lowess(df_long_cc$VWC, df_long_cc$NEE_umol), col="red")
abline(lm(NEE_umol ~ VWC, data=df_long_cc), col="blue")

# Thaw depth vs NEE
plot(df_long_cc$thawdepth, df_long_cc$NEE_umol, 
     main="NEE vs Thaw Depth", xlab="Thaw Depth", ylab="NEE")
lines(lowess(df_long_cc$thawdepth, df_long_cc$NEE_umol), col="red")
abline(lm(NEE_umol ~ thawdepth, data=df_long_cc), col="blue")
```
```{r}
# Check linearity using correlation tests
response_vars <- c("soil_temp_10_cm", "air_temp", "VWC", "thawdepth", "NEE_umol")
predictor_vars <- c("soil_temp_10_cm", "air_temp", "VWC", "thawdepth", 
                   "lp_slope", "lp_upland", "inundated_numeric")

# Compare linear correlation (pearson) vs rank correlation (non-parametric, spearman)
# If they're similar, relationship is likely linear
for (resp in response_vars) {
  cat("\nChecking linearity for response:", resp, "\n")
  
  for (pred in predictor_vars) {
    if (resp != pred) {
      pearson_r <- cor(df_long_cc[[resp]], df_long_cc[[pred]], method="pearson", use="complete.obs")
      spearman_r <- cor(df_long_cc[[resp]], df_long_cc[[pred]], method="spearman", use="complete.obs")
      
      cat(pred, "- Pearson r:", round(pearson_r, 3), 
          "Spearman r:", round(spearman_r, 3),
          "Diff:", round(abs(pearson_r - spearman_r), 3), "\n")
      
      # Large differences suggest non-linearity
      if (abs(pearson_r - spearman_r) > 0.2) {
        cat("  Potential non-linear relationship detected\n")
      }
    }
  }
}

#Potential non-linear relationships - pearson vs spearman correlations with a difference > 0.1:
```
#bootstrapping and piecewise SEM - code from Claude 

#### NEE using rlm
```{r}
library(boot)
library(piecewiseSEM)
library(MASS)

# Simplified bootstrap function focusing just on direct and indirect effects
boot_psem_robust <- function(data, indices) {
  # Create bootstrap sample
  boot_data <- data[indices, ]
  
  # Fit models on bootstrap sample - using robust regression here 
  b_soil <- rlm(soil_temp_10_cm ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_air <- rlm(air_temp ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_vwc <- rlm(VWC ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_thaw <- rlm(thawdepth ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_nee <- rlm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
             lp_slope + lp_upland + inundated_numeric, data = boot_data)
  
  # Direct effects
  direct_slope <- coef(b_nee)["lp_slope"]
  direct_upland <- coef(b_nee)["lp_upland"]
  direct_inundated <- coef(b_nee)["inundated_numeric"]
  
  # Environmental effects on NEE
  effect_soil <- coef(b_nee)["soil_temp_10_cm"]
  effect_air <- coef(b_nee)["air_temp"]
  effect_vwc <- coef(b_nee)["VWC"]
  effect_thaw <- coef(b_nee)["thawdepth"]
  
  # Landscape effects on environment
  slope_soil <- coef(b_soil)["lp_slope"]
  slope_air <- coef(b_air)["lp_slope"]
  slope_vwc <- coef(b_vwc)["lp_slope"]
  slope_thaw <- coef(b_thaw)["lp_slope"]
  
  upland_soil <- coef(b_soil)["lp_upland"]
  upland_air <- coef(b_air)["lp_upland"]
  upland_vwc <- coef(b_vwc)["lp_upland"]
  upland_thaw <- coef(b_thaw)["lp_upland"]
  
  inundated_soil <- coef(b_soil)["inundated_numeric"]
  inundated_air <- coef(b_air)["inundated_numeric"]
  inundated_vwc <- coef(b_vwc)["inundated_numeric"]
  inundated_thaw <- coef(b_thaw)["inundated_numeric"]
  
  # Individual indirect paths
  slope_via_soil <- slope_soil * effect_soil
  slope_via_air <- slope_air * effect_air
  slope_via_vwc <- slope_vwc * effect_vwc
  slope_via_thaw <- slope_thaw * effect_thaw
  
  upland_via_soil <- upland_soil * effect_soil
  upland_via_air <- upland_air * effect_air
  upland_via_vwc <- upland_vwc * effect_vwc
  upland_via_thaw <- upland_thaw * effect_thaw
  
  inundated_via_soil <- inundated_soil * effect_soil
  inundated_via_air <- inundated_air * effect_air
  inundated_via_vwc <- inundated_vwc * effect_vwc
  inundated_via_thaw <- inundated_thaw * effect_thaw
  
  # Total indirect effects
  slope_indirect <- slope_via_soil + slope_via_air + slope_via_vwc + slope_via_thaw
  upland_indirect <- upland_via_soil + upland_via_air + upland_via_vwc + upland_via_thaw
  inundated_indirect <- inundated_via_soil + inundated_via_air + inundated_via_vwc + inundated_via_thaw
  
  # Total effects
  slope_total <- direct_slope + slope_indirect
  upland_total <- direct_upland + upland_indirect
  inundated_total <- direct_inundated + inundated_indirect
  
  # Return all values as a named vector
  return(c(
    # Direct effects on NEE
    direct_slope = direct_slope,
    direct_upland = direct_upland,
    direct_inundated = direct_inundated,
    effect_soil = effect_soil,
    effect_air = effect_air,
    effect_vwc = effect_vwc,
    effect_thaw = effect_thaw,
    
    # Indirect paths
    slope_via_soil = slope_via_soil,
    slope_via_air = slope_via_air,
    slope_via_vwc = slope_via_vwc,
    slope_via_thaw = slope_via_thaw,
    upland_via_soil = upland_via_soil,
    upland_via_air = upland_via_air,
    upland_via_vwc = upland_via_vwc,
    upland_via_thaw = upland_via_thaw,
    inundated_via_soil = inundated_via_soil,
    inundated_via_air = inundated_via_air,
    inundated_via_vwc = inundated_via_vwc,
    inundated_via_thaw = inundated_via_thaw,
    
    # Total indirect effects
    slope_indirect = slope_indirect,
    upland_indirect = upland_indirect,
    inundated_indirect = inundated_indirect,
    
    # Total effects
    slope_total = slope_total,
    upland_total = upland_total,
    inundated_total = inundated_total
  ))
}

# Run bootstrap with this simplified function
boot_results_robust <- boot(data = df_long_cc, 
                    statistic = boot_psem_robust, 
                    R = 1000)

# Create summary with robust error handling
effect_summary_robust <- data.frame(
  Path = names(boot_results_robust$t0),
  Estimate = boot_results_robust$t0,
  CI_Lower = NA_real_,
  CI_Upper = NA_real_,
  Significant = NA,
  stringsAsFactors = FALSE
)

# Calculate CIs with robust error handling
for (i in 1:nrow(effect_summary_robust)) {
  # Check for NAs in bootstrap samples
  na_count <- sum(is.na(boot_results_robust$t[,i]))
  if (na_count > 0) {
    cat("Parameter", effect_summary_robust$Path[i], 
        "has", na_count, "missing values out of", nrow(boot_results_robust$t), "\n")
    
    # Skip if too many NAs
    if (na_count > 0.5 * nrow(boot_results_robust$t)) {
      cat("Skipping CI calculation due to excessive missing values\n")
      next
    }
  }
  
  # Try to calculate CI
  tryCatch({
    ci <- boot.ci(boot_results_robust, type = "perc", index = i)
    if (!is.null(ci) && !is.null(ci$percent)) {
      effect_summary_robust$CI_Lower[i] <- ci$percent[4]
      effect_summary_robust$CI_Upper[i] <- ci$percent[5]
      effect_summary_robust$Significant[i] <- (ci$percent[4] * ci$percent[5] > 0)
    }
  }, error = function(e) {
    cat("Error calculating CI for parameter", effect_summary_robust$Path[i], ":", e$message, "\n")
  })
}

# Print summary of valid results
valid_results <- effect_summary_robust[!is.na(effect_summary_robust$Significant), ]
valid_results$Estimate <- round(valid_results$Estimate, 4)
valid_results$CI_Lower <- round(valid_results$CI_Lower, 4)
valid_results$CI_Upper <- round(valid_results$CI_Upper, 4)

# Sort by effect type
direct_effects <- valid_results[grep("^direct|^effect", valid_results$Path), ]
indirect_paths <- valid_results[grep("via", valid_results$Path), ]
indirect_totals <- valid_results[grep("indirect$", valid_results$Path), ]
total_effects <- valid_results[grep("total$", valid_results$Path), ]

# Print results
cat("\n--- DIRECT EFFECTS ---\n")
print(direct_effects[order(-abs(direct_effects$Estimate)), ])

cat("\n--- INDIRECT PATHWAYS ---\n")
print(indirect_paths[order(-abs(indirect_paths$Estimate)), ])

cat("\n--- TOTAL INDIRECT EFFECTS ---\n")
print(indirect_totals)

cat("\n--- TOTAL EFFECTS ---\n")
print(total_effects)


```
####adding in interaction effects
```{r}
# Modified bootstrap function with interaction effects
boot_psem_with_interactions <- function(data, indices) {
  # Create bootstrap sample
  boot_data <- data[indices, ]
  
  # Environmental models (can include interactions here too if needed)
  b_soil <- rlm(soil_temp_10_cm ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_air <- rlm(air_temp ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_vwc <- rlm(VWC ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_thaw <- rlm(thawdepth ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  
  # NEE model with key interactions
  b_nee <- rlm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
              lp_slope + lp_upland + inundated_numeric +
              # Add important interaction terms
              VWC:lp_slope + VWC:lp_upland +     # Effect of VWC varies by landscape position
              thawdepth:inundated_numeric +     # Effect of thaw depth varies by inundation
               thawdepth:air_temp +
                thawdepth:VWC +
                air_temp:soil_temp_10_cm,       
              
              data = boot_data)
  
  # Extract direct effects (now includes interaction terms)
  direct_effects <- c(
    #environmental effects on NEE 
    "effect_soil" = coef(b_nee)["soil_temp_10_cm"],
    "effect_air" = coef(b_nee)["air_temp"],
    "effect vwc" = coef(b_nee)["VWC"],        # Main effect (when lp_slope=0, lp_upland=0)
    "effect_thaw" = coef(b_nee)["thawdepth"],        # Main effect (when inundated_numeric=0)
    #landscape effects on NEE
    "direct_slope" = coef(b_nee)["lp_slope"],
    "direct_upland" = coef(b_nee)["lp_upland"],
    "direct_inundated" = coef(b_nee)["inundated_numeric"],
    # Interaction effects on NEE
    "vwc_x_slope" = coef(b_nee)["VWC:lp_slope"],
    "vwc_x_upland" = coef(b_nee)["VWC:lp_upland"],
    "thaw_x_inundated" = coef(b_nee)["thawdepth:inundated_numeric"],
    "thaw_x_VWC" = coef(b_nee)["thawdepth:VWC"],
    "thaw_x_airtemp" = coef(b_nee)["thawdepth:air_temp"],
    "airtemp_x_soiltemp" = coef(b_nee)["air_temp:soil_temp_10_cm"]
                                
  )
  
  # Landscape effects on environment
  slope_soil <- coef(b_soil)["lp_slope"]
  slope_air <- coef(b_air)["lp_slope"]
  slope_vwc <- coef(b_vwc)["lp_slope"]
  slope_thaw <- coef(b_thaw)["lp_slope"]
  
  upland_soil <- coef(b_soil)["lp_upland"]
  upland_air <- coef(b_air)["lp_upland"]
  upland_vwc <- coef(b_vwc)["lp_upland"]
  upland_thaw <- coef(b_thaw)["lp_upland"]
  
  inundated_soil <- coef(b_soil)["inundated_numeric"]
  inundated_air <- coef(b_air)["inundated_numeric"]
  inundated_vwc <- coef(b_vwc)["inundated_numeric"]
  inundated_thaw <- coef(b_thaw)["inundated_numeric"]
  
  # Individual indirect paths
  slope_via_soil <- slope_soil * effect_soil
  slope_via_air <- slope_air * effect_air
  slope_via_vwc <- slope_vwc * effect_vwc
  slope_via_thaw <- slope_thaw * effect_thaw
  
  upland_via_soil <- upland_soil * effect_soil
  upland_via_air <- upland_air * effect_air
  upland_via_vwc <- upland_vwc * effect_vwc
  upland_via_thaw <- upland_thaw * effect_thaw
  
  inundated_via_soil <- inundated_soil * effect_soil
  inundated_via_air <- inundated_air * effect_air
  inundated_via_vwc <- inundated_vwc * effect_vwc
  inundated_via_thaw <- inundated_thaw * effect_thaw
  
  # Total indirect effects
  slope_indirect <- slope_via_soil + slope_via_air + slope_via_vwc + slope_via_thaw
  upland_indirect <- upland_via_soil + upland_via_air + upland_via_vwc + upland_via_thaw
  inundated_indirect <- inundated_via_soil + inundated_via_air + inundated_via_vwc + inundated_via_thaw
  
  # Total effects
  slope_total <- direct_slope + slope_indirect
  upland_total <- direct_upland + upland_indirect
  inundated_total <- direct_inundated + inundated_indirect
  
  # Return all values as a named vector
  return(c( direct_effects,
    # Direct effects on NEE
    direct_slope = direct_slope,
    direct_upland = direct_upland,
    direct_inundated = direct_inundated,
    effect_soil = effect_soil,
    effect_air = effect_air,
    effect_vwc = effect_vwc,
    effect_thaw = effect_thaw,
    
    # Indirect paths
    slope_via_soil = slope_via_soil,
    slope_via_air = slope_via_air,
    slope_via_vwc = slope_via_vwc,
    slope_via_thaw = slope_via_thaw,
    upland_via_soil = upland_via_soil,
    upland_via_air = upland_via_air,
    upland_via_vwc = upland_via_vwc,
    upland_via_thaw = upland_via_thaw,
    inundated_via_soil = inundated_via_soil,
    inundated_via_air = inundated_via_air,
    inundated_via_vwc = inundated_via_vwc,
    inundated_via_thaw = inundated_via_thaw,
    
    # Total indirect effects
    slope_indirect = slope_indirect,
    upland_indirect = upland_indirect,
    inundated_indirect = inundated_indirect,
    
    # Total effects
    slope_total = slope_total,
    upland_total = upland_total,
    inundated_total = inundated_total
  ))
}

# Run bootstrap with this simplified function
boot_results_interact.robust <- boot(data = df_long_cc, 
                    statistic = boot_psem_with_interactions, 
                    R = 1000)

# Create summary with robust error handling
effect_summary_robust2 <- data.frame(
  Path = names(boot_results_interact.robust$t0),
  Estimate = boot_results_interact.robust$t0,
  CI_Lower = NA_real_,
  CI_Upper = NA_real_,
  Significant = NA,
  stringsAsFactors = FALSE
)

# Calculate CIs with robust error handling
for (i in 1:nrow(effect_summary_robust2)) {
  # Check for NAs in bootstrap samples
  na_count <- sum(is.na(boot_results_interact.robust$t[,i]))
  if (na_count > 0) {
    cat("Parameter", effect_summary_robust2$Path[i], 
        "has", na_count, "missing values out of", nrow(boot_results_interact.robust$t), "\n")
    
    # Skip if too many NAs
    if (na_count > 0.5 * nrow(boot_results_interact.robust$t)) {
      cat("Skipping CI calculation due to excessive missing values\n")
      next
    }
  }
  
  # Try to calculate CI
  tryCatch({
    ci <- boot.ci(boot_results_interact.robust, type = "perc", index = i)
    if (!is.null(ci) && !is.null(ci$percent)) {
      effect_summary_robust2$CI_Lower[i] <- ci$percent[4]
      effect_summary_robust2$CI_Upper[i] <- ci$percent[5]
      effect_summary_robust2$Significant[i] <- (ci$percent[4] * ci$percent[5] > 0)
    }
  }, error = function(e) {
    cat("Error calculating CI for parameter", effect_summary_robust2$Path[i], ":", e$message, "\n")
  })
}

# Print summary of valid results
valid_results2 <- effect_summary_robust2[!is.na(effect_summary_robust2$Significant), ]
valid_results2$Estimate <- round(valid_results2$Estimate, 4)
valid_results2$CI_Lower <- round(valid_results2$CI_Lower, 4)
valid_results2$CI_Upper <- round(valid_results2$CI_Upper, 4)

# Sort by effect type
direct_effects <- valid_results2[grep("^direct|^effect", valid_results2$Path), ]
indirect_paths <- valid_results2[grep("via", valid_results2$Path), ]
indirect_totals <- valid_results2[grep("indirect$", valid_results2$Path), ]
total_effects <- valid_results2[grep("total$", valid_results2$Path), ]

# Print results
cat("\n--- DIRECT EFFECTS ---\n")
print(direct_effects[order(-abs(direct_effects$Estimate)), ])

cat("\n--- INDIRECT PATHWAYS ---\n")
print(indirect_paths[order(-abs(indirect_paths$Estimate)), ])

cat("\n--- TOTAL INDIRECT EFFECTS ---\n")
print(indirect_totals)

cat("\n--- TOTAL EFFECTS ---\n")
print(total_effects)
  

```
```{r}
# Modified bootstrap function with interaction effects
boot_psem_with_interactions <- function(data, indices) {
  # Create bootstrap sample
  boot_data <- data[indices, ]
  
  # Environmental models
  b_soil <- rlm(soil_temp_10_cm ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_air <- rlm(air_temp ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_vwc <- rlm(VWC ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_thaw <- rlm(thawdepth ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  
  # NEE model with key interactions
  b_nee <- rlm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
              lp_slope + lp_upland + inundated_numeric +
              # Add important interaction terms
              VWC:lp_slope + VWC:lp_upland +     
              thawdepth:inundated_numeric +
              thawdepth:air_temp +
              thawdepth:VWC +
              air_temp:soil_temp_10_cm,
              data = boot_data)
  
  # Extract environmental effects on NEE
  env_effect_soil <- coef(b_nee)["soil_temp_10_cm"]
  env_effect_air <- coef(b_nee)["air_temp"]
  env_effect_vwc <- coef(b_nee)["VWC"]
  env_effect_thaw <- coef(b_nee)["thawdepth"]
  
  # Direct landscape effects on NEE
  direct_slope <- coef(b_nee)["lp_slope"]
  direct_upland <- coef(b_nee)["lp_upland"]
  direct_inundated <- coef(b_nee)["inundated_numeric"]
  
  # Extract interaction effects  
  int_vwc_slope <- coef(b_nee)["VWC:lp_slope"]
  int_vwc_upland <- coef(b_nee)["VWC:lp_upland"]
  int_thaw_inundated <- coef(b_nee)["thawdepth:inundated_numeric"]
  int_thaw_vwc <- coef(b_nee)["thawdepth:VWC"]
  int_thaw_airtemp <- coef(b_nee)["thawdepth:air_temp"]
  int_airtemp_soiltemp <- coef(b_nee)["air_temp:soil_temp_10_cm"]
  
  # Combine all direct effects
  direct_effects <- c(
    "effect_soil" = env_effect_soil,
    "effect_air" = env_effect_air,
    "effect_vwc" = env_effect_vwc,
    "effect_thaw" = env_effect_thaw,
    "direct_slope" = direct_slope,
    "direct_upland" = direct_upland, 
    "direct_inundated" = direct_inundated,
    "int_vwc_slope" = int_vwc_slope,
    "int_vwc_upland" = int_vwc_upland,
    "int_thaw_inundated" = int_thaw_inundated,
    "int_thaw_vwc" = int_thaw_vwc,
    "int_thaw_airtemp" = int_thaw_airtemp,
    "int_airtemp_soiltemp" = int_airtemp_soiltemp
  )
  
  # Landscape effects on environment
  slope_soil <- coef(b_soil)["lp_slope"]
  slope_air <- coef(b_air)["lp_slope"]
  slope_vwc <- coef(b_vwc)["lp_slope"]
  slope_thaw <- coef(b_thaw)["lp_slope"]
  
  upland_soil <- coef(b_soil)["lp_upland"]
  upland_air <- coef(b_air)["lp_upland"]
  upland_vwc <- coef(b_vwc)["lp_upland"]
  upland_thaw <- coef(b_thaw)["lp_upland"]
  
  inundated_soil <- coef(b_soil)["inundated_numeric"]
  inundated_air <- coef(b_air)["inundated_numeric"]
  inundated_vwc <- coef(b_vwc)["inundated_numeric"]
  inundated_thaw <- coef(b_thaw)["inundated_numeric"]
  
  # Individual indirect paths - now using the correctly named variables
  slope_via_soil <- slope_soil * env_effect_soil
  slope_via_air <- slope_air * env_effect_air
  slope_via_vwc <- slope_vwc * env_effect_vwc
  slope_via_thaw <- slope_thaw * env_effect_thaw
  
  upland_via_soil <- upland_soil * env_effect_soil
  upland_via_air <- upland_air * env_effect_air
  upland_via_vwc <- upland_vwc * env_effect_vwc
  upland_via_thaw <- upland_thaw * env_effect_thaw
  
  inundated_via_soil <- inundated_soil * env_effect_soil
  inundated_via_air <- inundated_air * env_effect_air
  inundated_via_vwc <- inundated_vwc * env_effect_vwc
  inundated_via_thaw <- inundated_thaw * env_effect_thaw
  
  # Total indirect effects
  slope_indirect <- slope_via_soil + slope_via_air + slope_via_vwc + slope_via_thaw
  upland_indirect <- upland_via_soil + upland_via_air + upland_via_vwc + upland_via_thaw
  inundated_indirect <- inundated_via_soil + inundated_via_air + inundated_via_vwc + inundated_via_thaw
  
  # Total effects
  slope_total <- direct_slope + slope_indirect
  upland_total <- direct_upland + upland_indirect
  inundated_total <- direct_inundated + inundated_indirect
  
  # Create a vector of all effects to return
  all_effects <- c(
    # Direct and interaction effects
    direct_effects,
    
    # Landscape-to-environment effects
    "slope_soil" = slope_soil,
    "slope_air" = slope_air,
    "slope_vwc" = slope_vwc,
    "slope_thaw" = slope_thaw,
    "upland_soil" = upland_soil,
    "upland_air" = upland_air,
    "upland_vwc" = upland_vwc,
    "upland_thaw" = upland_thaw,
    "inundated_soil" = inundated_soil,
    "inundated_air" = inundated_air,
    "inundated_vwc" = inundated_vwc,
    "inundated_thaw" = inundated_thaw,
    
    # Indirect pathways
    "slope_via_soil" = slope_via_soil,
    "slope_via_air" = slope_via_air,
    "slope_via_vwc" = slope_via_vwc,
    "slope_via_thaw" = slope_via_thaw,
    "upland_via_soil" = upland_via_soil,
    "upland_via_air" = upland_via_air,
    "upland_via_vwc" = upland_via_vwc,
    "upland_via_thaw" = upland_via_thaw,
    "inundated_via_soil" = inundated_via_soil,
    "inundated_via_air" = inundated_via_air,
    "inundated_via_vwc" = inundated_via_vwc,
    "inundated_via_thaw" = inundated_via_thaw,
    
    # Total indirect effects
    "slope_indirect" = slope_indirect,
    "upland_indirect" = upland_indirect,
    "inundated_indirect" = inundated_indirect,
    
    # Total effects
    "slope_total" = slope_total,
    "upland_total" = upland_total,
    "inundated_total" = inundated_total
  )
  
  return(all_effects)
}

# Run bootstrap with this simplified function
boot_results_interact.robust <- boot(data = df_long_cc, 
                    statistic = boot_psem_with_interactions, 
                    R = 1000)

# Create summary with robust error handling
effect_summary_robust2 <- data.frame(
  Path = names(boot_results_interact.robust$t0),
  Estimate = boot_results_interact.robust$t0,
  CI_Lower = NA_real_,
  CI_Upper = NA_real_,
  Significant = NA,
  stringsAsFactors = FALSE
)

# Calculate CIs with robust error handling
for (i in 1:nrow(effect_summary_robust2)) {
  # Check for NAs in bootstrap samples
  na_count <- sum(is.na(boot_results_interact.robust$t[,i]))
  if (na_count > 0) {
    cat("Parameter", effect_summary_robust2$Path[i], 
        "has", na_count, "missing values out of", nrow(boot_results_interact.robust$t), "\n")
    
    # Skip if too many NAs
    if (na_count > 0.5 * nrow(boot_results_interact.robust$t)) {
      cat("Skipping CI calculation due to excessive missing values\n")
      next
    }
  }
  
  # Try to calculate CI
  tryCatch({
    ci <- boot.ci(boot_results_interact.robust, type = "perc", index = i)
    if (!is.null(ci) && !is.null(ci$percent)) {
      effect_summary_robust2$CI_Lower[i] <- ci$percent[4]
      effect_summary_robust2$CI_Upper[i] <- ci$percent[5]
      effect_summary_robust2$Significant[i] <- (ci$percent[4] * ci$percent[5] > 0)
    }
  }, error = function(e) {
    cat("Error calculating CI for parameter", effect_summary_robust2$Path[i], ":", e$message, "\n")
  })
}

# Print summary of valid results
valid_results2 <- effect_summary_robust2[!is.na(effect_summary_robust2$Significant), ]
valid_results2$Estimate <- round(valid_results2$Estimate, 4)
valid_results2$CI_Lower <- round(valid_results2$CI_Lower, 4)
valid_results2$CI_Upper <- round(valid_results2$CI_Upper, 4)

# Sort by effect type
direct_effects <- valid_results2[grep("^direct|^effect", valid_results2$Path), ]
indirect_paths <- valid_results2[grep("via", valid_results2$Path), ]
indirect_totals <- valid_results2[grep("indirect$", valid_results2$Path), ]
total_effects <- valid_results2[grep("total$", valid_results2$Path), ]

# Print results
cat("\n--- DIRECT EFFECTS ---\n")
print(direct_effects[order(-abs(direct_effects$Estimate)), ])

cat("\n--- INDIRECT PATHWAYS ---\n")
print(indirect_paths[order(-abs(indirect_paths$Estimate)), ])

cat("\n--- TOTAL INDIRECT EFFECTS ---\n")
print(indirect_totals)

cat("\n--- TOTAL EFFECTS ---\n")
print(total_effects)

```
#TO DO: ASK claude about these errors *** for robust bootstrap with interactions 


```{r}
# Fit NEE model using different approaches
lm_model <- lm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
               lp_slope + lp_upland + inundated_numeric, data = df_long_cc)

rlm_model <- rlm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
                lp_slope + lp_upland + inundated_numeric, data = df_long_cc)

glm_model <- glm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
                lp_slope + lp_upland + inundated_numeric, 
                family = gaussian(link = "identity"), data = df_long_cc)



# Compare residuals
par(mfrow=c(2,3))
hist(residuals(lm_model), main="LM Residuals")
hist(residuals(rlm_model), main="RLM Residuals")
hist(residuals(glm_model), main="GLM Residuals")


# Compare QQ plots
qqnorm(residuals(lm_model), main="LM QQ Plot"); qqline(residuals(lm_model))
qqnorm(residuals(rlm_model), main="RLM QQ Plot"); qqline(residuals(rlm_model))
qqnorm(residuals(glm_model), main="GLM QQ Plot"); qqline(residuals(glm_model))

#all essentially the same with a strong positive skew, going with robust as it's best able to handle non-normal distributions

```


####using glm 
```{r}
library(boot)
library(piecewiseSEM)

# Simplified bootstrap function focusing just on direct and indirect effects
boot_psem_robust <- function(data, indices) {
  # Create bootstrap sample
  boot_data <- data[indices, ]
  
  # Fit models on bootstrap sample - using robust regression here 
  b_soil <- glm(soil_temp_10_cm ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_air <- glm(air_temp ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_vwc <- glm(VWC ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_thaw <- glm(thawdepth ~ lp_slope + lp_upland + inundated_numeric, data = boot_data)
  b_nee <- glm(NEE_umol ~ soil_temp_10_cm + air_temp + VWC + thawdepth + 
             lp_slope + lp_upland + inundated_numeric, data = boot_data)
  
  # Direct effects
  direct_slope <- coef(b_nee)["lp_slope"]
  direct_upland <- coef(b_nee)["lp_upland"]
  direct_inundated <- coef(b_nee)["inundated_numeric"]
  
  # Environmental effects on NEE
  effect_soil <- coef(b_nee)["soil_temp_10_cm"]
  effect_air <- coef(b_nee)["air_temp"]
  effect_vwc <- coef(b_nee)["VWC"]
  effect_thaw <- coef(b_nee)["thawdepth"]
  
  # Landscape effects on environment
  slope_soil <- coef(b_soil)["lp_slope"]
  slope_air <- coef(b_air)["lp_slope"]
  slope_vwc <- coef(b_vwc)["lp_slope"]
  slope_thaw <- coef(b_thaw)["lp_slope"]
  
  upland_soil <- coef(b_soil)["lp_upland"]
  upland_air <- coef(b_air)["lp_upland"]
  upland_vwc <- coef(b_vwc)["lp_upland"]
  upland_thaw <- coef(b_thaw)["lp_upland"]
  
  inundated_soil <- coef(b_soil)["inundated_numeric"]
  inundated_air <- coef(b_air)["inundated_numeric"]
  inundated_vwc <- coef(b_vwc)["inundated_numeric"]
  inundated_thaw <- coef(b_thaw)["inundated_numeric"]
  
  # Individual indirect paths
  slope_via_soil <- slope_soil * effect_soil
  slope_via_air <- slope_air * effect_air
  slope_via_vwc <- slope_vwc * effect_vwc
  slope_via_thaw <- slope_thaw * effect_thaw
  
  upland_via_soil <- upland_soil * effect_soil
  upland_via_air <- upland_air * effect_air
  upland_via_vwc <- upland_vwc * effect_vwc
  upland_via_thaw <- upland_thaw * effect_thaw
  
  inundated_via_soil <- inundated_soil * effect_soil
  inundated_via_air <- inundated_air * effect_air
  inundated_via_vwc <- inundated_vwc * effect_vwc
  inundated_via_thaw <- inundated_thaw * effect_thaw
  
  # Total indirect effects
  slope_indirect <- slope_via_soil + slope_via_air + slope_via_vwc + slope_via_thaw
  upland_indirect <- upland_via_soil + upland_via_air + upland_via_vwc + upland_via_thaw
  inundated_indirect <- inundated_via_soil + inundated_via_air + inundated_via_vwc + inundated_via_thaw
  
  # Total effects
  slope_total <- direct_slope + slope_indirect
  upland_total <- direct_upland + upland_indirect
  inundated_total <- direct_inundated + inundated_indirect
  
  # Return all values as a named vector
  return(c(
    # Direct effects on NEE
    direct_slope = direct_slope,
    direct_upland = direct_upland,
    direct_inundated = direct_inundated,
    effect_soil = effect_soil,
    effect_air = effect_air,
    effect_vwc = effect_vwc,
    effect_thaw = effect_thaw,
    
    # Indirect paths
    slope_via_soil = slope_via_soil,
    slope_via_air = slope_via_air,
    slope_via_vwc = slope_via_vwc,
    slope_via_thaw = slope_via_thaw,
    upland_via_soil = upland_via_soil,
    upland_via_air = upland_via_air,
    upland_via_vwc = upland_via_vwc,
    upland_via_thaw = upland_via_thaw,
    inundated_via_soil = inundated_via_soil,
    inundated_via_air = inundated_via_air,
    inundated_via_vwc = inundated_via_vwc,
    inundated_via_thaw = inundated_via_thaw,
    
    # Total indirect effects
    slope_indirect = slope_indirect,
    upland_indirect = upland_indirect,
    inundated_indirect = inundated_indirect,
    
    # Total effects
    slope_total = slope_total,
    upland_total = upland_total,
    inundated_total = inundated_total
  ))
}

# Run bootstrap with this simplified function
boot_results_robust <- boot(data = df_long_cc, 
                    statistic = boot_psem_robust, 
                    R = 1000)

# Create summary with robust error handling
effect_summary_robust <- data.frame(
  Path = names(boot_results_robust$t0),
  Estimate = boot_results_robust$t0,
  CI_Lower = NA_real_,
  CI_Upper = NA_real_,
  Significant = NA,
  stringsAsFactors = FALSE
)

# Calculate CIs with robust error handling
for (i in 1:nrow(effect_summary_robust)) {
  # Check for NAs in bootstrap samples
  na_count <- sum(is.na(boot_results_robust$t[,i]))
  if (na_count > 0) {
    cat("Parameter", effect_summary_robust$Path[i], 
        "has", na_count, "missing values out of", nrow(boot_results_robust$t), "\n")
    
    # Skip if too many NAs
    if (na_count > 0.5 * nrow(boot_results_robust$t)) {
      cat("Skipping CI calculation due to excessive missing values\n")
      next
    }
  }
  
  # Try to calculate CI
  tryCatch({
    ci <- boot.ci(boot_results_robust, type = "perc", index = i)
    if (!is.null(ci) && !is.null(ci$percent)) {
      effect_summary_robust$CI_Lower[i] <- ci$percent[4]
      effect_summary_robust$CI_Upper[i] <- ci$percent[5]
      effect_summary_robust$Significant[i] <- (ci$percent[4] * ci$percent[5] > 0)
    }
  }, error = function(e) {
    cat("Error calculating CI for parameter", effect_summary_robust$Path[i], ":", e$message, "\n")
  })
}

# Print summary of valid results
valid_results <- effect_summary_robust[!is.na(effect_summary_robust$Significant), ]
valid_resultst$Estimate <- round(valid_results$Estimate, 4)
valid_results$CI_Lower <- round(valid_results$CI_Lower, 4)
valid_results$CI_Upper <- round(valid_results$CI_Upper, 4)

# Sort by effect type
direct_effects <- valid_results[grep("^direct|^effect", valid_results$Path), ]
indirect_paths <- valid_results[grep("via", valid_results$Path), ]
indirect_totals <- valid_results[grep("indirect$", valid_results$Path), ]
total_effects <- valid_results[grep("total$", valid_results$Path), ]

# Print results
cat("\n--- DIRECT EFFECTS ---\n")
print(direct_effects[order(-abs(direct_effects$Estimate)), ])

cat("\n--- INDIRECT PATHWAYS ---\n")
print(indirect_paths[order(-abs(indirect_paths$Estimate)), ])

cat("\n--- TOTAL INDIRECT EFFECTS ---\n")
print(indirect_totals)

cat("\n--- TOTAL EFFECTS ---\n")
print(total_effects)

```
#### p-values from bootstrapping
```{r}
# Calculate p-values from bootstrap distribution
effect_summary$p_value <- NA

for (i in 1:nrow(effect_summary)) {
  if (!is.na(effect_summary$CI_Lower[i])) {
    # Count how many bootstrap samples had opposite sign from estimate
    opposite_sign <- sum((boot_results$t[,i] > 0) != (effect_summary$Estimate[i] > 0), na.rm = TRUE)
    total_samples <- sum(!is.na(boot_results$t[,i]))
    
    # Calculate p-value (two-tailed)
    effect_summary$p_value[i] <- 2 * (opposite_sign / total_samples)
    
    # Cap at 1 (can't have p > 1)
    effect_summary$p_value[i] <- min(effect_summary$p_value[i], 1)
  }
}

# Add significance stars
effect_summary$stars <- ""
effect_summary$stars[effect_summary$p_value < 0.05] <- "*"
effect_summary$stars[effect_summary$p_value < 0.01] <- "**"
effect_summary$stars[effect_summary$p_value < 0.001] <- "***"

# Print updated results with p-values
valid_results <- effect_summary[!is.na(effect_summary$p_value), ]
valid_results$p_value <- round(valid_results$p_value, 4)
print(valid_results[, c("Path", "Estimate", "CI_Lower", "CI_Upper", "p_value", "stars")])
```



#linear relationships 

####RECO vs temp relationship -lm with R2 and p 
```{r}

library(ggplot2)

# Create the plot with points and a linear regression line
soil_RECO <- ggplot(df_RECO, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship between RECO and Soil Temperature",
       x = "Soil Temperature at 10 cm",
       y = "RECO (Carbon Flux)")

# Print the plot
print(soil_RECO)



# Fit a linear model
model <- lm(flux_value ~ soil_temp_10_cm, data = df_RECO, na.action=na.exclude)

# Calculate R² value
r_squared <- summary(model)$r.squared
cat("R² value:", r_squared, "\n")

# Test for significance
summary(model)


#Add R2 to the figure 
# Fit a linear model
model <- lm(flux_value ~ soil_temp_10_cm, data = df_RECO)

# Calculate R² value
r_squared <- summary(model)$r.squared

# Create the plot with points, a linear regression line, and R² annotation
soil_RECO2 <- ggplot(df_RECO, aes(x = soil_temp_10_cm, y = flux_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  annotate("text", x = Inf, y = Inf, label = paste("R² =", round(r_squared, 2)),
           hjust = 1.1, vjust = 1.1, size = 5, color = "red") +
  labs(title = "Relationship between RECO and Soil Temperature",
       x = "Soil Temperature at 10 cm",
       y = "RECO (Carbon Flux)")+
  theme_minimal()

# Print the plot
print(soil_RECO2)

```


#TO DO: how enviro variables change by landpos, and from year to year?







